### 2019-12-27 manuscript/appendix.Rmd
first commit

---
header-includes:
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage{lineno}
  - \usepackage[width=\textwidth]{caption}
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: no
  html_document:
    fig_caption: yes
    force_captions: yes
    number_sections: yes
classoption: landscape
---

```{r setup, include=FALSE, message=FALSE}
  knitr::opts_chunk$set(echo = TRUE)
  library(orthodrc)
  library(landscapemetrics)
  library(raster)
  library(ggplot2)
  library(ggthemes)
  library(tidyverse)
  library(knitr)
  library(kableExtra)

  r <- raster("../data/orthomosaic/yangambi_forest_cover_difference_1958_2000.tif")

  # read site data
  psp <- read.table("../data/surveys/site_characteristics.csv",
                    header = TRUE,
                    sep = ",")

  # Define lat / lon projection.
  lat_lon <- CRS("+init=epsg:4326")

  # Read in the coordinates and assign them a projection
  ll <- sp::SpatialPoints(cbind(psp$lon,psp$lat), lat_lon)

  # stuff results into original dataframe
  lulcc_px <- raster::extract(r, ll, buffer = 100)

  psp$lulcc_summary <- lapply(lulcc_px, function(x){
    s <- table(x)
    s <- round(s/sum(s)*100)
    n <- names(s)
    paste(paste(s," (",n,")", sep = ""), collapse = ", ")
  })

  # flight data
  flight_data <- read.table("../data/orthomosaic/flight_path_meta-data.csv",
                    header = TRUE,
                    sep = ",")

  # select classes into their respective time frame
  historical <- (r == 2 | r == 4)
  contemporary <- (r == 3 | r == 4)

  # reproject to meter based mercator projection
  # should be fine around the equator
  historical <- projectRaster(historical,
                              crs = "+init=epsg:3395",
                              method = "ngb")

  contemporary <- projectRaster(contemporary,
                              crs = "+init=epsg:3395",
                              method = "ngb")

  frag_stats <- rbind(landscape_stats(historical),
                 landscape_stats(contemporary))
```

# Appendix

```{r echo = FALSE, eval = TRUE}
psp %>%
  select(-c("label")) %>%
  kable(
    caption="Site description of all the current permanent sampling plots in the greater Yangambi region. We list the forest type, plot number, geographic location (in decimal degrees), stem density, basal area Above Ground Carbon (AGC), species richness and Land-Use and Land-Cover Change (LULCC) classes in a radius of 100m around each plot location.",
    booktabs = TRUE,
    escape = FALSE,
    col.names = c(
         "type",
         "nr",
         "latitude",
         "longitude",
         "stem density ($ha\\textsuperscript{-1}$)",
         "basal area ($m\\textsuperscript{2}$ $ha\\textsuperscript{-1}$)",
         "AGC (Mg C $ha\\textsuperscript{-1}$)",
         "species richness",
         "LULCC class coverage (\\%)")) %>%
  kable_styling(latex_options = c("hold_position"))
```

```{r echo = FALSE, eval = TRUE}
flight_data %>%
  kable(
    caption="Flight path meta-data for the Isangi-Stanleyville aerial survey. Data provided consists of the flight path number, cardinal direction of the flight, the image numbers, and the duration of the flight provided by the start and end time of the acquisition. Data is sourced from Appendix Figure 2 and the sensor logs recorded in the margin of acquired images (see Figure 1 main manuscript).",
    booktabs = TRUE,
    escape = FALSE,
    col.names = c("path",
                  "direction",
                  "image \\#",
                  "start (H:M:S)",
                  "end (H:M:S)")
    ) %>%
  kable_styling(latex_options = c("hold_position"))
```

\pagebreak

```{r echo=FALSE, out.width='100%', fig.cap="Overview of the complete flight plan of the survey around Kisangani (then Stanleyville) stored in the archives at the Africa Museum."}
knitr::include_graphics('./figures/flight_paths_overview.jpg')
```

```{r echo=FALSE, out.height='75%', fig.cap="Overview of the complete flight plan meta-data as stored in the archives at the Africa Museum."}
knitr::include_graphics('./figures/flight_paths_meta_data.jpg')
```

```{r echo=FALSE, out.height='50%', fig.cap="Boxplots comparing the first principal component (PC1) of a site based FOTO analysis across different forest types for both contemporary (Geo-Eye) and historical orthomosaic data."}
knitr::include_graphics('./figures/foto_bplot_psp.png')
```

```{r echo = FALSE, out.width="75%", eval = TRUE, fig.cap="Scatterplot comparing Above Ground Carbon and the first principal component of the FOTO analysis for contemporary Geo-Eye (left) and the historical orthomosaic (right) data. Different forest types are plotted using closed circles, closed triangles, closed squares and this for fallow, mixed mon-dominant and young-regrowth forests, respectivelly "}
knitr::include_graphics('./figures/foto_pc1_agc.png')
```

```{r echo = FALSE, out.width="75%", eval = TRUE, fig.cap="Scatterplot comparing tree species richness and the first principal component of the FOTO analysis for contemporary Geo-Eye (left) and the historical orthomosaic (right) data. Different forest types are plotted using closed circles, closed triangles, closed squares and this for fallow, mixed mon-dominant and young-regrowth forests, respectivelly"}
knitr::include_graphics('./figures/foto_pc1_diversity.png')
```

```{r echo = FALSE, out.width="75%", eval = TRUE, fig.cap="Scatterplot comparing stem density (per ha) and the first principal component of the FOTO analysis for contemporary Geo-Eye (left) and the historical orthomosaic (right) data. Different forest types are plotted using closed circles, closed triangles, closed squares and this for fallow, mixed mon-dominant and young-regrowth forests, respectivelly"}
knitr::include_graphics('./figures/foto_pc1_stems.png')
```

### 2019-12-27 manuscript/appendix.tex
first commit

\documentclass[landscape]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx}
% grffile has become a legacy package: https://ctan.org/pkg/grffile
\IfFileExists{grffile.sty}{%
\usepackage{grffile}
}{}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\providecommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{}
    \pretitle{\vspace{\droptitle}}
  \posttitle{}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\usepackage{setspace}
\doublespacing
\usepackage{lineno}
\usepackage[width=\textwidth]{caption}

\begin{document}

\hypertarget{appendix}{%
\section{Appendix}\label{appendix}}

\begin{table}[!h]

\caption{\label{tab:unnamed-chunk-1}Site description of all the current permanent sampling plots in the greater Yangambi region. We list the forest type, plot number, geographic location (in decimal degrees), stem density, basal area Above Ground Carbon (AGC), species richness and Land-Use and Land-Cover Change (LULCC) classes in a radius of 100m around each plot location.}
\centering
\begin{tabular}[t]{lrrrrrrrl}
\toprule
type & nr & latitude & longitude & stem density ($ha\textsuperscript{-1}$) & basal area ($m\textsuperscript{2}$ $ha\textsuperscript{-1}$) & AGC (Mg C $ha\textsuperscript{-1}$) & species richness & LULCC class coverage (\%)\\
\midrule
fallow & 1 & 0.7959139 & 24.49423 & 350 & 5.39000 & 6.7000 & 26 & 37 (1), 63 (3)\\
fallow & 2 & 0.7921028 & 24.49717 & 132 & 2.06000 & 2.0400 & 22 & 68 (1), 32 (3)\\
young-regrowth & 1 & 0.7894028 & 24.51755 & 322 & 20.44000 & 46.5600 & 40 & 68 (1), 32 (3)\\
young-regrowth & 2 & 0.7948806 & 24.49193 & 447 & 16.63000 & 27.8300 & 25 & 56 (1), 44 (3)\\
young-regrowth & 3 & 0.7931083 & 24.49013 & 313 & 17.80000 & 37.1000 & 43 & 51 (1), 49 (3)\\
\addlinespace
old-regrowth & 1 & 0.8824444 & 24.49614 & 384 & 19.48000 & 81.7800 & 92 & 100 (1)\\
mixed & 1 & 0.8135486 & 24.51264 & 563 & 34.81000 & 168.0400 & 83 & 100 (1)\\
mixed & 2 & 0.7805000 & 24.52109 & 403 & 35.25000 & 174.5800 & 78 & 100 (1)\\
mixed & 3 & 0.7866181 & 24.52349 & 367 & 30.69000 & 150.3000 & 75 & 100 (1)\\
mixed & 4 & 0.8140333 & 24.49370 & 432 & 33.01000 & 168.6100 & 84 & 100 (1)\\
\addlinespace
mixed & 5 & 0.8025611 & 24.48750 & 329 & 25.20000 & 124.1800 & 80 & 100 (1)\\
mixed & 6 & 0.9925556 & 24.53906 & 490 & 32.63000 & 203.8100 & 86 & 100 (1)\\
mixed & 7 & 0.9898056 & 24.53853 & 598 & 31.47000 & 146.1100 & 92 & 100 (1)\\
mixed & 8 & 0.9866111 & 24.53856 & 556 & 29.20000 & 148.2100 & 90 & 100 (1)\\
mono-dominant & 1 & 0.8269306 & 24.52303 & 344 & 31.80000 & 183.0100 & 48 & 100 (1)\\
\addlinespace
mono-dominant & 2 & 0.8282708 & 24.53196 & 436 & 32.06000 & 173.5800 & 55 & 100 (1)\\
mono-dominant & 3 & 0.7965903 & 24.49779 & 376 & 30.57000 & 166.4400 & 62 & 100 (1)\\
mono-dominant & 4 & 0.8080889 & 24.52811 & 374 & 27.69000 & 145.5500 & 46 & 100 (1)\\
mono-dominant & 5 & 0.8682806 & 24.45655 & 217 & 27.19000 & 159.0000 & 46 & 95 (1), 5 (2)\\
mono-dominant & 6 & 0.7992278 & 24.49195 & 378 & 33.95000 & 165.4800 & 68 & 100 (1)\\
\addlinespace
edge & 1 & 0.7895486 & 24.51990 & 415 & 30.93601 & 152.2792 & 77 & 100 (1)\\
edge & 2 & 0.7979472 & 24.48885 & 368 & 31.82097 & 152.1736 & 72 & 95 (1), 5 (3)\\
edge & 3 & 0.8144250 & 24.48555 & 458 & 35.79959 & 197.4022 & 86 & 100 (1)\\
edge & 4 & 0.7659306 & 24.51125 & 320 & 29.72547 & 154.6380 & 75 & 98 (1), 2 (3)\\
edge & 5 & 0.8691778 & 24.47019 & 459 & 33.72462 & 158.6934 & 88 & 100 (1)\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[!h]

\caption{\label{tab:unnamed-chunk-2}Flight path meta-data for the Isangi-Stanleyville aerial survey. Data provided consists of the flight path number, cardinal direction of the flight, the image numbers, and the duration of the flight provided by the start and end time of the acquisition. Data is sourced from Appendix Figure 2 and the sensor logs recorded in the margin of acquired images (see Figure 1 main manuscript).}
\centering
\begin{tabular}[t]{rllll}
\toprule
path & direction & image \# & start (H:M:S) & end (H:M:S)\\
\midrule
1 & W – E & 04 / 78 – 04 / 127 & 10:15:00 & 10:40:00\\
2 & W – E & 05 / 01 – 05 / 56 & 10:15:00 & 10:40:00\\
3 & W – E & 06 / 01 – 06 / 48 & 08:55:00 & 09:20:00\\
4 & W – E & 07 / 01 – 07 / 20 & 09:30:00 & 09:45:00\\
5 & W – E & 05 / 167 – 05 / 187 & 08:55:00 & 09:10:00\\
\addlinespace
6 & W – E & 07 / 67 – 07 / 82 & 10:50:00 & 11:00:00\\
7 & E – W & 06 / 49 – 06 / 95 & 09:35:00 & 10:05:00\\
8 & E – W & 05 / 57 – 05 / 103 & 10:20:00 & 10:50:00\\
9 & E – W & 04 / 128 – 04 / 176 & 10:50:00 & 11:10:00\\
10 & W – E & 08 / 01 – 08 / 27 & 09:05:00 & 09:20:00\\
\addlinespace
11 & W – E & 06 / 96 – 06 / 139 & 10:25:00 & 10:45:00\\
\bottomrule
\end{tabular}
\end{table}

\pagebreak

\begin{figure}
\includegraphics[width=1\linewidth]{./figures/flight_paths_overview} \caption{Overview of the complete flight plan of the survey around Kisangani (then Stanleyville) stored in the archives at the Africa Museum.}\label{fig:unnamed-chunk-3}
\end{figure}

\begin{figure}
\includegraphics[width=82.67in,height=0.75\textheight]{./figures/flight_paths_meta_data} \caption{Overview of the complete flight plan meta-data as stored in the archives at the Africa Museum.}\label{fig:unnamed-chunk-4}
\end{figure}

\begin{figure}
\includegraphics[width=14.58in,height=0.5\textheight]{./figures/foto_bplot_psp} \caption{Boxplots comparing the first principal component (PC1) of a site based FOTO analysis across different forest types for both contemporary (Geo-Eye) and historical orthomosaic data.}\label{fig:unnamed-chunk-5}
\end{figure}

\begin{figure}
\includegraphics[width=0.75\linewidth]{./figures/foto_pc1_agc} \caption{Scatterplot comparing Above Ground Carbon and the first principal component of the FOTO analysis for contemporary Geo-Eye (left) and the historical orthomosaic (right) data. Different forest types are plotted using closed circles, closed triangles, closed squares and this for fallow, mixed mon-dominant and young-regrowth forests, respectivelly }\label{fig:unnamed-chunk-6}
\end{figure}

\begin{figure}
\includegraphics[width=0.75\linewidth]{./figures/foto_pc1_diversity} \caption{Scatterplot comparing tree species richness and the first principal component of the FOTO analysis for contemporary Geo-Eye (left) and the historical orthomosaic (right) data. Different forest types are plotted using closed circles, closed triangles, closed squares and this for fallow, mixed mon-dominant and young-regrowth forests, respectivelly}\label{fig:unnamed-chunk-7}
\end{figure}

\begin{figure}
\includegraphics[width=0.75\linewidth]{./figures/foto_pc1_stems} \caption{Scatterplot comparing stem density (per ha) and the first principal component of the FOTO analysis for contemporary Geo-Eye (left) and the historical orthomosaic (right) data. Different forest types are plotted using closed circles, closed triangles, closed squares and this for fallow, mixed mon-dominant and young-regrowth forests, respectivelly}\label{fig:unnamed-chunk-8}
\end{figure}

\end{document}

### 2019-12-27 manuscript/main.Rmd
first commit

---
title: Historical aerial surveys map long-term changes of forest cover and structure in the central Congo Basin

author: 
  - name: Koen Hufkens
    affil: 1,2,*
    orcid: 0000-0002-5070-8109
  - name: Thalès de Haulleville
    affil: 3
  - name: Elizabeth Kearsley
    affil: 1
  - name: Kim Jacobsen
    affil: 2,3
  - name: Hans Beeckman
    affil: 3
  - name: Piet Stoffelen
    affil: 4
  - name: Filip Vandelook
    affil: 4
  - name: Sofie Meeus
    affil: 4
  - name: Michael Amara
    affil: 5
  - name: Leen Van Hirtum
    affil: 5
  - name: Jan Van den Bulcke
    affil: 1
  - name: Hans Verbeeck
    affil: 1
  - name: Lisa Wingate
    affil: 2

affiliation:
  - num: 1
    address: Ghent University, Ghent, Belgium
  - num: 2
    address: INRA, UMR ISPA, Villenave d’Ornon, France
  - num: 3
    address: Royal Museum for Central Africa, Tervuren, Belgium
  - num: 4
    address: Botanic Garden Meise, Meise, Belgium
  - num: 5
    address: National Archives of Belgium, Brussels, Belgium

correspondence: koen.hufkens@gmail.com
journal: remote sensing
type: article
status: submit
bibliography: bibliography.bib

header-includes: >
  \usepackage{wrapfig}
  \usepackage{graphics}
  \usepackage{setspace}
  \doublespacing
  \usepackage{gensymb}
  \usepackage{booktabs}

abstract: |
   Land Use and Land Cover change (LULCC) of African rainforest contribute to global carbon emissions. Yet, most historical estimates of LULCC and their carbon emissions rely on non-spatially explicity data in the pre-satellite era (< 1972). Past studies have expanded this pre-satellite time frame with declassified satellite surveillance data or aerial surveys, but none cover the Congo Basin. Here, we use historical aerial survey photos to map the extent and structure of LULCC around Yangambi, central Congo Basin, in 1958. Our study leveraged Structure-from-Motion to generate a large orthomosaic covering 828 km^2^, counting 733 million pixels, at a sub meter resolution and geo-referenced to ~4.7 $\pm$ 4.3 m. Primary forest in the orthomosaic was classified with a Convolutional Neural Network Deep Learning approach. We used these data to quantify LULCC, landscape and canopy complexity and scale above ground biomass between historical and contemporary condition. We show a shift from historical highly structured industrial deforestation in 1958 (162 km^2^), to contemporary smallholder farming clearing (88 km^2^), increasing landscape fragmentation and forest edges. Efforts to quantify canopy texture and their link to carbon storage had limited to no success. Our analysis provides an insight in key spatial and temporal patterns of deforestation and reforestation at a multi-decadal scale and provide a historical context to land-cover and land-use change spatial patterns for past and ongoing field research in the area.
keywords: |
  Aerial survey, data recovery, CNN, Deep Learning, SfM, Congo Basin
acknowledgement: |
  This research was supported through the Belgian Science Policy office COBECORE project (BELSPO; grant BR/175/A3/COBECORE). KH acknowledges funding from the European Union Marie Skłodowska-Curie Action (project number 797668). I'm grateful for the support of Thales D'Haulleville volunteering his time in scanning the images.
authorcontributions: |
  K.H. conceived and designed the study, analyzed the data, prepared figures and tables, authored, reviewed and approved the final draft of the manuscript. T.d.H. scanned all image data. E.K. and T.d.H. provided plot based AGC estimates. T.D., K.J., E.K, H.B., P.S., F.V.,S.M., M.A., J.V.D.B., H.V and L.W. reviewed the final manuscript.
conflictsofinterest: |
  The authors declare no conflict of interest. The founding sponsors had no role in the design of the study; in the collection, analyses, or interpretation of data; in the writing of the manuscript, an in the decision to publish the results.
output: rticles::mdpi_article
---

```{r setup, include=FALSE, message=FALSE}
  knitr::opts_chunk$set(echo = TRUE)
  library(orthodrc)
  library(raster)
  library(ggplot2)
  library(ggthemes)
  library(tidyverse)
  library(knitr)
  library(kableExtra)
  library(landscapemetrics)

  # set kable NA values to blank
  options(knitr.kable.NA = '-')

  # landscape stats function
  landscape_stats <- function(r){

  df <- lsm_p_perim(r) %>%
    filter(class == 1) %>%
    select(-c(metric)) %>%
    rename(perimeter = value)

  df$perimeter <- df$perimeter

  area <- lsm_p_area(r) %>%
    filter(class == 1) %>%
    select(value) %>%
    rename(area = value)

  frac <- lsm_p_frac(r)  %>%
    filter(class == 1) %>%
    select(value) %>%
    rename(fractal_index = value)

  df <- data.frame(df, area = area, frac = frac)
  df$pa_ratio <- df$perimeter/(df$area*(100*100))

  df <- df %>%
    filter(area > 1) %>%
    summarize(
      perimeter = paste0(
        round(mean(perimeter)),
        " $\\pm$ ",
        round(sd(perimeter))),
      area = paste0(
        round(mean(area)),
        " $\\pm$ ",
        round(sd(area))),
      ratio = paste0(
        round(mean(pa_ratio),3),
        " $\\pm$ ",
        round(sd(pa_ratio),3)),
      fractal = paste0(
        round(mean(fractal_index),3),
        " $\\pm$ ",
        round(sd(fractal_index),3))
    )

  return(df)
}

  # read data
  change_map <- raster("../data/orthomosaic/yangambi_forest_cover_difference_1958_2000.tif")
  # 
  # read site data
  psp <- read.table("../data/surveys/site_characteristics.csv",
                    header = TRUE,
                    sep = ",")

  # patch metrics
  # select classes into their respective time frame
  historical <- (change_map == 2 | change_map == 4)
  contemporary <- (change_map == 3 | change_map == 4)

  # reproject to meter based mercator projection
  # should be fine around the equator
  historical <- projectRaster(historical,
                              crs = "+init=epsg:3395",
                              method = "ngb")

  contemporary <- projectRaster(contemporary,
                              crs = "+init=epsg:3395",
                              method = "ngb")

  patch_metrics <- rbind(landscape_stats(historical),
                 landscape_stats(contemporary))

  row.names(patch_metrics) <- c("historical","contemporary")

  # read change stats
  change_stats <- read.csv2("../data/surveys/lulcc_change_stats.csv",
                            header = TRUE,
                            sep = ",")
  change_stats[1:2,3] <- NA

  row.names(change_stats) <- c(
   "Forest",
   "  -  of which edges",
   "Regrowth > 1958 loss",
   "Loss > 2000",
   "Loss > 1958 (permanent)")

```

# Introduction

The functioning of a tropical forest, and its capacity to sequester carbon, is heavily influenced by anthropogenic disturbances such as selective logging, clear felling, or slash and burn practices, all increasing forest fragmentation. This increased pace in the loss of rainforest cover [@butsic2015] is predicted to negatively impact the function of these tropical forests, and their future capacity to sequester carbon, as anthropogenic disturbances such as selective logging, clear felling, or slash and burn practices collectively contribute to increasing forest fragmentation.

These forest fragments and their edges contain fewer large trees [@laurance2000], in part due to their hotter and brighter microclimates [@magnago2015]. Similarly, species composition and biodiversity [@poorter2006] are negatively affected. Spatial disturbances also have an explicit temporal component, where the time since disturbance affects the strength of the effects [@didham1999]. Deforestation and gap formation also influence the overall canopy structure, where pioneer species grow taller and with narrower crowns [@poorter2006; @peres2006] affecting the forest light regime and carbon dynamics [@stark2012]. As such, anthropogenic factors through land-use and land cover change (LULCC) heavily affect forest structure and functioning [@fauset2017; @brinck2017; @didham1999]. In the tropics the majority of emissions originate from deforestation and forest degradation [@hansen2013; @houghton2012; @tyukavina2015] contributing to 10-15% of the total global emissions [@vanderwerf2009] and concomitant losses in biodiversity losses [@barlow2016; @vandeperre2018] and how LULCC is evolving over time. Although the African rainforest is the second largest on Earth and covers ~630 million ha, representing up to 66 Pg of carbon storage [@Lewis2009b], and currently loses forest at an increasing pace [@butsic2015]. Yet, the tropics is now becoming a large source of CO_2 to the atmosphere as deforestation and forest degradation [@hansen2013; @houghton2012; @tyukavina2015] contribute up to 10-15% of the total global CO2 emissions [@vanderwerf2009]. Furthermore, recent estimates show that 31% of recent carbon emissions are caused by edge effects alone [@brinck2017]. In this context, there are very few long term forest inventory plots that presently exist in the central Congo Basin [@Verbeeck2011a; @lewis2013]. This further limits the accuracy of current biomass estimates and challenges our understanding of how forest structure and function changes in the short and long term in these ecosystems [@Kearsley2013; @Doetterl2015].

Historical estimates of LULCC, and their carbon emissions, exist [@houghton2012; @ramankutty1999; @defries2002; @sader1988], however they generally rely on non-spatially explicit data in the pre-satellite era (< 1972) [@houghton2010]. These analysis often use FAO statistics, historical land-use sources, and/or population and socio-economic data [@houghton2012], as limited earlier spatial data sources are available [@hansen2013; @achard2014]. For example, the @houghton2010 long-term study used FAO and national censuses alongside a bookkeeping model to estimate carbon emissions from LULCC. Other studies use remote sensing data, but are limited in time to recent decades [@defries2002; @houghton2012]. These proxy based studies have a limited view on long-term direct and indirect spatial effects of deforestation. As such, both the spatial structure and the fragmentation pattern of a forest combined with its life history has a profound effect on carbon emssisions. However, in this context remote sensing assessments provide more accurate spatial information to help estimate carbon emissions [@mitchard2018].

Past studies expanded knowledge during the pre-satellite time frame through the use of now declassified satellite surveillance data to map deforestation in the US, central Brazil, and post world war II Germany [@song2015; @nita2018]. In some cases these records could extend back in time even further using historical aerial survey images. In northern Europe especially these data are well known and used, for example to map the extent of forest cover across ~100 km^2^ regularly, and at times using automated workflows [@giordano2018]. Survey data across the African continent is less common or inaccessible or both. Some studies do exist, both @buitenwerf2012 and @okeke2006 used aerial survey images to map vegetation changes in South African savannas and the Adulam Nature Reserve, respectively. Whilst @frankl2015 used 20 aerial survey images to map the Ethiopian highlands of 1935. High resolution aerial or surveillance satellite images provide scientists tools to monitor forest extent and structure over long time frames. Moreover, canopy texture from aerial survey images has been successfully linked to above ground biomass [@Ploton2012; @Couteron2005; @Barbier2010]. Thus using texture based metrics obtained from (high resolution) remote sensing data opens up possibilities to link historical aerial footage to carbon density across large spatial extents [@Couteron2005].

Given the impact of LULCC on the structure and functioning of tropical forests, and their influence on both carbon dynamics [@Kearsley2013] and biodiversity [@vandeperre2018], accurate long term reporting of historical forest cover in the pre-satellite era is an imperative [@willcock2016]. In this study we use a combination of historical aerial photography and contemporary remote sensing data to map long-term changes in the extent and structure of the tropical forest surrounding Yangambi, in the central Congo Basin at the beginning of 1958 and the start of the antrophocene [@lewis2015]. Our analysis provides a historical insight into important LULCC spatial patterns in Yangambi, such as fragmentation and edge complexity. Our analysis further contextualises the influence of changes in the forest's life history on past and current research into Above Ground Carbon (AGC) storage [@Kearsley2013] and biodiversity [@vandeperre2018] in the central Congo Basin.

# Methods

## Historical data acquisition

Data for the central Congo Basin region, surrounding Kisangani, were collected in several flights during the dry season of 1958 and 1959 (from 8/01/1958 to 20/02/1958 and from 28/12/1958 to 9/01/1959 respectively, see Appendix Figure 1) to generate topographic maps of the area, supervised by the "Institut Géographique du Congo Belge" in Kinshasa (then Léopoldville). Black-and-White infrared images (0.4 - 0.9 $\mu$m) were gathered along flight paths running mostly from west to east, between 9 - 11h local time. Along a flight path continuous images were taken using a Wild Heerburgg RC5a (currently Leica Geosystems) with an Aviogon lens assembly (114.83mm / f 5.6, with a 90\degree view angle) resulting in square photo negative of 180 by 180 mm. Flights were flown at an average absolute altitude of ~5200 m above sea level, covering roughly 18 530 km^2^ at an approximate scale of 1/40 000. The use of the integrated autograph system ensured timely acquisition of pictures with a precise overlap (~1/3) between images. This large overlap between images together with flight parameters would allow post-processing, using stereographs, to create accurate topographic maps. Original data from this campaign are stored in the [Royal Museum for Central Africa](https://www.africamuseum.be/en) in Tervuren, Belgium (Figure 1).

```{r echo=FALSE, out.width='100%', fig.cap="A box of historical aerial photographs (left) and a single aerial photograph (right) showing part of the Congo river. Note the meta-data provided in the margins of the image such as acquisition time and fleight height and the solar glare on the Congo river."}
knitr::include_graphics('./figures/print_examples.jpg')
```

## Site selection

```{r echo=FALSE, out.width='100%', message=FALSE, warning=FALSE, fig.cap="Overview of the historical flight paths during aerial photo acquisition and ancillary data used in this study. The bounding box of the orthomosaic data presented in this study is shown as a rectangle (23x36 km). The outline of a recent high-resolution Geo-eye panchromatic image is shown as a dashed dark grey rectangle (10x10km). The location of various permanent sampling plots are shown as x, +, and open squares and triangles for the mixed, mono-dominant and edge plots respectively. The grey polygon delineates the current day Yangambi Man and Biosphere reserve. The inset, top right, situates the greater Yangabmi region with the DR Congo. The full flight plan and details are shown in Appendix Figure 1 and 2"}
knitr::include_graphics('./figures/flight_paths.png')
```

We prioritised flight paths and images that contained current day permanent sampling plots, larger protected areas, and past agricultural and forest research facilities (Figure 2). This selection provides a comprehensive mapping of the Yangambi area and the life history of the forest surrounding it. Thereafter, we selected flight paths 1 through 11 for digitization. From this larger dataset of 334 images we selected 74 survey images for orthomosaic compositing and further analysis. All the selected images stem from the flight campaign made during January and February of 1958. The area includes the Yangambi village, 20 contemporary permanent sampling plots [@Kearsley2013], past and present agricultural experimental plots [@bauters2015] and large sections of the Yangambi [UNESCO Man and Biosphere reserve](http://www.unesco.org/new/en/natural-sciences/environment/ecological-sciences/biosphere-reserves/africa/democratic-republic-of-the-congo/yangambi/) surrounding to the west and east of the village. Although not formally mosaicked we provided a full dataset of pre-processed images using the cropping and normalization routines described below. The latter data was not used in subsequent LULCC analysis, but has been archived and made available to the public separately (see code & data availability statement below).

## Digitization and data processing

All selected images, covering the Yangambi area, were contact prints as original negatives of the prints were not available. Images were scanned at a resolution exceeding their original resolution (or grain) at the maximal physical resolution of an Epson A3 flatbed scanner (i.e. 2400 dpi or 160MP per image) and saved as lossless tiff images. Data were normalized using contrast limited histogram equalization [@zuiderveld1994] with a window size of 32 and a clip limit of 1.5. Fiduciary marks were used to rectify and downsample the images into square 7700x7700 pixel images (~1200 dpi, 81MP). This resulted in a dataset with digital images at a resolution that remained above the visible grain of the photographs, whilst the reduced image size facilitated easier file handling and processing speed.

Data was processed into a georeferenced orthomosaic using a Structure from motion (SfM, @ullman1979) approach implemented in [Agisoft Metashape](http://www.agisoft.com) version 1.5.2 (Agisoft LLC, St. Petersburg, Russia). An orthomosaic corrects remote sensing data to represent a perfectly downward looking image, free from perspective distortions due to topography and camera tilt. Using the SfM technique features, areas in images with a large degree of similarity, are matched across various images to reconstruct a three dimensional scene (topography) from two-dimensional image sequences. During the SfM analysis we masked clouds, glare or large water bodies such as the Congo river.

We calculated the orthomosaic using a low resolution point cloud and digital elevation map (DEM). Additional ground control points were provided to assist in the referencing of image and constrain the optimization routine used in the SfM algorithm. Ground control points consisted of rooftop edges of permanent structures which could be verified in both old and new aerial imagery (i.e. ESRI World Imagery). Although clouds were removed during the SfM routine we did not mask all clouds in the final orthomosaic to maximize forest coverage. The final scene was cropped, to provide consistent wall to wall coverage of the reconstructed scene. The orthomosaic was exported as a geotiff for further georeferencing in QGIS [@qgisdevelopmentteam2018] using the georeferencer plugin (version 3.1.9) and additional ESRI World Imagery high resolution reference data. We used 3^th^ degree polynomial and 16 ground control points to correct the final image. Ground control points, raw image data and final processed image are provided in addition to measures of uncertainty such as mean root mean squared (RMSE), mean and median error across all ground control points. All subsequent analysis are executed on the final geo-referenced orthomosaic or subsets of it.

## Land-Use and Land-Cover Change

### Classifying forest cover

We automatically delineated all natural forest in the historical data, thus excluding tree plantations, thinned or deteriorated forest stands, fields and buildings. We used the Unet Convolutional Neural Net (CNN, @ronneberger2015) architecture implemented in Keras [@chollet2015] with an efficientnetb3 backbone [@yakubovskiy2019] running on TensorFlow [@martinabadi2015] to train a binary classifier (i.e. forest or non-forested). This methodology is increasingly being used to automate pixel-level classification in (color) digital photography data [@buscombe2018]. Training data were collected from the orthomosaic by randomly selecting 513 pixel square tiles from homogeneous forested or non-forested areas within the historical orthomosaic. Homogeneous tiles were combined in synthetic landscapes using a random gaussian field based binary mask (Figure 3). We generated 5000 synthetic landscapes for training, while 500 landscapes were generated for both the validation and the testing dataset. Source tiles did not repeat across datasets to limit overfitting. In order to limit stitch line misclassifications, along the seams of mosaicked images, we created synthetic landscapes with different forest tiles to mimick forest texture transitions. We applied this technique to 10% of the generated synthetic landscapes. The CNN model was trained for 100 epochs on a graphics processing unit (GPU) maximizing the Intersect-over-Union (IoU) using additional data augmentation. Data augmentation included random cropping to 320 pixel squares, random orientation, scaling, perspective, contrast and brightness shifts and image blurring. During final model evaluation we report the IoU of our out-of-sample test datasets.  The optimized model was used to classify the complete orthomosaic using a moving window approach with a step size of 110 pixels and a majority vote across overlapping areas to limit segmentation edge effects. We refer to Figure 4 for a synoptic overview of the full deep learning learning workflow.

More details on the optimizer and other metrics etc.

```{r echo=FALSE, out.width='100%', fig.cap="orthomosaic", fig.cap="An example synthetic landscape, combining homogeneous forest and non-forest images into a patchy landscape using random gaussian field based masks. The left panel shows a combined synthetic landscape, while the right panel shows the corresponding forest (black) and non-forest (white) labels."}
knitr::include_graphics('./figures/synthetic_landscape.png')
```

```{r echo=FALSE, out.width='100%', fig.cap="orthomosaic", fig.cap="A diagram of the deep learning workflow followed in training a binary forest / non-forest cover convolutional neural net U-Net model to generate our forest cover map."}
knitr::include_graphics('./figures/cnn_diagram.pdf')
```

To map long term land-use and land-cover change in the Yangambi region we used the contemporary Global Forest Change version 1.6 data (GFC, tile 10N-020E) [@hansen2013]. Using this data we calculated the latest state of the forest with respect to the conditions at the start of 1958, 60 years earlier. In our analysis we only included forested pixels which recorded no loss throughout the whole 2000 - 2018 period. As the resolution of the historical forest classification exceeds that of the GFC map we downsampled our historical forest cover data to 30m GFC resolution. We masked out all water bodies using the Global Forest Change survey data mask layer, and limited the analysis to the right bank of the Congo river. Where the datasets overlap we provide summary statistics of deforestation, reforestation and afforestation.

### Landscape fragmentation & Above Ground Carbon estimates

To quantify changes in the structure of forest cover and its disturbances we used spatial pattern analysis landscape fragmentation metrics [@dale1999]. We report the ratio of edge to area and the fractal index to quantify landscape complexity of forest disturbances. Statistics were calculated for all forest disturbance patches larger than 1 ha using the R package landscapemetrics [@hesselbarth2019]. We provide mean and standard deviation on edge, area, their ratio and fractal dimension for both the historical and contemporary @hansen2013 forest cover maps.

We estimated above ground carbon (AGC) losses and gains over time using plot based averages of recent inventory data at permanent sampling plots in the area (Figure 2). We refer to @Kearsley2013 for the survey method and allometric relations used to scale the survey data. Unlike standard square 1 ha plots edge plots were set back 200m from forest edges and were 50x200m, with the 50m side of the plot along the forest edge and continuing 200m into the forest (Appendix Table 1). We further confirmed that forest edge plots did not show a significantly different AGC compared to those of non-edged / mixed forest plots (Mann Whitney U test, p < 0.05). Thus it was not necessary to explicitly quantify changes in AGC caused by edge effects. Moreover, we used the mean values of the mixed forest as representative for potential AGC losses. Despite the challenges inherent in quantifying AGC for forest edges we mapped the total extent of the edges in the contemporary landscape. To align our landscape analysis with exploratory analysis of the survey data we used a buffer of 200m to estimate the extent of forest edges and patches, up to the location of forest edge plots.

Surveys of old plantations show a large variation in AGC, depending on age and the crop type. For example, the AGC values varied from 168.67 to 86.55 g Mg C ha^-1^, for *Hevea brasiliensis* (rubber tree) and *Elaeis guineensis* (oil palm) plots respectively (@bustillo2018, personal communications). These higher values are in line with the mixed AGC estimates in the area, while the palm plantations resemble old-regrowth values (81.8 Mg C ha^-1^, see Table A1). We therefore use both the estimates of old-regrowth and mixed forest to estimate AGC for regrowth. We did not have sufficient data to account for individual changes in AGC across plantations.

## Canopy structure & FOTO texture analysis

We compared the structure of the canopy both visually and using Fourier Transform Textural Ordination (FOTO, @couteron2002). Fourier Transform Textural Ordination uses a principal component analysis (PCA) on radially averaged 2D Fourier spectra to characterize canopy (image) texture.  The FOTO technique was first described by @couteron2002 to quantify canopy stucture in relation to biomass and biodiversity, and can be used across multiple scenes using normalization [@Barbier2010].

We used an area of 400x400m (16 ha) around contemporary permanent sampling plot locations to quantify the general state of the canopy in both historical and contemporary remote sensing data. Contemporary data were resampled to the resolution of the historical data using a nearest neighbour interpolation for comparison. We used the first acquisition of a pan-chromatic Geo-Eye 1 stereo pair (GeoEye, 2012, Thornton, Colorado, U.S.A., order 737537, 2011-11-11 8:55h GMT or 9:55h local time). For permanent sampling plot locations in both historical and contemporary data a rectangular buffer of 200m around the centre of all plots was used to create small image subsets. All image subsets were processed using the R based FOTO implementation and package [@hufkens2019a] using global normalization, ensuring intercomparability of texture metrics across time and space [@Barbier2010]. We used a FOTO (moving) window size of 187 pixels (or ~150m), as described in @solorzano2018 and @Barbier2010, to ensure that multiple individual canopies could be included in the analysis. A buffer with a radius of 50m around the center of all permanent sampling plots was used to determine the average value of the first principle component (PC), explaining the majority of the variance in canopy texture. Where both the Geo-Eye data and the orthomosaic image overlapped we processed the intersecting region to explore a wall-to-wall comparison between past and current canopy texture metrics.

We used the standard FOTO methodology with fixed zones, instead of the moving window approach. The window size was set to the same size (187 pixels or ~150m) as used in the moving window analysis above. To account for illumination differences between the two scenes we applied histogram matching. No global normalization was applied, as the scene was processed as a whole. PC values from this analysis for all permanent sampling plots in both image scences were extracted using a buffer with a radius of 50m.

For both site based and scene analysis we correlate PC values with permanent sample plot inventory data such as stem density, above ground biomass and tree species richness. Additional comparisons are made between contemporary Geo-Eye data and the historical orthomosaic derived PC values. Due to the few available permanent sampling plots in both scenes we used a non-parametric paired signed rank (Wilcoxon) test to determine differences between the PC values of the Geo-Eye and historical orthomosaic image scenes across mono-dominant and mixed forest types. In all analysis, mono-dominant site 4 was removed from the analysis due to cloud contamination.

# Results

Our analysis provides a first spatially explicity historical composite of aerial survey images mapping LULCC within the Congo Basin. The use of high resolution historical images combined with structure from motion image processing techniques allowed us to mosaic old imagery across a large extent. The final orthomosaic composition of the Yangambi region resulted in an image scene covering approximately 733 million pixels across ~828 km^2^ (~23x36 km, Figure 2). The overall accuracy of the structure from motion orthomosaic composition was 0.88 m/pixel using the sparse cloud DEM for corrections at 45.8 m/pixel. The resulting georeferenced scene had a spatial accuracy of approximately 23m. Further georeferencing outside the SfM workflow reduced the mean error at the ground control points to 5.3 $\pm$ 4.9 px (~4.7 $\pm$ 4.3 m), with a median error of 2.9 px (2.6m). The orthomosaic served as input for all subsequent LULCC analysis with all derived maps provided with the manuscript repository (see data & code availability statements below).

```{r echo=FALSE, out.width='100%', fig.cap="orthomosaic", fig.cap="Overview of the final orthomosaic of the greater Yangambi region (A), a detailed inset (B) and the derived land use land cover change map displayed as the difference between a manual classification of the orthomosaic and a recent map by Hansen et al. 2013 (C) and the corresponding land cover map for inset B (D)."}
knitr::include_graphics('./figures/orthomosaic_maps.png')
```

## Land-use and Land-Cover Change & Above Ground Carbon

Our CNN deep learning classifier reached an Intersection-over-Union (IoU) accuracy of 97% on the detection of disturbed forest in out-of-sample test data. Scaling our classifier to the whole historical orthomosaic we detected 162 km^2^ (or ~20% of the scene) of disturbed forests. A large fraction of the disturbed area was restored in the period between the two acquisitions. In total, 99 km^2^, or little over half of the affected forest was restored (Figure 5C-D, dark blue). Recent deforested areas, as registered through satellite remote sensing, approximate 88 km^2^ (Table1, Figure 5C-D / light green).

Recent deforestation follows a distincly different pattern compared to historical patterns. Historical deforestation showed a classical fishbone pattern for forest clearing with very sharp edges, while current patterns are patchy and ad-hoc (figure 5D, dark blue and green colours respectively). These differences are reflected in the analysis of landscape metrics of forest loss. Between the historical and contemporary LULCC maps we see an increase in small disturbances, as indicated by the decreasing area of the mean patch size, down to 37 ha from 120 ha historically, while perimeter lengths remained approximately the same, with 8653m and 9902m for historical and contemporary maps respectively (Table 2). This shift in perimeter area ratio led to a similar change in the fractal index, increasing in value to 1.13 $\pm$ 0.07 from 1.10 $\pm$ 0.05 over time. Values closer to a fractal index of 2 suggest a more complex (fragmented) landscape.

A comparison of forest edge plots with mixed forest plots showed no signficant difference in AGC, or other reported values such as species richness, basal area or stem density (Mann Whitney U test, p < 0.05). Edge influence did not extend beyond 200 m from a forest edge, but still represented an area of 132 km^2^.

Changes in both land-use and land-cover led to concomitant changes in above ground carbon stocks. Recovery throughout the region was characterized for patches of forest and plantations. Assuming high density stands, based on previous work, this could amount to a carbon gains of 1592 Gg C, offsetting more recent losses of approximately 1408 Gg C. On the other hand, at the low end, if we assume a lower carbon density of 81.8 Mg C ha^-1^ this would result in a total carbon gain of 811 Gg C. Using our approach results indicate that overall deforestation around Yangambi has resulted in a loss of ~2416 Gg C in AGC stocks.

```{r echo = FALSE, eval = TRUE}
change_stats %>%
  kable(caption="Land use land cover change statistics of forest cover around Yangambi in the central Congo Basin. The data evaluates a difference between a historical (1958) aerial photography based survey and the Hansen et al. 2013 based satellite remote sensing data. Spatial coverage statistics are provided in square kilometers (km) and hectares (ha), rounded to the nearest integer as well as Above Ground Carbon (AGC) scaled using recent survey measurements.",
       booktabs = TRUE,
       digits = 3,
       row.names = TRUE,
       escape = FALSE,
       col.names = c("$km\\textsuperscript{2}$",
                     "ha",
                     "(Gg C $ha\\textsuperscript{-1}$")
         ) %>%
    add_header_above(c(" " = 3,"AGC" = 1)) %>%
    kable_styling(latex_options = c("hold_position"))
```

```{r echo = FALSE, eval = TRUE}
patch_metrics %>%
  kable(caption="Landscape metrics for historical and contemporary deforestation patterns. We report patch perimeter and area, their ratio and fractal dimension. Values are reportedpat as mean $\\pm$ standard deviation, across all deforestation patches.",
       booktabs = TRUE,
       row.names = TRUE,
       escape = FALSE,
              col.names = c("perimeter (m)",
                     "area (ha)",
                     "ratio ($m\\textsuperscript{-1}$)",
                     "fractal dimension")
         ) %>%
    kable_styling(latex_options = c("hold_position"))
```

## Canopy structure & FOTO texture analysis

Visual interpretation of the scenes provide evidence that most locations do not change dramatically with respect to canopy composition, except for the large areas of disturbances in contemporary fallow or young-regrowth plots. One marked difference is noted in the mono-dominant site 6 (Appendix Table 1). Here, the current mono-dominant *Brachystegia laurenttii* is a recent development, changing the canopy structure visibly during the last half century (Figure 6). The previous varied canopy structure gave way to a more dense and uniform canopy. This is reflected in a change of the FOTO PC value from 0.19 historically to its current value of 0.54. This historical value is similar to the mean of contemporary mono-dominant stands with PC averaging 0.34 $\pm$ 0.1, and is only slightly higher than historical values for a mixed forest (0.18 $\pm$ 0.08). The reverse pattern is seen in the contemporary PC values. Here, the value of 0.54 exceeds those of most mono-dominant stands (0.35 $\pm$ 0.08), and is even further removed from the values noted for mixed forests (0.12 $\pm$ 0.03).

```{r echo=FALSE, out.width='100%', fig.cap="Visual comparison between a historical (A) and contemporary (B) permanent sampling plot. The site is currently listed as a mono-dominant Brachystegia laurenttii stand. Note the structural differences with a 'courser' canopy structure in the historical image compared to the more closed contemporary stand."}
knitr::include_graphics('./figures/visual_comparison_psp.png')
```

Using only small subsets around existing permanent sampling plots we show distinct differences between forest types, with PC values in both historical and contemporary imagery markedly higher for the mono-dominant forest types compared to all others (Appendix Figure 3). Provided that the young-regrowth and fallow permanent sampling plots have seen recent disturbance the Wilcoxon signed rank test on the mixed and mono-dominant plots between the historical and contemporary PC values did not show a significant difference (p > 0.05). Similarly, no significant different using PC values extracted from the whole scene analysis was noted (p > 0.05). Any relationships between contemporary Geo-Eye data and permanent sampling plot measurements of Above Ground Biomass, stem density and species richness were non-significant (p > 0.05, Appendix Figures 4-6).

Furthermore, visual inspection of the scene wide analysis suggests historical scences do not show landscape wide canopy features (Figure 7 A-B), unlike the contemporary scene (Figure 7 C-D). Where the FOTO algorithm picks up landscape features such as changes in texture across the contemporary Geo-Eye scene (e.g. the river valley as a diagonal line in Figure 7D), however, no corresponding landscape patterns are found by the FOTO algorithm in the historical orthomosaic.

```{r echo=FALSE, out.width='100%', fig.cap="RGB visualizations of the first three principal components of scene wide FOTO texture analysis of historical and current (Geo-Eye) imagery. Current permanent sampling plots of mono-dominant, mixed, fallow and young (edge) forest plots are marked with open triangles, open circles, open squares and crosses, respectively."}
knitr::include_graphics('./figures/foto_maps.png')
```

# Discussion

Finely grained spatial data sources, such as remote sensing imagery, are rare before the satellite era (<1970). This lack of data limits our understanding of how forest structure has varied over longer time periods in especially remote areas. Long term assessment can be extended by using large inventories of historical aerial survey data [@kadmon1999; @song2015; @nita2018]. However, across the Congo Basin this data is rare, or difficult to access. Despite the difficulties in assessing hidden disturbances caused by deforestation, poorly recorded official assessments or simply those invisible to standard remote sensing techniques [@peres2006], remote sensing generally remains the best way to map and quantify LULCC [@houghton2012]. Despite these challenges in recovery and validation we present a first analysis of LULCC using historical aerial survey data for the central Congo Basin. We used a combination of Structure-from-Motion and a convolutional neural net deep learning approach to map deforestation across a large extent, providing a long term view of LULCC in the central Congo Basin

## Data recovery challenges

In our study the archive data recovered was limited to contact prints and therefore did not represent the true resolution of the original negative. In addition, analogue photography clearly produces a distinct softness compared to digital imagery (Figure 4). Despite favourable nadir image acquistions [@verhoeven2017] image softness combined with illumination effects between flight paths, and the self-similar nature of vast canopy expanses [@park2019; @simini2010; @sole1995], limited our ability to provide wall-to-wall coverage of the entire dataset containing 334 images. Few man made features in the scenes also made georeferencing challenging. Although the village of Yangambi provided a range of buildings as (hard-edge) references, other areas within the central Congo Basin might have fewer permanent structures and would require the use of soft-edged landscape features (e.g. trees, river outflows). Research has shown that soft-edged features can help georeference scenes even when containing few man-made features [@hughes2006], however accuracy becomes compromised. Our two step georeferencing approach resulted in a referencing accuracy of ~4.7 $\pm$ 4.3 m across reference points. However, it shoud be noted that referencing accuracy of the final scene is less constrained toward the edges of the scene.

## Land-use, Land-Cover change & Above Ground Carbon

When classifying our orthomosaic into forest and non-forest states we favoured a deep learning supervised classification using a Convolutional Neural Network over manual segmentation to guarantee an "apples-to-apples" comparison between the historical and the contemporary GFC forest cover maps. We acknowledge that both classifications use different features, i.e. spectral or spatial data, but attain a similarly high accuracy of 97% and 99% [@hansen2013], for the GFC and our CNN map respectively. Despite a number of challenges we were able to map and quantify a vast continuous area of tropical forest, with limited manual annotation requirements.

Our analysis shows that the majority of deforestation around Yangambi happened toward the late 1950's (162 km^2^). Considerable regrowth has occured since the aerial survey was executed (99 km^2^), and socio-economic instability prevented further large scale forest exploitation. In particular, many plantations have reached maturity and forest has re-established in previously cleared or disturbed areas. The majority of this forest recovery takes the form of isolated patches of forest but is offset by further deforestation of previously untouched forest. Generally, the function and structure of forests can be influenced by forest edges that are located up to 1km away however most effects are pronounced within the first 300m from the edge [@gascon2000]. Our analysis of edge effects on AGC has shown that the influence is negligible 200m away from the edge. @phillips2006 have shown similar weak responses to edge effects in the Amazon forest. Due to a lack of data on the extent (depth) of edge effects and their influence on AGC beyond 200m we did not include any estimates of carbon loss or gain within these zones. However, it must be stated that the influence of edges throughout the landscape was not marginal. as these areas would account for 132 km^2^. Thus edges could have a substantial negative [@brinck2017] or positive [@reinmann2017] influence on AGC. Similarly, uncertainties in how to explicitly correct for plantations in the landscape present a further challenge. Thus although our estimates are only indicative they do underscore the important influence of landscape structure when carbon accounting. However, our findings do not indicate that deforestation in Congo basin is declining, on the contrary.

Over the past half century there has been a clear shift in land use in Yangambi (Figure 5). Land use has shifted away from for example a regular fishbone deforestation pattern that emerges when (large scale) agricultural interests dominate the landscape [@arima2015], to a more fragmented landscape (Figure 5D). The latter former? is consistent with historical land management at the time of the aerial survey [@bustillo2018]. These regular patterns have since been reversed because of a decrease in large scale intensive agriculture, replaced by ad-hoc small scale subsistence farming with large perimeter to area relationships (i.e. ragged edges). Consequently, edge effects in the current landscape are far more pronounced than in the historical scene.

Visual inspection of the images also suggests that reforestation within the historically cleared areas and experimental plots is not necessarily limited to areas far removed from more densely populated areas. For example, large reforested areas exist close to the Congo stream and Yangambi village itself (Fig. 5). Here, regional political components, such as land leases and large scale ownership could have played a role in safeguarding some of these areas for rewilding or sustainable management [@arima2014; @larson2011]. Despite widespread anthropogenic influences throughout the tropics [@lewis2015] the retention of these forested areas show the potential of explicit or implicit protective policy measures (e.g. INERA concessions, @bustillo2018) on a multi-decadal time scale. Forest regrowth in non-continuous areas within Yangambi could increase landscape connectivity and help increase biodiversity [@vandeperre2018].

Our analysis therefore provides an opportunity to highlight and study those regions that have previously suffered confirmed long-term disturbances, and those that have been restored since. Assessments of old plantations and recovering clear-cut forests can serve as a guide to help estimate carbon storage capacity and forest recovery rates in managed and unmanaged conditions [@gourlet-fleury2013; @sader1988; @achard2014] over the mid- to long-term, in support of rewilding and general forest restoration [@arima2014; @larson2011; @vandeperre2018]. In addition, mapping long-term edge effects can further support research into issues such as receding forest edges [@gascon2000].

## Canopy structure & FOTO texture analysis

Finally the FOTO technique used to quantify relationships between canopy structure and forest characteristics rendered no valuable insights of either the historical orthomosaic or recent Geo-Eye scene. Similarly weak correlations were found previously by @solorzano2018. In contrast site based texture metric statistics did show correspondence between historical and contemporary satellite imagery. However, none of them were either consistent or significant. Although visual interpretation shows distinctly different canopy structures (Figure 5) the differences in how resolution is defined and issues related to image quality prevented us from quantifying these further. Unlike large scale studies by @Ploton2012 we could not scale this technique to historical data. We advise that future valorisation efforts should preferentially work from foto negatives (if available) to ensure optimal data quality in resolution, contrast and overall sharpness.

# Conclusion

Given the impact of tropical forest disturbances on atmospheric CO_2 emissions, biodiversity and ecosystem productivity accurate long term reporting on land-cover and land-use change especially in the pre-satellite era is an imperative. Our analysis of historical aerial survey images of the Central Congo Basin provides a window into the state of the forest at the start of the anthropocene. Efforts to quantify canopy texture and their link to carbon storage had limited to no success.We have shown that historical aerial survey data can be used to quantify land-use and land-cover change using a highly automated workflow to quantify the structure and extent of forest cover change that can help assess the impact of fragmentation on above ground carbon stocks. We also show a shift from previously highly structured industrial deforestation of large areas, to discrete smallholder clearing for farming, increasing landscape fragmentation but also opportuntiy for substantial regrowth. Efforts to quantify canopy texture and their link to carbon storage had limited to no success. As such, our analysis provides insights into the state of rarely studied tropical forests and the rate at which deforestation and reforestation has taken place over a multi-decadal scale in the central Congo basin providing a useful historical context to land-cover and land-use change spatial patterns for interpreting past and ongoing field research in the area.

# Additional Information and Declarations
## Data availability

Hufkens et al. (2019). A curated dataset of aerial survey images over the central Congo Basin, 1958. Zenodo: doi.org/10.5281/zenodo.3547767. All data not included in the latter repository can be found bundled with the analysis code as listed below. Proprietary datasets (i.e. Geo-Eye data) are not shared, but purchase order numbers allow for acquisition of these datasets to ensure reproducibility. The @hansen2013 image data is freely available from [the GFC project website](https://earthenginepartners.appspot.com/science-2013-global-forest/download_v1.6.html).

## Code availability

All analysis code is available as an R / python [@rcoreteam2019] projects (https://khufkens.github.io/orthodrc & https://khufkens.github.io/orthodrc_cnn/). The analysis relied heavily on the 'raster' [@hijmans2019], 'RStoolbox' [@leutner2019], and 'landscapemetrics' [@hesselbarth2019] packages, while post-processing and plotting was facilitated by the 'tidyverse' ecosystem [@wickham2017], 'ggthemes' [@arnold2019], 'scales' [@wickham2018] and 'cowplot' [@wilke2019]. Additional plotting elements were formatted or provided by 'sf' [@pebesma2018] and 'rnaturalearth' [@south2017] packages, respectively. I'm grateful for the contributions to the scientific community by the developers of these packages.

### 2019-12-27 manuscript/main.tex
first commit

%  LaTeX support: latex@mdpi.com
%  In case you need support, please attach all files that are necessary for compiling as well as the log file, and specify the details of your LaTeX setup (which operating system and LaTeX version / tools you are using).

% You need to save the "mdpi.cls" and "mdpi.bst" files into the same folder as this template file.

%=================================================================
\documentclass[remote sensing,article,submit,moreauthors,pdftex,10pt,a4paper]{mdpi}

%% Some pieces required from the pandoc template
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{4pt}}
\setlist[itemize]{leftmargin=*,labelsep=5.8mm}
\setlist[enumerate]{leftmargin=*,labelsep=4.9mm}

\usepackage{longtable}

% see https://stackoverflow.com/a/47122900

%
%--------------------
% Class Options:
%--------------------
% journal
%----------
% Choose between the following MDPI journals:
% actuators, addictions, admsci, aerospace, agriculture, agronomy, algorithms, animals, antibiotics, antibodies, antioxidants, applsci, arts, asi, atmosphere, atoms, axioms, batteries, bdcc, behavsci, beverages, bioengineering, biology, biomedicines, biomimetics, biomolecules, biosensors, brainsci, buildings, carbon, cancers, catalysts, cells, ceramics, challenges, chemengineering, chemosensors, children, chromatography, climate, coatings, colloids, computation, computers, condensedmatter, cosmetics, cryptography, crystals, cybersecurity, data, dentistry, designs, diagnostics, diseases, diversity, econometrics, economies, education, electrochemistry, electronics, energies, entropy, environments, epigenomes, est, fermentation, fibers, fire, fishes, fluids, foods, forests, fractalfract, futureinternet, galaxies, games, gastrointestdisord, gels, genealogy, genes, geosciences, geriatrics, hazardousmatters, healthcare, heritage, highthroughput, horticulturae, humanities, hydrology, informatics, information, infrastructures, inorganics, insects, instruments, ijerph, ijfs, ijms, ijgi, ijtpp, inventions, j, jcdd, jcm, jcs, jdb, jfb, jfmk, jimaging, jof, jintelligence, jlpea, jmmp, jmse, jpm, jrfm, jsan, land, languages, laws, life, literature, logistics, lubricants, machines, magnetochemistry, make, marinedrugs, materials, mathematics, mca, mti, medsci, medicines, membranes, metabolites, metals, microarrays, micromachines, microorganisms, minerals, modelling, molbank, molecules, mps, nanomaterials, ncrna, neonatalscreening, neuroglia, nitrogen, nutrients, ohbm, particles, pathogens, pharmaceuticals, pharmaceutics, pharmacy, philosophies, photonics, plants, plasma, polymers, polysaccharides, preprints, proceedings, processes, proteomes, publications, quaternary, qubs, recycling, religions, remotesensing, reports, resources, risks, robotics, safety, scipharm, sensors, separations, sexes, sinusitis, socsci, societies, soils, sports, standards, stats, surgeries, sustainability, symmetry, systems, technologies, toxics, toxins, tropicalmed, universe, urbansci, vaccines, vetsci, vibration, viruses, vision, water, wem
%---------
% article
%---------
% The default type of manuscript is article, but can be replaced by:
% abstract, addendum, article, benchmark, book, bookreview, briefreport, casereport, changes, comment, commentary, communication, conceptpaper, correction, conferenceproceedings, conferencereport, expressionofconcern, meetingreport, creative, datadescriptor, discussion, editorial, essay, erratum, hypothesis, interestingimages, letter, meetingreport, newbookreceived, opinion, obituary, projectreport, reply, reprint, retraction, review, perspective, protocol, shortnote, supfile, technicalnote, viewpoint
% supfile = supplementary materials
% protocol: If you are preparing a "Protocol" paper, please refer to http://www.mdpi.com/journal/mps/instructions for details on its expected structure and content.
%----------
% submit
%----------
% The class option "submit" will be changed to "accept" by the Editorial Office when the paper is accepted. This will only make changes to the frontpage (e.g. the logo of the journal will get visible), the headings, and the copyright information. Also, line numbering will be removed. Journal info and pagination for accepted papers will also be assigned by the Editorial Office.
%------------------
% moreauthors
%------------------
% If there is only one author the class option oneauthor should be used. Otherwise use the class option moreauthors.
%---------
% pdftex
%---------
% The option pdftex is for use with pdfLaTeX. If eps figures are used, remove the option pdftex and use LaTeX and dvi2pdf.

%=================================================================
\firstpage{1}
\makeatletter
\setcounter{page}{\@firstpage}
\makeatother
\articlenumber{x}
\doinum{10.3390/------}
\pubvolume{xx}
\pubyear{2018}
\copyrightyear{2018}
\externaleditor{Academic Editor: name}
\history{Received: date; Accepted: date; Published: date}

%------------------------------------------------------------------
% The following line should be uncommented if the LaTeX file is uploaded to arXiv.org
%\pdfoutput=1

%=================================================================
% Add packages and commands here. The following packages are loaded in our class file: fontenc, calc, indentfirst, fancyhdr, graphicx, lastpage, ifthen, lineno, float, amsmath, setspace, enumitem, mathpazo, booktabs, titlesec, etoolbox, amsthm, hyphenat, natbib, hyperref, footmisc, geometry, caption, url, mdframed, tabto, soul, multirow, microtype, tikz

%=================================================================
%% Please use the following mathematics environments: Theorem, Lemma, Corollary, Proposition, Characterization, Property, Problem, Example, ExamplesandDefinitions, Hypothesis, Remark, Definition
%% For proofs, please use the proof environment (the amsthm package is loaded by the MDPI class).

%=================================================================
% Full title of the paper (Capitalized)
\Title{Historical aerial surveys map long-term changes of forest cover and
structure in the central Congo Basin}

% Authors, for the paper (add full first names)
\Author{Koen Hufkens$^{1,2,*}$\href{https://orcid.org/0000-0002-5070-8109}{\orcidicon}, Thalès de Haulleville$^{3}$, Elizabeth Kearsley$^{1}$, Kim Jacobsen$^{2,3}$, Hans Beeckman$^{3}$, Piet Stoffelen$^{4}$, Filip Vandelook$^{4}$, Sofie Meeus$^{4}$, Michael Amara$^{5}$, Leen Van Hirtum$^{5}$, Jan Van den Bulcke$^{1}$, Hans Verbeeck$^{1}$, Lisa Wingate$^{2}$}

% Authors, for metadata in PDF
\AuthorNames{Koen Hufkens, Thalès de Haulleville, Elizabeth Kearsley, Kim Jacobsen, Hans Beeckman, Piet Stoffelen, Filip Vandelook, Sofie Meeus, Michael Amara, Leen Van Hirtum, Jan Van den Bulcke, Hans Verbeeck, Lisa Wingate}

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad Ghent University, Ghent, Belgium; \\
$^{2}$ \quad INRA, UMR ISPA, Villenave d'Ornon, France; \\
$^{3}$ \quad Royal Museum for Central Africa, Tervuren, Belgium; \\
$^{4}$ \quad Botanic Garden Meise, Meise, Belgium; \\
$^{5}$ \quad National Archives of Belgium, Brussels, Belgium; \\
}
% Contact information of the corresponding author
\corres{Correspondence: \href{mailto:koen.hufkens@gmail.com}{\nolinkurl{koen.hufkens@gmail.com}}}

% Current address and/or shared authorship

% The commands \thirdnote{} till \eighthnote{} are available for further notes

% Simple summary

% Abstract (Do not insert blank lines, i.e. \\)
\abstract{Land Use and Land Cover change (LULCC) of African rainforest contribute
to global carbon emissions. Yet, most historical estimates of LULCC and
their carbon emissions rely on non-spatially explicity data in the
pre-satellite era (\textless{} 1972). Past studies have expanded this
pre-satellite time frame with declassified satellite surveillance data
or aerial surveys, but none cover the Congo Basin. Here, we use
historical aerial survey photos to map the extent and structure of LULCC
around Yangambi, central Congo Basin, in 1958. Our study leveraged
Structure-from-Motion to generate a large orthomosaic covering 828
km\textsuperscript{2}, counting 733 million pixels, at a sub meter
resolution and geo-referenced to \textasciitilde{}4.7 \(\pm\) 4.3 m.
Primary forest in the orthomosaic was classified with a Convoluted
Neural Network Deep Learning approach. We used these data to quantify
LULCC, landscape and canopy complexity and scale above ground biomass
between historical and contemporary condition. We show a shift from
historical highly structured industrial deforestation in 1958 (162
km\textsuperscript{2}), to contemporary smallholder farming clearing (88
km\textsuperscript{2}), increasing landscape fragmentation and forest
edges. Efforts to quantify canopy texture and their link to carbon
storage had limited to no success. Our analysis provides an insight in
key spatial and temporal patterns of deforestation and reforestation at
a multi-decadal scale and provide a historical context to land-cover and
land-use change spatial patterns for past and ongoing field research in
the area.}

% Keywords
\keyword{Aerial survey, data recovery, CNN, Deep Learning, SfM, Congo Basin}

% The fields PACS, MSC, and JEL may be left empty or commented out if not applicable
%\PACS{J0101}
%\MSC{}
%\JEL{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Applied Sciences:
%\featuredapplication{Authors are encouraged to provide a concise description of the specific application or a potential application of the work. This section is not mandatory.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Data:
%\dataset{DOI number or link to the deposited data set in cases where the data set is published or set to be published separately. If the data set is submitted and will be published as a supplement to this paper in the journal Data, this field will be filled by the editors of the journal. In this case, please make sure to submit the data set as a supplement when entering your manuscript into our manuscript editorial system.}

%\datasetlicense{license under which the data set is made available (CC0, CC-BY, CC-BY-SA, CC-BY-NC, etc.)}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Toxins
%\keycontribution{The breakthroughs or highlights of the manuscript. Authors can write one or two sentences to describe the most important part of the paper.}

%\setcounter{secnumdepth}{4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage{wrapfig} \usepackage{graphics} \usepackage{setspace}

\doublespacing

\usepackage{gensymb} \usepackage{booktabs}

\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Only for the journal Gels: Please place the Experimental Section after the Conclusions

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

The functioning of a tropical forest, and its capacity to sequester
carbon, is heavily influenced by anthropogenic disturbances such as
selective logging, clear felling, or slash and burn practices, all
increasing forest fragmentation. This increased pace in the loss of
rainforest cover \citep{butsic2015} is predicted to negatively impact
the function of these tropical forests, and their future capacity to
sequester carbon, as anthropogenic disturbances such as selective
logging, clear felling, or slash and burn practices collectively
contribute to increasing forest fragmentation.

These forest fragments and their edges contain fewer large trees
\citep{laurance2000}, in part due to their hotter and brighter
microclimates \citep{magnago2015}. Similarly, species composition and
biodiversity \citep{poorter2006} are negatively affected. Spatial
disturbances also have an explicit temporal component, where the time
since disturbance affects the strength of the effects
\citep{didham1999}. Deforestation and gap formation also influence the
overall canopy structure, where pioneer species grow taller and with
narrower crowns \citep{poorter2006, peres2006} affecting the forest
light regime and carbon dynamics \citep{stark2012}. As such,
anthropogenic factors through land-use and land cover change (LULCC)
heavily affect forest structure and functioning
\citep{fauset2017, brinck2017, didham1999}. In the tropics the majority
of emissions originate from deforestation and forest degradation
\citep{hansen2013, houghton2012, tyukavina2015} contributing to 10-15\%
of the total global emissions \citep{vanderwerf2009} and concomitant
losses in biodiversity losses \citep{barlow2016, vandeperre2018} and how
LULCC is evolving over time. Although the African rainforest is the
second largest on Earth and covers \textasciitilde{}630 million ha,
representing up to 66 Pg of carbon storage \citep{Lewis2009b}, and
currently loses forest at an increasing pace \citep{butsic2015}. Yet,
the tropics is now becoming a large source of CO\_2 to the atmosphere as
deforestation and forest degradation
\citep{hansen2013, houghton2012, tyukavina2015} contribute up to 10-15\%
of the total global CO2 emissions \citep{vanderwerf2009}. Furthermore,
recent estimates show that 31\% of recent carbon emissions are caused by
edge effects alone \citep{brinck2017}. In this context, there are very
few long term forest inventory plots that presently exist in the central
Congo Basin \citep{Verbeeck2011a, lewis2013}. This further limits the
accuracy of current biomass estimates and challenges our understanding
of how forest structure and function changes in the short and long term
in these ecosystems \citep{Kearsley2013, Doetterl2015}.

Historical estimates of LULCC, and their carbon emissions, exist
\citep{houghton2012, ramankutty1999, defries2002, sader1988}, however
they generally rely on non-spatially explicit data in the pre-satellite
era (\textless{} 1972) \citep{houghton2010}. These analysis often use
FAO statistics, historical land-use sources, and/or population and
socio-economic data \citep{houghton2012}, as limited earlier spatial
data sources are available \citep{hansen2013, achard2014}. For example,
the \citet{houghton2010} long-term study used FAO and national censuses
alongside a bookkeeping model to estimate carbon emissions from LULCC.
Other studies use remote sensing data, but are limited in time to recent
decades \citep{defries2002, houghton2012}. These proxy based studies
have a limited view on long-term direct and indirect spatial effects of
deforestation. As such, both the spatial structure and the fragmentation
pattern of a forest combined with its life history has a profound effect
on carbon emssisions. However, in this context remote sensing
assessments provide more accurate spatial information to help estimate
carbon emissions \citep{mitchard2018}.

Past studies expanded knowledge during the pre-satellite time frame
through the use of now declassified satellite surveillance data to map
deforestation in the US, central Brazil, and post world war II Germany
\citep{song2015, nita2018}. In some cases these records could extend
back in time even further using historical aerial survey images. In
northern Europe especially these data are well known and used, for
example to map the extent of forest cover across \textasciitilde{}100
km\textsuperscript{2} regularly, and at times using automated workflows
\citep{giordano2018}. Survey data across the African continent is less
common or inaccessible or both. Some studies do exist, both
\citet{buitenwerf2012} and \citet{okeke2006} used aerial survey images
to map vegetation changes in South African savannas and the Adulam
Nature Reserve, respectively. Whilst \citet{frankl2015} used 20 aerial
survey images to map the Ethiopian highlands of 1935. High resolution
aerial or surveillance satellite images provide scientists tools to
monitor forest extent and structure over long time frames. Moreover,
canopy texture from aerial survey images has been successfully linked to
above ground biomass \citep{Ploton2012, Couteron2005, Barbier2010}. Thus
using texture based metrics obtained from (high resolution) remote
sensing data opens up possibilities to link historical aerial footage to
carbon density across large spatial extents \citep{Couteron2005}.

Given the impact of LULCC on the structure and functioning of tropical
forests, and their influence on both carbon dynamics
\citep{Kearsley2013} and biodiversity \citep{vandeperre2018}, accurate
long term reporting of historical forest cover in the pre-satellite era
is an imperative \citep{willcock2016}. In this study we use a
combination of historical aerial photography and contemporary remote
sensing data to map long-term changes in the extent and structure of the
tropical forest surrounding Yangambi, in the central Congo Basin at the
beginning of 1958 and the start of the antrophocene \citep{lewis2015}.
Our analysis provides a historical insight into important LULCC spatial
patterns in Yangambi, such as fragmentation and edge complexity. Our
analysis further contextualises the influence of changes in the forest's
life history on past and current research into Above Ground Carbon (AGC)
storage \citep{Kearsley2013} and biodiversity \citep{vandeperre2018} in
the central Congo Basin.

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{historical-data-acquisition}{%
\subsection{Historical data
acquisition}\label{historical-data-acquisition}}

Data for the central Congo Basin region, surrounding Kisangani, were
collected in several flights during the dry season of 1958 and 1959
(from 8/01/1958 to 20/02/1958 and from 28/12/1958 to 9/01/1959
respectively, see Appendix Figure 1) to generate topographic maps of the
area, supervised by the ``Institut Géographique du Congo Belge'' in
Kinshasa (then Léopoldville). Black-and-White infrared images (0.4 - 0.9
\(\mu\)m) were gathered along flight paths running mostly from west to
east, between 9 - 11h local time. Along a flight path continuous images
were taken using a Wild Heerburgg RC5a (currently Leica Geosystems) with
an Aviogon lens assembly (114.83mm / f 5.6, with a 90\degree view angle)
resulting in square photo negative of 180 by 180 mm. Flights were flown
at an average absolute altitude of \textasciitilde{}5200 m above sea
level, covering roughly 18 530 km\textsuperscript{2} at an approximate
scale of 1/40 000. The use of the integrated autograph system ensured
timely acquisition of pictures with a precise overlap
(\textasciitilde{}1/3) between images. This large overlap between images
together with flight parameters would allow post-processing, using
stereographs, to create accurate topographic maps. Original data from
this campaign are stored in the
\href{https://www.africamuseum.be/en}{Royal Museum for Central Africa}
in Tervuren, Belgium (Figure 1).

\begin{figure}
\includegraphics[width=1\linewidth]{./figures/print_examples} \caption{A box of historical aerial photographs (left) and a single aerial photograph (right) showing part of the Congo river. Note the meta-data provided in the margins of the image such as acquisition time and fleight height and the solar glare on the Congo river.}\label{fig:unnamed-chunk-1}
\end{figure}

\hypertarget{site-selection}{%
\subsection{Site selection}\label{site-selection}}

\begin{figure}
\includegraphics[width=1\linewidth]{./figures/flight_paths} \caption{Overview of the historical flight paths during aerial photo acquisition and ancillary data used in this study. The bounding box of the orthomosaic data presented in this study is shown as a rectangle (23x36 km). The outline of a recent high-resolution Geo-eye panchromatic image is shown as a dashed dark grey rectangle (10x10km). The location of various permanent sampling plots are shown as x, +, and open squares and triangles for the mixed, mono-dominant and edge plots respectively. The grey polygon delineates the current day Yangambi Man and Biosphere reserve. The inset, top right, situates the greater Yangabmi region with the DR Congo. The full flight plan and details are shown in Appendix Figure 1 and 2}\label{fig:unnamed-chunk-2}
\end{figure}

We prioritised flight paths and images that contained current day
permanent sampling plots, larger protected areas, and past agricultural
and forest research facilities (Figure 2). This selection provides a
comprehensive mapping of the Yangambi area and the life history of the
forest surrounding it. Thereafter, we selected flight paths 1 through 11
for digitization. From this larger dataset of 334 images we selected 74
survey images for orthomosaic compositing and further analysis. All the
selected images stem from the flight campaign made during January and
February of 1958. The area includes the Yangambi village, 20
contemporary permanent sampling plots \citep{Kearsley2013}, past and
present agricultural experimental plots \citep{bauters2015} and large
sections of the Yangambi
\href{http://www.unesco.org/new/en/natural-sciences/environment/ecological-sciences/biosphere-reserves/africa/democratic-republic-of-the-congo/yangambi/}{UNESCO
Man and Biosphere reserve} surrounding to the west and east of the
village. Although not formally mosaicked we provided a full dataset of
pre-processed images using the cropping and normalization routines
described below. The latter data was not used in subsequent LULCC
analysis, but has been archived and made available to the public
separately (see code \& data availability statement below).

\hypertarget{digitization-and-data-processing}{%
\subsection{Digitization and data
processing}\label{digitization-and-data-processing}}

All selected images, covering the Yangambi area, were contact prints as
original negatives of the prints were not available. Images were scanned
at a resolution exceeding their original resolution (or grain) at the
maximal physical resolution of an Epson A3 flatbed scanner (i.e.~2400
dpi or 160MP per image) and saved as lossless tiff images. Data were
normalized using contrast limited histogram equalization
\citep{zuiderveld1994} with a window size of 32 and a clip limit of 1.5.
Fiduciary marks were used to rectify and downsample the images into
square 7700x7700 pixel images (\textasciitilde{}1200 dpi, 81MP). This
resulted in a dataset with digital images at a resolution that remained
above the visible grain of the photographs, whilst the reduced image
size facilitated easier file handling and processing speed.

Data was processed into a georeferenced orthomosaic using a Structure
from motion (SfM, \citet{ullman1979}) approach implemented in
\href{http://www.agisoft.com}{Agisoft Metashape} version 1.5.2 (Agisoft
LLC, St.~Petersburg, Russia). An orthomosaic corrects remote sensing
data to represent a perfectly downward looking image, free from
perspective distortions due to topography and camera tilt. Using the SfM
technique features, areas in images with a large degree of similarity,
are matched across various images to reconstruct a three dimensional
scene (topography) from two-dimensional image sequences. During the SfM
analysis we masked clouds, glare or large water bodies such as the Congo
river.

We calculated the orthomosaic using a low resolution point cloud and
digital elevation map (DEM). Additional ground control points were
provided to assist in the referencing of image and constrain the
optimization routine used in the SfM algorithm. Ground control points
consisted of rooftop edges of permanent structures which could be
verified in both old and new aerial imagery (i.e.~ESRI World Imagery).
Although clouds were removed during the SfM routine we did not mask all
clouds in the final orthomosaic to maximize forest coverage. The final
scene was cropped, to provide consistent wall to wall coverage of the
reconstructed scene. The orthomosaic was exported as a geotiff for
further georeferencing in QGIS \citep{qgisdevelopmentteam2018} using the
georeferencer plugin (version 3.1.9) and additional ESRI World Imagery
high resolution reference data. We used 3\textsuperscript{th} degree
polynomial and 16 ground control points to correct the final image.
Ground control points, raw image data and final processed image are
provided in addition to measures of uncertainty such as mean root mean
squared (RMSE), mean and median error across all ground control points.
All subsequent analysis are executed on the final geo-referenced
orthomosaic or subsets of it.

\hypertarget{land-use-and-land-cover-change}{%
\subsection{Land-Use and Land-Cover
Change}\label{land-use-and-land-cover-change}}

\hypertarget{classifying-forest-cover}{%
\subsubsection{Classifying forest
cover}\label{classifying-forest-cover}}

We automatically delineated all natural forest in the historical data,
thus excluding tree plantations, thinned or deteriorated forest stands,
fields and buildings. We used the Unet Convoluted Neural Net (CNN,
\citet{ronneberger2015}) architecture implemented in Keras
\citep{chollet2015} with an efficientnetb3 backbone
\citep{yakubovskiy2019} running on TensorFlow \citep{martinabadi2015} to
train a binary classifier (i.e.~forest or non-forested). This
methodology is increasingly being used to automate pixel-level
classification in (color) digital photography data \citep{buscombe2018}.
Training data were collected from the orthomosaic by randomly selecting
513 pixel square tiles from homogeneous forested or non-forested areas
within the historical orthomosaic. Homogeneous tiles were combined in
synthetic landscapes using a random gaussian field based binary mask
(Figure 3). We generated 5000 synthetic landscapes for training, while
500 landscapes were generated for both the validation and the testing
dataset. Source tiles did not repeat across datasets to limit
overfitting. In order to limit stitch line misclassifications, along the
seams of mosaicked images, we created synthetic landscapes with
different forest tiles to mimick forest texture transitions. We applied
this technique to 10\% of the generated synthetic landscapes. The CNN
model was trained for 100 epochs on a graphics processing unit (GPU)
maximizing the Intersect-over-Union (IoU) using additional data
augmentation. Data augmentation included random cropping to 320 pixel
squares, random orientation, scaling, perspective, contrast and
brightness shifts and image blurring. During final model evaluation we
report the IoU of our out-of-sample test datasets. The optimized model
was used to classify the complete orthomosaic using a moving window
approach with a step size of 110 pixels and a majority vote across
overlapping areas to limit segmentation edge effects. We refer to Figure
4 for a synoptic overview of the full deep learning learning workflow.

\begin{figure}
\includegraphics[width=1\linewidth]{./figures/synthetic_landscape} \caption{An example synthetic landscape, combining homogeneous forest and non-forest images into a patchy landscape using random gaussian field based masks. The left panel shows a combined synthetic landscape, while the right panel shows the corresponding forest (black) and non-forest (white) labels.}\label{fig:unnamed-chunk-3}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth]{./figures/cnn_diagram} \caption{A diagram of the deep learning workflow followed in training a binary forest / non-forest cover convolutional neural net U-Net model to generate our forest cover map.}\label{fig:unnamed-chunk-4}
\end{figure}

To map long term land-use and land-cover change in the Yangambi region
we used the contemporary Global Forest Change version 1.6 data (GFC,
tile 10N-020E) \citep{hansen2013}. Using this data we calculated the
latest state of the forest with respect to the conditions at the start
of 1958, 60 years earlier. In our analysis we only included forested
pixels which recorded no loss throughout the whole 2000 - 2018 period.
As the resolution of the historical forest classification exceeds that
of the GFC map we downsampled our historical forest cover data to 30m
GFC resolution. We masked out all water bodies using the Global Forest
Change survey data mask layer, and limited the analysis to the right
bank of the Congo river. Where the datasets overlap we provide summary
statistics of deforestation, reforestation and afforestation.

\hypertarget{landscape-fragmentation-above-ground-carbon-estimates}{%
\subsubsection{Landscape fragmentation \& Above Ground Carbon
estimates}\label{landscape-fragmentation-above-ground-carbon-estimates}}

To quantify changes in the structure of forest cover and its
disturbances we used spatial pattern analysis landscape fragmentation
metrics \citep{dale1999}. We report the ratio of edge to area and the
fractal index to quantify landscape complexity of forest disturbances.
Statistics were calculated for all forest disturbance patches larger
than 1 ha using the R package landscapemetrics \citep{hesselbarth2019}.
We provide mean and standard deviation on edge, area, their ratio and
fractal dimension for both the historical and contemporary
\citet{hansen2013} forest cover maps.

We estimated above ground carbon (AGC) losses and gains over time using
plot based averages of recent inventory data at permanent sampling plots
in the area (Figure 2). We refer to \citet{Kearsley2013} for the survey
method and allometric relations used to scale the survey data. Unlike
standard square 1 ha plots edge plots were set back 200m from forest
edges and were 50x200m, with the 50m side of the plot along the forest
edge and continuing 200m into the forest (Appendix Table 1). We further
confirmed that forest edge plots did not show a significantly different
AGC compared to those of non-edged / mixed forest plots (Mann Whitney U
test, p \textless{} 0.05). Thus it was not necessary to explicitly
quantify changes in AGC caused by edge effects. Moreover, we used the
mean values of the mixed forest as representative for potential AGC
losses. Despite the challenges inherent in quantifying AGC for forest
edges we mapped the total extent of the edges in the contemporary
landscape. To align our landscape analysis with exploratory analysis of
the survey data we used a buffer of 200m to estimate the extent of
forest edges and patches, up to the location of forest edge plots.

Surveys of old plantations show a large variation in AGC, depending on
age and the crop type. For example, the AGC values varied from 168.67 to
86.55 g Mg C ha\textsuperscript{-1}, for \emph{Hevea brasiliensis}
(rubber tree) and \emph{Elaeis guineensis} (oil palm) plots respectively
(\citet{bustillo2018}, personal communications). These higher values are
in line with the mixed AGC estimates in the area, while the palm
plantations resemble old-regrowth values (81.8 Mg C
ha\textsuperscript{-1}, see Table A1). We therefore use both the
estimates of old-regrowth and mixed forest to estimate AGC for regrowth.
We did not have sufficient data to account for individual changes in AGC
across plantations.

\hypertarget{canopy-structure-foto-texture-analysis}{%
\subsection{Canopy structure \& FOTO texture
analysis}\label{canopy-structure-foto-texture-analysis}}

We compared the structure of the canopy both visually and using Fourier
Transform Textural Ordination (FOTO, \citet{couteron2002}). Fourier
Transform Textural Ordination uses a principal component analysis (PCA)
on radially averaged 2D Fourier spectra to characterize canopy (image)
texture. The FOTO technique was first described by \citet{couteron2002}
to quantify canopy stucture in relation to biomass and biodiversity, and
can be used across multiple scenes using normalization
\citep{Barbier2010}.

We used an area of 400x400m (16 ha) around contemporary permanent
sampling plot locations to quantify the general state of the canopy in
both historical and contemporary remote sensing data. Contemporary data
were resampled to the resolution of the historical data using a nearest
neighbour interpolation for comparison. We used the first acquisition of
a pan-chromatic Geo-Eye 1 stereo pair (GeoEye, 2012, Thornton, Colorado,
U.S.A., order 737537, 2011-11-11 8:55h GMT or 9:55h local time). For
permanent sampling plot locations in both historical and contemporary
data a rectangular buffer of 200m around the centre of all plots was
used to create small image subsets. All image subsets were processed
using the R based FOTO implementation and package \citep{hufkens2019a}
using global normalization, ensuring intercomparability of texture
metrics across time and space \citep{Barbier2010}. We used a FOTO
(moving) window size of 187 pixels (or \textasciitilde{}150m), as
described in \citet{solorzano2018} and \citet{Barbier2010}, to ensure
that multiple individual canopies could be included in the analysis. A
buffer with a radius of 50m around the center of all permanent sampling
plots was used to determine the average value of the first principle
component (PC), explaining the majority of the variance in canopy
texture. Where both the Geo-Eye data and the orthomosaic image
overlapped we processed the intersecting region to explore a
wall-to-wall comparison between past and current canopy texture metrics.

We used the standard FOTO methodology with fixed zones, instead of the
moving window approach. The window size was set to the same size (187
pixels or \textasciitilde{}150m) as used in the moving window analysis
above. To account for illumination differences between the two scenes we
applied histogram matching. No global normalization was applied, as the
scene was processed as a whole. PC values from this analysis for all
permanent sampling plots in both image scences were extracted using a
buffer with a radius of 50m.

For both site based and scene analysis we correlate PC values with
permanent sample plot inventory data such as stem density, above ground
biomass and tree species richness. Additional comparisons are made
between contemporary Geo-Eye data and the historical orthomosaic derived
PC values. Due to the few available permanent sampling plots in both
scenes we used a non-parametric paired signed rank (Wilcoxon) test to
determine differences between the PC values of the Geo-Eye and
historical orthomosaic image scenes across mono-dominant and mixed
forest types. In all analysis, mono-dominant site 4 was removed from the
analysis due to cloud contamination.

\hypertarget{results}{%
\section{Results}\label{results}}

Our analysis provides a first spatially explicity historical composite
of aerial survey images mapping LULCC within the Congo Basin. The use of
high resolution historical images combined with structure from motion
image processing techniques allowed us to mosaic old imagery across a
large extent. The final orthomosaic composition of the Yangambi region
resulted in an image scene covering approximately 733 million pixels
across \textasciitilde{}828 km\textsuperscript{2}
(\textasciitilde{}23x36 km, Figure 2). The overall accuracy of the
structure from motion orthomosaic composition was 0.88 m/pixel using the
sparse cloud DEM for corrections at 45.8 m/pixel. The resulting
georeferenced scene had a spatial accuracy of approximately 23m. Further
georeferencing outside the SfM workflow reduced the mean error at the
ground control points to 5.3 \(\pm\) 4.9 px (\textasciitilde{}4.7
\(\pm\) 4.3 m), with a median error of 2.9 px (2.6m). The orthomosaic
served as input for all subsequent LULCC analysis with all derived maps
provided with the manuscript repository (see data \& code availability
statements below).

\begin{figure}
\includegraphics[width=1\linewidth]{./figures/orthomosaic_maps} \caption{Overview of the final orthomosaic of the greater Yangambi region (A), a detailed inset (B) and the derived land use land cover change map displayed as the difference between a manual classification of the orthomosaic and a recent map by Hansen et al. 2013 (C) and the corresponding land cover map for inset B (D).}\label{fig:unnamed-chunk-5}
\end{figure}

\hypertarget{land-use-and-land-cover-change-above-ground-carbon}{%
\subsection{Land-use and Land-Cover Change \& Above Ground
Carbon}\label{land-use-and-land-cover-change-above-ground-carbon}}

Our CNN deep learning classifier reached an Intersection-over-Union
(IoU) accuracy of 97\% on the detection of disturbed forest in
out-of-sample test data. Scaling our classifier to the whole historical
orthomosaic we detected 162 km\textsuperscript{2} (or
\textasciitilde{}20\% of the scene) of disturbed forests. A large
fraction of the disturbed area was restored in the period between the
two acquisitions. In total, 99 km\textsuperscript{2}, or little over
half of the affected forest was restored (Figure 5C-D, dark blue).
Recent deforested areas, as registered through satellite remote sensing,
approximate 88 km\textsuperscript{2} (Table1, Figure 5C-D / light
green).

Recent deforestation follows a distincly different pattern compared to
historical patterns. Historical deforestation showed a classical
fishbone pattern for forest clearing with very sharp edges, while
current patterns are patchy and ad-hoc (figure 5D, dark blue and green
colours respectively). These differences are reflected in the analysis
of landscape metrics of forest loss. Between the historical and
contemporary LULCC maps we see an increase in small disturbances, as
indicated by the decreasing area of the mean patch size, down to 37 ha
from 120 ha historically, while perimeter lengths remained approximately
the same, with 8653m and 9902m for historical and contemporary maps
respectively (Table 2). This shift in perimeter area ratio led to a
similar change in the fractal index, increasing in value to 1.13 \(\pm\)
0.07 from 1.10 \(\pm\) 0.05 over time. Values closer to a fractal index
of 2 suggest a more complex (fragmented) landscape.

A comparison of forest edge plots with mixed forest plots showed no
signficant difference in AGC, or other reported values such as species
richness, basal area or stem density (Mann Whitney U test, p \textless{}
0.05). Edge influence did not extend beyond 200 m from a forest edge,
but still represented an area of 132 km\textsuperscript{2}.

Changes in both land-use and land-cover led to concomitant changes in
above ground carbon stocks. Recovery throughout the region was
characterized for patches of forest and plantations. Assuming high
density stands, based on previous work, this could amount to a carbon
gains of 1592 Gg C, offsetting more recent losses of approximately 1408
Gg C. On the other hand, at the low end, if we assume a lower carbon
density of 81.8 Mg C ha\textsuperscript{-1} this would result in a total
carbon gain of 811 Gg C. Using our approach results indicate that
overall deforestation around Yangambi has resulted in a loss of
\textasciitilde{}2416 Gg C in AGC stocks.

\begin{table}[!h]

\caption{\label{tab:unnamed-chunk-6}Land use land cover change statistics of forest cover around Yangambi in the central Congo Basin. The data evaluates a difference between a historical (1958) aerial photography based survey and the Hansen et al. 2013 based satellite remote sensing data. Spatial coverage statistics are provided in square kilometers (km) and hectares (ha), rounded to the nearest integer as well as Above Ground Carbon (AGC) scaled using recent survey measurements.}
\centering
\begin{tabular}[t]{lrrl}
\toprule
\multicolumn{3}{c}{ } & \multicolumn{1}{c}{AGC} \\
\cmidrule(l{3pt}r{3pt}){4-4}
  & $km\textsuperscript{2}$ & ha & (Gg C $ha\textsuperscript{-1}$\\
\midrule
Forest & 685 & 68455 & -\\
-  of which edges & 132 & 13151 & -\\
Regrowth > 1958 loss & 99 & 9918 & 811 - 1592\\
Loss > 2000 & 88 & 8776 & 1408\\
Loss > 1958 (permanent) & 63 & 6282 & 1008\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[!h]

\caption{\label{tab:unnamed-chunk-7}Landscape metrics for historical and contemporary deforestation patterns. We report patch perimeter and area, their ratio and fractal dimension. Values are reportedpat as mean $\pm$ standard deviation, across all deforestation patches.}
\centering
\begin{tabular}[t]{lllll}
\toprule
  & perimeter (m) & area (ha) & ratio ($m\textsuperscript{-1}$) & fractal dimension\\
\midrule
historical & 8653 $\pm$ 35676 & 120 $\pm$ 659 & 0.031 $\pm$ 0.014 & 1.099 $\pm$ 0.052\\
contemporary & 9902 $\pm$ 56975 & 37 $\pm$ 226 & 0.046 $\pm$ 0.014 & 1.13 $\pm$ 0.071\\
\bottomrule
\end{tabular}
\end{table}

\hypertarget{canopy-structure-foto-texture-analysis-1}{%
\subsection{Canopy structure \& FOTO texture
analysis}\label{canopy-structure-foto-texture-analysis-1}}

Visual interpretation of the scenes provide evidence that most locations
do not change dramatically with respect to canopy composition, except
for the large areas of disturbances in contemporary fallow or
young-regrowth plots. One marked difference is noted in the
mono-dominant site 6 (Appendix Table 1). Here, the current mono-dominant
\emph{Brachystegia laurenttii} is a recent development, changing the
canopy structure visibly during the last half century (Figure 6). The
previous varied canopy structure gave way to a more dense and uniform
canopy. This is reflected in a change of the FOTO PC value from 0.19
historically to its current value of 0.54. This historical value is
similar to the mean of contemporary mono-dominant stands with PC
averaging 0.34 \(\pm\) 0.1, and is only slightly higher than historical
values for a mixed forest (0.18 \(\pm\) 0.08). The reverse pattern is
seen in the contemporary PC values. Here, the value of 0.54 exceeds
those of most mono-dominant stands (0.35 \(\pm\) 0.08), and is even
further removed from the values noted for mixed forests (0.12 \(\pm\)
0.03).

\begin{figure}
\includegraphics[width=1\linewidth]{./figures/visual_comparison_psp} \caption{Visual comparison between a historical (A) and contemporary (B) permanent sampling plot. The site is currently listed as a mono-dominant Brachystegia laurenttii stand. Note the structural differences with a 'courser' canopy structure in the historical image compared to the more closed contemporary stand.}\label{fig:unnamed-chunk-8}
\end{figure}

Using only small subsets around existing permanent sampling plots we
show distinct differences between forest types, with PC values in both
historical and contemporary imagery markedly higher for the
mono-dominant forest types compared to all others (Appendix Figure 3).
Provided that the young-regrowth and fallow permanent sampling plots
have seen recent disturbance the Wilcoxon signed rank test on the mixed
and mono-dominant plots between the historical and contemporary PC
values did not show a significant difference (p \textgreater{} 0.05).
Similarly, no significant different using PC values extracted from the
whole scene analysis was noted (p \textgreater{} 0.05). Any
relationships between contemporary Geo-Eye data and permanent sampling
plot measurements of Above Ground Biomass, stem density and species
richness were non-significant (p \textgreater{} 0.05, Appendix Figures
4-6).

Furthermore, visual inspection of the scene wide analysis suggests
historical scences do not show landscape wide canopy features (Figure 7
A-B), unlike the contemporary scene (Figure 7 C-D). Where the FOTO
algorithm picks up landscape features such as changes in texture across
the contemporary Geo-Eye scene (e.g.~the river valley as a diagonal line
in Figure 7D), however, no corresponding landscape patterns are found by
the FOTO algorithm in the historical orthomosaic.

\begin{figure}
\includegraphics[width=1\linewidth]{./figures/foto_maps} \caption{RGB visualizations of the first three principal components of scene wide FOTO texture analysis of historical and current (Geo-Eye) imagery. Current permanent sampling plots of mono-dominant, mixed, fallow and young (edge) forest plots are marked with open triangles, open circles, open squares and crosses, respectively.}\label{fig:unnamed-chunk-9}
\end{figure}

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

Finely grained spatial data sources, such as remote sensing imagery, are
rare before the satellite era (\textless{}1970). This lack of data
limits our understanding of how forest structure has varied over longer
time periods in especially remote areas. Long term assessment can be
extended by using large inventories of historical aerial survey data
\citep{kadmon1999, song2015, nita2018}. However, across the Congo Basin
this data is rare, or difficult to access. Despite the difficulties in
assessing hidden disturbances caused by deforestation, poorly recorded
official assessments or simply those invisible to standard remote
sensing techniques \citep{peres2006}, remote sensing generally remains
the best way to map and quantify LULCC \citep{houghton2012}. Despite
these challenges in recovery and validation we present a first analysis
of LULCC using historical aerial survey data for the central Congo
Basin. We used a combination of Structure-from-Motion and a
convolutional neural net deep learning approach to map deforestation
across a large extent, providing a long term view of LULCC in the
central Congo Basin

\hypertarget{data-recovery-challenges}{%
\subsection{Data recovery challenges}\label{data-recovery-challenges}}

In our study the archive data recovered was limited to contact prints
and therefore did not represent the true resolution of the original
negative. In addition, analogue photography clearly produces a distinct
softness compared to digital imagery (Figure 4). Despite favourable
nadir image acquistions \citep{verhoeven2017} image softness combined
with illumination effects between flight paths, and the self-similar
nature of vast canopy expanses \citep{park2019, simini2010, sole1995},
limited our ability to provide wall-to-wall coverage of the entire
dataset containing 334 images. Few man made features in the scenes also
made georeferencing challenging. Although the village of Yangambi
provided a range of buildings as (hard-edge) references, other areas
within the central Congo Basin might have fewer permanent structures and
would require the use of soft-edged landscape features (e.g.~trees,
river outflows). Research has shown that soft-edged features can help
georeference scenes even when containing few man-made features
\citep{hughes2006}, however accuracy becomes compromised. Our two step
georeferencing approach resulted in a referencing accuracy of
\textasciitilde{}4.7 \(\pm\) 4.3 m across reference points. However, it
shoud be noted that referencing accuracy of the final scene is less
constrained toward the edges of the scene.

\hypertarget{land-use-land-cover-change-above-ground-carbon}{%
\subsection{Land-use, Land-Cover change \& Above Ground
Carbon}\label{land-use-land-cover-change-above-ground-carbon}}

When classifying our orthomosaic into forest and non-forest states we
favoured a deep learning supervised classification using a Convoluted
Neural Network over manual segmentation to guarantee an
``apples-to-apples'' comparison between the historical and the
contemporary GFC forest cover maps. We acknowledge that both
classifications use different features, i.e.~spectral or spatial data,
but attain a similarly high accuracy of 97\% and 99\%
\citep{hansen2013}, for the GFC and our CNN map respectively. Despite a
number of challenges we were able to map and quantify a vast continuous
area of tropical forest, with limited manual annotation requirements.

Our analysis shows that the majority of deforestation around Yangambi
happened toward the late 1950's (162 km\textsuperscript{2}).
Considerable regrowth has occured since the aerial survey was executed
(99 km\textsuperscript{2}), and socio-economic instability prevented
further large scale forest exploitation. In particular, many plantations
have reached maturity and forest has re-established in previously
cleared or disturbed areas. The majority of this forest recovery takes
the form of isolated patches of forest but is offset by further
deforestation of previously untouched forest. Generally, the function
and structure of forests can be influenced by forest edges that are
located up to 1km away however most effects are pronounced within the
first 300m from the edge \citep{gascon2000}. Our analysis of edge
effects on AGC has shown that the influence is negligible 200m away from
the edge. \citet{phillips2006} have shown similar weak responses to edge
effects in the Amazon forest. Due to a lack of data on the extent
(depth) of edge effects and their influence on AGC beyond 200m we did
not include any estimates of carbon loss or gain within these zones.
However, it must be stated that the influence of edges throughout the
landscape was not marginal. as these areas would account for 132
km\textsuperscript{2}. Thus edges could have a substantial negative
\citep{brinck2017} or positive \citep{reinmann2017} influence on AGC.
Similarly, uncertainties in how to explicitly correct for plantations in
the landscape present a further challenge. Thus although our estimates
are only indicative they do underscore the important influence of
landscape structure when carbon accounting. However, our findings do not
indicate that deforestation in Congo basin is declining, on the
contrary.

Over the past half century there has been a clear shift in land use in
Yangambi (Figure 5). Land use has shifted away from for example a
regular fishbone deforestation pattern that emerges when (large scale)
agricultural interests dominate the landscape \citep{arima2015}, to a
more fragmented landscape (Figure 5D). The latter former? is consistent
with historical land management at the time of the aerial survey
\citep{bustillo2018}. These regular patterns have since been reversed
because of a decrease in large scale intensive agriculture, replaced by
ad-hoc small scale subsistence farming with large perimeter to area
relationships (i.e.~ragged edges). Consequently, edge effects in the
current landscape are far more pronounced than in the historical scene.

Visual inspection of the images also suggests that reforestation within
the historically cleared areas and experimental plots is not necessarily
limited to areas far removed from more densely populated areas. For
example, large reforested areas exist close to the Congo stream and
Yangambi village itself (Fig. 5). Here, regional political components,
such as land leases and large scale ownership could have played a role
in safeguarding some of these areas for rewilding or sustainable
management \citep{arima2014, larson2011}. Despite widespread
anthropogenic influences throughout the tropics \citep{lewis2015} the
retention of these forested areas show the potential of explicit or
implicit protective policy measures (e.g.~INERA concessions,
\citet{bustillo2018}) on a multi-decadal time scale. Forest regrowth in
non-continuous areas within Yangambi could increase landscape
connectivity and help increase biodiversity \citep{vandeperre2018}.

Our analysis therefore provides an opportunity to highlight and study
those regions that have previously suffered confirmed long-term
disturbances, and those that have been restored since. Assessments of
old plantations and recovering clear-cut forests can serve as a guide to
help estimate carbon storage capacity and forest recovery rates in
managed and unmanaged conditions
\citep{gourlet-fleury2013, sader1988, achard2014} over the mid- to
long-term, in support of rewilding and general forest restoration
\citep{arima2014, larson2011, vandeperre2018}. In addition, mapping
long-term edge effects can further support research into issues such as
receding forest edges \citep{gascon2000}.

\hypertarget{canopy-structure-foto-texture-analysis-2}{%
\subsection{Canopy structure \& FOTO texture
analysis}\label{canopy-structure-foto-texture-analysis-2}}

Finally the FOTO technique used to quantify relationships between canopy
structure and forest characteristics rendered no valuable insights of
either the historical orthomosaic or recent Geo-Eye scene. Similarly
weak correlations were found previously by \citet{solorzano2018}. In
contrast site based texture metric statistics did show correspondence
between historical and contemporary satellite imagery. However, none of
them were either consistent or significant. Although visual
interpretation shows distinctly different canopy structures (Figure 5)
the differences in how resolution is defined and issues related to image
quality prevented us from quantifying these further. Unlike large scale
studies by \citet{Ploton2012} we could not scale this technique to
historical data. We advise that future valorisation efforts should
preferentially work from foto negatives (if available) to ensure optimal
data quality in resolution, contrast and overall sharpness.

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

Given the impact of tropical forest disturbances on atmospheric CO\_2
emissions, biodiversity and ecosystem productivity accurate long term
reporting on land-cover and land-use change especially in the
pre-satellite era is an imperative. Our analysis of historical aerial
survey images of the Central Congo Basin provides a window into the
state of the forest at the start of the anthropocene. Efforts to
quantify canopy texture and their link to carbon storage had limited to
no success.We have shown that historical aerial survey data can be used
to quantify land-use and land-cover change using a highly automated
workflow to quantify the structure and extent of forest cover change
that can help assess the impact of fragmentation on above ground carbon
stocks. We also show a shift from previously highly structured
industrial deforestation of large areas, to discrete smallholder
clearing for farming, increasing landscape fragmentation but also
opportuntiy for substantial regrowth. Efforts to quantify canopy texture
and their link to carbon storage had limited to no success. As such, our
analysis provides insights into the state of rarely studied tropical
forests and the rate at which deforestation and reforestation has taken
place over a multi-decadal scale in the central Congo basin providing a
useful historical context to land-cover and land-use change spatial
patterns for interpreting past and ongoing field research in the area.

\hypertarget{additional-information-and-declarations}{%
\section{Additional Information and
Declarations}\label{additional-information-and-declarations}}

\hypertarget{data-availability}{%
\subsection{Data availability}\label{data-availability}}

Hufkens et al.~(2019). A curated dataset of aerial survey images over
the central Congo Basin, 1958. Zenodo: doi.org/10.5281/zenodo.3547767.
All data not included in the latter repository can be found bundled with
the analysis code as listed below. Proprietary datasets (i.e.~Geo-Eye
data) are not shared, but purchase order numbers allow for acquisition
of these datasets to ensure reproducibility. The \citet{hansen2013}
image data is freely available from
\href{https://earthenginepartners.appspot.com/science-2013-global-forest/download_v1.6.html}{the
GFC project website}.

\hypertarget{code-availability}{%
\subsection{Code availability}\label{code-availability}}

All analysis code is available as an R \citep{rcoreteam2019} project
(\url{https://github.com/khufkens/orthodrc}). The analysis relied
heavily on the `raster' \citep{hijmans2019}, `RStoolbox'
\citep{leutner2019}, and `landscapemetrics' \citep{hesselbarth2019}
packages, while post-processing and plotting was facilitated by the
`tidyverse' ecosystem \citep{wickham2017}, `ggthemes'
\citep{arnold2019}, `scales' \citep{wickham2018} and `cowplot'
\citep{wilke2019}. Additional plotting elements were formatted or
provided by `sf' \citep{pebesma2018} and `rnaturalearth'
\citep{south2017} packages, respectively. I'm grateful for the
contributions to the scientific community by the developers of these
packages.

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% %% optional
% \supplementary{The following are available online at www.mdpi.com/link, Figure S1: title, Table S1: title, Video S1: title.}
%
% % Only for the journal Methods and Protocols:
% % If you wish to submit a video article, please do so with any other supplementary material.
% % \supplementary{The following are available at www.mdpi.com/link: Figure S1: title, Table S1: title, Video S1: title. A supporting video article is available at doi: link.}

\vspace{6pt}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\acknowledgments{This research was supported through the Belgian Science Policy office
COBECORE project (BELSPO; grant BR/175/A3/COBECORE). KH acknowledges
funding from the European Union Marie Skłodowska-Curie Action (project
number 797668). I'm grateful for the support of Thales D'Haulleville
volunteering his time in scanning the images.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\authorcontributions{K.H. conceived and designed the study, analyzed the data, prepared
figures and tables, authored, reviewed and approved the final draft of
the manuscript. T.d.H. scanned all image data. E.K. and T.d.H. provided
plot based AGC estimates. T.D., K.J., E.K, H.B., P.S., F.V.,S.M., M.A.,
J.V.D.B., H.V and L.W. reviewed the final manuscript.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\conflictsofinterest{The authors declare no conflict of interest. The founding sponsors had
no role in the design of the study; in the collection, analyses, or
interpretation of data; in the writing of the manuscript, an in the
decision to publish the results.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Citations and References in Supplementary files are permitted provided that they also appear in the reference list here.

%=====================================
% References, variant A: internal bibliography
%=====================================
% \reftitle{References}
% \begin{thebibliography}{999}
% % Reference 1
% \bibitem[Author1(year)]{ref-journal}
% Author1, T. The title of the cited article. {\em Journal Abbreviation} {\bf 2008}, {\em 10}, 142-149, DOI.
% % Reference 2
% \bibitem[Author2(year)]{ref-book}
% Author2, L. The title of the cited contribution. In {\em The Book Title}; Editor1, F., Editor2, A., Eds.; Publishing House: City, Country, 2007; pp. 32-58, ISBN.
% \end{thebibliography}

% The following MDPI journals use author-date citation: Arts, Econometrics, Economies, Genealogy, Humanities, IJFS, JRFM, Laws, Religions, Risks, Social Sciences. For those journals, please follow the formatting guidelines on http://www.mdpi.com/authors/references
% To cite two works by the same author: \citeauthor{ref-journal-1a} (\citeyear{ref-journal-1a}, \citeyear{ref-journal-1b}). This produces: Whittaker (1967, 1975)
% To cite two works by the same author with specific pages: \citeauthor{ref-journal-3a} (\citeyear{ref-journal-3a}, p. 328; \citeyear{ref-journal-3b}, p.475). This produces: Wong (1999, p. 328; 2000, p. 475)

%=====================================
% References, variant B: external bibliography
%=====================================
\reftitle{References}
\externalbibliography{yes}
\bibliography{bibliography.bib}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}

### 2019-12-27 analysis/00_lulcc_analysis.R

@@ -1,0 +1,63 @@
+# This script calculates the difference in forest cover
+# between the manual classification of the historical
+# orthomosaic data and the recent Hansen et al. 2013
+# global forest change map.
+#
+# I only consider true untouched forest in the period
+# 2000 - 2018 in this analysis.
+
+# load required libraries
+library(raster)
+
+# read outline of the mosaic
+m <- raster("data/orthomosaic/yangambi_orthomosaic_mask_resampled.tif")
+
+# read in forest mask (historical)
+yangambi <- raster("data/orthomosaic/yangambi_forest_mask_resampled.tif")
+yangambi[yangambi == 2] <- 5
+yangambi[is.na(yangambi)] <- 8
+yangambi <- mask(yangambi, m, maskvalue = NA)
+
+# read in Hansen map, crop/mask to fit the historical data for speed
+hansen <- raster("data/Hansen_et_al/Hansen_GFC_lossyear_yangambi.tif")
+hansen <- crop(hansen, extent(yangambi))
+hansen <- mask(hansen, m, maskvalue = NA)
+
+# select forests with no loss between 2000 and 2018
+hansen[hansen > 0] <- 1
+
+# difference between the maps
+# and reclassify the data
+change_map <- yangambi - hansen
+
+# in the final map values are,
+# 1: remaining untouched forest
+# 2: reforestation 1958 - 2000
+# 3: deforestation in period 2000 - 2018
+# 4: deforestation in 1958 until now
+change_map <- reclassify(change_map, c(3,4,3,
+                                       4,5,1,
+                                       7,8,2,
+                                       6,7,4))
+
+# print the map frequency statistics
+change_stats <- as.data.frame(freq(change_map, useNA='no'))
+
+# convert from pixel count to square km/m and ha
+# assuming a pixel size of ~30m
+change_stats$sq_m <- change_stats$count * 30^2
+change_stats$sq_km <- change_stats$count * 0.03^2
+change_stats$ha <- change_stats$sq_m / 10000
+
+change_stats$labels <- c(" no change",
+         " forest regrowth > 1958",
+         " forest loss > 2000",
+         " forest loss > 1958")
+
+# print the forest change stats to console
+print(change_stats)
+
+# write data to disk for further analysis
+writeRaster(change_map,
+"data/orthomosaic/yangambi_forest_cover_difference_1958_2000.tif",
+ overwrite = TRUE)

### 2019-12-27 analysis/01_lulcc_summary_statistics.R

@@ -1,0 +1,37 @@
+# some summary statistics on land use land cover change
+# in particular quickly figure out what the status was
+# of the sites in the old imagery
+library(raster)
+
+# read in site loctions and the lulcc map
+# to cross reference the state of the forest
+# within a historical perspective
+
+locations <- read.table("data/surveys/site_characteristics.csv",
+                        sep = ",",
+                        header = TRUE)
+
+lulcc <- raster("data/orthomosaic/yangambi_forest_cover_difference_1958_2000.tif")
+
+# Define lat / lon projection.
+lat_lon <- CRS("+init=epsg:4326")
+
+# Read in the coordinates and assign them a projection
+ll <- SpatialPoints(cbind(locations$lon,locations$lat), lat_lon)
+
+# stuff results into original dataframe
+locations$lulcc_100 <- raster::extract(lulcc, ll, buffer = 100)
+locations$lulcc_summary <- lapply(locations$lulcc_100, function(x){
+  s <- table(x)
+  s <- round(s/sum(s)*100)
+  n <- names(s)
+  paste(paste(s," (",n,")", sep = ""), collapse = ", ")
+  })
+
+# results basically show that everything was forest in 1958
+# this includes old and young regrowth forests
+# indicating that old regrowth is older than 60 years while
+# the young regrowth is younger than 60 years in age
+print(locations)
+
+

### 2019-12-27 analysis/02_foto_analysis_psp.R

@@ -1,0 +1,149 @@
+# Foto analysis of canopy texture
+library(raster)
+library(foto)
+library(tools)
+library(RStoolbox)
+
+# read in site location + other info
+locations <- read.table("./data/surveys/site_characteristics.csv",
+                        sep = ",",
+                        header = TRUE)
+
+# downsample the geo-eye image to the resolution
+# of the historical data
+historical <- raster("./data/orthomosaic/yangambi_orthomosaic_modified.tif")
+
+if(!file.exists("./data/geo-eye/geo-eye_panchromatic_latlon_resampled.tif")){
+  message("resampling geo-eye data")
+  geoeye <- raster("./data/geo-eye/geo-eye_panchromatic_latlon.tif")
+  geoeye_r <- resample(geoeye, historical)
+  geoeye_r <- crop(geoeye_r,extent(geoeye))
+  writeRaster(geoeye_r,
+              "./data/geo-eye/geo-eye_panchromatic_latlon_resampled.tif")
+} else {
+  message("reading in geo-eye data")
+  geoeye_r <- raster("./data/geo-eye/geo-eye_panchromatic_latlon_resampled.tif")
+}
+
+# histogram matching
+geoeye_r <- histMatch(geoeye_r, historical)
+
+# reproject to work in meters not degrees
+
+# create subsets of at least 500x500 m
+# then apply a global normalized analysis in foto using
+# a window size equivalent to roughly 150m or 150px
+# moving window to assure good coverage
+# then extract buffered extraction of these values
+# for 50m around the site location to get a metric
+# of
+
+# offset of approximately ~200m in degrees
+offset <- 0.00182
+
+# calculate the historical subsets
+apply(locations,1,function(l){
+  x <- as.numeric(l['lon'])
+  y <- as.numeric(l['lat'])
+
+  filename <- paste0("./data/foto/psp/",
+                     l["type"],
+                     "_",
+                     l["nr"],
+                     "_historical.tif")
+
+  if(!file.exists(filename)){
+
+    subset <- crop(historical, extent(c(x - offset,
+                                        x + offset,
+                                        y - offset,
+                                        y + offset)))
+
+    writeRaster(subset,
+                filename,
+                options = c("COMPRESS=DEFLATE"),
+                overwrite = TRUE,
+                datatype='INT1U')
+  }
+})
+
+# calculate the contemporary subsets
+apply(locations,1,function(l){
+
+  x <- as.numeric(l['lon'])
[…]

### 2019-12-27 analysis/03_foto_analysis_scene.R

@@ -1,0 +1,77 @@
+# Foto analysis of canopy texture
+library(raster)
+library(foto)
+library(tools)
+library(RStoolbox)
+
+# read in site location + other info
+locations <- read.table("./data/surveys/site_characteristics.csv",
+                        sep = ",",
+                        header = TRUE)
+
+# downsample the geo-eye image to the resolution
+# of the historical data
+historical <- raster("./data/orthomosaic/yangambi_orthomosaic_modified.tif")
+
+if(!file.exists("./data/geo-eye/geo-eye_panchromatic_latlon_resampled.tif")){
+  message("resampling geo-eye data")
+  geoeye <- raster("./data/geo-eye/geo-eye_panchromatic_latlon.tif")
+  geoeye_r <- resample(geoeye, historical)
+  geoeye_r <- crop(geoeye_r,extent(geoeye))
+  writeRaster(geoeye_r,
+              "./data/geo-eye/geo-eye_panchromatic_latlon_resampled.tif")
+} else {
+  message("reading in geo-eye data")
+  geoeye_r <- raster("./data/geo-eye/geo-eye_panchromatic_latlon_resampled.tif")
+}
+
+# calculate FOTO pca values
+if(!file.exists("./data/foto/scene/pca_values.rds")){
+
+  # crop orthomosaic
+  historical <- crop(historical, extent(geoeye_r))
+
+  # histogram matching
+  geoeye_r <- histMatch(geoeye_r, historical)
+
+  # write to file for batch processing
+  dir.create(file.path(tempdir(),"/foto"))
+  writeRaster(geoeye_r,file.path(tempdir(),"/foto/geoeye.tif"),
+              overwrite = TRUE)
+  writeRaster(historical,
+              file.path(tempdir(),"/foto/historical.tif"),
+              overwrite = TRUE)
+
+  pca_values <- foto_batch(file.path(tempdir(),"/foto/"),
+                           window_size = 187,
+                           method = "zones",
+                           cores = 2)
+
+  # save as rds file
+  saveRDS(pca_values, "./data/foto/scene/pca_values.rds")
+} else {
+  pca_values <- readRDS("./data/foto/scene/pca_values.rds")
+}
+
+# extract historical forest status
+forest_cover <- raster("./data/orthomosaic/yangambi_forest_cover_difference_1958_2000.tif")
+locations$forest_cover <- raster::extract(forest_cover,
+                                          cbind(locations$lon,
+                                                locations$lat),
[…]

### 2019-12-27 analysis/04_foto_summary_statistics.R

@@ -1,0 +1,175 @@
+library(tidyverse)
+library(raster)
+library(sf)
+
+#---- analysis psp snippets
+psp <- readRDS("./data/foto/psp/pc1_values.rds")
+psp$nr <- as.numeric(psp$nr)
+psp <- psp[!(psp$type == "mono-dominant" & psp$nr == 4),]
+
+# remove old regrowth
+psp <- psp %>%
+  filter(type != "old-regrowth")
+
+p <- ggplot(psp, aes(x=type, y=pc, fill = era)) +
+  geom_boxplot() +
+  theme_minimal() +
+  labs(y = "PC 1",
+       x = "") +
+  theme(legend.title = element_blank(),
+        legend.position="bottom",
+        text = element_text(size=20)) +
+  scale_fill_manual(
+    values=c("#999999",
+              rgb(255, 173, 70, 100, maxColorValue = 255))) +
+  coord_flip()
+
+ggsave("manuscript/figures/foto_bplot_psp.png",
+       p,
+       width = 7,
+       dpi = 150)
+
+# AGC plot
+p <- ggplot(psp) +
+  geom_point(aes(x = pc,
+                 y = above.ground.C.stock..Mg.C.ha.1.,
+                 pch = type),
+             cex = 4) +
+  labs(y = expression("Above Ground Carbon (Mg C ha"^-1 * ")"),
+       x = "PC 1") +
+  scale_shape_discrete(name = "Forest Type") +
+  theme_minimal() +
+  theme(axis.text = element_text(size = 15),
+        axis.title = element_text(size = 15),
+        legend.text = element_text(size = 15),
+        legend.title = element_text(size = 15),
+        strip.text = element_text(size = 15)) +
+  facet_wrap(~ era)
+
+ggsave("./manuscript/figures/foto_pc1_agc.png",
+       p,
+       width = 12,
+       height = 6,
+       dpi = 150)
+
+print(summary(lm(psp$pc ~ psp$above.ground.C.stock..Mg.C.ha.1.)))
+
+# richness plot
+p <- ggplot(psp) +
+  geom_point(aes(x = pc,
+                 y = species.richness,
+                 pch = type),
+             cex = 4) +
+  labs(y = "Species richness",
+       x = "PC 1") +
+  scale_shape_discrete(name = "Forest Type") +
+  theme_minimal() +
+  theme(axis.text = element_text(size = 15),
+        axis.title = element_text(size = 15),
[…]

### 2019-12-27 analysis/05_flight_paths.R

@@ -1,0 +1,122 @@
+# This routine generates a map of both the
+# orthomosaic as well as the derived land cover
+# change analysis (difference between Hansen et al. 2013 data
+# and the manual classification based on the mosaic).
+# Output is written to the manuscript directory for
+# integration in the final manuscript. Can be run standalone.
+
+# load libraries
+library(tidyverse)
+library(raster)
+library(rnaturalearth)
+library(rnaturalearthdata)
+library(sf)
+library(cowplot)
+library(rgdal)
+
+flight_paths = "./data/maps/flight_path_detail.tif"
+orthomosaic = "./data/orthomosaic/yangambi_orthomosaic_modified_resampled.tif"
+geoeye = "./data/geo-eye/geo-eye_panchromatic_latlon.tif"
+mab_outline = "./data/maps/"
+
+nc <- sf::st_read("data/maps/mab_ll.shp", quiet = TRUE)
+r <- brick(flight_paths)
+
+s <- read.table("./data/surveys/site_characteristics.csv",
+                sep = ",",
+                header = TRUE)
+
+y <- raster(orthomosaic)
+y_e <- st_as_sfc(st_bbox(y))
+bbox <- st_bbox(y)
+
+g <- raster(geoeye)
+g_e <- st_as_sfc(st_bbox(g))
+
+# load yangambi reserve boundaries
+e <- extent(c(24.35, 24.7, 0.7, 1.1))
+r <- crop(r, e)
+r_e <- st_as_sfc(st_bbox(r))
+nc <- sf::st_intersection(nc, r_e)
+
+# convert raster map to dataframe
+r_df <- r %>%
+  rasterToPoints %>%
+  as.data.frame() %>%
+  `colnames<-`(c("x", "y", "val"))
+
+# constructing raster map
+r_map <- ggplot()+
+  geom_tile(data = r_df, aes(x=x,y=y,fill = val)) +
+  scale_fill_gradient(low = "#1f78b4", high = "white") +
+  coord_fixed(ratio = 1)
+
+# adding vector data
+r_map <- r_map +
+  geom_sf(data = nc,
+          aes(alpha = 0.8),
+          fill = "grey",
+          colour = NA) +
+  geom_sf(data = y_e,
+          colour = "grey25",
+          fill = NA,
+          size = 1) +
+  geom_sf(data = g_e,
+          colour = "grey25",
+          fill = NA,
+          linetype = "dashed",
+          size = 1) +
+  geom_point(data = s,
[…]

### 2019-12-27 analysis/06_lulcc_map.R

@@ -1,0 +1,143 @@
+# This routine generates a map of both the
+# orthomosaic as well as the derived land cover
+# change analysis (difference between Hansen et al. 2013 data
+# and the manual classification based on the mosaic).
+# Output is written to the manuscript directory for
+# integration in the final manuscript. Can be run standalone.
+
+# load libraries
+library(tidyverse)
+library(raster)
+library(rnaturalearth)
+library(sf)
+library(cowplot)
+
+# load the forest change map
+forest_change <- raster(
+  "data/orthomosaic/yangambi_forest_cover_difference_1958_2000.tif")
+
+forest_change_subset <- crop(forest_change,
+                             extent(c(24.4356, 24.5165,
+                                       0.9156,0.9839)))
+# load the orthomosaic
+orthomosaic <- brick("data/orthomosaic/yangambi_orthomosaic_modified_resampled.tif")
+orthomosaic[orthomosaic == 255] <- NA
+orthomosaic_subset <-
+  crop(raster("data/orthomosaic/yangambi_orthomosaic_modified_resampled.tif"),
+                           extent(c(24.4356, 24.5165,
+                                    0.9156,0.9839)))
+
+# grab subset bounding box
+bb <- st_as_sfc(st_bbox(orthomosaic_subset))
+
+# convert gridded raster data dataframe
+forest_change_df <- forest_change %>%
+  rasterToPoints %>%
+  as.data.frame() %>%
+  `colnames<-`(c("x", "y", "val"))
+
+forest_change_subset_df <- forest_change_subset %>%
+  rasterToPoints %>%
+  as.data.frame() %>%
+  `colnames<-`(c("x", "y", "val"))
+
+orthomosaic_df <- orthomosaic %>%
+  rasterToPoints %>%
+  as.data.frame() %>%
+  `colnames<-`(c("x", "y", "val"))
+
+orthomosaic_subset_df <- orthomosaic_subset %>%
+  rasterToPoints %>%
+  as.data.frame() %>%
+  `colnames<-`(c("x", "y", "val"))
+
+# maps
+forest_change_map <- ggplot()+
+  geom_tile(data = forest_change_df, aes(x=x,y=y,fill=as.factor(val))) +
+  scale_fill_brewer(palette = "Paired",
+                    labels=c(" no change",
+                             " forest regrowth > 1958",
+                             " forest loss > 2000",
+                             " forest loss > 1958")) +
+  coord_fixed(ratio = 1) +
[…]

### 2019-12-27 analysis/07_foto_map.R

@@ -1,0 +1,226 @@
+# This routine generates a map of both the
+# orthomosaic as well as the derived land cover
+# change analysis (difference between Hansen et al. 2013 data
+# and the manual classification based on the mosaic).
+# Output is written to the manuscript directory for
+# integration in the final manuscript. Can be run standalone.
+
+# load libraries
+library(tidyverse)
+library(raster)
+library(rnaturalearth)
+library(sf)
+library(cowplot)
+library(RStoolbox)
+library(scales)
+
+# downsample the geo-eye image to the resolution
+# of the historical data
+historical <- raster("./data/orthomosaic/yangambi_orthomosaic_modified.tif")
+
+s <- read.table("./data/surveys/site_characteristics.csv",
+                sep = ",",
+                header = TRUE)
+
+if(!file.exists("./data/geo-eye/geo-eye_panchromatic_latlon_resampled.tif")){
+  message("resampling geo-eye data")
+  geoeye <- raster("./data/geo-eye/geo-eye_panchromatic_latlon.tif")
+  geoeye_r <- resample(geoeye, historical)
+  geoeye_r <- crop(geoeye_r,extent(geoeye))
+  writeRaster(geoeye_r,
+              "./data/geo-eye/geo-eye_panchromatic_latlon_resampled.tif")
+} else {
+  message("reading in geo-eye data")
+  geoeye_r <- raster("./data/geo-eye/geo-eye_panchromatic_latlon_resampled.tif")
+}
+
+# aggregate (lower resolution for plotting)
+geoeye_r <- aggregate(geoeye_r, fact = 4)
+historical <- crop(historical, extent(geoeye_r))
+historical <- aggregate(historical, fact = 4)
+geoeye_r <- histMatch(geoeye_r, historical)
+
+# grab extent of the data and crop out locations
+# based on this (to avoid growing the region of interest)
+e <- st_as_sfc(st_bbox(historical))
+s <- st_as_sf(s,
+              coords = c(5,4),
+              crs = 4326,
+              agr = "constant")
+s <- sf::st_intersection(s, e)
+
+# read pca values
+pca_values <- readRDS("./data/foto/scene/pca_values.rds")
+
+# convert gridded raster data dataframe
+geoeye_df <- geoeye_r %>%
+  rasterToPoints %>%
+  as.data.frame() %>%
+  `colnames<-`(c("x", "y", "val"))
+
[…]

### 2019-12-27 analysis/08_AGC_analysis.R

@@ -1,0 +1,84 @@
+library(raster)
+library(tidyverse)
+
+# load raster
+change_map <- raster("./data/orthomosaic/yangambi_forest_cover_difference_1958_2000.tif")
+change_map <- projectRaster(change_map, crs = "+init=epsg:3395",method = "ngb")
+change_map <- trim(change_map)
+
+# read site data estimates of AGC
+psp <- read.table("./data/surveys/site_characteristics.csv",
+                  header = TRUE,
+                  sep = ",")
+agc <- psp %>%
+  group_by(type) %>%
+  summarize(mean = mean(above.ground.C.stock..Mg.C.ha.1.),
+            median = median(above.ground.C.stock..Mg.C.ha.1.),
+            sd = sd(above.ground.C.stock..Mg.C.ha.1.))
+
+# test if the two plot sets (edge and mixed) come from
+# significantly different populations or not
+df <- psp %>%
+  filter(type == "edge" | type == "mixed") %>%
+  mutate(species.richness = as.double(species.richness))
+wilcox.test(above.ground.C.stock..Mg.C.ha.1. ~ type, data = df)
+wilcox.test(stem.density..ha.1. ~ type, data = df)
+wilcox.test(basal.area....m2.ha.1. ~ type, data = df)
+wilcox.test(species.richness ~ type, data = df)
+# p > 0.05, same population
+
+# select patches
+p <- change_map >= 3
+p[p==0] <- NA
+
+# buffer the non-forest areas by 400m (extending into forested areas)
+# See phillips et al. 2006 / Gascon et al. 2000 for rational
+# (few changes after 400m in biomass) and the survey methodology
+#pe <- buffer(p, 400, doEdge=TRUE)
+
+# report 200m in the final paper, as AGC not significantly different
+# for edge plots (rerun)
+pe <- buffer(p, 200, doEdge=TRUE)
+
+# calculate edge forest
+edge_values <- change_map == 1 & pe == 1
+edge_values[is.na(edge_values)] <- 0
+
+# print the map frequency statistics
+change_stats <- as.data.frame(freq(change_map, useNA='no'))
+
+# add a line for edges
+change_stats <- rbind(change_stats,c(5,freq(edge_values)[2,2]))
+
+# convert from pixel count to square km/m and ha
+# assuming a pixel size of ~30m
+change_stats$sq_km <- change_stats$count * 0.03^2
[…]

### 2019-12-27 analysis/09_random_image_selection.R

@@ -1,0 +1,45 @@
+# select CNN training site from shapefile
+
+# load libraries
+library(raster)
+library(rgdal)
+
+# load orthomosaic to subset
+img <- raster("./data/orthomosaic/yangambi_orthomosaic_modified.tif")
+
+# read in shapefile location digitized in QGIS
+loc <- readOGR("data/cnn/","cnn_data")
+
+# grab the cell values and convert to coordinates (row/col)
+cell_values <- extract(img, loc, cellnumbers = TRUE)[,'cells']
+coords <- data.frame(rowColFromCell(img, cell_values),
+                type = as.character(as.data.frame(loc)$type))
+
+img_size <- 256
+
+#nrow(coords)
+for (i in 1:nrow(coords)){
+  row <- coords[i,'row']
+  col <- coords[i,'col']
+  type <- coords[i,'type']
+
+  img_subset <- crop(img, extent(img,
+                                 row - (img_size + 1),
+                                 row + img_size,
+                                 col - (img_size + 1),
+                                 col + img_size))
+  img_subset[is.na(img_subset)] <- 0
+
+  jpeg(file.path("./data/cnn/images",paste0(type,"_",i,".jpg")),
+      (img_size * 2) + 1,
+      (img_size * 2) + 1,
+      quality = 100)
+  par(oma=rep(0,4), mar=rep(0,4))
+  image(img_subset, col = grey(1:255/256),
+        xaxt = 'n',
+        yaxt = 'n',
+        xlab = '',
+        ylab = '',
+        bbox = 'n')
+  dev.off()
+}

### 2019-12-27 analysis/10_cnn_gfc_geo-eye_validation_map.R

@@ -1,0 +1,87 @@
+# plot comparison of spectral and spatial
+# classifiers
+
+# load raster library
+library(raster)
+library(tidyverse)
+library(cowplot)
+
+# read in cnn forest mask
+m <- raster("./data/geo-eye/geo-eye_forest_mask_cnn.tif")
+
+# hansen
+hans <- raster("./data/Hansen_et_al/Hansen_GFC_lossyear_yangambi.tif")
+
+# read in Hansen map (get all forest loss before 2013)
+hans <- raster("data/Hansen_et_al/Hansen_GFC_lossyear_yangambi.tif")
+hans <- crop(hans, extent(m))
+loss <- (hans <= 12) + (hans == 0)
+loss <- (loss > 1)
+loss[is.na(loss)] <- 4
+loss[loss == 1] <- 5
+
+# combine maps
+loss_final <- loss + m
+
+# read in geo-eye downsampled data
+r <- raster("data/geo-eye/geo-eye_panchromatic_latlon_resampled_8bit.tif")
+geoeye_r <- aggregate(r, fact = 6)
+
+geoeye_df <- geoeye_r %>%
+  rasterToPoints %>%
+  as.data.frame() %>%
+  `colnames<-`(c("x", "y", "val"))
+
+class_map <- loss_final %>%
+  rasterToPoints %>%
+  as.data.frame() %>%
+  `colnames<-`(c("x", "y", "val"))
+
+geoeye_map <- ggplot()+
+  geom_tile(data = geoeye_df, aes(x=x,y=y,fill=val)) +
+  scale_fill_gradient(low = "black", high = "white") +
+  coord_fixed(ratio = 1) +
+  theme_minimal() +
+  theme(legend.position="none",
+        axis.text = element_text(size = 13),
+        axis.text.x = element_text(angle = 45, hjust = 1),
+        panel.grid.major = element_line(colour="grey"),
+        panel.grid.minor = element_line(colour="grey")) +
+  labs(
+    x = "",
+    y = "")
+
+spatial_map <- ggplot()+
+  geom_tile(data = class_map,
+            aes(x=x,y=y,fill=as.factor(val))) +
+  scale_fill_brewer(palette = "Paired",
+                    labels=c("GFC + CNN",
+                             "GFC",
+                             "CNN",
+                             "forest")) +
+  coord_fixed(ratio = 1) +
+  theme_minimal() +
+  theme(legend.title = element_blank(),
+        legend.position="bottom",
[…]

### 2019-12-27 src/clahe.py

@@ -1,0 +1,49 @@
+#!/usr/bin/env python
+
+import numpy as np
+import cv2
+import argparse
+
+# argument parser
+def getArgs():
+
+   parser = argparse.ArgumentParser(
+    description = '''Applies local histogram equalization to an image''')
+
+   parser.add_argument('-i',
+                       '--input',
+                       help = 'input file')
+
+   parser.add_argument('-o',
+                       '--output',
+                       help = 'where to store the file')
+                       
+   parser.add_argument('-w',
+                       '--windowsize',
+                       help = 'window size for analysis',
+                       default = 32)
+                       
+   parser.add_argument('-c',
+                       '--cliplimit',
+                       help = 'clip limit',
+                       default = 1.5)
+                       
+   return parser.parse_args()
+
+
+# main routine
+if __name__ == '__main__':
+  
+  # parse arguments
+  args = getArgs()
+
+  # read input image
+  img = cv2.imread(args.input,0)
+
+  # create a CLAHE object (Arguments are optional).
+  clahe = cv2.createCLAHE(clipLimit = float(args.cliplimit),
+   tileGridSize = (int(args.windowsize),int(args.windowsize)))
+  cl1 = clahe.apply(img)
+  
+  # write output to file
+  cv2.imwrite(args.output,cl1)

### 2019-12-27 src/enhance_contrast.sh

@@ -1,0 +1,10 @@
+#!/bin/bash
+
+files=`find /scratch/cobecore/aerial_photographs/png -name "*.png"`
+
+for i in $files;do
+    file=`basename $i`
+    output=`echo /scratch/cobecore/aerial_photographs/png_clahe/$file`
+    echo processing: $output
+    ./clahe.py -i $i -o $output -w 32 -c 1.5
+done

### 2019-12-27 src/fiducial_detection_hough.py

@@ -1,0 +1,187 @@
+#!/usr/bin/env python
+
+# Import necessary libraries
+import os, argparse, glob
+from os.path import basename
+import cv2
+import numpy as np
+from tqdm import tqdm
+
+# resized images have 7700 pixels between fudicial points
+# corners can be masked by a 350x350 section
+
+# argument parser
+def getArgs():
+
+   parser = argparse.ArgumentParser(
+    description = '''Find fudicial marks based upon hough transforms.''',
+    epilog = '''post bug reports to the github repository''')
+    
+   parser.add_argument('-f',
+                       '--files',
+                       help = 'location of a file to process') 
+    
+   parser.add_argument('-d',
+                       '--directory',
+                       help = 'location of the data to match',
+                       default = '/scratch/cobecore/aerial_photographs/png/path_4/')
+
+   parser.add_argument('-o',
+                       '--output_directory',
+                       help = 'location where to store the data',
+                       default = '/scratch/cobecore/aerial_photographs/auto_crop/path_4/')
+
+   parser.add_argument('-c',
+                       '--clahe',
+                       help = 'apply CLAHE histogram normalization',
+                       default = True)
+
+   parser.add_argument('-w',
+                       '--window',
+                       help = 'CLAHE window size',
+                       default = 32)
+                       
+   parser.add_argument('-m',
+                       '--mask',
+                       help = 'corner mask size, will return mask_file.png',
+                       default = 350)                       
+
+   parser.add_argument('-b',
+                       '--border',
+                       help = 'border to retain for evaluation, rest will be set to 0',
+                       default = 1000)
+
+   parser.add_argument('-s',
+                       '--size',
+                       help = 'resize value',
+                       default = 7700)
+                       
+   parser.add_argument('-minr',
+                       '--min_radius',
+                       help = 'hough min radius value',
+                       default = 5)
+                       
+   parser.add_argument('-maxr',
+                       '--max_radius',
+                       help = 'hough min radius value',
+                       default = 11)
+                       
+   parser.add_argument('-sm',
+                       '--smooth',
+                       help = 'smoothing / bluring value',
+                       default = 7)
+   return parser.parse_args()
+
+if __name__ == '__main__':
+
+  # parse arguments
+  args = getArgs()
+  
+  if args.files is not None:
+   files = [args.files]
+  else:
+   # list image files to be processed
+   files = sorted(glob.glob(args.directory + "*.png"))
[…]

### 2019-12-27 src/generate_corner_mask.py

@@ -1,0 +1,48 @@
+#!/usr/bin/env python
+
+# Import necessary libraries
+import cv2, argparse
+import numpy as np
+
+# Generates corner mask in the current working directory
+
+# argument parser
+def getArgs():
+
+   parser = argparse.ArgumentParser(
+    description = '''Create corner mask.''',
+    epilog = '''post bug reports to the github repository''')
+
+   parser.add_argument('-s',
+                       '--size',
+                       help = 'resize value',
+                       default = 7700)
+
+   parser.add_argument('-m',
+                       '--mask',
+                       help = 'corner mask size, will return mask_file.png',
+                       default = 350)
+   
+   return parser.parse_args()        
+
+if __name__ == '__main__':
+
+  # parse arguments
+  args = getArgs()
+    
+  # First create the image with alpha channel
+  mask = np.full((args.size, args.size), 0)
+                
+  # set start and end columns of data columns
+  start = args.mask
+  end = args.size - args.mask
+  mask[start:end,:] = 255
+  mask[:,start:end] = 255
+  
+  # set correct type
+  mask = mask.astype(np.uint8)
+  
+  # write mask to file
+  cv2.imwrite("mask_file.png",
+              mask,
+              [cv2.IMWRITE_PNG_COMPRESSION, 0])

### 2019-12-27 src/georeference_maps.sh

@@ -1,0 +1,15 @@
+#!/bin/bash
+
+# conversion georeference scripts using ground control points
+# on buildings, river crossings and constrained by river intersections at
+# the outer boundaries nearest neighbour using a polynomial of order 3
+
+# Polynomial order 3
+
+# TPS mean error
+# orthomosaic: 11.5477 m / pixels
+# DEM: 0.22134 m / pixels
+
+gdal_translate -of GTiff -gcp 24.4884 -0.754953 24.4885 0.754908 -gcp 24.4631 -0.762446 24.4631 0.762428 -gcp 24.4449 -0.765972 24.4449 0.766006 -gcp 24.3953 -0.774552 24.3957 0.774467 -gcp 24.4565 -0.800831 24.4565 0.800806 -gcp 24.4549 -0.780972 24.455 0.780964 -gcp 24.5942 -0.761086 24.5942 0.761027 -gcp 24.4586 -0.831196 24.4586 0.831191 -gcp 24.4424 -0.834813 24.4424 0.834928 -gcp 24.4638 -0.868746 24.4638 0.868703 -gcp 24.4885 -1.04527 24.4886 1.04534 -gcp 24.4932 -1.04606 24.4933 1.04611 -gcp 24.4073 -1.01719 24.4073 1.01752 -gcp 24.4428 -0.767977 24.4427 0.767981 -gcp 24.3699 -0.925589 24.3705 0.925278 -gcp 24.6132 -0.92235 24.6127 0.922348 -gcp 24.4341 -0.731394 24.4341 0.731337 "yangambi_orthomosaic.tif" "/tmp/yangambi_orthomosaic.tif"
+gdalwarp -r near -order 3 -co COMPRESS=DEFLATE  "/tmp/yangambi_orthomosaic.tif" "yangambi_orthomosaic_modified_poly3.tif"
+

### 2019-12-30 analysis/00_lulcc_analysis.R
zenodo

@@ -9,6 +9,13 @@
 # load required libraries
 library(raster)
 
+# download full resolution file from Zenodo repo
+# if the data does not exist (cloned project from github)
+if(!file.exists("data/orthomosaic/yangambi_orthomosaic_modified.tif")){
+download.file("download. https://zenodo.org/api/files/8039915b-30ea-4b2c-a799-04ef0aadf0ec/yangambi_orthomosaic.tif?versionId=7003f8f2-a826-496e-b4f1-698d2e89b76e",
+              "data/orthomosaic/yangambi_orthomosaic_modified.tif")
+}
+
 # read outline of the mosaic
 m <- raster("data/orthomosaic/yangambi_orthomosaic_mask_resampled.tif")
 

### 2019-12-30 manuscript/appendix.Rmd
appendix changes

@@
 
-  frag_stats <- rbind(landscape_stats(historical),
-                 landscape_stats(contemporary))
 ```
@@
 
+\pagebreak
+
 ```{r echo=FALSE, out.height='75%', fig.cap="Overview of the complete flight plan meta-data as stored in the archives at the Africa Museum."}
@@
 ```
+
+\pagebreak
 
@@
 ```
+
+\pagebreak
+
+```{r echo=FALSE, out.height='75%', fig.cap="CNN classifier results as run on a recent (2012) Geo-Eye panchromatic image. Results are compared with the Landsat based ."}
+knitr::include_graphics('./figures/visual_comparison_classifiers.png')
+```
+
+```{r echo = FALSE, eval = TRUE}
+# this data will not be available in the github repo due to missing geo-eye data
+
+# read in cnn forest mask
+m <- raster("../data/geo-eye/geo-eye_forest_mask_cnn.tif")
+
+# hansen
+hans <- raster("../data/Hansen_et_al/Hansen_GFC_lossyear_yangambi.tif")
+
+# read in Hansen map (get all forest loss before 2013)
+hans <- raster("../data/Hansen_et_al/Hansen_GFC_lossyear_yangambi.tif")
+hans <- crop(hans, extent(m))
+loss <- (hans <= 12) + (hans == 0)
+loss <- (loss > 1)
+
+ct <- crosstab(m, loss)
+ct <- round(ct/sum(ct),2)
+rownames(ct) <- c("non-forest","forest")
+
+ct %>%
+   kable(
+    caption="Contingency table between the two forest / non-forest maps generated from a Geo-Eye 1 pan-chromatic image using a Convolutional Neural Network (CNN) generated and the Global Forest Cover map (Hansen et al.
+2013).",
+    booktabs = TRUE,
+    col.names =  c("non-forest","forest")) %>%
+  add_header_above(c("CNN" = 1, "Global Forest Cover" = 2)) %>%
+  kable_styling(latex_options = c("hold_position"))
+
+```

### 2019-12-30 manuscript/appendix.tex
appendix changes

@@
 \urlstyle{same}  % don't use monospace font for urls
-\usepackage{graphicx}
-% grffile has become a legacy package: https://ctan.org/pkg/grffile
-\IfFileExists{grffile.sty}{%
-\usepackage{grffile}
-}{}
+\usepackage{graphicx,grffile}
 \makeatletter
@@
 
+\pagebreak
+
 \begin{figure}
-\includegraphics[width=82.67in,height=0.75\textheight]{./figures/flight_paths_meta_data} \caption{Overview of the complete flight plan meta-data as stored in the archives at the Africa Museum.}\label{fig:unnamed-chunk-4}
+\includegraphics[height=0.75\textheight]{./figures/flight_paths_meta_data} \caption{Overview of the complete flight plan meta-data as stored in the archives at the Africa Museum.}\label{fig:unnamed-chunk-4}
 \end{figure}
+
+\pagebreak
 
@@
 
+\pagebreak
+
+\begin{figure}
+\includegraphics[width=31.25in,height=0.75\textheight]{./figures/visual_comparison_classifiers} \caption{CNN classifier results as run on a recent (2012) Geo-Eye panchromatic image.
+Results are compared with the Landsat based .}\label{fig:unnamed-chunk-9}
+\end{figure}
+
+\begin{table}[!h]
+
+\caption{\label{tab:unnamed-chunk-10}Contingency table between the two forest / non-forest maps generated from a Geo-Eye 1 pan-chromatic image using a Convolutional Neural Network (CNN) generated and the Global Forest Cover map (Hansen et al.
+2013).}
+\centering
+\begin{tabular}[t]{lrr}
+\toprule
+\multicolumn{1}{c}{CNN} & \multicolumn{2}{c}{Global Forest Cover} \\
+\cmidrule(l{3pt}r{3pt}){1-1} \cmidrule(l{3pt}r{3pt}){2-3}
+  & non-forest & forest\\
+\midrule
+non-forest & 0.10 & 0.09\\
+forest & 0.04 & 0.77\\
+\bottomrule
+\end{tabular}
+\end{table}
+
 \end{document}

### 2019-12-30 manuscript/main.Rmd
appendix changes

@@
    "Loss > 1958 (permanent)")
-
 ```
@@
 We automatically delineated all natural forest in the historical data, thus excluding tree plantations, thinned or deteriorated forest stands, fields and buildings.
-We used the Unet Convolutional Neural Net (CNN, @ronneberger2015) architecture implemented in Keras [@chollet2015] with an efficientnetb3 backbone [@yakubovskiy2019] running on TensorFlow [@martinabadi2015] to train a binary classifier (i.e.
+We used the Unet Convolutional Neural Net (CNN, @ronneberger2015) architecture implemented in Keras [@chollet2015] with an efficientnetb3 pre-processing backbone [@yakubovskiy2019] running on TensorFlow [@martinabadi2015] to train a binary classifier (i.e.
 forest or non-forested).
@@
 We applied this technique to 10% of the generated synthetic landscapes.
-The CNN model was trained for 100 epochs on a graphics processing unit (GPU) maximizing the Intersect-over-Union (IoU) using additional data augmentation.
+The CNN model was trained for 100 epochs with a batch size of 30 using Adam optimization [@kingma2017], maximizing the Intersect-over-Union (IoU) using Sørensen–Dice and categorical cross-entropy loss functions.
 Data augmentation included random cropping to 320 pixel squares, random orientation, scaling, perspective, contrast and brightness shifts and image blurring.
@@
 We refer to Figure 4 for a synoptic overview of the full deep learning learning workflow.
-
-More details on the optimizer and other metrics etc.
+In addition, we used the first acquisition of a recent pan-chromatic Geo-Eye 1 stereo pair (GeoEye, 2012, Thornton, Colorado, U.S.A., order 737537, 2011-11-11 8:55h GMT or 9:55h local time) to classify and visually assess the robustness of the algorithm on contemporary remote sensing imagery with the same moving window approach.
 
@@
 Where the datasets overlap we provide summary statistics of deforestation, reforestation and afforestation.
+The same methodology was used to quantify CNN model performance across the Geo-Eye 1 scene.
 
@@
 
-We used an area of 400x400m (16 ha) around contemporary permanent sampling plot locations to quantify the general state of the canopy in both historical and contemporary remote sensing data.
-Contemporary data were resampled to the resolution of the historical data using a nearest neighbour interpolation for comparison.
-We used the first acquisition of a pan-chromatic Geo-Eye 1 stereo pair (GeoEye, 2012, Thornton, Colorado, U.S.A., order 737537, 2011-11-11 8:55h GMT or 9:55h local time).
+We used an area of 400x400m (16 ha) around contemporary permanent sampling plot locations to quantify the general state of the canopy in both historical and contemporary pan-chromatic Geo-Eye 1 remote sensing data (GeoEye, 2012, Thornton, Colorado, U.S.A., order 737537, 2011-11-11 8:55h GMT or 9:55h local time).
+Geo-Eye data were resampled to the resolution of the historical data using a nearest neighbour interpolation for comparison.
 For permanent sampling plot locations in both historical and contemporary data a rectangular buffer of 200m around the centre of all plots was used to create small image subsets.
@@
 Our CNN deep learning classifier reached an Intersection-over-Union (IoU) accuracy of 97% on the detection of disturbed forest in out-of-sample test data.
+A visual comparison on recent pan-chromatic Geo-Eye 1 data shows good agreement between the landsat based GFC data and downscaled CNN results (Appendix Figure 7).
+
+[NOTE STATS]
+
 Scaling our classifier to the whole historical orthomosaic we detected 162 km^2^ (or ~20% of the scene) of disturbed forests.

### 2019-12-30 analysis/10_cnn_gfc_geo-eye_validation_map.R

@@ -55,16 +55,16 @@ spatial_map <- ggplot()+
   geom_tile(data = class_map,
             aes(x=x,y=y,fill=as.factor(val))) +
   scale_fill_brewer(palette = "Paired",
-                    labels=c("GFC + CNN",
-                             "GFC",
-                             "CNN",
-                             "forest")) +
+                    labels=c("non-forest (GFC + CNN)",
+                             "non-forest (GFC)",
+                             "non-forest (CNN)",
+                             "forest (GFC + CNN)")) +
   coord_fixed(ratio = 1) +
   theme_minimal() +
   theme(legend.title = element_blank(),
         legend.position="bottom",
         legend.direction="horizontal",
-        axis.text = element_text(size = 13),
+        axis.text = element_text(size = 8),
         axis.text.x = element_text(angle = 45, hjust = 1),
         panel.grid.major = element_line(colour="grey"),
         panel.grid.minor = element_line(colour="grey")) +

### 2019-12-30 manuscript/main.Rmd
patch metrics, upper quantile threshold to limit undue influence of very large patches

@@
   df <- df %>%
-    filter(area > 1) %>%
+    filter(area > 1,
+           area < quantile(area, 0.95)) %>%
     summarize(
       perimeter = paste0(
-        round(mean(perimeter)),
+        round(mean(perimeter),2),
         " $\\pm$ ",
-        round(sd(perimeter))),
+        round(sd(perimeter),2)),
       area = paste0(
-        round(mean(area)),
+        round(mean(area), 2),
         " $\\pm$ ",
-        round(sd(area))),
+        round(sd(area), 2)),
       ratio = paste0(
-        round(mean(pa_ratio),3),
+        round(mean(pa_ratio), 2),
         " $\\pm$ ",
-        round(sd(pa_ratio),3)),
+        round(sd(pa_ratio), 2)),
       fractal = paste0(
-        round(mean(fractal_index),3),
+        round(mean(fractal_index), 2),
         " $\\pm$ ",
-        round(sd(fractal_index),3))
+        round(sd(fractal_index), 2))
     )
@@
 We report the ratio of edge to area and the fractal index to quantify landscape complexity of forest disturbances.
-Statistics were calculated for all forest disturbance patches larger than 1 ha using the R package landscapemetrics [@hesselbarth2019].
+Statistics were calculated for all forest disturbance patches larger than 1 ha and smaller than the 95^th^ percentile of the patch size distribution using the R package landscapemetrics [@hesselbarth2019].
 We provide mean and standard deviation on edge, area, their ratio and fractal dimension for both the historical and contemporary @hansen2013 forest cover maps.
@@
 Our CNN deep learning classifier reached an Intersection-over-Union (IoU) accuracy of 97% on the detection of disturbed forest in out-of-sample test data.
-A visual comparison on recent pan-chromatic Geo-Eye 1 data shows good agreement between the landsat based GFC data and downscaled CNN results (Appendix Figure 7).
-
-[NOTE STATS]
-
+A visual comparison on recent pan-chromatic Geo-Eye 1 data shows good agreement, for a total of 87% of all pixels, between the landsat based GFC data and downscaled CNN results (Appendix Table 3 and Figure 7).
 Scaling our classifier to the whole historical orthomosaic we detected 162 km^2^ (or ~20% of the scene) of disturbed forests.
@@
 These differences are reflected in the analysis of landscape metrics of forest loss.
-Between the historical and contemporary LULCC maps we see an increase in small disturbances, as indicated by the decreasing area of the mean patch size, down to 37 ha from 120 ha historically, while perimeter lengths remained approximately the same, with 8653m and 9902m for historical and contemporary maps respectively (Table 2).
-This shift in perimeter area ratio led to a similar change in the fractal index, increasing in value to 1.13 $\pm$ 0.07 from 1.10 $\pm$ 0.05 over time.
+Between the historical and contemporary LULCC maps we see an increase in small disturbances, as indicated by the decreasing area of the mean patch size, down to ~1.86 $\pm$ 0.75 ha from ~5.25 $\pm$ 5.02 ha historically.
+Perimeter lengths were longer historically, at 1451.58 $\pm$ 943.27 m,  compared to contemporary landscapes ~921.74 $\pm$ 362.59 m (Table 2).
+This shift in perimeter area ratio led to a similar change in the fractal index, increasing in value to 1.1 $\pm$ 0.05 from 1.09 $\pm$ 0.04 over time.
 Values closer to a fractal index of 2 suggest a more complex (fragmented) landscape.
@@
 Recovery throughout the region was characterized for patches of forest and plantations.
-Assuming high density stands, based on previous work, this could amount to a carbon gains of 1592 Gg C, offsetting more recent losses of approximately 1408 Gg C.
+Assuming high density stands, based on previous work, this could amount to a carbon gains of 1592 Gg C across our study area, offsetting more recent losses of approximately 1408 Gg C.
 On the other hand, at the low end, if we assume a lower carbon density of 81.8 Mg C ha^-1^ this would result in a total carbon gain of 811 Gg C.

### 2019-12-30 manuscript/main.tex
patch metrics, upper quantile threshold to limit undue influence of very large patches

@@
 resolution and geo-referenced to \textasciitilde{}4.7 \(\pm\) 4.3 m.
-Primary forest in the orthomosaic was classified with a Convoluted
+Primary forest in the orthomosaic was classified with a Convolutional
 Neural Network Deep Learning approach. We used these data to quantify
@@
 thus excluding tree plantations, thinned or deteriorated forest stands,
-fields and buildings. We used the Unet Convoluted Neural Net (CNN,
+fields and buildings. We used the Unet Convolutional Neural Net (CNN,
 \citet{ronneberger2015}) architecture implemented in Keras
-\citep{chollet2015} with an efficientnetb3 backbone
+\citep{chollet2015} with an efficientnetb3 pre-processing backbone
 \citep{yakubovskiy2019} running on TensorFlow \citep{martinabadi2015} to
@@
 this technique to 10\% of the generated synthetic landscapes. The CNN
-model was trained for 100 epochs on a graphics processing unit (GPU)
-maximizing the Intersect-over-Union (IoU) using additional data
-augmentation. Data augmentation included random cropping to 320 pixel
-squares, random orientation, scaling, perspective, contrast and
-brightness shifts and image blurring. During final model evaluation we
-report the IoU of our out-of-sample test datasets. The optimized model
-was used to classify the complete orthomosaic using a moving window
-approach with a step size of 110 pixels and a majority vote across
-overlapping areas to limit segmentation edge effects. We refer to Figure
-4 for a synoptic overview of the full deep learning learning workflow.
+model was trained for 100 epochs with a batch size of 30 using Adam
+optimization \citep{kingma2017}, maximizing the Intersect-over-Union
+(IoU) using Sørensen--Dice and categorical cross-entropy loss functions.
+Data augmentation included random cropping to 320 pixel squares, random
+orientation, scaling, perspective, contrast and brightness shifts and
+image blurring. During final model evaluation we report the IoU of our
+out-of-sample test datasets. The optimized model was used to classify
+the complete orthomosaic using a moving window approach with a step size
+of 110 pixels and a majority vote across overlapping areas to limit
+segmentation edge effects. We refer to Figure 4 for a synoptic overview
+of the full deep learning learning workflow. In addition, we used the
+first acquisition of a recent pan-chromatic Geo-Eye 1 stereo pair
+(GeoEye, 2012, Thornton, Colorado, U.S.A., order 737537, 2011-11-11
+8:55h GMT or 9:55h local time) to classify and visually assess the
+robustness of the algorithm on contemporary remote sensing imagery with
+the same moving window approach.
 
@@
 bank of the Congo river. Where the datasets overlap we provide summary
-statistics of deforestation, reforestation and afforestation.
+statistics of deforestation, reforestation and afforestation. The same
+methodology was used to quantify CNN model performance across the
+Geo-Eye 1 scene.
 
@@
 Statistics were calculated for all forest disturbance patches larger
-than 1 ha using the R package landscapemetrics \citep{hesselbarth2019}.
-We provide mean and standard deviation on edge, area, their ratio and
-fractal dimension for both the historical and contemporary
-\citet{hansen2013} forest cover maps.
+than 1 ha and smaller than the 95\textsuperscript{th} percentile of the
+patch size distribution using the R package landscapemetrics
+\citep{hesselbarth2019}. We provide mean and standard deviation on edge,
+area, their ratio and fractal dimension for both the historical and
+contemporary \citet{hansen2013} forest cover maps.
 
@@
 sampling plot locations to quantify the general state of the canopy in
-both historical and contemporary remote sensing data. Contemporary data
-were resampled to the resolution of the historical data using a nearest
-neighbour interpolation for comparison. We used the first acquisition of
-a pan-chromatic Geo-Eye 1 stereo pair (GeoEye, 2012, Thornton, Colorado,
-U.S.A., order 737537, 2011-11-11 8:55h GMT or 9:55h local time). For
-permanent sampling plot locations in both historical and contemporary
-data a rectangular buffer of 200m around the centre of all plots was
-used to create small image subsets. All image subsets were processed
-using the R based FOTO implementation and package \citep{hufkens2019a}
-using global normalization, ensuring intercomparability of texture
-metrics across time and space \citep{Barbier2010}. We used a FOTO
-(moving) window size of 187 pixels (or \textasciitilde{}150m), as
-described in \citet{solorzano2018} and \citet{Barbier2010}, to ensure
-that multiple individual canopies could be included in the analysis. A
-buffer with a radius of 50m around the center of all permanent sampling
-plots was used to determine the average value of the first principle
-component (PC), explaining the majority of the variance in canopy
-texture. Where both the Geo-Eye data and the orthomosaic image
-overlapped we processed the intersecting region to explore a
-wall-to-wall comparison between past and current canopy texture metrics.
+both historical and contemporary pan-chromatic Geo-Eye 1 remote sensing
+data (GeoEye, 2012, Thornton, Colorado, U.S.A., order 737537, 2011-11-11
+8:55h GMT or 9:55h local time). Geo-Eye data were resampled to the
+resolution of the historical data using a nearest neighbour
+interpolation for comparison. For permanent sampling plot locations in
+both historical and contemporary data a rectangular buffer of 200m
+around the centre of all plots was used to create small image subsets.
+All image subsets were processed using the R based FOTO implementation
+and package \citep{hufkens2019a} using global normalization, ensuring
+intercomparability of texture metrics across time and space
+\citep{Barbier2010}. We used a FOTO (moving) window size of 187 pixels
+(or \textasciitilde{}150m), as described in \citet{solorzano2018} and
+\citet{Barbier2010}, to ensure that multiple individual canopies could
+be included in the analysis. A buffer with a radius of 50m around the
+center of all permanent sampling plots was used to determine the average
+value of the first principle component (PC), explaining the majority of
+the variance in canopy texture. Where both the Geo-Eye data and the
+orthomosaic image overlapped we processed the intersecting region to
+explore a wall-to-wall comparison between past and current canopy
+texture metrics.
 
@@
 (IoU) accuracy of 97\% on the detection of disturbed forest in
-out-of-sample test data. Scaling our classifier to the whole historical
+out-of-sample test data. A visual comparison on recent pan-chromatic
+Geo-Eye 1 data shows good agreement, for a total of 87\% of all pixels,
+between the landsat based GFC data and downscaled CNN results (Appendix
+Table 3 and Figure 7). Scaling our classifier to the whole historical
 orthomosaic we detected 162 km\textsuperscript{2} (or
@@
 contemporary LULCC maps we see an increase in small disturbances, as
-indicated by the decreasing area of the mean patch size, down to 37 ha
-from 120 ha historically, while perimeter lengths remained approximately
-the same, with 8653m and 9902m for historical and contemporary maps
-respectively (Table 2). This shift in perimeter area ratio led to a
-similar change in the fractal index, increasing in value to 1.13 \(\pm\)
-0.07 from 1.10 \(\pm\) 0.05 over time. Values closer to a fractal index
-of 2 suggest a more complex (fragmented) landscape.
+indicated by the decreasing area of the mean patch size, down to
+\textasciitilde{}1.86 \(\pm\) 0.75 ha from \textasciitilde{}5.25 \(\pm\)
+5.02 ha historically. Perimeter lengths were longer historically, at
+1451.58 \(\pm\) 943.27 m, compared to contemporary landscapes
+\textasciitilde{}921.74 \(\pm\) 362.59 m (Table 2). This shift in
+perimeter area ratio led to a similar change in the fractal index,
+increasing in value to 1.1 \(\pm\) 0.05 from 1.09 \(\pm\) 0.04 over
+time. Values closer to a fractal index of 2 suggest a more complex
+(fragmented) landscape.
 
@@
 density stands, based on previous work, this could amount to a carbon
-gains of 1592 Gg C, offsetting more recent losses of approximately 1408
-Gg C. On the other hand, at the low end, if we assume a lower carbon
-density of 81.8 Mg C ha\textsuperscript{-1} this would result in a total
-carbon gain of 811 Gg C. Using our approach results indicate that
-overall deforestation around Yangambi has resulted in a loss of
-\textasciitilde{}2416 Gg C in AGC stocks.
+gains of 1592 Gg C across our study area, offsetting more recent losses
+of approximately 1408 Gg C. On the other hand, at the low end, if we
+assume a lower carbon density of 81.8 Mg C ha\textsuperscript{-1} this
+would result in a total carbon gain of 811 Gg C. Using our approach
+results indicate that overall deforestation around Yangambi has resulted
+in a loss of \textasciitilde{}2416 Gg C in AGC stocks.
 
@@
 \midrule
-historical & 8653 $\pm$ 35676 & 120 $\pm$ 659 & 0.031 $\pm$ 0.014 & 1.099 $\pm$ 0.052\\
-contemporary & 9902 $\pm$ 56975 & 37 $\pm$ 226 & 0.046 $\pm$ 0.014 & 1.13 $\pm$ 0.071\\
+historical & 1451.58 $\pm$ 943.27 & 5.25 $\pm$ 5.02 & 0.03 $\pm$ 0.01 & 1.09 $\pm$ 0.04\\
+contemporary & 921.74 $\pm$ 362.59 & 1.86 $\pm$ 0.75 & 0.05 $\pm$ 0.01 & 1.1 $\pm$ 0.05\\
 \bottomrule
@@
 When classifying our orthomosaic into forest and non-forest states we
-favoured a deep learning supervised classification using a Convoluted
+favoured a deep learning supervised classification using a Convolutional
 Neural Network over manual segmentation to guarantee an
@@
 
-All analysis code is available as an R \citep{rcoreteam2019} project
-(\url{https://github.com/khufkens/orthodrc}). The analysis relied
+All analysis code is available as an R / python \citep{rcoreteam2019}
+projects (\url{https://khufkens.github.io/orthodrc} \&
+\url{https://khufkens.github.io/orthodrc_cnn/}). The analysis relied
 heavily on the `raster' \citep{hijmans2019}, `RStoolbox'

### 2019-12-30 manuscript/appendix.Rmd
CNN stats and plot of Geo-Eye panchrom image

@@
 ct <- crosstab(m, loss)
-ct <- round(ct/sum(ct),2)
+ct <- round(ct/sum(ct),2) * 100
 rownames(ct) <- c("non-forest","forest")
@@
    kable(
-    caption="Contingency table between the two forest / non-forest maps generated from a Geo-Eye 1 pan-chromatic image using a Convolutional Neural Network (CNN) generated and the Global Forest Cover map (Hansen et al.
+    caption="Contingency table in agreement (%) between forest / non-forest classes as generated from a Geo-Eye 1 pan-chromatic image using a Convolutional Neural Network (CNN) generated and the Global Forest Cover map (Hansen et al.
 2013).",

### 2019-12-30 manuscript/appendix.tex
CNN stats and plot of Geo-Eye panchrom image

@@
 
-\begin{table}[!h]
+\textbackslash{}begin\{table\}{[}!h{]}
 
-\caption{\label{tab:unnamed-chunk-10}Contingency table between the two forest / non-forest maps generated from a Geo-Eye 1 pan-chromatic image using a Convolutional Neural Network (CNN) generated and the Global Forest Cover map (Hansen et al.
-2013).}
+\textbackslash{}caption\{\label{tab:unnamed-chunk-10}Contingency table
+in agreement (\%) between forest / non-forest classes as generated from
+a Geo-Eye 1 pan-chromatic image using a Convolutional Neural Network
+(CNN) generated and the Global Forest Cover map (Hansen et al.~2013).\}
 \centering
+
 \begin{tabular}[t]{lrr}
@@
 \midrule
-non-forest & 0.10 & 0.09\\
-forest & 0.04 & 0.77\\
+non-forest & 10 & 9\\
+forest & 4 & 77\\
 \bottomrule
 \end{tabular}
-\end{table}
+
+\textbackslash{}end\{table\}
 

### 2019-12-30 analysis/07_foto_map.R
fat symbols

@@ -35,9 +35,9 @@ if(!file.exists("./data/geo-eye/geo-eye_panchromatic_latlon_resampled.tif")){
 }
 
 # aggregate (lower resolution for plotting)
-geoeye_r <- aggregate(geoeye_r, fact = 4)
+geoeye_r <- aggregate(geoeye_r, fact = 6)
 historical <- crop(historical, extent(geoeye_r))
-historical <- aggregate(historical, fact = 4)
+historical <- aggregate(historical, fact = 6)
 geoeye_r <- histMatch(geoeye_r, historical)
 
 # grab extent of the data and crop out locations
@@ -77,7 +77,8 @@ historical_map <- ggplot()+
   scale_fill_gradient(low = "black", high = "white") +
   coord_fixed(ratio = 1) +
   geom_sf(data = s,
-          aes(shape = type, size = 3), colour = "white") +
+          aes(shape = type,
+              size = 3), stroke = 2, colour = "white") +
   scale_shape_manual(values = c(0, 1, 2, 3, 4)) +
   theme_minimal() +
   theme(legend.position="none",
@@ -94,7 +95,7 @@ geoeye_map <- ggplot()+
   scale_fill_gradient(low = "black", high = "white") +
   coord_fixed(ratio = 1) +
   geom_sf(data = s,
-          aes(shape = type, size = 3), colour = "white") +
+          aes(shape = type, size = 3), stroke = 2,colour = "white") +
   scale_shape_manual(values = c(0, 1, 2, 3, 4)) +
   theme_minimal() +
   theme(legend.position="none",
@@ -112,7 +113,7 @@ historical_foto_map <- ggplot()+
   scale_fill_identity() +
   coord_fixed(ratio = 1) +
   geom_sf(data = s,
-          aes(shape = type, size = 3), colour = "white") +
+          aes(shape = type, size = 3), stroke = 2, colour = "white") +
   scale_shape_manual(values = c(0, 1, 2, 3, 4)) +
   theme_minimal()  +
   theme(legend.position="none",
@@ -130,7 +131,7 @@ geoeye_foto_map <- ggplot()+
   scale_fill_identity() +
   coord_fixed(ratio = 1) +
   geom_sf(data = s,
-          aes(shape = type, size = 3), colour = "white") +
+          aes(shape = type, size = 3), stroke = 2, colour = "white") +
   scale_shape_manual(values = c(0, 1, 2, 3, 4)) +
   theme_minimal() +
   theme(legend.position="none",

### 2019-12-31 manuscript/main.Rmd
affiliation correction

@@
   - name: Kim Jacobsen
-    affil: 2,3
+    affil: 1,3
   - name: Hans Beeckman

### 2019-12-31 manuscript/main.tex
affiliation correction

@@
 % Authors, for the paper (add full first names)
-\Author{Koen Hufkens$^{1,2,*}$\href{https://orcid.org/0000-0002-5070-8109}{\orcidicon}, Thalès de Haulleville$^{3}$, Elizabeth Kearsley$^{1}$, Kim Jacobsen$^{2,3}$, Hans Beeckman$^{3}$, Piet Stoffelen$^{4}$, Filip Vandelook$^{4}$, Sofie Meeus$^{4}$, Michael Amara$^{5}$, Leen Van Hirtum$^{5}$, Jan Van den Bulcke$^{1}$, Hans Verbeeck$^{1}$, Lisa Wingate$^{2}$}
+\Author{Koen Hufkens$^{1,2,*}$\href{https://orcid.org/0000-0002-5070-8109}{\orcidicon}, Thalès de Haulleville$^{3}$, Elizabeth Kearsley$^{1}$, Kim Jacobsen$^{1,3}$, Hans Beeckman$^{3}$, Piet Stoffelen$^{4}$, Filip Vandelook$^{4}$, Sofie Meeus$^{4}$, Michael Amara$^{5}$, Leen Van Hirtum$^{5}$, Jan Van den Bulcke$^{1}$, Hans Verbeeck$^{1}$, Lisa Wingate$^{2}$}
 

### 2020-01-23 manuscript/appendix.Rmd
revision

--- 2019-12-27 manuscript/appendix.Rmd
@@
 
-```{r echo=FALSE, out.width='100%', fig.cap="Overview of the complete flight plan of the survey around Kisangani (then Stanleyville) stored in the archives at the Africa Museum."}
+```{r echo=FALSE, out.width='100%', fig.align="center", fig.cap="Overview of the complete flight plan of the survey around Kisangani (then Stanleyville) stored in the archives at the Africa Museum."}
 knitr::include_graphics('./figures/flight_paths_overview.jpg')
@@
 
-```{r echo=FALSE, out.height='75%', fig.cap="Overview of the complete flight plan meta-data as stored in the archives at the Africa Museum."}
+```{r echo=FALSE, out.height='75%', fig.align="center", fig.cap="Overview of the complete flight plan meta-data as stored in the archives at the Africa Museum."}
 knitr::include_graphics('./figures/flight_paths_meta_data.jpg')
@@
 
-```{r echo=FALSE, out.height='50%', fig.cap="Boxplots comparing the first principal component (PC1) of a site based FOTO analysis across different forest types for both contemporary (Geo-Eye) and historical orthomosaic data."}
+```{r echo=FALSE, out.height='75%', fig.align="center",fig.cap="Shelves with aerial survey images at the Africa Museum archives."}
+knitr::include_graphics('./figures/archives_aerial_photos_2.jpg')
+```
+
+```{r echo=FALSE, out.height='50%', fig.align="center",fig.cap="Boxplots comparing the first principal component (PC1) of a site based FOTO analysis across different forest types for both contemporary (Geo-Eye) and historical orthomosaic data."}
 knitr::include_graphics('./figures/foto_bplot_psp.png')
@@
 
-```{r echo = FALSE, out.width="75%", eval = TRUE, fig.cap="Scatterplot comparing Above Ground Carbon and the first principal component of the FOTO analysis for contemporary Geo-Eye (left) and the historical orthomosaic (right) data.
+```{r echo = FALSE, out.width="75%", fig.align="center",eval = TRUE, fig.cap="Scatterplot comparing Above Ground Carbon and the first principal component of the FOTO analysis for contemporary Geo-Eye (left) and the historical orthomosaic (right) data.
 Different forest types are plotted using closed circles, closed triangles, closed squares and this for fallow, mixed mon-dominant and young-regrowth forests, respectivelly "}
@@
 
-```{r echo = FALSE, out.width="75%", eval = TRUE, fig.cap="Scatterplot comparing tree species richness and the first principal component of the FOTO analysis for contemporary Geo-Eye (left) and the historical orthomosaic (right) data.
+```{r echo = FALSE, out.width="75%", fig.align="center",eval = TRUE, fig.cap="Scatterplot comparing tree species richness and the first principal component of the FOTO analysis for contemporary Geo-Eye (left) and the historical orthomosaic (right) data.
 Different forest types are plotted using closed circles, closed triangles, closed squares and this for fallow, mixed mon-dominant and young-regrowth forests, respectivelly"}
@@
 
-```{r echo = FALSE, out.width="75%", eval = TRUE, fig.cap="Scatterplot comparing stem density (per ha) and the first principal component of the FOTO analysis for contemporary Geo-Eye (left) and the historical orthomosaic (right) data.
+```{r echo = FALSE, out.width="75%", fig.align="center",eval = TRUE, fig.cap="Scatterplot comparing stem density (per ha) and the first principal component of the FOTO analysis for contemporary Geo-Eye (left) and the historical orthomosaic (right) data.
 Different forest types are plotted using closed circles, closed triangles, closed squares and this for fallow, mixed mon-dominant and young-regrowth forests, respectivelly"}

### 2020-01-23 manuscript/appendix.tex
revision

--- 2019-12-27 manuscript/appendix.tex
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{./figures/flight_paths_overview} \caption{Overview of the complete flight plan of the survey around Kisangani (then Stanleyville) stored in the archives at the Africa Museum.}\label{fig:unnamed-chunk-3}
+
+{\centering \includegraphics[width=1\linewidth]{./figures/flight_paths_overview}
+
+}
+
+\caption{Overview of the complete flight plan of the survey around Kisangani (then Stanleyville) stored in the archives at the Africa Museum.}\label{fig:unnamed-chunk-3}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=82.67in,height=0.75\textheight]{./figures/flight_paths_meta_data} \caption{Overview of the complete flight plan meta-data as stored in the archives at the Africa Museum.}\label{fig:unnamed-chunk-4}
+
+{\centering \includegraphics[width=82.67in,height=0.75\textheight]{./figures/flight_paths_meta_data}
+
+}
+
+\caption{Overview of the complete flight plan meta-data as stored in the archives at the Africa Museum.}\label{fig:unnamed-chunk-4}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=14.58in,height=0.5\textheight]{./figures/foto_bplot_psp} \caption{Boxplots comparing the first principal component (PC1) of a site based FOTO analysis across different forest types for both contemporary (Geo-Eye) and historical orthomosaic data.}\label{fig:unnamed-chunk-5}
+
+{\centering \includegraphics[width=14.39in,height=0.75\textheight]{./figures/archives_aerial_photos_2}
+
+}
+
+\caption{Shelves with aerial survey images at the Africa Museum archives.}\label{fig:unnamed-chunk-5}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=0.75\linewidth]{./figures/foto_pc1_agc} \caption{Scatterplot comparing Above Ground Carbon and the first principal component of the FOTO analysis for contemporary Geo-Eye (left) and the historical orthomosaic (right) data.
-Different forest types are plotted using closed circles, closed triangles, closed squares and this for fallow, mixed mon-dominant and young-regrowth forests, respectivelly }\label{fig:unnamed-chunk-6}
+
+{\centering \includegraphics[width=14.58in,height=0.5\textheight]{./figures/foto_bplot_psp}
+
+}
+
+\caption{Boxplots comparing the first principal component (PC1) of a site based FOTO analysis across different forest types for both contemporary (Geo-Eye) and historical orthomosaic data.}\label{fig:unnamed-chunk-6}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=0.75\linewidth]{./figures/foto_pc1_diversity} \caption{Scatterplot comparing tree species richness and the first principal component of the FOTO analysis for contemporary Geo-Eye (left) and the historical orthomosaic (right) data.
-Different forest types are plotted using closed circles, closed triangles, closed squares and this for fallow, mixed mon-dominant and young-regrowth forests, respectivelly}\label{fig:unnamed-chunk-7}
+
+{\centering \includegraphics[width=0.75\linewidth]{./figures/foto_pc1_agc}
+
+}
+
+\caption{Scatterplot comparing Above Ground Carbon and the first principal component of the FOTO analysis for contemporary Geo-Eye (left) and the historical orthomosaic (right) data.
+Different forest types are plotted using closed circles, closed triangles, closed squares and this for fallow, mixed mon-dominant and young-regrowth forests, respectivelly }\label{fig:unnamed-chunk-7}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=0.75\linewidth]{./figures/foto_pc1_stems} \caption{Scatterplot comparing stem density (per ha) and the first principal component of the FOTO analysis for contemporary Geo-Eye (left) and the historical orthomosaic (right) data.
+
+{\centering \includegraphics[width=0.75\linewidth]{./figures/foto_pc1_diversity}
+
+}
+
+\caption{Scatterplot comparing tree species richness and the first principal component of the FOTO analysis for contemporary Geo-Eye (left) and the historical orthomosaic (right) data.
 Different forest types are plotted using closed circles, closed triangles, closed squares and this for fallow, mixed mon-dominant and young-regrowth forests, respectivelly}\label{fig:unnamed-chunk-8}
@@
 
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{./figures/foto_pc1_stems}
+
+}
+
+\caption{Scatterplot comparing stem density (per ha) and the first principal component of the FOTO analysis for contemporary Geo-Eye (left) and the historical orthomosaic (right) data.
+Different forest types are plotted using closed circles, closed triangles, closed squares and this for fallow, mixed mon-dominant and young-regrowth forests, respectivelly}\label{fig:unnamed-chunk-9}
+\end{figure}
+
 \end{document}

### 2020-01-23 manuscript/main.Rmd
revision

@@
 Here, we use historical aerial survey photos to map the extent and structure of LULCC around Yangambi, central Congo Basin, in 1958.
-Our study leveraged Structure-from-Motion to generate a large orthomosaic covering 828 km^2^, counting 733 million pixels, at a sub meter resolution and geo-referenced to ~4.7 $\pm$ 4.3 m.
+Our study leveraged Structure-from-Motion to generate a large orthomosaic covering 82800 ha, counting 733 million pixels, at a sub meter resolution and geo-referenced to ~4.7 $\pm$ 4.3 m.
 Primary forest in the orthomosaic was classified with a Convolutional Neural Network Deep Learning approach.
 We used these data to quantify LULCC, landscape and canopy complexity and scale above ground biomass between historical and contemporary condition.
-We show a shift from historical highly structured industrial deforestation in 1958 (162 km^2^), to contemporary smallholder farming clearing (88 km^2^), increasing landscape fragmentation and forest edges.
+We show a shift from historical highly structured industrial deforestation in 1958 (16200 ha), to contemporary smallholder farming clearing (8800 ha), increasing landscape fragmentation and forest edges.
 Efforts to quantify canopy texture and their link to carbon storage had limited to no success.
@@
 
-```{r setup, include=FALSE, message=FALSE}
+```{r setup, include=FALSE, message=FALSE, warning=FALSE}
   knitr::opts_chunk$set(echo = TRUE)
@@
   library(raster)
+  library(sf)
+  library(rgdal)
   library(ggplot2)
@@
   # set kable NA values to blank
-  options(knitr.kable.NA = '-')
+  options(knitr.kable.NA = '')
 
@@
    "Loss > 1958 (permanent)")
+
+  # drop square km values per reviewer request
+  change_stats <- change_stats[,2:3]
+
+# this data will not be available in the github repo due to missing geo-eye data
+
+# read in cnn forest mask
+m <- raster("../data/geo-eye/geo-eye_forest_mask_cnn.tif")
+
+# hansen
+hans <- raster("../data/Hansen_et_al/Hansen_GFC_lossyear_yangambi.tif")
+
+# cloud mask
+cloud_mask <- st_read("../data/maps/geo_eye_cloud_mask.shp")
+
+# read in Hansen map (get all forest loss before 2013)
+hans <- raster("../data/Hansen_et_al/Hansen_GFC_lossyear_yangambi.tif")
+hans <- crop(hans, extent(m))
+loss <- (hans <= 11) + (hans == 0)
+loss <- (loss > 1)
+
+# mask clouds
+m <- mask(m, cloud_mask, inverse = TRUE)
+loss <- mask(loss, cloud_mask, inverse = TRUE)
+
+ct <- crosstab(m, loss)
+ct <- round(ct/sum(ct) * 100,2)
+rownames(ct) <- c("non-forest","forest")
+
+# read in shapefile location digitized in QGIS
+loc <- st_read("../data/cnn/sample_locations.shp")
+disturbed <- loc %>% select(type) %>% filter(type == "dist")
+forest <- loc %>% select(type) %>% filter(type == "forest")
+
+# forest cover map
+r <- raster("../data/orthomosaic/yangambi_forest_mask_qa_resampled.tif")
+r <- (r > 0.5) * 1
+r[is.na(r)] <- 3
+
+# calculate the stats for the whole raster
+forest_stats <- as.data.frame(na.omit(freq(mask(r, forest))))
+forest_stats$count <- round(forest_stats$count/sum(forest_stats$count) * 100,2)
+
+disturbed_stats <- as.data.frame(na.omit(freq(mask(r, disturbed))))
+disturbed_stats$count <- round(disturbed_stats$count/sum(disturbed_stats$count) * 100,2)
+
+valstats <- cbind(disturbed_stats$count, forest_stats$count)
+rownames(valstats) <- c("non-forest","forest")
+
+# calculate the stats for the validation
+# area only
+validation_forest <- loc %>%
+  filter(cnn == "validate",
+         type == "forest") %>%
+  select(type)
+
+validation_non_forest <- loc %>%
+  filter(cnn == "validate",
+         type == "dist") %>%
+  select(type)
+
+forest_stats <- as.data.frame(na.omit(freq(mask(r, validation_forest))))
+forest_stats$count <- round(forest_stats$count/sum(forest_stats$count) * 100,2)
+
+# add a row if forest cover is 100% accurate
+if(forest_stats$count[1] == 100){
+  if(forest_stats$value == 1){
+   forest_stats <- rbind(c(0,0),forest_stats)
+  }
+}
+
+disturbed_stats <- as.data.frame(na.omit(freq(mask(r, validation_non_forest))))
+disturbed_stats$count <- round(disturbed_stats$count/sum(disturbed_stats$count) * 100,2)
+
+valstats_subset <- cbind(disturbed_stats$count, forest_stats$count)
+
+acc_ct <- sum(diag(ct))/sum(ct)
+acc_valstats <- sum(diag(valstats))/sum(valstats)
+acc_valstats_subset <- sum(diag(valstats_subset))/sum(valstats_subset)
+
+acc <- round(c(NA,acc_ct,NA,acc_valstats,NA,acc_valstats_subset)*100,2)
+valstats <- cbind(ct, valstats, valstats_subset)
+valstats <- rbind(valstats, acc)
+rownames(valstats) <- c("non-forest","forest","")
+
 ```
@@
 
-The functioning of a tropical forest, and its capacity to sequester carbon, is heavily influenced by anthropogenic disturbances such as selective logging, clear felling, or slash and burn practices, all increasing forest fragmentation.
-This increased pace in the loss of rainforest cover [@butsic2015] is predicted to negatively impact the function of these tropical forests, and their future capacity to sequester carbon, as anthropogenic disturbances such as selective logging, clear felling, or slash and burn practices collectively contribute to increasing forest fragmentation.
+Tropical ecosystem services are severely impacted by deforestation and forest degradation [@hansen2013; @houghton2012; @tyukavina2015].
+Not only does tropical forest Land-Use and Land-Cover Change (LULCC) constitute 10 to 15% of the total global emissions [@vanderwerf2009], changes in forest fragmentation affect the forest structure and function.
+Strong fragmentation effects decreases the number of large trees along forest edges [@laurance2000; @magnago2015], while species composition and biodiversity are equally negatively affected [@poorter2006; @barlow2016; @vandeperre2018].
+Estimates show that 31% of carbon emissions are caused by edge effects alone [@brinck2017].
 
-These forest fragments and their edges contain fewer large trees [@laurance2000], in part due to their hotter and brighter microclimates [@magnago2015].
-Similarly, species composition and biodiversity [@poorter2006] are negatively affected.
-Spatial disturbances also have an explicit temporal component, where the time since disturbance affects the strength of the effects [@didham1999].
-Deforestation and gap formation also influence the overall canopy structure, where pioneer species grow taller and with narrower crowns [@poorter2006; @peres2006] affecting the forest light regime and carbon dynamics [@stark2012].
-As such, anthropogenic factors through land-use and land cover change (LULCC) heavily affect forest structure and functioning [@fauset2017; @brinck2017; @didham1999].
-In the tropics the majority of emissions originate from deforestation and forest degradation [@hansen2013; @houghton2012; @tyukavina2015] contributing to 10-15% of the total global emissions [@vanderwerf2009] and concomitant losses in biodiversity losses [@barlow2016; @vandeperre2018] and how LULCC is evolving over time.
-Although the African rainforest is the second largest on Earth and covers ~630 million ha, representing up to 66 Pg of carbon storage [@Lewis2009b], and currently loses forest at an increasing pace [@butsic2015].
-Yet, the tropics is now becoming a large source of CO_2 to the atmosphere as deforestation and forest degradation [@hansen2013; @houghton2012; @tyukavina2015] contribute up to 10-15% of the total global CO2 emissions [@vanderwerf2009].
-Furthermore, recent estimates show that 31% of recent carbon emissions are caused by edge effects alone [@brinck2017].
-In this context, there are very few long term forest inventory plots that presently exist in the central Congo Basin [@Verbeeck2011a; @lewis2013].
-This further limits the accuracy of current biomass estimates and challenges our understanding of how forest structure and function changes in the short and long term in these ecosystems [@Kearsley2013; @Doetterl2015].
+As such, anthropogenic factors through LULCC heavily affect forest structure and functioning [@fauset2017; @brinck2017; @didham1999].
+Accurate estimates of LULCC and forest canopy structure are therefore imperative to estimate carbon emissions and other ecosystem services [@houghton2012; @hansen2013].
+Remote sensing products have been key input as they provide accurate spatial information to help estimate carbon emissions [@mitchard2018; @hansen2013].
+High resolution aerial images provide scientists tools to monitor forest extent, structure and carbon emissions as canopy texture is linked to above ground biomass [@Ploton2012; @Couteron2005; @Barbier2010].
+Yet, these estimates are often limited in time to recent decades [@defries2002; @houghton2012; @hansen2013; @achard2014].
 
-Historical estimates of LULCC, and their carbon emissions, exist [@houghton2012; @ramankutty1999; @defries2002; @sader1988], however they generally rely on non-spatially explicit data in the pre-satellite era (< 1972) [@houghton2010].
-These analysis often use FAO statistics, historical land-use sources, and/or population and socio-economic data [@houghton2012], as limited earlier spatial data sources are available [@hansen2013; @achard2014].
-For example, the @houghton2010 long-term study used FAO and national censuses alongside a bookkeeping model to estimate carbon emissions from LULCC.
-Other studies use remote sensing data, but are limited in time to recent decades [@defries2002; @houghton2012].
-These proxy based studies have a limited view on long-term direct and indirect spatial effects of deforestation.
-As such, both the spatial structure and the fragmentation pattern of a forest combined with its life history has a profound effect on carbon emssisions.
-However, in this context remote sensing assessments provide more accurate spatial information to help estimate carbon emissions [@mitchard2018].
+Historical estimates **of Land-Use and Land-Cover (LULC)**, in the pre-satellite era (< 1972), exist but generally rely on non-spatially explicit data (i.e socio-economic data) [@houghton2012; @ramankutty1999; @defries2002; @sader1988].
+Efforts have been made to use other geo-spatial data sources such as historical maps [@willcock2016], declassified CORONA satellite surveillance data across the US and central Brazil [@song2015] as well as aerial surveys in post world war II German [@nita2018].
+Survey data across the African continent is less common, inaccessible or both.
+Some studies do exist, as @buitenwerf2012, @hudak1998 and @okeke2006 used aerial survey images to map vegetation changes in South African savannas and the Adulam Nature Reserve, respectively, whilst @frankl2015 and @nyssen2016 mapped the Ethiopian highlands of 1930's.
 
-Past studies expanded knowledge during the pre-satellite time frame through the use of now declassified satellite surveillance data to map deforestation in the US, central Brazil, and post world war II Germany [@song2015; @nita2018].
-In some cases these records could extend back in time even further using historical aerial survey images.
-In northern Europe especially these data are well known and used, for example to map the extent of forest cover across ~100 km^2^ regularly, and at times using automated workflows [@giordano2018].
-Survey data across the African continent is less common or inaccessible or both.
-Some studies do exist, both @buitenwerf2012 and @okeke2006 used aerial survey images to map vegetation changes in South African savannas and the Adulam Nature Reserve, respectively.
-Whilst @frankl2015 used 20 aerial survey images to map the Ethiopian highlands of 1935.
-High resolution aerial or surveillance satellite images provide scientists tools to monitor forest extent and structure over long time frames.
-Moreover, canopy texture from aerial survey images has been successfully linked to above ground biomass [@Ploton2012; @Couteron2005; @Barbier2010].
-Thus using texture based metrics obtained from (high resolution) remote sensing data opens up possibilities to link historical aerial footage to carbon density across large spatial extents [@Couteron2005].
+Despite the existence of large historical archives of aerial survey imagery for the central Congo Basin (Figure 1, Appendix Figure 3) as of yet no studies have valorized these data.
+The lack of a consistent valorization effort is unfortunate as the African rainforest is the second largest on Earth and covers ~630 million ha, representing up to 66 Pg of carbon storage [@Lewis2009b], and currently loses forest at an increasing pace [@butsic2015].
+Given the impact of LULCC on the structure and functioning of central African tropical forests, and their influence on both carbon dynamics [@Kearsley2013] and biodiversity [@vandeperre2018], accurate long term reporting of historical forest cover warrants more attention [@willcock2016].
 
-Given the impact of LULCC on the structure and functioning of tropical forests, and their influence on both carbon dynamics [@Kearsley2013] and biodiversity [@vandeperre2018], accurate long term reporting of historical forest cover in the pre-satellite era is an imperative [@willcock2016].
-In this study we use a combination of historical aerial photography and contemporary remote sensing data to map long-term changes in the extent and structure of the tropical forest surrounding Yangambi, in the central Congo Basin at the beginning of 1958 and the start of the antrophocene [@lewis2015].
-Our analysis provides a historical insight into important LULCC spatial patterns in Yangambi, such as fragmentation and edge complexity.
-Our analysis further contextualises the influence of changes in the forest's life history on past and current research into Above Ground Carbon (AGC) storage [@Kearsley2013] and biodiversity [@vandeperre2018] in the central Congo Basin.
+Here, we use a combination of historical aerial photography and contemporary remote sensing data to map long-term changes in the extent and structure of the tropical forest surrounding Yangambi, in the central Congo Basin, effectively linking the start of the antrophocene [@lewis2015] with current assessments.
+Yangambi was, and remains, a focal center of forest and agricultural research and development in the central Congo Basin.
+Past research in the region allows for thorough assessment of land-use and land-cover change from a multi-disciplinary point of view, confronting us with complex deforestation and land-use patterns.
+
+We leverage Structure-from-Motion to generate a large orthomosaic of historical imagery and develop a Convolutional Neural Network based forest cover mapping approach, based upon a semi-supervised generated dataset extensively leveraging data augmentation.
+Our methodology aims to provides a historical insight into important LULCC spatial patterns in Yangambi, such as fragmentation and edge complexity.
+We further contextualize the influence of changes in the forest's life history on past and current research into Above Ground Carbon (AGC) storage [@Kearsley2013] and biodiversity [@vandeperre2018] in the central Congo Basin.
+Our fast scalable mapping approach for historical aerial survey data, using limited supervised input, would further support long-term land-use and land-cover change analysis across the central Congo Basin.
 
@@
 Data for the central Congo Basin region, surrounding Kisangani, were collected in several flights during the dry season of 1958 and 1959 (from 8/01/1958 to 20/02/1958 and from 28/12/1958 to 9/01/1959 respectively, see Appendix Figure 1) to generate topographic maps of the area, supervised by the "Institut Géographique du Congo Belge" in Kinshasa (then Léopoldville).
-Black-and-White infrared images (0.4 - 0.9 $\mu$m) were gathered along flight paths running mostly from west to east, between 9 - 11h local time.
+Black-and-White infrared images (0.4 - 0.9 $\mu$m) were gathered along flight paths running mostly from west to east, between **09:00 and 11:00** local time.
 Along a flight path continuous images were taken using a Wild Heerburgg RC5a (currently Leica Geosystems) with an Aviogon lens assembly (114.83mm / f 5.6, with a 90\degree view angle) resulting in square photo negative of 180 by 180 mm.
@@
 
-```{r echo=FALSE, out.width='100%', fig.cap="A box of historical aerial photographs (left) and a single aerial photograph (right) showing part of the Congo river.
+```{r echo=FALSE, out.width='75%', fig.align="center", fig.cap="A box of historical aerial photographs (left) and a single aerial photograph (right) showing part of the Congo river.
 Note the meta-data provided in the margins of the image such as acquisition time and fleight height and the solar glare on the Congo river."}
@@
 
-```{r echo=FALSE, out.width='100%', message=FALSE, warning=FALSE, fig.cap="Overview of the historical flight paths during aerial photo acquisition and ancillary data used in this study.
+```{r echo=FALSE, out.width='75%', fig.align="center", message=FALSE, warning=FALSE, fig.cap="Overview of the historical flight paths during aerial photo acquisition and ancillary data used in this study.
 The bounding box of the orthomosaic data presented in this study is shown as a rectangle (23x36 km).
@@
 The grey polygon delineates the current day Yangambi Man and Biosphere reserve.
-The inset, top right, situates the greater Yangabmi region with the DR Congo.
+The inset, top right, situates the greater Yangabmi region (white rectangle) with the DR Congo.
 The full flight plan and details are shown in Appendix Figure 1 and 2"}
@@
 Images were scanned at a resolution exceeding their original resolution (or grain) at the maximal physical resolution of an Epson A3 flatbed scanner (i.e.
-2400 dpi or 160MP per image) and saved as lossless tiff images.
+2400 dpi or 160 MP per image) and saved as lossless tiff images.
 Data were normalized using contrast limited histogram equalization [@zuiderveld1994] with a window size of 32 and a clip limit of 1.5.
-Fiduciary marks were used to rectify and downsample the images into square 7700x7700 pixel images (~1200 dpi, 81MP).
+Fiduciary marks were used to rectify and downsample the images into square 7700x7700 pixel images (~1200 dpi, 81 MP).
 This resulted in a dataset with digital images at a resolution that remained above the visible grain of the photographs, whilst the reduced image size facilitated easier file handling and processing speed.
@@
 Additional ground control points were provided to assist in the referencing of image and constrain the optimization routine used in the SfM algorithm.
-Ground control points consisted of rooftop edges of permanent structures which could be verified in both old and new aerial imagery (i.e.
-ESRI World Imagery).
-Although clouds were removed during the SfM routine we did not mask all clouds in the final orthomosaic to maximize forest coverage.
+Ground control points consisted of permanent structures which could be verified in both old and new aerial imagery (i.e.
+ESRI World Imagery) **and consisted of corner points of build structures (e.g.
+a building, bridge or swimming pool etc.).** Although clouds were removed during the SfM routine we did not mask all clouds in the final orthomosaic to maximize forest coverage.
 The final scene was cropped, to provide consistent wall to wall coverage of the reconstructed scene.
@@
 We used 3^th^ degree polynomial and 16 ground control points to correct the final image.
-Ground control points, raw image data and final processed image are provided in addition to measures of uncertainty such as mean root mean squared (RMSE), mean and median error across all ground control points.
+Ground control points, raw image data and final processed image are provided in addition to measures of uncertainty such as mean and median error across all ground control points.
 All subsequent analysis are executed on the final geo-referenced orthomosaic or subsets of it.
@@
 
-### Classifying forest cover
+### Classifying Land-Use and Land-Cover
 
-We automatically delineated all natural forest in the historical data, thus excluding tree plantations, thinned or deteriorated forest stands, fields and buildings.
+#### *Model training*
+
+We automatically delineated all natural forest in the historical data, thus excluding tree plantations, thinned or deteriorated forest stands which **showed visible canopy cover loss**, fields and buildings.
 We used the Unet Convolutional Neural Net (CNN, @ronneberger2015) architecture implemented in Keras [@chollet2015] with an efficientnetb3 pre-processing backbone [@yakubovskiy2019] running on TensorFlow [@martinabadi2015] to train a binary classifier (i.e.
 forest or non-forested).
-This methodology is increasingly being used to automate pixel-level classification in (color) digital photography data [@buscombe2018].
-Training data were collected from the orthomosaic by randomly selecting 513 pixel square tiles from homogeneous forested or non-forested areas within the historical orthomosaic.
-Homogeneous tiles were combined in synthetic landscapes using a random gaussian field based binary mask (Figure 3).
-We generated 5000 synthetic landscapes for training, while 500 landscapes were generated for both the validation and the testing dataset.
-Source tiles did not repeat across datasets to limit overfitting.
+Training data were collected from the orthomosaic by randomly selecting 513 pixel square tiles from **locations within** homogeneous forested or non-forested polygons in the historical orthomosaic **(Figure 5).
+Separate polygons were selected for training, testing and validation purposes.
+Validation polygons were sampled 300 times, while both testing and validation polygons were sampled at 100 random locations.
+Tiles extracted from locations close to the polygon border at times contained mixed cover types.
+Tiles with mixed cover types were removed from the list of source tiles (Table 1).** Homogeneous **source** tiles were combined in synthetic landscapes using a random gaussian field based binary mask (Figure 3).
+We generated 5000 synthetic landscapes for training, while 500 landscapes were generated for both the validation and the testing datasets for a total of 6000 synthetic landscapes.
 In order to limit stitch line misclassifications, along the seams of mosaicked images, we created synthetic landscapes with different forest tiles to mimick forest texture transitions.
-We applied this technique to 10% of the generated synthetic landscapes.
-The CNN model was trained for 100 epochs with a batch size of 30 using Adam optimization [@kingma2017], maximizing the Intersect-over-Union (IoU) using Sørensen–Dice and categorical cross-entropy loss functions.
-Data augmentation included random cropping to 320 pixel squares, random orientation, scaling, perspective, contrast and brightness shifts and image blurring.
-During final model evaluation we report the IoU of our out-of-sample test datasets.
-The optimized model was used to classify the complete orthomosaic using a moving window approach with a step size of 110 pixels and a majority vote across overlapping areas to limit segmentation edge effects.
-We refer to Figure 4 for a synoptic overview of the full deep learning learning workflow.
-In addition, we used the first acquisition of a recent pan-chromatic Geo-Eye 1 stereo pair (GeoEye, 2012, Thornton, Colorado, U.S.A., order 737537, 2011-11-11 8:55h GMT or 9:55h local time) to classify and visually assess the robustness of the algorithm on contemporary remote sensing imagery with the same moving window approach.
+We applied this technique to 10% of the generated synthetic landscapes **(across training, validation and testing data)**.
 
-```{r echo=FALSE, out.width='100%', fig.cap="orthomosaic", fig.cap="An example synthetic landscape, combining homogeneous forest and non-forest images into a patchy landscape using random gaussian field based masks.
+```{r echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE}
+
+training <- c(300, 294)
+testing <- c(100, 92)
+validation <- c(100, 84)
+
+source_img <- data.frame(training, testing, validation)
+rownames(source_img) <- c("forest","non-forest")
+
+source_img %>%
+   kable(
+    caption="Number of source tiles used for the generation of synthetic landscapes.",
+    booktabs = TRUE,
+    escape = TRUE) %>%
+  kable_styling(latex_options = c("hold_position"))
+
+```
+
+```{r echo=FALSE, out.width='100%', fig.align="center", fig.cap="orthomosaic", fig.cap="An example synthetic landscape, combining homogeneous forest and non-forest images into a patchy landscape using random gaussian field based masks.
 The left panel shows a combined synthetic landscape, while the right panel shows the corresponding forest (black) and non-forest (white) labels."}
@@
 
-```{r echo=FALSE, out.width='100%', fig.cap="orthomosaic", fig.cap="A diagram of the deep learning workflow followed in training a binary forest / non-forest cover convolutional neural net U-Net model to generate our forest cover map."}
+The CNN model was trained for 100 epochs with a batch size of 30 using Adam optimization [@kingma2017], maximizing the Intersect-over-Union (IoU) using Sørensen–Dice and categorical cross-entropy loss functions.
+Data augmentation included random cropping to 320 pixel squares, random orientation, scaling, perspective, contrast and brightness shifts and image blurring.
+The optimized model was used to classify the complete orthomosaic using a moving window approach with a step size of 110 pixels and a majority vote **(> 50% agreement)** across overlapping areas to limit segmentation edge effects.
+**In addition we provide, raw pixel level classification agreement data for quality control purposes (see Data availability below).** We refer to Figure 4 for a synoptic overview of the full deep learning learning workflow.
+
+```{r echo=FALSE, out.width='100%', fig.align="center", fig.cap="orthomosaic", fig.cap="A diagram of the deep learning workflow followed in training a binary forest / non-forest cover convolutional neural net U-Net model to generate our forest cover map."}
 knitr::include_graphics('./figures/cnn_diagram.pdf')
@@
 
-To map long term land-use and land-cover change in the Yangambi region we used the contemporary Global Forest Change version 1.6 data (GFC, tile 10N-020E) [@hansen2013].
-Using this data we calculated the latest state of the forest with respect to the conditions at the start of 1958, 60 years earlier.
-In our analysis we only included forested pixels which recorded no loss throughout the whole 2000 - 2018 period.
+#### *Model validation*
+
+**We report the CNN accuracy based upon the IoU of our out-of-sample validation dataset of synthetic landscapes.
+In addition, we report confusion matrices for all pixels across the homogeneous validation polygons, as well as the training and testing polygons (see Figure 5).
+Furthermore, we used the first acquisition of a recent pan-chromatic Geo-Eye 1 stereo pair (Geo-Eye, Thornton, Colorado, U.S.A., order 737537, 2011-11-11 08:55 GMT or 09:55 local time) to classify and assess the robustness of the CNN algorithm on contemporary remote sensing imagery with similar spectral and spatial characteristics.
+We used the Global Forest Change version 1.6 (GFC, tile 10N-020E) [@hansen2013] map data, capturing forest loss up to 2011, to quantify accuracy on downsampled CNN Geo-Eye classification results.
+Once more, we report the confusion matrix of between the GFC and CNN derived forest cover maps, masking clouds and cloud shadows.
+To summarize confusion matrices we report accuracy as:**
+
+$Accuracy = \frac{(TP + TN)}{(TP + TN + FP + FN)}$
+
+**in which TP, TN, FP, FN are True Positive, True Negative, False Positive and False Negative, respectivelly.**
+
+### Characterizing long term change
+
+To map long term land-use and land-cover change in the Yangambi region we used the contemporary Global Forest Change version 1.6 (GFC, tile 10N-020E) map data [@hansen2013].
+Using the **GFC** data we calculated the latest state of the forest with respect to the conditions at the start of 1958, 60 years earlier.
+In our analysis we only included **GFC** pixels which recorded no **forest** loss throughout the whole 2000 - 2018 period.
 As the resolution of the historical forest classification exceeds that of the GFC map we downsampled our historical forest cover data to 30m GFC resolution.
@@
 Where the datasets overlap we provide summary statistics of deforestation, reforestation and afforestation.
-The same methodology was used to quantify CNN model performance across the Geo-Eye 1 scene.
 
@@
 For example, the AGC values varied from 168.67 to 86.55 g Mg C ha^-1^, for *Hevea brasiliensis* (rubber tree) and *Elaeis guineensis* (oil palm) plots respectively (@bustillo2018, personal communications).
-These higher values are in line with the mixed AGC estimates in the area, while the palm plantations resemble old-regrowth values (81.8 Mg C ha^-1^, see Table A1).
+These higher values are in line with the mixed AGC estimates in the area, while the palm plantations resemble old-regrowth values (81.8 Mg C ha^-1^, see **Appendix Table 1**).
 We therefore use both the estimates of old-regrowth and mixed forest to estimate AGC for regrowth.
@@
 
-We used an area of 400x400m (16 ha) around contemporary permanent sampling plot locations to quantify the general state of the canopy in both historical and contemporary pan-chromatic Geo-Eye 1 remote sensing data (GeoEye, 2012, Thornton, Colorado, U.S.A., order 737537, 2011-11-11 8:55h GMT or 9:55h local time).
-Geo-Eye data were resampled to the resolution of the historical data using a nearest neighbour interpolation for comparison.
-For permanent sampling plot locations in both historical and contemporary data a rectangular buffer of 200m around the centre of all plots was used to create small image subsets.
-All image subsets were processed using the R based FOTO implementation and package [@hufkens2019a] using global normalization, ensuring intercomparability of texture metrics across time and space [@Barbier2010].
-We used a FOTO (moving) window size of 187 pixels (or ~150m), as described in @solorzano2018 and @Barbier2010, to ensure that multiple individual canopies could be included in the analysis.
-A buffer with a radius of 50m around the center of all permanent sampling plots was used to determine the average value of the first principle component (PC), explaining the majority of the variance in canopy texture.
-Where both the Geo-Eye data and the orthomosaic image overlapped we processed the intersecting region to explore a wall-to-wall comparison between past and current canopy texture metrics.
-
 We used the standard FOTO methodology with fixed zones, instead of the moving window approach.
-The window size was set to the same size (187 pixels or ~150m) as used in the moving window analysis above.
+The window size was set to the same size (187 pixels or ~150 m) as used in the moving window analysis above.
 To account for illumination differences between the two scenes we applied histogram matching.
 No global normalization was applied, as the scene was processed as a whole.
-PC values from this analysis for all permanent sampling plots in both image scences were extracted using a buffer with a radius of 50m.
+PC values from this analysis for all permanent sampling plots in both image scences were extracted using a buffer with a radius of 50 m.
 
@@
 
-Our analysis provides a first spatially explicity historical composite of aerial survey images mapping LULCC within the Congo Basin.
+## Orthomosaic construction
+
+Our analysis provides a first spatially explicity historical composite of aerial survey images **in support of mapping land-use and land-cover** within the Congo Basin.
 The use of high resolution historical images combined with structure from motion image processing techniques allowed us to mosaic old imagery across a large extent.
-The final orthomosaic composition of the Yangambi region resulted in an image scene covering approximately 733 million pixels across ~828 km^2^ (~23x36 km, Figure 2).
-The overall accuracy of the structure from motion orthomosaic composition was 0.88 m/pixel using the sparse cloud DEM for corrections at 45.8 m/pixel.
-The resulting georeferenced scene had a spatial accuracy of approximately 23m.
-Further georeferencing outside the SfM workflow reduced the mean error at the ground control points to 5.3 $\pm$ 4.9 px (~4.7 $\pm$ 4.3 m), with a median error of 2.9 px (2.6m).
+The final orthomosaic composition of the Yangambi region **provided** an image scene covering approximately 733 million pixels across **~82800 ha** (~23x36 km, Figure 2).
+The overall accuracy of the SfM orthomosaic composition was 0.88 m/pixel using the sparse cloud DEM for corrections at 45.8 m/pixel.
+The resulting georeferenced scene had a spatial accuracy of approximately **23 m**.
+Further georeferencing outside the SfM workflow reduced the mean error at the ground control points to 5.3 $\pm$ 4.9 px (~4.7 $\pm$ 4.3 m), with a median error of 2.9 px (2.6 m).
 The orthomosaic served as input for all subsequent LULCC analysis with all derived maps provided with the manuscript repository (see data & code availability statements below).
 
-```{r echo=FALSE, out.width='100%', fig.cap="orthomosaic", fig.cap="Overview of the final orthomosaic of the greater Yangambi region (A), a detailed inset (B) and the derived land use land cover change map displayed as the difference between a manual classification of the orthomosaic and a recent map by Hansen et al.
-2013 (C) and the corresponding land cover map for inset B (D)."}
+```{r echo=FALSE, out.width='100%', fig.align="center", fig.cap="orthomosaic", fig.cap="Overview of the final orthomosaic of the greater Yangambi region (A) as well as the forest cover classification uncertainty (B) used to generate the final Land Land Cover map (see Figure 6).
+Homogeneous polygons used in the generation of the synthetic landscape for Convolutional Neural Network training, testing and validation are marked as dashed and full lines for forest and non-forest regions, respectively.
+Training, testing and validation regions are denoted in panel B in green, red and orange, respectivelly.
+Black polygon outlines denote cloud and image stitch line regions which were manually excluded from analysis, but retained in calculation of validation statistics (see Table 2)."}
 knitr::include_graphics('./figures/orthomosaic_maps.png')
@@
 
-## Land-use and Land-Cover Change & Above Ground Carbon
+## Land-use and Land-Cover Classification
 
-Our CNN deep learning classifier reached an Intersection-over-Union (IoU) accuracy of 97% on the detection of disturbed forest in out-of-sample test data.
-A visual comparison on recent pan-chromatic Geo-Eye 1 data shows good agreement, for a total of 87% of all pixels, between the landsat based GFC data and downscaled CNN results (Appendix Table 3 and Figure 7).
-Scaling our classifier to the whole historical orthomosaic we detected 162 km^2^ (or ~20% of the scene) of disturbed forests.
-A large fraction of the disturbed area was restored in the period between the two acquisitions.
-In total, 99 km^2^, or little over half of the affected forest was restored (Figure 5C-D, dark blue).
-Recent deforested areas, as registered through satellite remote sensing, approximate 88 km^2^ (Table1, Figure 5C-D / light green).
+### CNN model validation
+
+**The CNN deep learning classifier reached an Intersection-over-Union (IoU) accuracy of 97% on the detection of disturbed forest in the out-of-sample (validation) synthetic landscape data.
+Using all pixels within the validatoin polygons (Figure 5) showed a similar accuracy value of ~98%.
+Using all polygons across the scene, including those used in the generation of testing and training synthetic landscapes, increased the accuracy to ~99% (Table 2).
+A comparison with recent pan-chromatic Geo-Eye data shows good agreement, with an accuracy of ~87% across all pixels, between the landsat based GFC data and downscaled CNN results (Table 2 and Figure 7).**
+
+```{r echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE}
+valstats %>%
+   kable(
+    caption="Confusion matrix showing \\% agreement between forest / non-forest classes using a Convolutional Neural Network (CNN) across previously selected homogenous areas.",
+    booktabs = TRUE,
+    escape = TRUE,
+    col.names =  c("non-forest","forest","non-forest","forest","non-forest","forest")) %>%
+    pack_rows("Accuracy",3,3) %>%
+    add_header_above(c("CNN" = 1, "Geo-Eye (full scene)" = 2, "Aerial Survey (all polygons)" = 2, "Aerial Survey (validation polygons)" = 2)) %>%
+  kable_styling(latex_options = c("hold_position"))
+```
+
+```{r echo=FALSE, fig.align="center", out.width='100%', fig.cap="orthomosaic", fig.cap="Overview of the final Land Use Land Cover Change map, a detailed inset of both the underlying orthomosaic (B) and the derived land use land cover change map displayed as the difference between the Convolutional Neural Network based classification orthomosaic and the recent Landsat based forest cover map by Hansen et al.
+2013 (C)."}
+knitr::include_graphics('./figures/forest_cover_map.png')
+```
+
+```{r echo=FALSE, fig.align="center", out.width='100%', fig.cap="Convolutional Neural Network (CNN) based forest cover classification results (B) as run on a recent (2011) Geo-Eye panchromatic image (A).
+We show the difference between the Convolutional Neural Network based classification and a recent Landsat based forest cover map by Hansen et al.
+2013.
+Full white outlines denote cloud contamination, the dashed rectangle shows a location where the CNN outperforms the Landsat based forest classification."}
+knitr::include_graphics('./figures/visual_comparison_classifiers.png')
+```
+
+## Long term LULC changes and Above Ground Carbon
+
+Scaling our classifier to the whole historical orthomosaic we detected **16200 ha** (or ~20% of the scene) of disturbed forests.
+A large fraction of the disturbed area was restored in the period between the two **acquisition periods**.
+In total, **9900 ha**, or little over half of the affected forest was restored (Figure 5C-D, dark blue).
+Recent deforested areas, as registered through satellite remote sensing, approximate **8800 ha (Table 3, Figure 6 - light green)**.
 
 Recent deforestation follows a distincly different pattern compared to historical patterns.
-Historical deforestation showed a classical fishbone pattern for forest clearing with very sharp edges, while current patterns are patchy and ad-hoc (figure 5D, dark blue and green colours respectively).
+Historical deforestation showed a classical fishbone pattern for forest clearing with very sharp edges, while current patterns are patchy and ad-hoc (**figure 6C**, dark blue and green colours respectively).
 These differences are reflected in the analysis of landscape metrics of forest loss.
 Between the historical and contemporary LULCC maps we see an increase in small disturbances, as indicated by the decreasing area of the mean patch size, down to ~1.86 $\pm$ 0.75 ha from ~5.25 $\pm$ 5.02 ha historically.
-Perimeter lengths were longer historically, at 1451.58 $\pm$ 943.27 m,  compared to contemporary landscapes ~921.74 $\pm$ 362.59 m (Table 2).
+Perimeter lengths were longer historically, at 1451 $\pm$ 943 m,  compared to contemporary landscapes ~921 $\pm$ 362 m (Table 3).
 This shift in perimeter area ratio led to a similar change in the fractal index, increasing in value to 1.1 $\pm$ 0.05 from 1.09 $\pm$ 0.04 over time.
@@
 A comparison of forest edge plots with mixed forest plots showed no signficant difference in AGC, or other reported values such as species richness, basal area or stem density (Mann Whitney U test, p < 0.05).
-Edge influence did not extend beyond 200 m from a forest edge, but still represented an area of 132 km^2^.
+Edge influence did not extend beyond 200 m from a forest edge, but still represented an area of **13200 ha**.
 
@@
        escape = FALSE,
-       col.names = c("$km\\textsuperscript{2}$",
-                     "ha",
+       col.names = c("ha",
                      "(Gg C $ha\\textsuperscript{-1}$")
          ) %>%
-    add_header_above(c(" " = 3,"AGC" = 1)) %>%
+    add_header_above(c(" " = 2,"AGC" = 1)) %>%
     kable_styling(latex_options = c("hold_position"))
@@
 We report patch perimeter and area, their ratio and fractal dimension.
-Values are reportedpat as mean $\\pm$ standard deviation, across all deforestation patches.",
+Values are reported as mean $\\pm$ standard deviation, across all deforestation patches.",
        booktabs = TRUE,
@@
 One marked difference is noted in the mono-dominant site 6 (Appendix Table 1).
-Here, the current mono-dominant *Brachystegia laurenttii* is a recent development, changing the canopy structure visibly during the last half century (Figure 6).
+Here, the current mono-dominant *Brachystegia laurenttii* is a recent development, changing the canopy structure visibly during the last half century (**Figure 8**).
 The previous varied canopy structure gave way to a more dense and uniform canopy.
@@
 
-```{r echo=FALSE, out.width='100%', fig.cap="Visual comparison between a historical (A) and contemporary (B) permanent sampling plot.
+```{r echo=FALSE, out.width='100%', fig.align="center", fig.cap="Visual comparison between a historical (A) and contemporary (B) permanent sampling plot.
 The site is currently listed as a mono-dominant Brachystegia laurenttii stand.
@@
 
-Using only small subsets around existing permanent sampling plots we show distinct differences between forest types, with PC values in both historical and contemporary imagery markedly higher for the mono-dominant forest types compared to all others (Appendix Figure 3).
+Using only small subsets around existing permanent sampling plots we show distinct differences between forest types, with PC values in both historical and contemporary imagery markedly higher for the mono-dominant forest types compared to all others (**Appendix Figure 4**).
 Provided that the young-regrowth and fallow permanent sampling plots have seen recent disturbance the Wilcoxon signed rank test on the mixed and mono-dominant plots between the historical and contemporary PC values did not show a significant difference (p > 0.05).
 Similarly, no significant different using PC values extracted from the whole scene analysis was noted (p > 0.05).
-Any relationships between contemporary Geo-Eye data and permanent sampling plot measurements of Above Ground Biomass, stem density and species richness were non-significant (p > 0.05, Appendix Figures 4-6).
+Any relationships between contemporary Geo-Eye data and permanent sampling plot measurements of Above Ground Biomass, stem density and species richness were non-significant (p > 0.05, Appendix **Figures 5-7**).
 
-Furthermore, visual inspection of the scene wide analysis suggests historical scences do not show landscape wide canopy features (Figure 7 A-B), unlike the contemporary scene (Figure 7 C-D).
+Furthermore, visual inspection of the scene wide analysis suggests historical scences do not show landscape wide canopy features (**Figure 9 A-B**), unlike the contemporary scene (**Figure 9 C-D**).
 Where the FOTO algorithm picks up landscape features such as changes in texture across the contemporary Geo-Eye scene (e.g.
-the river valley as a diagonal line in Figure 7D), however, no corresponding landscape patterns are found by the FOTO algorithm in the historical orthomosaic.
+the river valley as a diagonal line in **Figure 9D**), however, no corresponding landscape patterns are found by the FOTO algorithm in the historical orthomosaic.
 
-```{r echo=FALSE, out.width='100%', fig.cap="RGB visualizations of the first three principal components of scene wide FOTO texture analysis of historical and current (Geo-Eye) imagery.
+```{r echo=FALSE, out.width='100%', fig.align="center", fig.cap="RGB visualizations of the first three principal components of scene wide FOTO texture analysis of historical and current (Geo-Eye) imagery.
 Current permanent sampling plots of mono-dominant, mixed, fallow and young (edge) forest plots are marked with open triangles, open circles, open squares and crosses, respectively."}
@@
 
-Finely grained spatial data sources, such as remote sensing imagery, are rare before the satellite era (<1970).
-This lack of data limits our understanding of how forest structure has varied over longer time periods in especially remote areas.
+Finely grained spatial data sources, such as remote sensing imagery, are rare before the satellite era (<1972).
+This lack of data limits our understanding of how forest structure has varied over longer time periods in remote areas.
 Long term assessment can be extended by using large inventories of historical aerial survey data [@kadmon1999; @song2015; @nita2018].
-However, across the Congo Basin this data is rare, or difficult to access.
-Despite the difficulties in assessing hidden disturbances caused by deforestation, poorly recorded official assessments or simply those invisible to standard remote sensing techniques [@peres2006], remote sensing generally remains the best way to map and quantify LULCC [@houghton2012].
-Despite these challenges in recovery and validation we present a first analysis of LULCC using historical aerial survey data for the central Congo Basin.
-We used a combination of Structure-from-Motion and a convolutional neural net deep learning approach to map deforestation across a large extent, providing a long term view of LULCC in the central Congo Basin
+Despite the difficulties in recovering this data and its limitations, such as invisible disturbances [@peres2006], remote sensing generally remains the best way to map and quantify LULCC [@houghton2012].
+In our study we used novel numerical remote sensing techniques to valorize, for the first time, historical remote sensing data in order to quantify (long term) land-use and land-cover change and canopy structural properties in the central Congo Basin.
+Despite these successes some methodological and research considerations remain.
 
-## Data recovery challenges
+## Methodological considerations
+
+### Data recovery challenges
 
 In our study the archive data recovered was limited to contact prints and therefore did not represent the true resolution of the original negative.
-In addition, analogue photography clearly produces a distinct softness compared to digital imagery (Figure 4).
+In addition, analogue photography clearly produces a distinct softness compared to digital imagery (**Figure 8**).
 Despite favourable nadir image acquistions [@verhoeven2017] image softness combined with illumination effects between flight paths, and the self-similar nature of vast canopy expanses [@park2019; @simini2010; @sole1995], limited our ability to provide wall-to-wall coverage of the entire dataset containing 334 images.
@@
 trees, river outflows).
-Research has shown that soft-edged features can help georeference scenes even when containing few man-made features [@hughes2006], however accuracy becomes compromised.
+Research has shown that soft-edged features can help georeference scenes even when containing few man-made features [@hughes2006].
 Our two step georeferencing approach resulted in a referencing accuracy of ~4.7 $\pm$ 4.3 m across reference points.
@@
 
+### Land-use, Land-Cover change classification & validation
+
+**When classifying the orthomosaic into forest and non-forest states we favoured a deep learning supervised classification using a CNN over manual segmentation to guarantee an "apples-to-apples" comparison between the historical and the contemporary GFC forest cover map used in our analysis.
+We acknowledge that both the CNN and GFC land-use and land-cover maps use different underlying features, i.e.
+spatial or spectral data, yet attain a similarly high accuracy of up to 99% [@hansen2013].
+More so, when validating our CNN classifier against GFC data (Figure 7) for a contemporary high resolution Geo-Eye panchromatic image we reach an accuracy of ~87%, despite a time difference of almost 60 years.
+Visual inspection of the classification data in Figure 7 suggests that the GFC map more often than not classifies non-forest areas as forest.
+Actual classification accuracy of our algorithm might therefore be higher than our reported value.**
+
+### Scaling opportunities
+
+**Our approach uses broadly defined homogeneous polygons to construct a balanced dataset of synthethic landscapes.
+The methodology is analoguous to the use of sparse labelling as used by [@buscombe2018] and contrasts with the standard methodologies which generally extract pixel (windows) [@song2015] or delineate land cover classes [@nita2018] to drive a classifier or analysis.
+More so, the use of heavy image augmentation during model training sidesteps texture representation issues which affect classification of image scenes with inconsistent illumination or sharpness [@hudak1998] or ad-hoc feature engineering [@song2015].
+The use of synthetic landscapes allowed us to account for most, but not all, of the variability within our orthomosaic.
+More so, our analysis has shown that despite being trained on historical data our model could map contemporary forest cover in remote sensing data with similar spatial and spectral characteristics (Figure 7), suggesting that the classifier consistently across works both space and time.
+We acknowledge that our approach will speed up labelling, but is limited by the available homogeneous areas to sample from and the number of classes.
+Yet, the latter should not be a constrained as previous research efforts have focussed on simple forest loss maps [@hansen2013].**
+
+## Research context
+
 ## Land-use, Land-Cover change & Above Ground Carbon
 
-When classifying our orthomosaic into forest and non-forest states we favoured a deep learning supervised classification using a Convolutional Neural Network over manual segmentation to guarantee an "apples-to-apples" comparison between the historical and the contemporary GFC forest cover maps.
-We acknowledge that both classifications use different features, i.e.
-spectral or spatial data, but attain a similarly high accuracy of 97% and 99% [@hansen2013], for the GFC and our CNN map respectively.
-Despite a number of challenges we were able to map and quantify a vast continuous area of tropical forest, with limited manual annotation requirements.
-
-Our analysis shows that the majority of deforestation around Yangambi happened toward the late 1950's (162 km^2^).
-Considerable regrowth has occured since the aerial survey was executed (99 km^2^), and socio-economic instability prevented further large scale forest exploitation.
+Our analysis shows that the majority of deforestation around Yangambi happened toward the late 1950's (**16200 ha**).
+Considerable regrowth has occured since the aerial survey was executed (**~9900 ha**), and socio-economic instability prevented further large scale forest exploitation.
 In particular, many plantations have reached maturity and forest has re-established in previously cleared or disturbed areas.
@@
 However, it must be stated that the influence of edges throughout the landscape was not marginal.
-as these areas would account for 132 km^2^.
+as these areas would account for **13200 ha**.
 Thus edges could have a substantial negative [@brinck2017] or positive [@reinmann2017] influence on AGC.
@@
 
-Finally the FOTO technique used to quantify relationships between canopy structure and forest characteristics rendered no valuable insights of either the historical orthomosaic or recent Geo-Eye scene.
+Finally, the FOTO technique used to quantify relationships between canopy structure and forest characteristics rendered no valuable insights of either the historical orthomosaic or recent Geo-Eye scene.
 Similarly weak correlations were found previously by @solorzano2018.
-In contrast site based texture metric statistics did show correspondence between historical and contemporary satellite imagery.
-However, none of them were either consistent or significant.
+In contrast, site based texture metric statistics did show correspondence between historical and contemporary satellite imagery.
+None of them were either consistent or significant.
 Although visual interpretation shows distinctly different canopy structures (Figure 5) the differences in how resolution is defined and issues related to image quality prevented us from quantifying these further.
 Unlike large scale studies by @Ploton2012 we could not scale this technique to historical data.
-We advise that future valorisation efforts should preferentially work from foto negatives (if available) to ensure optimal data quality in resolution, contrast and overall sharpness.
+**The successful use of our CNN classification model on a contemporary remote sensing data does suggest that texture can be used to consistently capture canopy properties 60 years apart.
+This suggests inflexibilit on part of the FOTO technique in dealing with non-standardized data.** We advise that future valorisation efforts should preferentially work from foto negatives (if available) to ensure optimal data quality in resolution, contrast and sharpness.
 
@@
 
-Given the impact of tropical forest disturbances on atmospheric CO_2 emissions, biodiversity and ecosystem productivity accurate long term reporting on land-cover and land-use change especially in the pre-satellite era is an imperative.
+Given the impact of tropical forest disturbances on atmospheric CO~2~ emissions, biodiversity and ecosystem productivity accurate long term reporting on land-cover and land-use change especially in the pre-satellite era is an imperative.
 Our analysis of historical aerial survey images of the Central Congo Basin provides a window into the state of the forest at the start of the anthropocene.
 Efforts to quantify canopy texture and their link to carbon storage had limited to no success.We have shown that historical aerial survey data can be used to quantify land-use and land-cover change using a highly automated workflow to quantify the structure and extent of forest cover change that can help assess the impact of fragmentation on above ground carbon stocks.
-We also show a shift from previously highly structured industrial deforestation of large areas, to discrete smallholder clearing for farming, increasing landscape fragmentation but also opportuntiy for substantial regrowth.
+We also show a shift from previously highly structured industrial deforestation of large areas **for plantation purposes**, to discrete smallholder clearing for farming, increasing landscape fragmentation but also opportuntiy for substantial regrowth.
 Efforts to quantify canopy texture and their link to carbon storage had limited to no success.
@@
 Geo-Eye data) are not shared, but purchase order numbers allow for acquisition of these datasets to ensure reproducibility.
-The @hansen2013 image data is freely available from [the GFC project website](https://earthenginepartners.appspot.com/science-2013-global-forest/download_v1.6.html).
 

### 2020-01-23 manuscript/main.tex
revision

@@
 
-% You need to save the "mdpi.cls" and "mdpi.bst" files into the same folder as this template file.
-
 %=================================================================
-\documentclass[remote sensing,article,submit,moreauthors,pdftex,10pt,a4paper]{mdpi}
+\documentclass[remote sensing,article,submit,moreauthors,pdftex]{mdpi}
+
+% If you would like to post an early version of this manuscript as a preprint, you may use preprint as the journal and change 'submit' to 'accept'.
+The document class line would be, e.g., \documentclass[preprints,article,accept,moreauthors,pdftex]{mdpi}.
+This is especially recommended for submission to arXiv, where line numbers should be removed before posting.
+For preprints.org, the editorial staff will make this change immediately prior to posting.
 
@@
 
-%
 %--------------------
@@
 %--------------------
+%----------
 % journal
@@
 % Choose between the following MDPI journals:
-% actuators, addictions, admsci, aerospace, agriculture, agronomy, algorithms, animals, antibiotics, antibodies, antioxidants, applsci, arts, asi, atmosphere, atoms, axioms, batteries, bdcc, behavsci, beverages, bioengineering, biology, biomedicines, biomimetics, biomolecules, biosensors, brainsci, buildings, carbon, cancers, catalysts, cells, ceramics, challenges, chemengineering, chemosensors, children, chromatography, climate, coatings, colloids, computation, computers, condensedmatter, cosmetics, cryptography, crystals, cybersecurity, data, dentistry, designs, diagnostics, diseases, diversity, econometrics, economies, education, electrochemistry, electronics, energies, entropy, environments, epigenomes, est, fermentation, fibers, fire, fishes, fluids, foods, forests, fractalfract, futureinternet, galaxies, games, gastrointestdisord, gels, genealogy, genes, geosciences, geriatrics, hazardousmatters, healthcare, heritage, highthroughput, horticulturae, humanities, hydrology, informatics, information, infrastructures, inorganics, insects, instruments, ijerph, ijfs, ijms, ijgi, ijtpp, inventions, j, jcdd, jcm, jcs, jdb, jfb, jfmk, jimaging, jof, jintelligence, jlpea, jmmp, jmse, jpm, jrfm, jsan, land, languages, laws, life, literature, logistics, lubricants, machines, magnetochemistry, make, marinedrugs, materials, mathematics, mca, mti, medsci, medicines, membranes, metabolites, metals, microarrays, micromachines, microorganisms, minerals, modelling, molbank, molecules, mps, nanomaterials, ncrna, neonatalscreening, neuroglia, nitrogen, nutrients, ohbm, particles, pathogens, pharmaceuticals, pharmaceutics, pharmacy, philosophies, photonics, plants, plasma, polymers, polysaccharides, preprints, proceedings, processes, proteomes, publications, quaternary, qubs, recycling, religions, remotesensing, reports, resources, risks, robotics, safety, scipharm, sensors, separations, sexes, sinusitis, socsci, societies, soils, sports, standards, stats, surgeries, sustainability, symmetry, systems, technologies, toxics, toxins, tropicalmed, universe, urbansci, vaccines, vetsci, vibration, viruses, vision, water, wem
+% acoustics, actuators, addictions, admsci, aerospace, agriculture, agriengineering, agronomy, algorithms, animals, antibiotics, antibodies, antioxidants, applsci, arts, asc, asi, atmosphere, atoms, axioms, batteries, bdcc, behavsci , beverages, bioengineering, biology, biomedicines, biomimetics, biomolecules, biosensors, brainsci , buildings, cancers, carbon , catalysts, cells, ceramics, challenges, chemengineering, chemistry, chemosensors, children, cleantechnol, climate, clockssleep, cmd, coatings, colloids, computation, computers, condensedmatter, cosmetics, cryptography, crystals, dairy, data, dentistry, designs , diagnostics, diseases, diversity, drones, econometrics, economies, education, electrochem, electronics, energies, entropy, environments, epigenomes, est, fermentation, fibers, fire, fishes, fluids, foods, forecasting, forests, fractalfract, futureinternet, futurephys, galaxies, games, gastrointestdisord, gels, genealogy, genes, geohazards, geosciences, geriatrics, hazardousmatters, healthcare, heritage, highthroughput, horticulturae, humanities, hydrology, ijerph, ijfs, ijgi, ijms, ijns, ijtpp, informatics, information, infrastructures, inorganics, insects, instruments, inventions, iot, j, jcdd, jcm, jcp, jcs, jdb, jfb, jfmk, jimaging, jintelligence, jlpea, jmmp, jmse, jnt, jof, joitmc, jpm, jrfm, jsan, land, languages, laws, life, literature, logistics, lubricants, machines, magnetochemistry, make, marinedrugs, materials, mathematics, mca, medicina, medicines, medsci, membranes, metabolites, metals, microarrays, micromachines, microorganisms, minerals, modelling, molbank, molecules, mps, mti, nanomaterials, ncrna, neuroglia, nitrogen, notspecified, nutrients, ohbm, particles, pathogens, pharmaceuticals, pharmaceutics, pharmacy, philosophies, photonics, physics, plants, plasma, polymers, polysaccharides, preprints , proceedings, processes, proteomes, psych, publications, quantumrep, quaternary, qubs, reactions, recycling, religions, remotesensing, reports, resources, risks, robotics, safety, sci, scipharm, sensors, separations, sexes, signals, sinusitis, smartcities, sna, societies, socsci, soilsystems, sports, standards, stats, surfaces, surgeries, sustainability, symmetry, systems, technologies, test, toxics, toxins, tropicalmed, universe, urbansci, vaccines, vehicles, vetsci, vibration, viruses, vision, water, wem, wevj
+
 %---------
@@
 %---------
-% The default type of manuscript is article, but can be replaced by:
-% abstract, addendum, article, benchmark, book, bookreview, briefreport, casereport, changes, comment, commentary, communication, conceptpaper, correction, conferenceproceedings, conferencereport, expressionofconcern, meetingreport, creative, datadescriptor, discussion, editorial, essay, erratum, hypothesis, interestingimages, letter, meetingreport, newbookreceived, opinion, obituary, projectreport, reply, reprint, retraction, review, perspective, protocol, shortnote, supfile, technicalnote, viewpoint
+% The default type of manuscript is "article", but can be replaced by:
+% abstract, addendum, article, benchmark, book, bookreview, briefreport, casereport, changes, comment, commentary, communication, conceptpaper, conferenceproceedings, correction, conferencereport, expressionofconcern, extendedabstract, meetingreport, creative, datadescriptor, discussion, editorial, essay, erratum, hypothesis, interestingimages, letter, meetingreport, newbookreceived, obituary, opinion, projectreport, reply, retraction, review, perspective, protocol, shortnote, supfile, technicalnote, viewpoint
 % supfile = supplementary materials
-% protocol: If you are preparing a "Protocol" paper, please refer to http://www.mdpi.com/journal/mps/instructions for details on its expected structure and content.
+
 %----------
@@
 % The class option "submit" will be changed to "accept" by the Editorial Office when the paper is accepted.
-This will only make changes to the frontpage (e.g.
-the logo of the journal will get visible), the headings, and the copyright information.
+This will only make changes to the frontpage (e.g., the logo of the journal will get visible), the headings, and the copyright information.
 Also, line numbering will be removed.
 Journal info and pagination for accepted papers will also be assigned by the Editorial Office.
+
 %------------------
@@
 % If there is only one author the class option oneauthor should be used. Otherwise use the class option moreauthors.
+
 %---------
@@
 \makeatother
-\articlenumber{x}
-\doinum{10.3390/------}
 \pubvolume{xx}
-\pubyear{2018}
-\copyrightyear{2018}
-\externaleditor{Academic Editor: name}
+\issuenum{1}
+\articlenumber{5}
+\pubyear{2019}
+\copyrightyear{2019}
+%\externaleditor{Academic Editor: name}
 \history{Received: date; Accepted: date; Published: date}
+\updates{yes} % If there is an update available, un-comment this line
+
+%% MDPI internal command: uncomment if new journal that already uses continuous page numbers
+%\continuouspages{yes}
 
@@
 around Yangambi, central Congo Basin, in 1958. Our study leveraged
-Structure-from-Motion to generate a large orthomosaic covering 828
-km\textsuperscript{2}, counting 733 million pixels, at a sub meter
-resolution and geo-referenced to \textasciitilde{}4.7 \(\pm\) 4.3 m.
-Primary forest in the orthomosaic was classified with a Convolutional
-Neural Network Deep Learning approach. We used these data to quantify
-LULCC, landscape and canopy complexity and scale above ground biomass
-between historical and contemporary condition. We show a shift from
-historical highly structured industrial deforestation in 1958 (162
-km\textsuperscript{2}), to contemporary smallholder farming clearing (88
-km\textsuperscript{2}), increasing landscape fragmentation and forest
-edges. Efforts to quantify canopy texture and their link to carbon
-storage had limited to no success. Our analysis provides an insight in
-key spatial and temporal patterns of deforestation and reforestation at
-a multi-decadal scale and provide a historical context to land-cover and
-land-use change spatial patterns for past and ongoing field research in
-the area.}
+Structure-from-Motion to generate a large orthomosaic covering 82800 ha,
+counting 733 million pixels, at a sub meter resolution and
+geo-referenced to \textasciitilde{}4.7 \(\pm\) 4.3 m. Primary forest in
+the orthomosaic was classified with a Convolutional Neural Network Deep
+Learning approach. We used these data to quantify LULCC, landscape and
+canopy complexity and scale above ground biomass between historical and
+contemporary condition. We show a shift from historical highly
+structured industrial deforestation in 1958 (16200 ha), to contemporary
+smallholder farming clearing (8800 ha), increasing landscape
+fragmentation and forest edges. Efforts to quantify canopy texture and
+their link to carbon storage had limited to no success. Our analysis
+provides an insight in key spatial and temporal patterns of
+deforestation and reforestation at a multi-decadal scale and provide a
+historical context to land-cover and land-use change spatial patterns
+for past and ongoing field research in the area.}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Only for the journal Diversity
+%\LSID{\url{http://}}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % Only for the journal Applied Sciences:
@@
 
-\usepackage{wrapfig} \usepackage{graphics} \usepackage{setspace}
-
-\doublespacing
-
-\usepackage{gensymb} \usepackage{booktabs}
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
+
+\usepackage{wrapfig} \usepackage{graphics} \usepackage{setspace} \doublespacing \usepackage{gensymb} \usepackage{booktabs}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Only for the journal Gels: Please place the Experimental Section after the Conclusions
 
@@
 
-The functioning of a tropical forest, and its capacity to sequester
-carbon, is heavily influenced by anthropogenic disturbances such as
-selective logging, clear felling, or slash and burn practices, all
-increasing forest fragmentation. This increased pace in the loss of
-rainforest cover \citep{butsic2015} is predicted to negatively impact
-the function of these tropical forests, and their future capacity to
-sequester carbon, as anthropogenic disturbances such as selective
-logging, clear felling, or slash and burn practices collectively
-contribute to increasing forest fragmentation.
-
-These forest fragments and their edges contain fewer large trees
-\citep{laurance2000}, in part due to their hotter and brighter
-microclimates \citep{magnago2015}. Similarly, species composition and
-biodiversity \citep{poorter2006} are negatively affected. Spatial
-disturbances also have an explicit temporal component, where the time
-since disturbance affects the strength of the effects
-\citep{didham1999}. Deforestation and gap formation also influence the
-overall canopy structure, where pioneer species grow taller and with
-narrower crowns \citep{poorter2006, peres2006} affecting the forest
-light regime and carbon dynamics \citep{stark2012}. As such,
-anthropogenic factors through land-use and land cover change (LULCC)
-heavily affect forest structure and functioning
-\citep{fauset2017, brinck2017, didham1999}. In the tropics the majority
-of emissions originate from deforestation and forest degradation
-\citep{hansen2013, houghton2012, tyukavina2015} contributing to 10-15\%
-of the total global emissions \citep{vanderwerf2009} and concomitant
-losses in biodiversity losses \citep{barlow2016, vandeperre2018} and how
-LULCC is evolving over time. Although the African rainforest is the
+Tropical ecosystem services are severely impacted by deforestation and
+forest degradation \citep{hansen2013, houghton2012, tyukavina2015}. Not
+only does tropical forest Land-Use and Land-Cover Change (LULCC)
+constitute 10 to 15\% of the total global emissions
+\citep{vanderwerf2009}, changes in forest fragmentation affect the
+forest structure and function. Strong fragmentation effects decreases
+the number of large trees along forest edges
+\citep{laurance2000, magnago2015}, while species composition and
+biodiversity are equally negatively affected
+\citep{poorter2006, barlow2016, vandeperre2018}. Estimates show that
+31\% of carbon emissions are caused by edge effects alone
+\citep{brinck2017}.
+
+As such, anthropogenic factors through LULCC heavily affect forest
+structure and functioning \citep{fauset2017, brinck2017, didham1999}.
+Accurate estimates of LULCC and forest canopy structure are therefore
+imperative to estimate carbon emissions and other ecosystem services
+\citep{houghton2012, hansen2013}. Remote sensing products have been key
+input as they provide accurate spatial information to help estimate
+carbon emissions \citep{mitchard2018, hansen2013}. High resolution
+aerial images provide scientists tools to monitor forest extent,
+structure and carbon emissions as canopy texture is linked to above
+ground biomass \citep{Ploton2012, Couteron2005, Barbier2010}. Yet, these
+estimates are often limited in time to recent decades
+\citep{defries2002, houghton2012, hansen2013, achard2014}.
+
+Historical estimates \textbf{of Land-Use and Land-Cover (LULC)}, in the
+pre-satellite era (\textless{} 1972), exist but generally rely on
+non-spatially explicit data (i.e socio-economic data)
+\citep{houghton2012, ramankutty1999, defries2002, sader1988}. Efforts
+have been made to use other geo-spatial data sources such as historical
+maps \citep{willcock2016}, declassified CORONA satellite surveillance
+data across the US and central Brazil \citep{song2015} as well as aerial
+surveys in post world war II German \citep{nita2018}. Survey data across
+the African continent is less common, inaccessible or both. Some studies
+do exist, as \citet{buitenwerf2012}, \citet{hudak1998} and
+\citet{okeke2006} used aerial survey images to map vegetation changes in
+South African savannas and the Adulam Nature Reserve, respectively,
+whilst \citet{frankl2015} and \citet{nyssen2016} mapped the Ethiopian
+highlands of 1930's.
+
+Despite the existence of large historical archives of aerial survey
+imagery for the central Congo Basin (Figure 1, Appendix Figure 3) as of
+yet no studies have valorized these data. The lack of a consistent
+valorization effort is unfortunate as the African rainforest is the
 second largest on Earth and covers \textasciitilde{}630 million ha,
 representing up to 66 Pg of carbon storage \citep{Lewis2009b}, and
-currently loses forest at an increasing pace \citep{butsic2015}. Yet,
-the tropics is now becoming a large source of CO\_2 to the atmosphere as
-deforestation and forest degradation
-\citep{hansen2013, houghton2012, tyukavina2015} contribute up to 10-15\%
-of the total global CO2 emissions \citep{vanderwerf2009}. Furthermore,
-recent estimates show that 31\% of recent carbon emissions are caused by
-edge effects alone \citep{brinck2017}. In this context, there are very
-few long term forest inventory plots that presently exist in the central
-Congo Basin \citep{Verbeeck2011a, lewis2013}. This further limits the
-accuracy of current biomass estimates and challenges our understanding
-of how forest structure and function changes in the short and long term
-in these ecosystems \citep{Kearsley2013, Doetterl2015}.
-
-Historical estimates of LULCC, and their carbon emissions, exist
-\citep{houghton2012, ramankutty1999, defries2002, sader1988}, however
-they generally rely on non-spatially explicit data in the pre-satellite
-era (\textless{} 1972) \citep{houghton2010}. These analysis often use
-FAO statistics, historical land-use sources, and/or population and
-socio-economic data \citep{houghton2012}, as limited earlier spatial
-data sources are available \citep{hansen2013, achard2014}. For example,
-the \citet{houghton2010} long-term study used FAO and national censuses
-alongside a bookkeeping model to estimate carbon emissions from LULCC.
-Other studies use remote sensing data, but are limited in time to recent
-decades \citep{defries2002, houghton2012}. These proxy based studies
-have a limited view on long-term direct and indirect spatial effects of
-deforestation. As such, both the spatial structure and the fragmentation
-pattern of a forest combined with its life history has a profound effect
-on carbon emssisions. However, in this context remote sensing
-assessments provide more accurate spatial information to help estimate
-carbon emissions \citep{mitchard2018}.
-
-Past studies expanded knowledge during the pre-satellite time frame
-through the use of now declassified satellite surveillance data to map
-deforestation in the US, central Brazil, and post world war II Germany
-\citep{song2015, nita2018}. In some cases these records could extend
-back in time even further using historical aerial survey images. In
-northern Europe especially these data are well known and used, for
-example to map the extent of forest cover across \textasciitilde{}100
-km\textsuperscript{2} regularly, and at times using automated workflows
-\citep{giordano2018}. Survey data across the African continent is less
-common or inaccessible or both. Some studies do exist, both
-\citet{buitenwerf2012} and \citet{okeke2006} used aerial survey images
-to map vegetation changes in South African savannas and the Adulam
-Nature Reserve, respectively. Whilst \citet{frankl2015} used 20 aerial
-survey images to map the Ethiopian highlands of 1935. High resolution
-aerial or surveillance satellite images provide scientists tools to
-monitor forest extent and structure over long time frames. Moreover,
-canopy texture from aerial survey images has been successfully linked to
-above ground biomass \citep{Ploton2012, Couteron2005, Barbier2010}. Thus
-using texture based metrics obtained from (high resolution) remote
-sensing data opens up possibilities to link historical aerial footage to
-carbon density across large spatial extents \citep{Couteron2005}.
-
-Given the impact of LULCC on the structure and functioning of tropical
-forests, and their influence on both carbon dynamics
+currently loses forest at an increasing pace \citep{butsic2015}. Given
+the impact of LULCC on the structure and functioning of central African
+tropical forests, and their influence on both carbon dynamics
 \citep{Kearsley2013} and biodiversity \citep{vandeperre2018}, accurate
-long term reporting of historical forest cover in the pre-satellite era
-is an imperative \citep{willcock2016}. In this study we use a
-combination of historical aerial photography and contemporary remote
-sensing data to map long-term changes in the extent and structure of the
-tropical forest surrounding Yangambi, in the central Congo Basin at the
-beginning of 1958 and the start of the antrophocene \citep{lewis2015}.
-Our analysis provides a historical insight into important LULCC spatial
-patterns in Yangambi, such as fragmentation and edge complexity. Our
-analysis further contextualises the influence of changes in the forest's
-life history on past and current research into Above Ground Carbon (AGC)
-storage \citep{Kearsley2013} and biodiversity \citep{vandeperre2018} in
-the central Congo Basin.
+long term reporting of historical forest cover warrants more attention
+\citep{willcock2016}.
+
+Here, we use a combination of historical aerial photography and
+contemporary remote sensing data to map long-term changes in the extent
+and structure of the tropical forest surrounding Yangambi, in the
+central Congo Basin, effectively linking the start of the antrophocene
+\citep{lewis2015} with current assessments. Yangambi was, and remains, a
+focal center of forest and agricultural research and development in the
+central Congo Basin. Past research in the region allows for thorough
+assessment of land-use and land-cover change from a multi-disciplinary
+point of view, confronting us with complex deforestation and land-use
+patterns.
+
+We leverage Structure-from-Motion to generate a large orthomosaic of
+historical imagery and develop a Convolutional Neural Network based
+forest cover mapping approach, based upon a semi-supervised generated
+dataset extensively leveraging data augmentation. Our methodology aims
+to provides a historical insight into important LULCC spatial patterns
+in Yangambi, such as fragmentation and edge complexity. We further
+contextualize the influence of changes in the forest's life history on
+past and current research into Above Ground Carbon (AGC) storage
+\citep{Kearsley2013} and biodiversity \citep{vandeperre2018} in the
+central Congo Basin. Our fast scalable mapping approach for historical
+aerial survey data, using limited supervised input, would further
+support long-term land-use and land-cover change analysis across the
+central Congo Basin.
 
@@
 \(\mu\)m) were gathered along flight paths running mostly from west to
-east, between 9 - 11h local time. Along a flight path continuous images
-were taken using a Wild Heerburgg RC5a (currently Leica Geosystems) with
-an Aviogon lens assembly (114.83mm / f 5.6, with a 90\degree view angle)
-resulting in square photo negative of 180 by 180 mm. Flights were flown
-at an average absolute altitude of \textasciitilde{}5200 m above sea
-level, covering roughly 18 530 km\textsuperscript{2} at an approximate
-scale of 1/40 000. The use of the integrated autograph system ensured
-timely acquisition of pictures with a precise overlap
-(\textasciitilde{}1/3) between images. This large overlap between images
-together with flight parameters would allow post-processing, using
-stereographs, to create accurate topographic maps. Original data from
-this campaign are stored in the
+east, between \textbf{09:00 and 11:00} local time. Along a flight path
+continuous images were taken using a Wild Heerburgg RC5a (currently
+Leica Geosystems) with an Aviogon lens assembly (114.83mm / f 5.6, with
+a 90\degree view angle) resulting in square photo negative of 180 by 180
+mm. Flights were flown at an average absolute altitude of
+\textasciitilde{}5200 m above sea level, covering roughly 18 530
+km\textsuperscript{2} at an approximate scale of 1/40 000. The use of
+the integrated autograph system ensured timely acquisition of pictures
+with a precise overlap (\textasciitilde{}1/3) between images. This large
+overlap between images together with flight parameters would allow
+post-processing, using stereographs, to create accurate topographic
+maps. Original data from this campaign are stored in the
 \href{https://www.africamuseum.be/en}{Royal Museum for Central Africa}
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{./figures/print_examples} \caption{A box of historical aerial photographs (left) and a single aerial photograph (right) showing part of the Congo river.
+
+{\centering \includegraphics[width=0.75\linewidth]{./figures/print_examples}
+
+}
+
+\caption{A box of historical aerial photographs (left) and a single aerial photograph (right) showing part of the Congo river.
 Note the meta-data provided in the margins of the image such as acquisition time and fleight height and the solar glare on the Congo river.}\label{fig:unnamed-chunk-1}
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{./figures/flight_paths} \caption{Overview of the historical flight paths during aerial photo acquisition and ancillary data used in this study.
+
+{\centering \includegraphics[width=0.75\linewidth]{./figures/flight_paths}
+
+}
+
+\caption{Overview of the historical flight paths during aerial photo acquisition and ancillary data used in this study.
 The bounding box of the orthomosaic data presented in this study is shown as a rectangle (23x36 km).
@@
 The grey polygon delineates the current day Yangambi Man and Biosphere reserve.
-The inset, top right, situates the greater Yangabmi region with the DR Congo.
+The inset, top right, situates the greater Yangabmi region (white rectangle) with the DR Congo.
 The full flight plan and details are shown in Appendix Figure 1 and 2}\label{fig:unnamed-chunk-2}
@@
 maximal physical resolution of an Epson A3 flatbed scanner (i.e.~2400
-dpi or 160MP per image) and saved as lossless tiff images. Data were
+dpi or 160 MP per image) and saved as lossless tiff images. Data were
 normalized using contrast limited histogram equalization
@@
 Fiduciary marks were used to rectify and downsample the images into
-square 7700x7700 pixel images (\textasciitilde{}1200 dpi, 81MP). This
+square 7700x7700 pixel images (\textasciitilde{}1200 dpi, 81 MP). This
 resulted in a dataset with digital images at a resolution that remained
@@
 optimization routine used in the SfM algorithm. Ground control points
-consisted of rooftop edges of permanent structures which could be
-verified in both old and new aerial imagery (i.e.~ESRI World Imagery).
-Although clouds were removed during the SfM routine we did not mask all
-clouds in the final orthomosaic to maximize forest coverage. The final
-scene was cropped, to provide consistent wall to wall coverage of the
-reconstructed scene. The orthomosaic was exported as a geotiff for
-further georeferencing in QGIS \citep{qgisdevelopmentteam2018} using the
-georeferencer plugin (version 3.1.9) and additional ESRI World Imagery
-high resolution reference data. We used 3\textsuperscript{th} degree
-polynomial and 16 ground control points to correct the final image.
-Ground control points, raw image data and final processed image are
-provided in addition to measures of uncertainty such as mean root mean
-squared (RMSE), mean and median error across all ground control points.
-All subsequent analysis are executed on the final geo-referenced
+consisted of permanent structures which could be verified in both old
+and new aerial imagery (i.e.~ESRI World Imagery) \textbf{and consisted
+of corner points of build structures (e.g.~a building, bridge or
+swimming pool etc.).} Although clouds were removed during the SfM
+routine we did not mask all clouds in the final orthomosaic to maximize
+forest coverage. The final scene was cropped, to provide consistent wall
+to wall coverage of the reconstructed scene. The orthomosaic was
+exported as a geotiff for further georeferencing in QGIS
+\citep{qgisdevelopmentteam2018} using the georeferencer plugin (version
+3.1.9) and additional ESRI World Imagery high resolution reference data.
+We used 3\textsuperscript{th} degree polynomial and 16 ground control
+points to correct the final image. Ground control points, raw image data
+and final processed image are provided in addition to measures of
+uncertainty such as mean and median error across all ground control
+points. All subsequent analysis are executed on the final geo-referenced
 orthomosaic or subsets of it.
@@
 
-\hypertarget{classifying-forest-cover}{%
-\subsubsection{Classifying forest
-cover}\label{classifying-forest-cover}}
+\hypertarget{classifying-land-use-and-land-cover}{%
+\subsubsection{Classifying Land-Use and
+Land-Cover}\label{classifying-land-use-and-land-cover}}
+
+\hypertarget{model-training}{%
+\paragraph{\texorpdfstring{\emph{Model
+training}}{Model training}}\label{model-training}}
 
 We automatically delineated all natural forest in the historical data,
-thus excluding tree plantations, thinned or deteriorated forest stands,
-fields and buildings. We used the Unet Convolutional Neural Net (CNN,
-\citet{ronneberger2015}) architecture implemented in Keras
-\citep{chollet2015} with an efficientnetb3 pre-processing backbone
-\citep{yakubovskiy2019} running on TensorFlow \citep{martinabadi2015} to
-train a binary classifier (i.e.~forest or non-forested). This
-methodology is increasingly being used to automate pixel-level
-classification in (color) digital photography data \citep{buscombe2018}.
-Training data were collected from the orthomosaic by randomly selecting
-513 pixel square tiles from homogeneous forested or non-forested areas
-within the historical orthomosaic. Homogeneous tiles were combined in
+thus excluding tree plantations, thinned or deteriorated forest stands
+which \textbf{showed visible canopy cover loss}, fields and buildings.
+We used the Unet Convolutional Neural Net (CNN, \citet{ronneberger2015})
+architecture implemented in Keras \citep{chollet2015} with an
+efficientnetb3 pre-processing backbone \citep{yakubovskiy2019} running
+on TensorFlow \citep{martinabadi2015} to train a binary classifier
+(i.e.~forest or non-forested). Training data were collected from the
+orthomosaic by randomly selecting 513 pixel square tiles from
+\textbf{locations within} homogeneous forested or non-forested polygons
+in the historical orthomosaic \textbf{(Figure 5). Separate polygons were
+selected for training, testing and validation purposes. Validation
+polygons were sampled 300 times, while both testing and validation
+polygons were sampled at 100 random locations. Tiles extracted from
+locations close to the polygon border at times contained mixed cover
+types. Tiles with mixed cover types were removed from the list of source
+tiles (Table 1).} Homogeneous \textbf{source} tiles were combined in
 synthetic landscapes using a random gaussian field based binary mask
@@
 500 landscapes were generated for both the validation and the testing
-dataset. Source tiles did not repeat across datasets to limit
-overfitting. In order to limit stitch line misclassifications, along the
-seams of mosaicked images, we created synthetic landscapes with
-different forest tiles to mimick forest texture transitions. We applied
-this technique to 10\% of the generated synthetic landscapes. The CNN
-model was trained for 100 epochs with a batch size of 30 using Adam
-optimization \citep{kingma2017}, maximizing the Intersect-over-Union
-(IoU) using Sørensen--Dice and categorical cross-entropy loss functions.
-Data augmentation included random cropping to 320 pixel squares, random
-orientation, scaling, perspective, contrast and brightness shifts and
-image blurring. During final model evaluation we report the IoU of our
-out-of-sample test datasets. The optimized model was used to classify
-the complete orthomosaic using a moving window approach with a step size
-of 110 pixels and a majority vote across overlapping areas to limit
-segmentation edge effects. We refer to Figure 4 for a synoptic overview
-of the full deep learning learning workflow. In addition, we used the
-first acquisition of a recent pan-chromatic Geo-Eye 1 stereo pair
-(GeoEye, 2012, Thornton, Colorado, U.S.A., order 737537, 2011-11-11
-8:55h GMT or 9:55h local time) to classify and visually assess the
-robustness of the algorithm on contemporary remote sensing imagery with
-the same moving window approach.
+datasets for a total of 6000 synthetic landscapes. In order to limit
+stitch line misclassifications, along the seams of mosaicked images, we
+created synthetic landscapes with different forest tiles to mimick
+forest texture transitions. We applied this technique to 10\% of the
+generated synthetic landscapes \textbf{(across training, validation and
+testing data)}.
+
+\begin{table}[!h]
+
+\caption{\label{tab:unnamed-chunk-3}Number of source tiles used for the generation of synthetic landscapes.}
+\centering
+\begin{tabular}[t]{lrrr}
+\toprule
+  & training & testing & validation\\
+\midrule
+forest & 300 & 100 & 100\\
+non-forest & 294 & 92 & 84\\
+\bottomrule
+\end{tabular}
+\end{table}
 
 \begin{figure}
-\includegraphics[width=1\linewidth]{./figures/synthetic_landscape} \caption{An example synthetic landscape, combining homogeneous forest and non-forest images into a patchy landscape using random gaussian field based masks.
-The left panel shows a combined synthetic landscape, while the right panel shows the corresponding forest (black) and non-forest (white) labels.}\label{fig:unnamed-chunk-3}
+
+{\centering \includegraphics[width=1\linewidth]{./figures/synthetic_landscape}
+
+}
+
+\caption{An example synthetic landscape, combining homogeneous forest and non-forest images into a patchy landscape using random gaussian field based masks.
+The left panel shows a combined synthetic landscape, while the right panel shows the corresponding forest (black) and non-forest (white) labels.}\label{fig:unnamed-chunk-4}
 \end{figure}
 
+The CNN model was trained for 100 epochs with a batch size of 30 using
+Adam optimization \citep{kingma2017}, maximizing the
+Intersect-over-Union (IoU) using Sørensen--Dice and categorical
+cross-entropy loss functions. Data augmentation included random cropping
+to 320 pixel squares, random orientation, scaling, perspective, contrast
+and brightness shifts and image blurring. The optimized model was used
+to classify the complete orthomosaic using a moving window approach with
+a step size of 110 pixels and a majority vote \textbf{(\textgreater{}
+50\% agreement)} across overlapping areas to limit segmentation edge
+effects. \textbf{In addition we provide, raw pixel level classification
+agreement data for quality control purposes (see Data availability
+below).} We refer to Figure 4 for a synoptic overview of the full deep
+learning learning workflow.
+
 \begin{figure}
-\includegraphics[width=1\linewidth]{./figures/cnn_diagram} \caption{A diagram of the deep learning workflow followed in training a binary forest / non-forest cover convolutional neural net U-Net model to generate our forest cover map.}\label{fig:unnamed-chunk-4}
+
+{\centering \includegraphics[width=1\linewidth]{./figures/cnn_diagram}
+
+}
+
+\caption{A diagram of the deep learning workflow followed in training a binary forest / non-forest cover convolutional neural net U-Net model to generate our forest cover map.}\label{fig:unnamed-chunk-5}
 \end{figure}
 
+\hypertarget{model-validation}{%
+\paragraph{\texorpdfstring{\emph{Model
+validation}}{Model validation}}\label{model-validation}}
+
+\textbf{We report the CNN accuracy based upon the IoU of our
+out-of-sample validation dataset of synthetic landscapes. In addition,
+we report confusion matrices for all pixels across the homogeneous
+validation polygons, as well as the training and testing polygons (see
+Figure 5). Furthermore, we used the first acquisition of a recent
+pan-chromatic Geo-Eye 1 stereo pair (Geo-Eye, Thornton, Colorado,
+U.S.A., order 737537, 2011-11-11 08:55 GMT or 09:55 local time) to
+classify and assess the robustness of the CNN algorithm on contemporary
+remote sensing imagery with similar spectral and spatial
+characteristics. We used the Global Forest Change version 1.6 (GFC, tile
+10N-020E) \citep{hansen2013} map data, capturing forest loss up to 2011,
+to quantify accuracy on downsampled CNN Geo-Eye classification results.
+Once more, we report the confusion matrix of between the GFC and CNN
+derived forest cover maps, masking clouds and cloud shadows. To
+summarize confusion matrices we report accuracy as:}
+
+\(Accuracy = \frac{(TP + TN)}{(TP + TN + FP + FN)}\)
+
+\textbf{in which TP, TN, FP, FN are True Positive, True Negative, False
+Positive and False Negative, respectivelly.}
+
+\hypertarget{characterizing-long-term-change}{%
+\subsubsection{Characterizing long term
+change}\label{characterizing-long-term-change}}
+
 To map long term land-use and land-cover change in the Yangambi region
-we used the contemporary Global Forest Change version 1.6 data (GFC,
-tile 10N-020E) \citep{hansen2013}. Using this data we calculated the
-latest state of the forest with respect to the conditions at the start
-of 1958, 60 years earlier. In our analysis we only included forested
-pixels which recorded no loss throughout the whole 2000 - 2018 period.
-As the resolution of the historical forest classification exceeds that
-of the GFC map we downsampled our historical forest cover data to 30m
-GFC resolution. We masked out all water bodies using the Global Forest
-Change survey data mask layer, and limited the analysis to the right
-bank of the Congo river. Where the datasets overlap we provide summary
-statistics of deforestation, reforestation and afforestation. The same
-methodology was used to quantify CNN model performance across the
-Geo-Eye 1 scene.
+we used the contemporary Global Forest Change version 1.6 (GFC, tile
+10N-020E) map data \citep{hansen2013}. Using the \textbf{GFC} data we
+calculated the latest state of the forest with respect to the conditions
+at the start of 1958, 60 years earlier. In our analysis we only included
+\textbf{GFC} pixels which recorded no \textbf{forest} loss throughout
+the whole 2000 - 2018 period. As the resolution of the historical forest
+classification exceeds that of the GFC map we downsampled our historical
+forest cover data to 30m GFC resolution. We masked out all water bodies
+using the Global Forest Change survey data mask layer, and limited the
+analysis to the right bank of the Congo river. Where the datasets
+overlap we provide summary statistics of deforestation, reforestation
+and afforestation.
 
@@
 plantations resemble old-regrowth values (81.8 Mg C
-ha\textsuperscript{-1}, see Table A1). We therefore use both the
-estimates of old-regrowth and mixed forest to estimate AGC for regrowth.
-We did not have sufficient data to account for individual changes in AGC
-across plantations.
+ha\textsuperscript{-1}, see \textbf{Appendix Table 1}). We therefore use
+both the estimates of old-regrowth and mixed forest to estimate AGC for
+regrowth. We did not have sufficient data to account for individual
+changes in AGC across plantations.
 
@@
 
-We used an area of 400x400m (16 ha) around contemporary permanent
-sampling plot locations to quantify the general state of the canopy in
-both historical and contemporary pan-chromatic Geo-Eye 1 remote sensing
-data (GeoEye, 2012, Thornton, Colorado, U.S.A., order 737537, 2011-11-11
-8:55h GMT or 9:55h local time). Geo-Eye data were resampled to the
-resolution of the historical data using a nearest neighbour
-interpolation for comparison. For permanent sampling plot locations in
-both historical and contemporary data a rectangular buffer of 200m
-around the centre of all plots was used to create small image subsets.
-All image subsets were processed using the R based FOTO implementation
-and package \citep{hufkens2019a} using global normalization, ensuring
-intercomparability of texture metrics across time and space
-\citep{Barbier2010}. We used a FOTO (moving) window size of 187 pixels
-(or \textasciitilde{}150m), as described in \citet{solorzano2018} and
-\citet{Barbier2010}, to ensure that multiple individual canopies could
-be included in the analysis. A buffer with a radius of 50m around the
-center of all permanent sampling plots was used to determine the average
-value of the first principle component (PC), explaining the majority of
-the variance in canopy texture. Where both the Geo-Eye data and the
-orthomosaic image overlapped we processed the intersecting region to
-explore a wall-to-wall comparison between past and current canopy
-texture metrics.
-
 We used the standard FOTO methodology with fixed zones, instead of the
 moving window approach. The window size was set to the same size (187
-pixels or \textasciitilde{}150m) as used in the moving window analysis
+pixels or \textasciitilde{}150 m) as used in the moving window analysis
 above. To account for illumination differences between the two scenes we
@@
 permanent sampling plots in both image scences were extracted using a
-buffer with a radius of 50m.
+buffer with a radius of 50 m.
 
@@
 
+\hypertarget{orthomosaic-construction}{%
+\subsection{Orthomosaic construction}\label{orthomosaic-construction}}
+
 Our analysis provides a first spatially explicity historical composite
-of aerial survey images mapping LULCC within the Congo Basin. The use of
-high resolution historical images combined with structure from motion
-image processing techniques allowed us to mosaic old imagery across a
-large extent. The final orthomosaic composition of the Yangambi region
-resulted in an image scene covering approximately 733 million pixels
-across \textasciitilde{}828 km\textsuperscript{2}
-(\textasciitilde{}23x36 km, Figure 2). The overall accuracy of the
-structure from motion orthomosaic composition was 0.88 m/pixel using the
-sparse cloud DEM for corrections at 45.8 m/pixel. The resulting
-georeferenced scene had a spatial accuracy of approximately 23m. Further
-georeferencing outside the SfM workflow reduced the mean error at the
-ground control points to 5.3 \(\pm\) 4.9 px (\textasciitilde{}4.7
-\(\pm\) 4.3 m), with a median error of 2.9 px (2.6m). The orthomosaic
-served as input for all subsequent LULCC analysis with all derived maps
-provided with the manuscript repository (see data \& code availability
-statements below).
+of aerial survey images \textbf{in support of mapping land-use and
+land-cover} within the Congo Basin. The use of high resolution
+historical images combined with structure from motion image processing
+techniques allowed us to mosaic old imagery across a large extent. The
+final orthomosaic composition of the Yangambi region \textbf{provided}
+an image scene covering approximately 733 million pixels across
+\textbf{\textasciitilde{}82800 ha} (\textasciitilde{}23x36 km, Figure
+2). The overall accuracy of the SfM orthomosaic composition was 0.88
+m/pixel using the sparse cloud DEM for corrections at 45.8 m/pixel. The
+resulting georeferenced scene had a spatial accuracy of approximately
+\textbf{23 m}. Further georeferencing outside the SfM workflow reduced
+the mean error at the ground control points to 5.3 \(\pm\) 4.9 px
+(\textasciitilde{}4.7 \(\pm\) 4.3 m), with a median error of 2.9 px (2.6
+m). The orthomosaic served as input for all subsequent LULCC analysis
+with all derived maps provided with the manuscript repository (see data
+\& code availability statements below).
 
 \begin{figure}
-\includegraphics[width=1\linewidth]{./figures/orthomosaic_maps} \caption{Overview of the final orthomosaic of the greater Yangambi region (A), a detailed inset (B) and the derived land use land cover change map displayed as the difference between a manual classification of the orthomosaic and a recent map by Hansen et al.
-2013 (C) and the corresponding land cover map for inset B (D).}\label{fig:unnamed-chunk-5}
+
+{\centering \includegraphics[width=1\linewidth]{./figures/orthomosaic_maps}
+
+}
+
+\caption{Overview of the final orthomosaic of the greater Yangambi region (A) as well as the forest cover classification uncertainty (B) used to generate the final Land Land Cover map (see Figure 6).
+Homogeneous polygons used in the generation of the synthetic landscape for Convolutional Neural Network training, testing and validation are marked as dashed and full lines for forest and non-forest regions, respectively.
+Training, testing and validation regions are denoted in panel B in green, red and orange, respectivelly.
+Black polygon outlines denote cloud and image stitch line regions which were manually excluded from analysis, but retained in calculation of validation statistics (see Table 2).}\label{fig:unnamed-chunk-6}
 \end{figure}
 
-\hypertarget{land-use-and-land-cover-change-above-ground-carbon}{%
-\subsection{Land-use and Land-Cover Change \& Above Ground
-Carbon}\label{land-use-and-land-cover-change-above-ground-carbon}}
-
-Our CNN deep learning classifier reached an Intersection-over-Union
-(IoU) accuracy of 97\% on the detection of disturbed forest in
-out-of-sample test data. A visual comparison on recent pan-chromatic
-Geo-Eye 1 data shows good agreement, for a total of 87\% of all pixels,
-between the landsat based GFC data and downscaled CNN results (Appendix
-Table 3 and Figure 7). Scaling our classifier to the whole historical
-orthomosaic we detected 162 km\textsuperscript{2} (or
-\textasciitilde{}20\% of the scene) of disturbed forests. A large
-fraction of the disturbed area was restored in the period between the
-two acquisitions. In total, 99 km\textsuperscript{2}, or little over
-half of the affected forest was restored (Figure 5C-D, dark blue).
-Recent deforested areas, as registered through satellite remote sensing,
-approximate 88 km\textsuperscript{2} (Table1, Figure 5C-D / light
-green).
+\hypertarget{land-use-and-land-cover-classification}{%
+\subsection{Land-use and Land-Cover
+Classification}\label{land-use-and-land-cover-classification}}
+
+\hypertarget{cnn-model-validation}{%
+\subsubsection{CNN model validation}\label{cnn-model-validation}}
+
+\textbf{The CNN deep learning classifier reached an
+Intersection-over-Union (IoU) accuracy of 97\% on the detection of
+disturbed forest in the out-of-sample (validation) synthetic landscape
+data. Using all pixels within the validatoin polygons (Figure 5) showed
+a similar accuracy value of \textasciitilde{}98\%. Using all polygons
+across the scene, including those used in the generation of testing and
+training synthetic landscapes, increased the accuracy to
+\textasciitilde{}99\% (Table 2). A comparison with recent pan-chromatic
+Geo-Eye data shows good agreement, with an accuracy of
+\textasciitilde{}87\% across all pixels, between the landsat based GFC
+data and downscaled CNN results (Table 2 and Figure 7).}
+
+\begin{table}[!h]
+
+\caption{\label{tab:unnamed-chunk-7}Confusion matrix showing \% agreement between forest / non-forest classes using a Convolutional Neural Network (CNN) across previously selected homogenous areas.}
+\centering
+\begin{tabular}[t]{lrrrrrr}
+\toprule
+\multicolumn{1}{c}{CNN} & \multicolumn{2}{c}{Geo-Eye (full scene)} & \multicolumn{2}{c}{Aerial Survey (all polygons)} & \multicolumn{2}{c}{Aerial Survey (validation polygons)} \\
+\cmidrule(l{3pt}r{3pt}){1-1} \cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7}
+  & non-forest & forest & non-forest & forest & non-forest & forest\\
+\midrule
+non-forest & 10.17 & 8.55 & 97.4 & 0.19 & 98.59 & 0.0\\
+forest & 3.75 & 77.52 & 2.6 & 99.81 & 1.41 & 100.0\\
+\addlinespace[0.3em]
+\multicolumn{7}{l}{\textbf{Accuracy}}\\
+\hspace{1em} &  & 87.70 &  & 98.61 &  & 99.3\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{./figures/forest_cover_map}
+
+}
+
+\caption{Overview of the final Land Use Land Cover Change map, a detailed inset of both the underlying orthomosaic (B) and the derived land use land cover change map displayed as the difference between the Convolutional Neural Network based classification orthomosaic and the recent Landsat based forest cover map by Hansen et al.
+2013 (C).}\label{fig:unnamed-chunk-8}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{./figures/visual_comparison_classifiers}
+
+}
+
+\caption{Convolutional Neural Network (CNN) based forest cover classification results (B) as run on a recent (2011) Geo-Eye panchromatic image (A).
+We show the difference between the Convolutional Neural Network based classification and a recent Landsat based forest cover map by Hansen et al.
+2013.
+Full white outlines denote cloud contamination, the dashed rectangle shows a location where the CNN outperforms the Landsat based forest classification.}\label{fig:unnamed-chunk-9}
+\end{figure}
+
+\hypertarget{long-term-lulc-changes-and-above-ground-carbon}{%
+\subsection{Long term LULC changes and Above Ground
+Carbon}\label{long-term-lulc-changes-and-above-ground-carbon}}
+
+Scaling our classifier to the whole historical orthomosaic we detected
+\textbf{16200 ha} (or \textasciitilde{}20\% of the scene) of disturbed
+forests. A large fraction of the disturbed area was restored in the
+period between the two \textbf{acquisition periods}. In total,
+\textbf{9900 ha}, or little over half of the affected forest was
+restored (Figure 5C-D, dark blue). Recent deforested areas, as
+registered through satellite remote sensing, approximate \textbf{8800 ha
+(Table 3, Figure 6 - light green)}.
 
@@
 fishbone pattern for forest clearing with very sharp edges, while
-current patterns are patchy and ad-hoc (figure 5D, dark blue and green
-colours respectively). These differences are reflected in the analysis
-of landscape metrics of forest loss. Between the historical and
+current patterns are patchy and ad-hoc (\textbf{figure 6C}, dark blue
+and green colours respectively). These differences are reflected in the
+analysis of landscape metrics of forest loss. Between the historical and
 contemporary LULCC maps we see an increase in small disturbances, as
@@
 5.02 ha historically. Perimeter lengths were longer historically, at
-1451.58 \(\pm\) 943.27 m, compared to contemporary landscapes
-\textasciitilde{}921.74 \(\pm\) 362.59 m (Table 2). This shift in
-perimeter area ratio led to a similar change in the fractal index,
-increasing in value to 1.1 \(\pm\) 0.05 from 1.09 \(\pm\) 0.04 over
-time. Values closer to a fractal index of 2 suggest a more complex
-(fragmented) landscape.
+1451 \(\pm\) 943 m, compared to contemporary landscapes
+\textasciitilde{}921 \(\pm\) 362 m (Table 3). This shift in perimeter
+area ratio led to a similar change in the fractal index, increasing in
+value to 1.1 \(\pm\) 0.05 from 1.09 \(\pm\) 0.04 over time. Values
+closer to a fractal index of 2 suggest a more complex (fragmented)
+landscape.
 
@@
 0.05). Edge influence did not extend beyond 200 m from a forest edge,
-but still represented an area of 132 km\textsuperscript{2}.
+but still represented an area of \textbf{13200 ha}.
 
@@
 
-\caption{\label{tab:unnamed-chunk-6}Land use land cover change statistics of forest cover around Yangambi in the central Congo Basin.
+\caption{\label{tab:unnamed-chunk-10}Land use land cover change statistics of forest cover around Yangambi in the central Congo Basin.
 The data evaluates a difference between a historical (1958) aerial photography based survey and the Hansen et al.
@@
 \centering
-\begin{tabular}[t]{lrrl}
+\begin{tabular}[t]{lrl}
 \toprule
-\multicolumn{3}{c}{ } & \multicolumn{1}{c}{AGC} \\
-\cmidrule(l{3pt}r{3pt}){4-4}
-  & $km\textsuperscript{2}$ & ha & (Gg C $ha\textsuperscript{-1}$\\
+\multicolumn{2}{c}{ } & \multicolumn{1}{c}{AGC} \\
+\cmidrule(l{3pt}r{3pt}){3-3}
+  & ha & (Gg C $ha\textsuperscript{-1}$\\
 \midrule
-Forest & 685 & 68455 & -\\
--  of which edges & 132 & 13151 & -\\
-Regrowth > 1958 loss & 99 & 9918 & 811 - 1592\\
-Loss > 2000 & 88 & 8776 & 1408\\
-Loss > 1958 (permanent) & 63 & 6282 & 1008\\
+Forest & 68455 & \\
+-  of which edges & 13151 & \\
+Regrowth > 1958 loss & 9918 & 811 - 1592\\
+Loss > 2000 & 8776 & 1408\\
+Loss > 1958 (permanent) & 6282 & 1008\\
 \bottomrule
@@
 
-\caption{\label{tab:unnamed-chunk-7}Landscape metrics for historical and contemporary deforestation patterns.
+\caption{\label{tab:unnamed-chunk-11}Landscape metrics for historical and contemporary deforestation patterns.
 We report patch perimeter and area, their ratio and fractal dimension.
-Values are reportedpat as mean $\pm$ standard deviation, across all deforestation patches.}
+Values are reported as mean $\pm$ standard deviation, across all deforestation patches.}
 \centering
@@
 \emph{Brachystegia laurenttii} is a recent development, changing the
-canopy structure visibly during the last half century (Figure 6). The
-previous varied canopy structure gave way to a more dense and uniform
-canopy. This is reflected in a change of the FOTO PC value from 0.19
-historically to its current value of 0.54. This historical value is
+canopy structure visibly during the last half century (\textbf{Figure
+8}). The previous varied canopy structure gave way to a more dense and
+uniform canopy. This is reflected in a change of the FOTO PC value from
+0.19 historically to its current value of 0.54. This historical value is
 similar to the mean of contemporary mono-dominant stands with PC
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{./figures/visual_comparison_psp} \caption{Visual comparison between a historical (A) and contemporary (B) permanent sampling plot.
+
+{\centering \includegraphics[width=1\linewidth]{./figures/visual_comparison_psp}
+
+}
+
+\caption{Visual comparison between a historical (A) and contemporary (B) permanent sampling plot.
 The site is currently listed as a mono-dominant Brachystegia laurenttii stand.
-Note the structural differences with a 'courser' canopy structure in the historical image compared to the more closed contemporary stand.}\label{fig:unnamed-chunk-8}
+Note the structural differences with a 'courser' canopy structure in the historical image compared to the more closed contemporary stand.}\label{fig:unnamed-chunk-12}
 \end{figure}
@@
 historical and contemporary imagery markedly higher for the
-mono-dominant forest types compared to all others (Appendix Figure 3).
-Provided that the young-regrowth and fallow permanent sampling plots
-have seen recent disturbance the Wilcoxon signed rank test on the mixed
-and mono-dominant plots between the historical and contemporary PC
-values did not show a significant difference (p \textgreater{} 0.05).
-Similarly, no significant different using PC values extracted from the
-whole scene analysis was noted (p \textgreater{} 0.05). Any
-relationships between contemporary Geo-Eye data and permanent sampling
-plot measurements of Above Ground Biomass, stem density and species
-richness were non-significant (p \textgreater{} 0.05, Appendix Figures
-4-6).
+mono-dominant forest types compared to all others (\textbf{Appendix
+Figure 4}). Provided that the young-regrowth and fallow permanent
+sampling plots have seen recent disturbance the Wilcoxon signed rank
+test on the mixed and mono-dominant plots between the historical and
+contemporary PC values did not show a significant difference (p
+\textgreater{} 0.05). Similarly, no significant different using PC
+values extracted from the whole scene analysis was noted (p
+\textgreater{} 0.05). Any relationships between contemporary Geo-Eye
+data and permanent sampling plot measurements of Above Ground Biomass,
+stem density and species richness were non-significant (p \textgreater{}
+0.05, Appendix \textbf{Figures 5-7}).
 
 Furthermore, visual inspection of the scene wide analysis suggests
-historical scences do not show landscape wide canopy features (Figure 7
-A-B), unlike the contemporary scene (Figure 7 C-D). Where the FOTO
-algorithm picks up landscape features such as changes in texture across
-the contemporary Geo-Eye scene (e.g.~the river valley as a diagonal line
-in Figure 7D), however, no corresponding landscape patterns are found by
-the FOTO algorithm in the historical orthomosaic.
+historical scences do not show landscape wide canopy features
+(\textbf{Figure 9 A-B}), unlike the contemporary scene (\textbf{Figure 9
+C-D}). Where the FOTO algorithm picks up landscape features such as
+changes in texture across the contemporary Geo-Eye scene (e.g.~the river
+valley as a diagonal line in \textbf{Figure 9D}), however, no
+corresponding landscape patterns are found by the FOTO algorithm in the
+historical orthomosaic.
 
 \begin{figure}
-\includegraphics[width=1\linewidth]{./figures/foto_maps} \caption{RGB visualizations of the first three principal components of scene wide FOTO texture analysis of historical and current (Geo-Eye) imagery.
-Current permanent sampling plots of mono-dominant, mixed, fallow and young (edge) forest plots are marked with open triangles, open circles, open squares and crosses, respectively.}\label{fig:unnamed-chunk-9}
+
+{\centering \includegraphics[width=1\linewidth]{./figures/foto_maps}
+
+}
+
+\caption{RGB visualizations of the first three principal components of scene wide FOTO texture analysis of historical and current (Geo-Eye) imagery.
+Current permanent sampling plots of mono-dominant, mixed, fallow and young (edge) forest plots are marked with open triangles, open circles, open squares and crosses, respectively.}\label{fig:unnamed-chunk-13}
 \end{figure}
@@
 Finely grained spatial data sources, such as remote sensing imagery, are
-rare before the satellite era (\textless{}1970). This lack of data
+rare before the satellite era (\textless{}1972). This lack of data
 limits our understanding of how forest structure has varied over longer
-time periods in especially remote areas. Long term assessment can be
-extended by using large inventories of historical aerial survey data
-\citep{kadmon1999, song2015, nita2018}. However, across the Congo Basin
-this data is rare, or difficult to access. Despite the difficulties in
-assessing hidden disturbances caused by deforestation, poorly recorded
-official assessments or simply those invisible to standard remote
-sensing techniques \citep{peres2006}, remote sensing generally remains
-the best way to map and quantify LULCC \citep{houghton2012}. Despite
-these challenges in recovery and validation we present a first analysis
-of LULCC using historical aerial survey data for the central Congo
-Basin. We used a combination of Structure-from-Motion and a
-convolutional neural net deep learning approach to map deforestation
-across a large extent, providing a long term view of LULCC in the
-central Congo Basin
+time periods in remote areas. Long term assessment can be extended by
+using large inventories of historical aerial survey data
+\citep{kadmon1999, song2015, nita2018}. Despite the difficulties in
+recovering this data and its limitations, such as invisible disturbances
+\citep{peres2006}, remote sensing generally remains the best way to map
+and quantify LULCC \citep{houghton2012}. In our study we used novel
+numerical remote sensing techniques to valorize, for the first time,
+historical remote sensing data in order to quantify (long term) land-use
+and land-cover change and canopy structural properties in the central
+Congo Basin. Despite these successes some methodological and research
+considerations remain.
+
+\hypertarget{methodological-considerations}{%
+\subsection{Methodological
+considerations}\label{methodological-considerations}}
 
 \hypertarget{data-recovery-challenges}{%
-\subsection{Data recovery challenges}\label{data-recovery-challenges}}
+\subsubsection{Data recovery
+challenges}\label{data-recovery-challenges}}
 
@@
 negative. In addition, analogue photography clearly produces a distinct
-softness compared to digital imagery (Figure 4). Despite favourable
-nadir image acquistions \citep{verhoeven2017} image softness combined
-with illumination effects between flight paths, and the self-similar
-nature of vast canopy expanses \citep{park2019, simini2010, sole1995},
-limited our ability to provide wall-to-wall coverage of the entire
-dataset containing 334 images. Few man made features in the scenes also
-made georeferencing challenging. Although the village of Yangambi
-provided a range of buildings as (hard-edge) references, other areas
-within the central Congo Basin might have fewer permanent structures and
-would require the use of soft-edged landscape features (e.g.~trees,
-river outflows). Research has shown that soft-edged features can help
-georeference scenes even when containing few man-made features
-\citep{hughes2006}, however accuracy becomes compromised. Our two step
-georeferencing approach resulted in a referencing accuracy of
-\textasciitilde{}4.7 \(\pm\) 4.3 m across reference points. However, it
-shoud be noted that referencing accuracy of the final scene is less
-constrained toward the edges of the scene.
+softness compared to digital imagery (\textbf{Figure 8}). Despite
+favourable nadir image acquistions \citep{verhoeven2017} image softness
+combined with illumination effects between flight paths, and the
+self-similar nature of vast canopy expanses
+\citep{park2019, simini2010, sole1995}, limited our ability to provide
+wall-to-wall coverage of the entire dataset containing 334 images. Few
+man made features in the scenes also made georeferencing challenging.
+Although the village of Yangambi provided a range of buildings as
+(hard-edge) references, other areas within the central Congo Basin might
+have fewer permanent structures and would require the use of soft-edged
+landscape features (e.g.~trees, river outflows). Research has shown that
+soft-edged features can help georeference scenes even when containing
+few man-made features \citep{hughes2006}. Our two step georeferencing
+approach resulted in a referencing accuracy of \textasciitilde{}4.7
+\(\pm\) 4.3 m across reference points. However, it shoud be noted that
+referencing accuracy of the final scene is less constrained toward the
+edges of the scene.
+
+\hypertarget{land-use-land-cover-change-classification-validation}{%
+\subsubsection{Land-use, Land-Cover change classification \&
+validation}\label{land-use-land-cover-change-classification-validation}}
+
+\textbf{When classifying the orthomosaic into forest and non-forest
+states we favoured a deep learning supervised classification using a CNN
+over manual segmentation to guarantee an ``apples-to-apples'' comparison
+between the historical and the contemporary GFC forest cover map used in
+our analysis. We acknowledge that both the CNN and GFC land-use and
+land-cover maps use different underlying features, i.e.~spatial or
+spectral data, yet attain a similarly high accuracy of up to 99\%
+\citep{hansen2013}. More so, when validating our CNN classifier against
+GFC data (Figure 7) for a contemporary high resolution Geo-Eye
+panchromatic image we reach an accuracy of \textasciitilde{}87\%,
+despite a time difference of almost 60 years. Visual inspection of the
+classification data in Figure 7 suggests that the GFC map more often
+than not classifies non-forest areas as forest. Actual classification
+accuracy of our algorithm might therefore be higher than our reported
+value.}
+
+\hypertarget{scaling-opportunities}{%
+\subsubsection{Scaling opportunities}\label{scaling-opportunities}}
+
+\textbf{Our approach uses broadly defined homogeneous polygons to
+construct a balanced dataset of synthethic landscapes. The methodology
+is analoguous to the use of sparse labelling as used by
+\citep{buscombe2018} and contrasts with the standard methodologies which
+generally extract pixel (windows) \citep{song2015} or delineate land
+cover classes \citep{nita2018} to drive a classifier or analysis. More
+so, the use of heavy image augmentation during model training sidesteps
+texture representation issues which affect classification of image
+scenes with inconsistent illumination or sharpness \citep{hudak1998} or
+ad-hoc feature engineering \citep{song2015}. The use of synthetic
+landscapes allowed us to account for most, but not all, of the
+variability within our orthomosaic. More so, our analysis has shown that
+despite being trained on historical data our model could map
+contemporary forest cover in remote sensing data with similar spatial
+and spectral characteristics (Figure 7), suggesting that the classifier
+consistently across works both space and time. We acknowledge that our
+approach will speed up labelling, but is limited by the available
+homogeneous areas to sample from and the number of classes. Yet, the
+latter should not be a constrained as previous research efforts have
+focussed on simple forest loss maps \citep{hansen2013}.}
+
+\hypertarget{research-context}{%
+\subsection{Research context}\label{research-context}}
 
@@
 
-When classifying our orthomosaic into forest and non-forest states we
-favoured a deep learning supervised classification using a Convolutional
-Neural Network over manual segmentation to guarantee an
-``apples-to-apples'' comparison between the historical and the
-contemporary GFC forest cover maps. We acknowledge that both
-classifications use different features, i.e.~spectral or spatial data,
-but attain a similarly high accuracy of 97\% and 99\%
-\citep{hansen2013}, for the GFC and our CNN map respectively. Despite a
-number of challenges we were able to map and quantify a vast continuous
-area of tropical forest, with limited manual annotation requirements.
-
 Our analysis shows that the majority of deforestation around Yangambi
-happened toward the late 1950's (162 km\textsuperscript{2}).
-Considerable regrowth has occured since the aerial survey was executed
-(99 km\textsuperscript{2}), and socio-economic instability prevented
-further large scale forest exploitation. In particular, many plantations
-have reached maturity and forest has re-established in previously
-cleared or disturbed areas. The majority of this forest recovery takes
-the form of isolated patches of forest but is offset by further
-deforestation of previously untouched forest. Generally, the function
-and structure of forests can be influenced by forest edges that are
-located up to 1km away however most effects are pronounced within the
-first 300m from the edge \citep{gascon2000}. Our analysis of edge
+happened toward the late 1950's (\textbf{16200 ha}). Considerable
+regrowth has occured since the aerial survey was executed
+(\textbf{\textasciitilde{}9900 ha}), and socio-economic instability
+prevented further large scale forest exploitation. In particular, many
+plantations have reached maturity and forest has re-established in
+previously cleared or disturbed areas. The majority of this forest
+recovery takes the form of isolated patches of forest but is offset by
+further deforestation of previously untouched forest. Generally, the
+function and structure of forests can be influenced by forest edges that
+are located up to 1km away however most effects are pronounced within
+the first 300m from the edge \citep{gascon2000}. Our analysis of edge
 effects on AGC has shown that the influence is negligible 200m away from
@@
 However, it must be stated that the influence of edges throughout the
-landscape was not marginal. as these areas would account for 132
-km\textsuperscript{2}. Thus edges could have a substantial negative
+landscape was not marginal. as these areas would account for
+\textbf{13200 ha}. Thus edges could have a substantial negative
 \citep{brinck2017} or positive \citep{reinmann2017} influence on AGC.
@@
 
-Finally the FOTO technique used to quantify relationships between canopy
-structure and forest characteristics rendered no valuable insights of
-either the historical orthomosaic or recent Geo-Eye scene. Similarly
-weak correlations were found previously by \citet{solorzano2018}. In
-contrast site based texture metric statistics did show correspondence
-between historical and contemporary satellite imagery. However, none of
-them were either consistent or significant. Although visual
-interpretation shows distinctly different canopy structures (Figure 5)
-the differences in how resolution is defined and issues related to image
-quality prevented us from quantifying these further. Unlike large scale
-studies by \citet{Ploton2012} we could not scale this technique to
-historical data. We advise that future valorisation efforts should
-preferentially work from foto negatives (if available) to ensure optimal
-data quality in resolution, contrast and overall sharpness.
+Finally, the FOTO technique used to quantify relationships between
+canopy structure and forest characteristics rendered no valuable
+insights of either the historical orthomosaic or recent Geo-Eye scene.
+Similarly weak correlations were found previously by
+\citet{solorzano2018}. In contrast, site based texture metric statistics
+did show correspondence between historical and contemporary satellite
+imagery. None of them were either consistent or significant. Although
+visual interpretation shows distinctly different canopy structures
+(Figure 5) the differences in how resolution is defined and issues
+related to image quality prevented us from quantifying these further.
+Unlike large scale studies by \citet{Ploton2012} we could not scale this
+technique to historical data. \textbf{The successful use of our CNN
+classification model on a contemporary remote sensing data does suggest
+that texture can be used to consistently capture canopy properties 60
+years apart. This suggests inflexibilit on part of the FOTO technique in
+dealing with non-standardized data.} We advise that future valorisation
+efforts should preferentially work from foto negatives (if available) to
+ensure optimal data quality in resolution, contrast and sharpness.
 
@@
 
-Given the impact of tropical forest disturbances on atmospheric CO\_2
-emissions, biodiversity and ecosystem productivity accurate long term
-reporting on land-cover and land-use change especially in the
-pre-satellite era is an imperative. Our analysis of historical aerial
-survey images of the Central Congo Basin provides a window into the
-state of the forest at the start of the anthropocene. Efforts to
-quantify canopy texture and their link to carbon storage had limited to
-no success.We have shown that historical aerial survey data can be used
-to quantify land-use and land-cover change using a highly automated
-workflow to quantify the structure and extent of forest cover change
-that can help assess the impact of fragmentation on above ground carbon
-stocks. We also show a shift from previously highly structured
-industrial deforestation of large areas, to discrete smallholder
-clearing for farming, increasing landscape fragmentation but also
-opportuntiy for substantial regrowth. Efforts to quantify canopy texture
-and their link to carbon storage had limited to no success. As such, our
-analysis provides insights into the state of rarely studied tropical
-forests and the rate at which deforestation and reforestation has taken
-place over a multi-decadal scale in the central Congo basin providing a
-useful historical context to land-cover and land-use change spatial
-patterns for interpreting past and ongoing field research in the area.
+Given the impact of tropical forest disturbances on atmospheric
+CO\textsubscript{2} emissions, biodiversity and ecosystem productivity
+accurate long term reporting on land-cover and land-use change
+especially in the pre-satellite era is an imperative. Our analysis of
+historical aerial survey images of the Central Congo Basin provides a
+window into the state of the forest at the start of the anthropocene.
+Efforts to quantify canopy texture and their link to carbon storage had
+limited to no success.We have shown that historical aerial survey data
+can be used to quantify land-use and land-cover change using a highly
+automated workflow to quantify the structure and extent of forest cover
+change that can help assess the impact of fragmentation on above ground
+carbon stocks. We also show a shift from previously highly structured
+industrial deforestation of large areas \textbf{for plantation
+purposes}, to discrete smallholder clearing for farming, increasing
+landscape fragmentation but also opportuntiy for substantial regrowth.
+Efforts to quantify canopy texture and their link to carbon storage had
+limited to no success. As such, our analysis provides insights into the
+state of rarely studied tropical forests and the rate at which
+deforestation and reforestation has taken place over a multi-decadal
+scale in the central Congo basin providing a useful historical context
+to land-cover and land-use change spatial patterns for interpreting past
+and ongoing field research in the area.
 
@@
 
-Hufkens et al.~(2019). A curated dataset of aerial survey images over
+Hufkens et al. (2019). A curated dataset of aerial survey images over
 the central Congo Basin, 1958. Zenodo: doi.org/10.5281/zenodo.3547767.
@@
 data) are not shared, but purchase order numbers allow for acquisition
-of these datasets to ensure reproducibility. The \citet{hansen2013}
-image data is freely available from
-\href{https://earthenginepartners.appspot.com/science-2013-global-forest/download_v1.6.html}{the
-GFC project website}.
+of these datasets to ensure reproducibility.
 
@@
 %=====================================
-% \reftitle{References}
-% \begin{thebibliography}{999}
-% % Reference 1
-% \bibitem[Author1(year)]{ref-journal}
-% Author1, T. The title of the cited article. {\em Journal Abbreviation} {\bf 2008}, {\em 10}, 142-149, DOI.
-% % Reference 2
-% \bibitem[Author2(year)]{ref-book}
-% Author2, L. The title of the cited contribution. In {\em The Book Title}; Editor1, F., Editor2, A., Eds.; Publishing House: City, Country, 2007; pp. 32-58, ISBN.
-% \end{thebibliography}
+%\reftitle{References}
+%\begin{thebibliography}{999}
+% Reference 1
+%\bibitem[Author1(year)]{ref-journal}
+%Author1, T. The title of the cited article. {\em Journal Abbreviation} {\bf 2008}, {\em 10}, 142--149.
+% Reference 2
+%\bibitem[Author2(year)]{ref-book}
+%Author2, L. The title of the cited contribution. In {\em The Book Title}; Editor1, F., Editor2, A., Eds.; Publishing House: City, Country, 2007; pp. 32--58.
+%\end{thebibliography}
 
@@
 
+%% for journal Sci
+%\reviewreports{\\
+%Reviewer 1 comments and authors’ response\\
+%Reviewer 2 comments and authors’ response\\
+%Reviewer 3 comments and authors’ response
+%}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2020-01-23 analysis/05_flight_paths.R

@@ -60,12 +60,12 @@ r_map <- r_map +
   geom_sf(data = y_e,
           colour = "grey25",
           fill = NA,
-          size = 1) +
+          size = 0.8) +
   geom_sf(data = g_e,
           colour = "grey25",
           fill = NA,
           linetype = "dashed",
-          size = 1) +
+          size = 0.8) +
   geom_point(data = s,
              aes(x = lon,
                  y = lat,
@@ -79,13 +79,12 @@ r_map <- r_map +
 r_map <- r_map +
   theme_minimal() +
   theme(legend.position="none",
+        axis.text = element_text(size = 13),
+        axis.text.x = element_text(angle = 45, hjust = 1),
         panel.grid.major = element_line(colour="grey"),
-        panel.grid.minor = element_line(colour="grey")
-        ) +
-  labs(
-    x = expression("Longitude (" * degree * ")"),
-    y = expression("Latitude (" * degree * ")")
-  )
+        panel.grid.minor = element_line(colour="grey")) +
+  labs(x = "",
+       y = "")
 
 world <- ne_countries(scale='medium', returnclass = 'sf')
 drc <- subset(world, admin == "Democratic Republic of the Congo")
@@ -97,6 +96,8 @@ drc_inset <- ggplot(data = drc) +
   theme(legend.position = "none",
         axis.text.x = element_blank(),
         axis.text.y = element_blank(),
+        panel.grid.major = element_blank(),
+        panel.grid.minor = element_blank(),
         panel.background = element_rect(fill="white",
                                         colour = "grey25",
                                         size = 1.5)) +
@@ -119,4 +120,4 @@ r_map <- r_map +
 
 ggsave("manuscript/figures/flight_paths.png",
         width = 5,
-        dpi = 150)
+        dpi = 300)

### 2020-01-23 analysis/06_lulcc_map.R

@@ -12,10 +12,22 @@ library(rnaturalearth)
 library(sf)
 library(cowplot)
 
+# homogeneous areas
+h_area <- na.omit(st_read("data/cnn/sample_locations.shp"))
+
+# read cloud and stitch-line mask
+cloud_mask <- st_read("data/maps/cloud_stitch_line_mask.shp")
+
 # load the forest change map
 forest_change <- raster(
   "data/orthomosaic/yangambi_forest_cover_difference_1958_2000.tif")
 
+# QA values
+qa <- raster(
+    "data/orthomosaic/yangambi_forest_mask_qa_resampled.tif")
+qa <- qa * 100 # percentages instead of fractions
+
+# subset
 forest_change_subset <- crop(forest_change,
                              extent(c(24.4356, 24.5165,
                                        0.9156,0.9839)))
@@ -46,6 +58,11 @@ orthomosaic_df <- orthomosaic %>%
   as.data.frame() %>%
   `colnames<-`(c("x", "y", "val"))
 
+qa_df <- qa %>%
+  rasterToPoints %>%
+  as.data.frame() %>%
+  `colnames<-`(c("x", "y", "val"))
+
 orthomosaic_subset_df <- orthomosaic_subset %>%
   rasterToPoints %>%
   as.data.frame() %>%
@@ -60,7 +77,7 @@ forest_change_map <- ggplot()+
                              " forest loss > 2000",
                              " forest loss > 1958")) +
   coord_fixed(ratio = 1) +
-  geom_sf(data = bb, colour = NA, fill = NA) +
+  geom_sf(data = bb, colour = "white", fill = NA, size = 0.8) +
   ylim(c(0.725,1.07)) +
   theme_minimal() +
   theme(legend.position="none",
@@ -85,7 +102,7 @@ forest_change_subset_map <- ggplot()+
   theme(legend.title = element_blank(),
         legend.position="bottom",
         legend.direction="horizontal",
-        axis.text = element_text(size = 9),
+        axis.text = element_text(size = 13),
         axis.text.x = element_text(angle = 45, hjust = 1),
         panel.grid.major = element_line(colour="grey"),
         panel.grid.minor = element_line(colour="grey")) +
@@ -96,13 +113,48 @@ orthomosaic_map <- ggplot()+
   geom_tile(data = orthomosaic_df, aes(x=x,y=y,fill=val)) +
   scale_fill_gradient(low = "black", high = "white") +
[…]

### 2020-01-23 analysis/07_foto_map.R

@@ -78,7 +78,7 @@ historical_map <- ggplot()+
   coord_fixed(ratio = 1) +
   geom_sf(data = s,
           aes(shape = type,
-              size = 3), stroke = 2, colour = "white") +
+              size = 3), stroke = 1.5, colour = "white") +
   scale_shape_manual(values = c(0, 1, 2, 3, 4)) +
   theme_minimal() +
   theme(legend.position="none",
@@ -95,7 +95,7 @@ geoeye_map <- ggplot()+
   scale_fill_gradient(low = "black", high = "white") +
   coord_fixed(ratio = 1) +
   geom_sf(data = s,
-          aes(shape = type, size = 3), stroke = 2,colour = "white") +
+          aes(shape = type, size = 3), stroke = 1.5,colour = "white") +
   scale_shape_manual(values = c(0, 1, 2, 3, 4)) +
   theme_minimal() +
   theme(legend.position="none",
@@ -113,7 +113,7 @@ historical_foto_map <- ggplot()+
   scale_fill_identity() +
   coord_fixed(ratio = 1) +
   geom_sf(data = s,
-          aes(shape = type, size = 3), stroke = 2, colour = "white") +
+          aes(shape = type, size = 3), stroke = 1.5, colour = "white") +
   scale_shape_manual(values = c(0, 1, 2, 3, 4)) +
   theme_minimal()  +
   theme(legend.position="none",
@@ -131,7 +131,7 @@ geoeye_foto_map <- ggplot()+
   scale_fill_identity() +
   coord_fixed(ratio = 1) +
   geom_sf(data = s,
-          aes(shape = type, size = 3), stroke = 2, colour = "white") +
+          aes(shape = type, size = 3), stroke = 1.5, colour = "white") +
   scale_shape_manual(values = c(0, 1, 2, 3, 4)) +
   theme_minimal() +
   theme(legend.position="none",
@@ -151,7 +151,7 @@ p <- plot_grid(historical_map,
                align = "hv",
                axis= "tblr",
                labels = c("A", "B", "C", "D"),
-               label_x = 0.9)
+               label_x = 0)
 
 save_plot("manuscript/figures/foto_maps.png",
           p,
@@ -218,10 +218,10 @@ p <- plot_grid(historical_map,
                align = "hv",
                axis= "tblr",
                labels = c("A", "B"),
-               label_x = 0.9)
+               label_x = 0)
 
 save_plot("manuscript/figures/visual_comparison_psp.png",
[…]

### 2020-01-23 analysis/09_random_image_selection.R

@@ -8,7 +8,7 @@ library(rgdal)
 img <- raster("./data/orthomosaic/yangambi_orthomosaic_modified.tif")
 
 # read in shapefile location digitized in QGIS
-loc <- readOGR("data/cnn/","cnn_data")
+loc <- readOGR("data/cnn/","sample_locations")
 
 # grab the cell values and convert to coordinates (row/col)
 cell_values <- extract(img, loc, cellnumbers = TRUE)[,'cells']

### 2020-01-23 analysis/10_cnn_gfc_geo-eye_validation_map.R

@@ -5,6 +5,10 @@
 library(raster)
 library(tidyverse)
 library(cowplot)
+library(sf)
+
+# read cloud mask
+cloud_mask <- st_read("data/maps/geo_eye_cloud_mask.shp")
 
 # read in cnn forest mask
 m <- raster("./data/geo-eye/geo-eye_forest_mask_cnn.tif")
@@ -12,20 +16,28 @@ m <- raster("./data/geo-eye/geo-eye_forest_mask_cnn.tif")
 # hansen
 hans <- raster("./data/Hansen_et_al/Hansen_GFC_lossyear_yangambi.tif")
 
-# read in Hansen map (get all forest loss before 2013)
+# read in Hansen map (get all forest loss before and in 2011)
 hans <- raster("data/Hansen_et_al/Hansen_GFC_lossyear_yangambi.tif")
 hans <- crop(hans, extent(m))
-loss <- (hans <= 12) + (hans == 0)
+loss <- (hans <= 11) + (hans == 0)
 loss <- (loss > 1)
 loss[is.na(loss)] <- 4
 loss[loss == 1] <- 5
+
+# box in detailed feature
+box <- st_as_sfc(st_bbox(c(
+  xmin = 24.47501,
+  xmax = 24.48995,
+  ymin = 0.80304,
+  ymax = 0.81505
+), crs = st_crs(4326)))
 
 # combine maps
 loss_final <- loss + m
 
 # read in geo-eye downsampled data
 r <- raster("data/geo-eye/geo-eye_panchromatic_latlon_resampled_8bit.tif")
-geoeye_r <- aggregate(r, fact = 6)
+geoeye_r <- aggregate(r, fact = 8)
 
 geoeye_df <- geoeye_r %>%
   rasterToPoints %>%
@@ -41,6 +53,16 @@ geoeye_map <- ggplot()+
   geom_tile(data = geoeye_df, aes(x=x,y=y,fill=val)) +
   scale_fill_gradient(low = "black", high = "white") +
   coord_fixed(ratio = 1) +
+  geom_sf(data = cloud_mask,
+          colour = "white",
+          fill = NA,
+          lty = 1,
+          size = 0.8) +
+  geom_sf(data = box,
+          colour = "white",
+          fill = NA,
+          lty = 2,
+          size = 0.8) +
   theme_minimal() +
   theme(legend.position="none",
         axis.text = element_text(size = 13),
[…]

### 2020-01-23 manuscript/main.Rmd
revision

@@
 In our analysis we only included **GFC** pixels which recorded no **forest** loss throughout the whole 2000 - 2018 period.
-As the resolution of the historical forest classification exceeds that of the GFC map we downsampled our historical forest cover data to 30m GFC resolution.
+As the resolution of the historical forest classification exceeds that of the GFC map we downsampled our historical forest cover data to 30 m GFC resolution **using a nearest neighbour approach**.
 We masked out all water bodies using the Global Forest Change survey data mask layer, and limited the analysis to the right bank of the Congo river.
-Where the datasets overlap we provide summary statistics of deforestation, reforestation and afforestation.
+We provide summary statistics of historical and contemporary forest loss and regrowth.
+
+deforestation (persistent forest loss after 1958), reforestation (regrowth after loss in 1958).
 
@@
 We refer to @Kearsley2013 for the survey method and allometric relations used to scale the survey data.
-Unlike standard square 1 ha plots edge plots were set back 200m from forest edges and were 50x200m, with the 50m side of the plot along the forest edge and continuing 200m into the forest (Appendix Table 1).
+Unlike standard square 1 ha plots edge plots were set back 200m from forest edges and were 50x200 m, with the 50 m side of the plot along the forest edge and continuing 200 m into the forest (Appendix Table 1).
 We further confirmed that forest edge plots did not show a significantly different AGC compared to those of non-edged / mixed forest plots (Mann Whitney U test, p < 0.05).
@@
 Despite the challenges inherent in quantifying AGC for forest edges we mapped the total extent of the edges in the contemporary landscape.
-To align our landscape analysis with exploratory analysis of the survey data we used a buffer of 200m to estimate the extent of forest edges and patches, up to the location of forest edge plots.
+To align our landscape analysis with exploratory analysis of the survey data we used a buffer of 200 m to estimate the extent of forest edges and patches, up to the location of forest edge plots.
 
@@
 
-Scaling our classifier to the whole historical orthomosaic we detected **16200 ha** (or ~20% of the scene) of disturbed forests.
+Scaling our classifier to the whole historical orthomosaic we detected **~16200 ha** (or ~20% of the scene) of disturbed forests.
 A large fraction of the disturbed area was restored in the period between the two **acquisition periods**.
-In total, **9900 ha**, or little over half of the affected forest was restored (Figure 5C-D, dark blue).
-Recent deforested areas, as registered through satellite remote sensing, approximate **8800 ha (Table 3, Figure 6 - light green)**.
+In total, **9918 ha**, or little over half of the affected forest was restored (Figure 5C-D, dark blue).
+Recent deforested areas, as registered through satellite remote sensing, approximate **8776 ha (Table 3, Figure 6 - light green)**.
 
@@
 A comparison of forest edge plots with mixed forest plots showed no signficant difference in AGC, or other reported values such as species richness, basal area or stem density (Mann Whitney U test, p < 0.05).
-Edge influence did not extend beyond 200 m from a forest edge, but still represented an area of **13200 ha**.
+Edge influence did not extend beyond 200 m from a forest edge, but still represented an area of **13151 ha**.
 
@@
 **Our approach uses broadly defined homogeneous polygons to construct a balanced dataset of synthethic landscapes.
-The methodology is analoguous to the use of sparse labelling as used by [@buscombe2018] and contrasts with the standard methodologies which generally extract pixel (windows) [@song2015] or delineate land cover classes [@nita2018] to drive a classifier or analysis.
+The methodology is analoguous to the use of sparse labelling as used by @buscombe2018 and contrasts with the standard methodologies which generally extract pixel (windows) [@song2015] or delineate land cover classes [@nita2018] to drive a classifier or analysis.
 More so, the use of heavy image augmentation during model training sidesteps texture representation issues which affect classification of image scenes with inconsistent illumination or sharpness [@hudak1998] or ad-hoc feature engineering [@song2015].
 The use of synthetic landscapes allowed us to account for most, but not all, of the variability within our orthomosaic.
-More so, our analysis has shown that despite being trained on historical data our model could map contemporary forest cover in remote sensing data with similar spatial and spectral characteristics (Figure 7), suggesting that the classifier consistently across works both space and time.
-We acknowledge that our approach will speed up labelling, but is limited by the available homogeneous areas to sample from and the number of classes.
+Our analysis has shown that despite being trained on historical data our model could map contemporary forest cover in remote sensing data with similar spatial and spectral characteristics (Figure 7), suggesting that the classifier consistently across works both space and time.
+We acknowledge that the use of synthetic landscapes is limited by the available homogeneous areas to sample from and the number of classes.
 Yet, the latter should not be a constrained as previous research efforts have focussed on simple forest loss maps [@hansen2013].**
@@
 
-Our analysis shows that the majority of deforestation around Yangambi happened toward the late 1950's (**16200 ha**).
-Considerable regrowth has occured since the aerial survey was executed (**~9900 ha**), and socio-economic instability prevented further large scale forest exploitation.
+Our analysis shows that the majority of deforestation around Yangambi happened toward the late 1950's (**~16200 ha**).
+Considerable regrowth has occured since the aerial survey was executed (**~9918 ha**), and socio-economic instability prevented further large scale forest exploitation.
 In particular, many plantations have reached maturity and forest has re-established in previously cleared or disturbed areas.
@@
 However, it must be stated that the influence of edges throughout the landscape was not marginal.
-as these areas would account for **13200 ha**.
+as these areas would account for **13151 ha**.
 Thus edges could have a substantial negative [@brinck2017] or positive [@reinmann2017] influence on AGC.
@@
 
-Given the impact of tropical forest disturbances on atmospheric CO~2~ emissions, biodiversity and ecosystem productivity accurate long term reporting on land-cover and land-use change especially in the pre-satellite era is an imperative.
+Given the impact of tropical forest disturbances on atmospheric CO~2~ emissions, biodiversity and ecosystem productivity accurate long term reporting on land-cover and land-use change in the pre-satellite era is an imperative.
 Our analysis of historical aerial survey images of the Central Congo Basin provides a window into the state of the forest at the start of the anthropocene.
-Efforts to quantify canopy texture and their link to carbon storage had limited to no success.We have shown that historical aerial survey data can be used to quantify land-use and land-cover change using a highly automated workflow to quantify the structure and extent of forest cover change that can help assess the impact of fragmentation on above ground carbon stocks.
+Efforts to quantify canopy texture and their link to carbon storage had limited to no success.
+We have shown that historical aerial survey data can be used to quantify land-use and land-cover change using a highly automated workflow to quantify the structure and extent of forest cover change that can help assess the impact of fragmentation on above ground carbon stocks.
 We also show a shift from previously highly structured industrial deforestation of large areas **for plantation purposes**, to discrete smallholder clearing for farming, increasing landscape fragmentation but also opportuntiy for substantial regrowth.

### 2020-01-23 manuscript/main.tex
revision

@@
 classification exceeds that of the GFC map we downsampled our historical
-forest cover data to 30m GFC resolution. We masked out all water bodies
-using the Global Forest Change survey data mask layer, and limited the
-analysis to the right bank of the Congo river. Where the datasets
-overlap we provide summary statistics of deforestation, reforestation
-and afforestation.
+forest cover data to 30 m GFC resolution \textbf{using a nearest
+neighbour approach}. We masked out all water bodies using the Global
+Forest Change survey data mask layer, and limited the analysis to the
+right bank of the Congo river. We provide summary statistics of
+historical and contemporary forest loss and regrowth.
+
+deforestation (persistent forest loss after 1958), reforestation
+(regrowth after loss in 1958).
 
@@
 standard square 1 ha plots edge plots were set back 200m from forest
-edges and were 50x200m, with the 50m side of the plot along the forest
-edge and continuing 200m into the forest (Appendix Table 1). We further
+edges and were 50x200 m, with the 50 m side of the plot along the forest
+edge and continuing 200 m into the forest (Appendix Table 1). We further
 confirmed that forest edge plots did not show a significantly different
@@
 landscape. To align our landscape analysis with exploratory analysis of
-the survey data we used a buffer of 200m to estimate the extent of
+the survey data we used a buffer of 200 m to estimate the extent of
 forest edges and patches, up to the location of forest edge plots.
@@
 Scaling our classifier to the whole historical orthomosaic we detected
-\textbf{16200 ha} (or \textasciitilde{}20\% of the scene) of disturbed
-forests. A large fraction of the disturbed area was restored in the
-period between the two \textbf{acquisition periods}. In total,
-\textbf{9900 ha}, or little over half of the affected forest was
+\textbf{\textasciitilde{}16200 ha} (or \textasciitilde{}20\% of the
+scene) of disturbed forests. A large fraction of the disturbed area was
+restored in the period between the two \textbf{acquisition periods}. In
+total, \textbf{9918 ha}, or little over half of the affected forest was
 restored (Figure 5C-D, dark blue). Recent deforested areas, as
-registered through satellite remote sensing, approximate \textbf{8800 ha
+registered through satellite remote sensing, approximate \textbf{8776 ha
 (Table 3, Figure 6 - light green)}.
@@
 0.05). Edge influence did not extend beyond 200 m from a forest edge,
-but still represented an area of \textbf{13200 ha}.
+but still represented an area of \textbf{13151 ha}.
 
@@
 is analoguous to the use of sparse labelling as used by
-\citep{buscombe2018} and contrasts with the standard methodologies which
+\citet{buscombe2018} and contrasts with the standard methodologies which
 generally extract pixel (windows) \citep{song2015} or delineate land
@@
 landscapes allowed us to account for most, but not all, of the
-variability within our orthomosaic. More so, our analysis has shown that
-despite being trained on historical data our model could map
-contemporary forest cover in remote sensing data with similar spatial
-and spectral characteristics (Figure 7), suggesting that the classifier
-consistently across works both space and time. We acknowledge that our
-approach will speed up labelling, but is limited by the available
-homogeneous areas to sample from and the number of classes. Yet, the
-latter should not be a constrained as previous research efforts have
-focussed on simple forest loss maps \citep{hansen2013}.}
+variability within our orthomosaic. Our analysis has shown that despite
+being trained on historical data our model could map contemporary forest
+cover in remote sensing data with similar spatial and spectral
+characteristics (Figure 7), suggesting that the classifier consistently
+across works both space and time. We acknowledge that the use of
+synthetic landscapes is limited by the available homogeneous areas to
+sample from and the number of classes. Yet, the latter should not be a
+constrained as previous research efforts have focussed on simple forest
+loss maps \citep{hansen2013}.}
 
@@
 Our analysis shows that the majority of deforestation around Yangambi
-happened toward the late 1950's (\textbf{16200 ha}). Considerable
-regrowth has occured since the aerial survey was executed
-(\textbf{\textasciitilde{}9900 ha}), and socio-economic instability
+happened toward the late 1950's (\textbf{\textasciitilde{}16200 ha}).
+Considerable regrowth has occured since the aerial survey was executed
+(\textbf{\textasciitilde{}9918 ha}), and socio-economic instability
 prevented further large scale forest exploitation. In particular, many
@@
 landscape was not marginal. as these areas would account for
-\textbf{13200 ha}. Thus edges could have a substantial negative
+\textbf{13151 ha}. Thus edges could have a substantial negative
 \citep{brinck2017} or positive \citep{reinmann2017} influence on AGC.
@@
 CO\textsubscript{2} emissions, biodiversity and ecosystem productivity
-accurate long term reporting on land-cover and land-use change
-especially in the pre-satellite era is an imperative. Our analysis of
-historical aerial survey images of the Central Congo Basin provides a
-window into the state of the forest at the start of the anthropocene.
-Efforts to quantify canopy texture and their link to carbon storage had
-limited to no success.We have shown that historical aerial survey data
-can be used to quantify land-use and land-cover change using a highly
-automated workflow to quantify the structure and extent of forest cover
-change that can help assess the impact of fragmentation on above ground
-carbon stocks. We also show a shift from previously highly structured
+accurate long term reporting on land-cover and land-use change in the
+pre-satellite era is an imperative. Our analysis of historical aerial
+survey images of the Central Congo Basin provides a window into the
+state of the forest at the start of the anthropocene. Efforts to
+quantify canopy texture and their link to carbon storage had limited to
+no success. We have shown that historical aerial survey data can be used
+to quantify land-use and land-cover change using a highly automated
+workflow to quantify the structure and extent of forest cover change
+that can help assess the impact of fragmentation on above ground carbon
+stocks. We also show a shift from previously highly structured
 industrial deforestation of large areas \textbf{for plantation

### 2020-01-24 manuscript/main.Rmd
revision

@@
 abstract: |
-   Land Use and Land Cover change (LULCC) of African rainforest contribute to global carbon emissions.
-Yet, most historical estimates of LULCC and their carbon emissions rely on non-spatially explicity data in the pre-satellite era (< 1972).
-Past studies have expanded this pre-satellite time frame with declassified satellite surveillance data or aerial surveys, but none cover the Congo Basin.
-Here, we use historical aerial survey photos to map the extent and structure of LULCC around Yangambi, central Congo Basin, in 1958.
-Our study leveraged Structure-from-Motion to generate a large orthomosaic covering 82800 ha, counting 733 million pixels, at a sub meter resolution and geo-referenced to ~4.7 $\pm$ 4.3 m.
-Primary forest in the orthomosaic was classified with a Convolutional Neural Network Deep Learning approach.
-We used these data to quantify LULCC, landscape and canopy complexity and scale above ground biomass between historical and contemporary condition.
-We show a shift from historical highly structured industrial deforestation in 1958 (16200 ha), to contemporary smallholder farming clearing (8800 ha), increasing landscape fragmentation and forest edges.
-Efforts to quantify canopy texture and their link to carbon storage had limited to no success.
-Our analysis provides an insight in key spatial and temporal patterns of deforestation and reforestation at a multi-decadal scale and provide a historical context to land-cover and land-use change spatial patterns for past and ongoing field research in the area.
+   Given the impact of tropical forest disturbances on atmospheric carbon emissions, biodiversity and ecosystem productivity accurate long term reporting of Land-Use and Land-Cover (LULC) change in the pre-satellite era (< 1972) is an imperative.
+Here, we use a combination of historical (1958) aerial photography and contemporary remote sensing data to map long-term changes in the extent and structure of the tropical forest surrounding Yangambi (DR Congo), in the central Congo Basin.
+Our study leveraged Structure-from-Motion and a novel Convolutional Neural Network based LULC classifier, using synthetic landscapes based image augmentation, to map historical forest cover across a large sub meter resolution orthomosaic (~82800 ha) geo-referenced to ~4.7 $\pm$ 4.3 m.
+A comparison with contemporary LULC data showed a shift from previously highly regular industrial deforestation of large areas, to discrete smallholder farming clearing, increasing landscape fragmentation but also providing opportunties for substantial forest regrowth.
+Efforts to quantify long term canopy texture changes and a link to above ground carbon had limited to no success.
+Our analysis provides methods and insights into key spatial and temporal patterns of deforestation and reforestation at a multi-decadal scale, providing a historical context for past and ongoing forest research in the area.
 keywords: |
@@
 KH acknowledges funding from the European Union Marie Skłodowska-Curie Action (project number 797668).
-I'm grateful for the support of Thales D'Haulleville volunteering his time in scanning the images.
+I'm grateful for the support of Thales de Haulleville volunteering his time in scanning the images.
 authorcontributions: |
   K.H.
-conceived and designed the study, analyzed the data, prepared figures and tables, authored, reviewed and approved the final draft of the manuscript.
+conceived and designed the study, analyzed the data, prepared figures, tables and authored the final draft of the manuscript.
 T.d.H.
@@
 Tropical ecosystem services are severely impacted by deforestation and forest degradation [@hansen2013; @houghton2012; @tyukavina2015].
-Not only does tropical forest Land-Use and Land-Cover Change (LULCC) constitute 10 to 15% of the total global emissions [@vanderwerf2009], changes in forest fragmentation affect the forest structure and function.
+Not only does tropical forest Land-Use and Land-Cover Change (LULCC) constitute 10 to 15% of the total global carbon emissions [@vanderwerf2009], changes in forest fragmentation affect the forest structure and function [@fauset2017; @brinck2017; @didham1999].
 Strong fragmentation effects decreases the number of large trees along forest edges [@laurance2000; @magnago2015], while species composition and biodiversity are equally negatively affected [@poorter2006; @barlow2016; @vandeperre2018].
@@
 
-As such, anthropogenic factors through LULCC heavily affect forest structure and functioning [@fauset2017; @brinck2017; @didham1999].
 Accurate estimates of LULCC and forest canopy structure are therefore imperative to estimate carbon emissions and other ecosystem services [@houghton2012; @hansen2013].
-Remote sensing products have been key input as they provide accurate spatial information to help estimate carbon emissions [@mitchard2018; @hansen2013].
-High resolution aerial images provide scientists tools to monitor forest extent, structure and carbon emissions as canopy texture is linked to above ground biomass [@Ploton2012; @Couteron2005; @Barbier2010].
-Yet, these estimates are often limited in time to recent decades [@defries2002; @houghton2012; @hansen2013; @achard2014].
+Remote sensing products have been key inputs in LULCC assessments as they provide accurate spatial information to help estimate carbon emissions [@mitchard2018; @hansen2013].
+More so, high resolution aerial images provide scientists tools to monitor forest extent, structure and carbon emissions as canopy texture is linked to above ground biomass [@Ploton2012; @Couteron2005; @Barbier2010].
+Yet, most of these estimates are limited in time to recent decades [@defries2002; @houghton2012; @hansen2013; @achard2014].
 
 Historical estimates **of Land-Use and Land-Cover (LULC)**, in the pre-satellite era (< 1972), exist but generally rely on non-spatially explicit data (i.e socio-economic data) [@houghton2012; @ramankutty1999; @defries2002; @sader1988].
-Efforts have been made to use other geo-spatial data sources such as historical maps [@willcock2016], declassified CORONA satellite surveillance data across the US and central Brazil [@song2015] as well as aerial surveys in post world war II German [@nita2018].
+Efforts have been made to use other geo-spatial data sources such as historical maps [@willcock2016], declassified CORONA satellite surveillance data across the US and central Brazil [@song2015] as well as aerial surveys in post World War II German [@nita2018].
 Survey data across the African continent is less common, inaccessible or both.
-Some studies do exist, as @buitenwerf2012, @hudak1998 and @okeke2006 used aerial survey images to map vegetation changes in South African savannas and the Adulam Nature Reserve, respectively, whilst @frankl2015 and @nyssen2016 mapped the Ethiopian highlands of 1930's.
+Some studies do exist, as @buitenwerf2012 and @hudak1998 used aerial survey images to map vegetation changes in South African savannas, whilst @frankl2015 and @nyssen2016 mapped the Ethiopian highlands of 1930's.
 
-Despite the existence of large historical archives of aerial survey imagery for the central Congo Basin (Figure 1, Appendix Figure 3) as of yet no studies have valorized these data.
+Across the central Congo Basin most of these historical images were collected within the context of national cartographic efforts by the "Institut Géographique du Congo Belge" in Kinshasa (then Léopoldville), DR Congo.
+Despite the existence of large archives of aerial survey imagery (Figure 1, Appendix Figure 3), as of yet, no studies have valorized these data.
 The lack of a consistent valorization effort is unfortunate as the African rainforest is the second largest on Earth and covers ~630 million ha, representing up to 66 Pg of carbon storage [@Lewis2009b], and currently loses forest at an increasing pace [@butsic2015].
@@
 
-Here, we use a combination of historical aerial photography and contemporary remote sensing data to map long-term changes in the extent and structure of the tropical forest surrounding Yangambi, in the central Congo Basin, effectively linking the start of the antrophocene [@lewis2015] with current assessments.
+Here, we use a combination of historical aerial photography and contemporary remote sensing data to map long-term changes in the extent and structure of the tropical forest surrounding Yangambi (DR Congo), in the central Congo Basin, effectively linking the start of the antrophocene [@lewis2015] with current assessments.
 Yangambi was, and remains, a focal center of forest and agricultural research and development in the central Congo Basin.
@@
 
-Data for the central Congo Basin region, surrounding Kisangani, were collected in several flights during the dry season of 1958 and 1959 (from 8/01/1958 to 20/02/1958 and from 28/12/1958 to 9/01/1959 respectively, see Appendix Figure 1) to generate topographic maps of the area, supervised by the "Institut Géographique du Congo Belge" in Kinshasa (then Léopoldville).
+Data for the central Congo Basin region, surrounding Kisangani, were collected in several flights during the dry season of 1958 and 1959 (from 8/01/1958 to 20/02/1958 and from 28/12/1958 to 9/01/1959 respectively, see Appendix Figure 1) to generate topographic maps of the area, supervised by the "Institut Géographique du Congo Belge" in Kinshasa, DR Congo (then Léopoldville).
 Black-and-White infrared images (0.4 - 0.9 $\mu$m) were gathered along flight paths running mostly from west to east, between **09:00 and 11:00** local time.
-Along a flight path continuous images were taken using a Wild Heerburgg RC5a (currently Leica Geosystems) with an Aviogon lens assembly (114.83mm / f 5.6, with a 90\degree view angle) resulting in square photo negative of 180 by 180 mm.
+Along a flight path continuous images were taken using a Wild Heerburgg RC5a (currently Leica Geosystems) with an Aviogon lens assembly (114.83mm / f 5.6, with a 90\degree view angle) resulting in square photo negatives of 180 by 180 mm.
 Flights were flown at an average absolute altitude of ~5200 m above sea level, covering roughly 18 530 km^2^ at an approximate scale of 1/40 000.
@@
 ```{r echo=FALSE, out.width='75%', fig.align="center", fig.cap="A box of historical aerial photographs (left) and a single aerial photograph (right) showing part of the Congo river.
-Note the meta-data provided in the margins of the image such as acquisition time and fleight height and the solar glare on the Congo river."}
+Note the meta-data provided in the right side margin of the image, such as acquisition time and fleight height."}
 knitr::include_graphics('./figures/print_examples.jpg')
@@
 
-Data was processed into a georeferenced orthomosaic using a Structure from motion (SfM, @ullman1979) approach implemented in [Agisoft Metashape](http://www.agisoft.com) version 1.5.2 (Agisoft LLC, St.
+Data was processed into a georeferenced orthomosaic using a Structure-from-motion (SfM, @ullman1979) approach implemented in [Agisoft Metashape](http://www.agisoft.com) version 1.5.2 (Agisoft LLC, St.
 Petersburg, Russia).
@@
 
-The CNN model was trained for 100 epochs with a batch size of 30 using Adam optimization [@kingma2017], maximizing the Intersect-over-Union (IoU) using Sørensen–Dice and categorical cross-entropy loss functions.
+The CNN model was trained for 100 epochs with a batch size of 30 using Adam optimization [@kingma2017], maximizing the Intersect-over-Union (IoU) using Sørensen–Dice [@dice1945] and categorical cross-entropy loss functions.
 Data augmentation included random cropping to 320 pixel squares, random orientation, scaling, perspective, contrast and brightness shifts and image blurring.
 The optimized model was used to classify the complete orthomosaic using a moving window approach with a step size of 110 pixels and a majority vote **(> 50% agreement)** across overlapping areas to limit segmentation edge effects.
-**In addition we provide, raw pixel level classification agreement data for quality control purposes (see Data availability below).** We refer to Figure 4 for a synoptic overview of the full deep learning learning workflow.
+**In addition, we provide raw pixel level classification agreement data for quality control purposes (see Data availability below).** We refer to Figure 4 for a synoptic overview of the full deep learning workflow.
 
@@
 
-To map long term land-use and land-cover change in the Yangambi region we used the contemporary Global Forest Change version 1.6 (GFC, tile 10N-020E) map data [@hansen2013].
+To map long term LULCC in the Yangambi region we used the contemporary Global Forest Change version 1.6 (GFC, tile 10N-020E) map data [@hansen2013].
 Using the **GFC** data we calculated the latest state of the forest with respect to the conditions at the start of 1958, 60 years earlier.
@@
 We provide summary statistics of historical and contemporary forest loss and regrowth.
-
-deforestation (persistent forest loss after 1958), reforestation (regrowth after loss in 1958).
+**We map persistent forest loss after 1958, regrowth after loss in 1958, recent forest loss and long term (stable) forest cover.**
 
@@
 We refer to @Kearsley2013 for the survey method and allometric relations used to scale the survey data.
-Unlike standard square 1 ha plots edge plots were set back 200m from forest edges and were 50x200 m, with the 50 m side of the plot along the forest edge and continuing 200 m into the forest (Appendix Table 1).
+Unlike standard square 1 ha plots, edge plots were set back 200 m from forest edges and were 50x200 m, with the 50 m side of the plot along the forest edge and continuing 200 m into the forest (Appendix Table 1).
 We further confirmed that forest edge plots did not show a significantly different AGC compared to those of non-edged / mixed forest plots (Mann Whitney U test, p < 0.05).
@@
 Surveys of old plantations show a large variation in AGC, depending on age and the crop type.
-For example, the AGC values varied from 168.67 to 86.55 g Mg C ha^-1^, for *Hevea brasiliensis* (rubber tree) and *Elaeis guineensis* (oil palm) plots respectively (@bustillo2018, personal communications).
+For example, the AGC values varied from 168.67 to 86.55 Mg C ha^-1^, for *Hevea brasiliensis* (rubber tree) and *Elaeis guineensis* (oil palm) plots respectively (@bustillo2018, personal communications).
 These higher values are in line with the mixed AGC estimates in the area, while the palm plantations resemble old-regrowth values (81.8 Mg C ha^-1^, see **Appendix Table 1**).
@@
 We compared the structure of the canopy both visually and using Fourier Transform Textural Ordination (FOTO, @couteron2002).
-Fourier Transform Textural Ordination uses a principal component analysis (PCA) on radially averaged 2D Fourier spectra to characterize canopy (image) texture.
+FOTO uses a principal component analysis (PCA) on radially averaged 2D Fourier spectra to characterize canopy (image) texture.
 The FOTO technique was first described by @couteron2002 to quantify canopy stucture in relation to biomass and biodiversity, and can be used across multiple scenes using normalization [@Barbier2010].
@@
 
-For both site based and scene analysis we correlate PC values with permanent sample plot inventory data such as stem density, above ground biomass and tree species richness.
+For both site based and scene analysis we correlated PC values with permanent sample plot inventory data such as stem density, above ground biomass and tree species richness.
 Additional comparisons are made between contemporary Geo-Eye data and the historical orthomosaic derived PC values.
-Due to the few available permanent sampling plots in both scenes we used a non-parametric paired signed rank (Wilcoxon) test to determine differences between the PC values of the Geo-Eye and historical orthomosaic image scenes across mono-dominant and mixed forest types.
+Due to the few available permanent sampling plots in both scenes we used a non-parametric paired signed rank (Wilcoxon) test [@wilcoxon1945] to determine differences between the PC values of the Geo-Eye and historical orthomosaic image scenes across mono-dominant and mixed forest types.
 In all analysis, mono-dominant site 4 was removed from the analysis due to cloud contamination.
@@
 Our analysis provides a first spatially explicity historical composite of aerial survey images **in support of mapping land-use and land-cover** within the Congo Basin.
-The use of high resolution historical images combined with structure from motion image processing techniques allowed us to mosaic old imagery across a large extent.
+The use of high resolution historical images combined with SfM image processing techniques allowed us to mosaic old imagery across a large extent.
 The final orthomosaic composition of the Yangambi region **provided** an image scene covering approximately 733 million pixels across **~82800 ha** (~23x36 km, Figure 2).
@@
 
-**The CNN deep learning classifier reached an Intersection-over-Union (IoU) accuracy of 97% on the detection of disturbed forest in the out-of-sample (validation) synthetic landscape data.
+**The CNN deep learning classifier reached an Intersection-over-Union accuracy of 97% on the detection of disturbed forest in the out-of-sample (validation) synthetic landscape data.
 Using all pixels within the validatoin polygons (Figure 5) showed a similar accuracy value of ~98%.
@@
 
-## Long term LULC changes and Above Ground Carbon
+## Long term changes in LULC and Above Ground Carbon
 
@@
 
-Recent deforestation follows a distincly different pattern compared to historical patterns.
+Recent deforestation follows a distinctly different pattern compared to historical patterns.
 Historical deforestation showed a classical fishbone pattern for forest clearing with very sharp edges, while current patterns are patchy and ad-hoc (**figure 6C**, dark blue and green colours respectively).
@@
 Perimeter lengths were longer historically, at 1451 $\pm$ 943 m,  compared to contemporary landscapes ~921 $\pm$ 362 m (Table 3).
-This shift in perimeter area ratio led to a similar change in the fractal index, increasing in value to 1.1 $\pm$ 0.05 from 1.09 $\pm$ 0.04 over time.
+This shift in perimeter area ratio led to a similar change in the fractal index, slightly increasing in value to 1.1 $\pm$ 0.05 from 1.09 $\pm$ 0.04 over time.
 Values closer to a fractal index of 2 suggest a more complex (fragmented) landscape.
@@
 
-Changes in both land-use and land-cover led to concomitant changes in above ground carbon stocks.
+Changes in both land-use and land-cover led to concomitant changes in AGC stocks.
 Recovery throughout the region was characterized for patches of forest and plantations.
@@
        col.names = c("ha",
-                     "(Gg C $ha\\textsuperscript{-1}$")
+                     "Gg C $ha\\textsuperscript{-1}$")
          ) %>%
@@
 Visual interpretation of the scenes provide evidence that most locations do not change dramatically with respect to canopy composition, except for the large areas of disturbances in contemporary fallow or young-regrowth plots.
-One marked difference is noted in the mono-dominant site 6 (Appendix Table 1).
-Here, the current mono-dominant *Brachystegia laurenttii* is a recent development, changing the canopy structure visibly during the last half century (**Figure 8**).
+One marked difference is noted in the mono-dominant plot 6 (Appendix Table 1).
+Here, the current mono-dominant *Brachystegia laurentii* is a recent development, changing the canopy structure visibly during the last half century (**Figure 8**).
 The previous varied canopy structure gave way to a more dense and uniform canopy.
 This is reflected in a change of the FOTO PC value from 0.19 historically to its current value of 0.54.
-This historical value is similar to the mean of contemporary mono-dominant stands with PC averaging 0.34 $\pm$ 0.1, and is only slightly higher than historical values for a mixed forest (0.18 $\pm$ 0.08).
+This historical value is similar to the mean of contemporary mono-dominant stands of *Gilbertiodendron dewevrei* with PC averaging 0.34 $\pm$ 0.1, and is only slightly higher than historical values for a mixed forest (0.18 $\pm$ 0.08).
 The reverse pattern is seen in the contemporary PC values.
@@
 Similarly, no significant different using PC values extracted from the whole scene analysis was noted (p > 0.05).
-Any relationships between contemporary Geo-Eye data and permanent sampling plot measurements of Above Ground Biomass, stem density and species richness were non-significant (p > 0.05, Appendix **Figures 5-7**).
+Any relationships between contemporary Geo-Eye data and permanent sampling plot measurements of Above Ground Carbon, stem density and species richness were non-significant (p > 0.05, Appendix **Figures 5-7**).
 
@@
 
-### Land-use, Land-Cover change classification & validation
+### LULC classification & validation
 
@@
 The use of synthetic landscapes allowed us to account for most, but not all, of the variability within our orthomosaic.
-Our analysis has shown that despite being trained on historical data our model could map contemporary forest cover in remote sensing data with similar spatial and spectral characteristics (Figure 7), suggesting that the classifier consistently across works both space and time.
+Our analysis has shown that despite being trained on historical data our model could map contemporary forest cover in remote sensing data with similar spatial and spectral characteristics (Figure 7), suggesting that the classifier consistently works across both space and time.
 We acknowledge that the use of synthetic landscapes is limited by the available homogeneous areas to sample from and the number of classes.
@@
 
-## Land-use, Land-Cover change & Above Ground Carbon
+### Long term changes in LULC and Above Ground Carbon
 
@@
 The majority of this forest recovery takes the form of isolated patches of forest but is offset by further deforestation of previously untouched forest.
-Generally, the function and structure of forests can be influenced by forest edges that are located up to 1km away however most effects are pronounced within the first 300m from the edge [@gascon2000].
-Our analysis of edge effects on AGC has shown that the influence is negligible 200m away from the edge.
+Generally, the function and structure of forests can be influenced by forest edges that are located up to 1km away however most effects are pronounced within the first 300 m from the edge [@gascon2000].
+Our analysis of edge effects on AGC has shown that the influence is negligible 200 m away from the edge.
 @phillips2006 have shown similar weak responses to edge effects in the Amazon forest.
-Due to a lack of data on the extent (depth) of edge effects and their influence on AGC beyond 200m we did not include any estimates of carbon loss or gain within these zones.
-However, it must be stated that the influence of edges throughout the landscape was not marginal.
-as these areas would account for **13151 ha**.
+Due to a lack of data on the extent (depth) of edge effects and their influence on AGC beyond 200 m we did not include any estimates of carbon loss or gain within these zones.
+However, it must be stated that the influence of edges throughout the landscape was not marginal as these areas would account for **13151 ha**.
 Thus edges could have a substantial negative [@brinck2017] or positive [@reinmann2017] influence on AGC.
 Similarly, uncertainties in how to explicitly correct for plantations in the landscape present a further challenge.
-Thus although our estimates are only indicative they do underscore the important influence of landscape structure when carbon accounting.
+Thus, although our estimates are only indicative they do underscore the important influence of landscape structure when carbon accounting.
 However, our findings do not indicate that deforestation in Congo basin is declining, on the contrary.
@@
 
-## Canopy structure & FOTO texture analysis
+### Canopy structure & FOTO texture analysis
 
@@
 **The successful use of our CNN classification model on a contemporary remote sensing data does suggest that texture can be used to consistently capture canopy properties 60 years apart.
-This suggests inflexibilit on part of the FOTO technique in dealing with non-standardized data.** We advise that future valorisation efforts should preferentially work from foto negatives (if available) to ensure optimal data quality in resolution, contrast and sharpness.
+This suggests inflexibility on part of the FOTO technique in dealing with non-standardized data.** We advise that future valorisation efforts should preferentially work from foto negatives (if available) to ensure optimal data quality in resolution, contrast and sharpness.
 
@@
 
-Given the impact of tropical forest disturbances on atmospheric CO~2~ emissions, biodiversity and ecosystem productivity accurate long term reporting on land-cover and land-use change in the pre-satellite era is an imperative.
-Our analysis of historical aerial survey images of the Central Congo Basin provides a window into the state of the forest at the start of the anthropocene.
-Efforts to quantify canopy texture and their link to carbon storage had limited to no success.
-We have shown that historical aerial survey data can be used to quantify land-use and land-cover change using a highly automated workflow to quantify the structure and extent of forest cover change that can help assess the impact of fragmentation on above ground carbon stocks.
-We also show a shift from previously highly structured industrial deforestation of large areas **for plantation purposes**, to discrete smallholder clearing for farming, increasing landscape fragmentation but also opportuntiy for substantial regrowth.
-Efforts to quantify canopy texture and their link to carbon storage had limited to no success.
-As such, our analysis provides insights into the state of rarely studied tropical forests and the rate at which deforestation and reforestation has taken place over a multi-decadal scale in the central Congo basin providing a useful historical context to land-cover and land-use change spatial patterns for interpreting past and ongoing field research in the area.
+Given the impact of tropical forest disturbances on atmospheric carbon emissions, biodiversity and ecosystem productivity accurate long term reporting of LULCC is an imperative.
+Our analysis of historical aerial survey images (1958) of the Central Congo Basin provides a window into the state of the forest at the start of the anthropocene.
+The use of a CNN based LULC classifier, using synthetic landscapes based image augmentation, provides a robust semi-supervised solution which scales across space and time, even for image scenes with inconsistent illumination or sharpness.
+Combined with contemporary remote sensing data we have shown that historical aerial survey data can be used to quantify long-term changes in LULC and AGC.
+We showed a shift from previously highly structured industrial deforestation of large areas **for plantation purposes**, to discrete smallholder clearing for farming, increasing landscape fragmentation but also opportunties for substantial regrowth.
+Efforts to quantify canopy texture features and their link to AGC had limited to no success.
+Our analysis provides insights into the rate at which deforestation and reforestation has taken place over a multi-decadal scale in the central Congo basin.
+As such, it provides a useful historical context while interpreting past and ongoing forest research in the area.
 
@@
 A curated dataset of aerial survey images over the central Congo Basin, 1958.
-Zenodo: doi.org/10.5281/zenodo.3547767.
+Zenodo: https://doi.org/10.5281/zenodo.3547767.
 All data not included in the latter repository can be found bundled with the analysis code as listed below.

### 2020-01-24 manuscript/main.tex
revision

@@
 % Abstract (Do not insert blank lines, i.e. \\)
-\abstract{Land Use and Land Cover change (LULCC) of African rainforest contribute
-to global carbon emissions. Yet, most historical estimates of LULCC and
-their carbon emissions rely on non-spatially explicity data in the
-pre-satellite era (\textless{} 1972). Past studies have expanded this
-pre-satellite time frame with declassified satellite surveillance data
-or aerial surveys, but none cover the Congo Basin. Here, we use
-historical aerial survey photos to map the extent and structure of LULCC
-around Yangambi, central Congo Basin, in 1958. Our study leveraged
-Structure-from-Motion to generate a large orthomosaic covering 82800 ha,
-counting 733 million pixels, at a sub meter resolution and
-geo-referenced to \textasciitilde{}4.7 \(\pm\) 4.3 m. Primary forest in
-the orthomosaic was classified with a Convolutional Neural Network Deep
-Learning approach. We used these data to quantify LULCC, landscape and
-canopy complexity and scale above ground biomass between historical and
-contemporary condition. We show a shift from historical highly
-structured industrial deforestation in 1958 (16200 ha), to contemporary
-smallholder farming clearing (8800 ha), increasing landscape
-fragmentation and forest edges. Efforts to quantify canopy texture and
-their link to carbon storage had limited to no success. Our analysis
-provides an insight in key spatial and temporal patterns of
-deforestation and reforestation at a multi-decadal scale and provide a
-historical context to land-cover and land-use change spatial patterns
-for past and ongoing field research in the area.}
+\abstract{Given the impact of tropical forest disturbances on atmospheric carbon
+emissions, biodiversity and ecosystem productivity accurate long term
+reporting of Land-Use and Land-Cover (LULC) change in the pre-satellite
+era (\textless{} 1972) is an imperative. Here, we use a combination of
+historical (1958) aerial photography and contemporary remote sensing
+data to map long-term changes in the extent and structure of the
+tropical forest surrounding Yangambi (DR Congo), in the central Congo
+Basin. Our study leveraged Structure-from-Motion and a novel
+Convolutional Neural Network based LULC classifier, using synthetic
+landscapes based image augmentation, to map historical forest cover
+across a large sub meter resolution orthomosaic (\textasciitilde{}82800
+ha) geo-referenced to \textasciitilde{}4.7 \(\pm\) 4.3 m. A comparison
+with contemporary LULC data showed a shift from previously highly
+regular industrial deforestation of large areas, to discrete smallholder
+farming clearing, increasing landscape fragmentation but also providing
+opportunties for substantial forest regrowth. Efforts to quantify long
+term canopy texture changes and a link to above ground carbon had
+limited to no success. Our analysis provides methods and insights into
+key spatial and temporal patterns of deforestation and reforestation at
+a multi-decadal scale, providing a historical context for past and
+ongoing forest research in the area.}
 
@@
 only does tropical forest Land-Use and Land-Cover Change (LULCC)
-constitute 10 to 15\% of the total global emissions
+constitute 10 to 15\% of the total global carbon emissions
 \citep{vanderwerf2009}, changes in forest fragmentation affect the
-forest structure and function. Strong fragmentation effects decreases
-the number of large trees along forest edges
+forest structure and function
+\citep{fauset2017, brinck2017, didham1999}. Strong fragmentation effects
+decreases the number of large trees along forest edges
 \citep{laurance2000, magnago2015}, while species composition and
@@
 
-As such, anthropogenic factors through LULCC heavily affect forest
-structure and functioning \citep{fauset2017, brinck2017, didham1999}.
 Accurate estimates of LULCC and forest canopy structure are therefore
@@
 \citep{houghton2012, hansen2013}. Remote sensing products have been key
-input as they provide accurate spatial information to help estimate
-carbon emissions \citep{mitchard2018, hansen2013}. High resolution
-aerial images provide scientists tools to monitor forest extent,
-structure and carbon emissions as canopy texture is linked to above
-ground biomass \citep{Ploton2012, Couteron2005, Barbier2010}. Yet, these
-estimates are often limited in time to recent decades
+inputs in LULCC assessments as they provide accurate spatial information
+to help estimate carbon emissions \citep{mitchard2018, hansen2013}. More
+so, high resolution aerial images provide scientists tools to monitor
+forest extent, structure and carbon emissions as canopy texture is
+linked to above ground biomass
+\citep{Ploton2012, Couteron2005, Barbier2010}. Yet, most of these
+estimates are limited in time to recent decades
 \citep{defries2002, houghton2012, hansen2013, achard2014}.
@@
 data across the US and central Brazil \citep{song2015} as well as aerial
-surveys in post world war II German \citep{nita2018}. Survey data across
+surveys in post World War II German \citep{nita2018}. Survey data across
 the African continent is less common, inaccessible or both. Some studies
-do exist, as \citet{buitenwerf2012}, \citet{hudak1998} and
-\citet{okeke2006} used aerial survey images to map vegetation changes in
-South African savannas and the Adulam Nature Reserve, respectively,
+do exist, as \citet{buitenwerf2012} and \citet{hudak1998} used aerial
+survey images to map vegetation changes in South African savannas,
 whilst \citet{frankl2015} and \citet{nyssen2016} mapped the Ethiopian
@@
 
-Despite the existence of large historical archives of aerial survey
-imagery for the central Congo Basin (Figure 1, Appendix Figure 3) as of
-yet no studies have valorized these data. The lack of a consistent
-valorization effort is unfortunate as the African rainforest is the
-second largest on Earth and covers \textasciitilde{}630 million ha,
-representing up to 66 Pg of carbon storage \citep{Lewis2009b}, and
-currently loses forest at an increasing pace \citep{butsic2015}. Given
-the impact of LULCC on the structure and functioning of central African
-tropical forests, and their influence on both carbon dynamics
-\citep{Kearsley2013} and biodiversity \citep{vandeperre2018}, accurate
-long term reporting of historical forest cover warrants more attention
-\citep{willcock2016}.
+Across the central Congo Basin most of these historical images were
+collected within the context of national cartographic efforts by the
+``Institut Géographique du Congo Belge'' in Kinshasa (then
+Léopoldville), DR Congo. Despite the existence of large archives of
+aerial survey imagery (Figure 1, Appendix Figure 3), as of yet, no
+studies have valorized these data. The lack of a consistent valorization
+effort is unfortunate as the African rainforest is the second largest on
+Earth and covers \textasciitilde{}630 million ha, representing up to 66
+Pg of carbon storage \citep{Lewis2009b}, and currently loses forest at
+an increasing pace \citep{butsic2015}. Given the impact of LULCC on the
+structure and functioning of central African tropical forests, and their
+influence on both carbon dynamics \citep{Kearsley2013} and biodiversity
+\citep{vandeperre2018}, accurate long term reporting of historical
+forest cover warrants more attention \citep{willcock2016}.
 
@@
 contemporary remote sensing data to map long-term changes in the extent
-and structure of the tropical forest surrounding Yangambi, in the
-central Congo Basin, effectively linking the start of the antrophocene
-\citep{lewis2015} with current assessments. Yangambi was, and remains, a
-focal center of forest and agricultural research and development in the
-central Congo Basin. Past research in the region allows for thorough
-assessment of land-use and land-cover change from a multi-disciplinary
-point of view, confronting us with complex deforestation and land-use
-patterns.
+and structure of the tropical forest surrounding Yangambi (DR Congo), in
+the central Congo Basin, effectively linking the start of the
+antrophocene \citep{lewis2015} with current assessments. Yangambi was,
+and remains, a focal center of forest and agricultural research and
+development in the central Congo Basin. Past research in the region
+allows for thorough assessment of land-use and land-cover change from a
+multi-disciplinary point of view, confronting us with complex
+deforestation and land-use patterns.
 
@@
 area, supervised by the ``Institut Géographique du Congo Belge'' in
-Kinshasa (then Léopoldville). Black-and-White infrared images (0.4 - 0.9
-\(\mu\)m) were gathered along flight paths running mostly from west to
-east, between \textbf{09:00 and 11:00} local time. Along a flight path
-continuous images were taken using a Wild Heerburgg RC5a (currently
-Leica Geosystems) with an Aviogon lens assembly (114.83mm / f 5.6, with
-a 90\degree view angle) resulting in square photo negative of 180 by 180
-mm. Flights were flown at an average absolute altitude of
+Kinshasa, DR Congo (then Léopoldville). Black-and-White infrared images
+(0.4 - 0.9 \(\mu\)m) were gathered along flight paths running mostly
+from west to east, between \textbf{09:00 and 11:00} local time. Along a
+flight path continuous images were taken using a Wild Heerburgg RC5a
+(currently Leica Geosystems) with an Aviogon lens assembly (114.83mm / f
+5.6, with a 90\degree view angle) resulting in square photo negatives of
+180 by 180 mm. Flights were flown at an average absolute altitude of
 \textasciitilde{}5200 m above sea level, covering roughly 18 530
@@
 \caption{A box of historical aerial photographs (left) and a single aerial photograph (right) showing part of the Congo river.
-Note the meta-data provided in the margins of the image such as acquisition time and fleight height and the solar glare on the Congo river.}\label{fig:unnamed-chunk-1}
+Note the meta-data provided in the right side margin of the image, such as acquisition time and fleight height.}\label{fig:unnamed-chunk-1}
 \end{figure}
@@
 
-Data was processed into a georeferenced orthomosaic using a Structure
-from motion (SfM, \citet{ullman1979}) approach implemented in
+Data was processed into a georeferenced orthomosaic using a
+Structure-from-motion (SfM, \citet{ullman1979}) approach implemented in
 \href{http://www.agisoft.com}{Agisoft Metashape} version 1.5.2 (Agisoft
@@
 Adam optimization \citep{kingma2017}, maximizing the
-Intersect-over-Union (IoU) using Sørensen--Dice and categorical
-cross-entropy loss functions. Data augmentation included random cropping
-to 320 pixel squares, random orientation, scaling, perspective, contrast
-and brightness shifts and image blurring. The optimized model was used
-to classify the complete orthomosaic using a moving window approach with
-a step size of 110 pixels and a majority vote \textbf{(\textgreater{}
-50\% agreement)} across overlapping areas to limit segmentation edge
-effects. \textbf{In addition we provide, raw pixel level classification
-agreement data for quality control purposes (see Data availability
-below).} We refer to Figure 4 for a synoptic overview of the full deep
-learning learning workflow.
+Intersect-over-Union (IoU) using Sørensen--Dice \citep{dice1945} and
+categorical cross-entropy loss functions. Data augmentation included
+random cropping to 320 pixel squares, random orientation, scaling,
+perspective, contrast and brightness shifts and image blurring. The
+optimized model was used to classify the complete orthomosaic using a
+moving window approach with a step size of 110 pixels and a majority
+vote \textbf{(\textgreater{} 50\% agreement)} across overlapping areas
+to limit segmentation edge effects. \textbf{In addition, we provide raw
+pixel level classification agreement data for quality control purposes
+(see Data availability below).} We refer to Figure 4 for a synoptic
+overview of the full deep learning workflow.
 
@@
 
-To map long term land-use and land-cover change in the Yangambi region
-we used the contemporary Global Forest Change version 1.6 (GFC, tile
-10N-020E) map data \citep{hansen2013}. Using the \textbf{GFC} data we
-calculated the latest state of the forest with respect to the conditions
-at the start of 1958, 60 years earlier. In our analysis we only included
-\textbf{GFC} pixels which recorded no \textbf{forest} loss throughout
-the whole 2000 - 2018 period. As the resolution of the historical forest
-classification exceeds that of the GFC map we downsampled our historical
-forest cover data to 30 m GFC resolution \textbf{using a nearest
-neighbour approach}. We masked out all water bodies using the Global
-Forest Change survey data mask layer, and limited the analysis to the
-right bank of the Congo river. We provide summary statistics of
-historical and contemporary forest loss and regrowth.
-
-deforestation (persistent forest loss after 1958), reforestation
-(regrowth after loss in 1958).
+To map long term LULCC in the Yangambi region we used the contemporary
+Global Forest Change version 1.6 (GFC, tile 10N-020E) map data
+\citep{hansen2013}. Using the \textbf{GFC} data we calculated the latest
+state of the forest with respect to the conditions at the start of 1958,
+60 years earlier. In our analysis we only included \textbf{GFC} pixels
+which recorded no \textbf{forest} loss throughout the whole 2000 - 2018
+period. As the resolution of the historical forest classification
+exceeds that of the GFC map we downsampled our historical forest cover
+data to 30 m GFC resolution \textbf{using a nearest neighbour approach}.
+We masked out all water bodies using the Global Forest Change survey
+data mask layer, and limited the analysis to the right bank of the Congo
+river. We provide summary statistics of historical and contemporary
+forest loss and regrowth. \textbf{We map persistent forest loss after
+1958, regrowth after loss in 1958, recent forest loss and long term
+(stable) forest cover.}
 
@@
 method and allometric relations used to scale the survey data. Unlike
-standard square 1 ha plots edge plots were set back 200m from forest
+standard square 1 ha plots, edge plots were set back 200 m from forest
 edges and were 50x200 m, with the 50 m side of the plot along the forest
@@
 age and the crop type. For example, the AGC values varied from 168.67 to
-86.55 g Mg C ha\textsuperscript{-1}, for \emph{Hevea brasiliensis}
-(rubber tree) and \emph{Elaeis guineensis} (oil palm) plots respectively
+86.55 Mg C ha\textsuperscript{-1}, for \emph{Hevea brasiliensis} (rubber
+tree) and \emph{Elaeis guineensis} (oil palm) plots respectively
 (\citet{bustillo2018}, personal communications). These higher values are
@@
 We compared the structure of the canopy both visually and using Fourier
-Transform Textural Ordination (FOTO, \citet{couteron2002}). Fourier
-Transform Textural Ordination uses a principal component analysis (PCA)
-on radially averaged 2D Fourier spectra to characterize canopy (image)
-texture. The FOTO technique was first described by \citet{couteron2002}
-to quantify canopy stucture in relation to biomass and biodiversity, and
-can be used across multiple scenes using normalization
-\citep{Barbier2010}.
+Transform Textural Ordination (FOTO, \citet{couteron2002}). FOTO uses a
+principal component analysis (PCA) on radially averaged 2D Fourier
+spectra to characterize canopy (image) texture. The FOTO technique was
+first described by \citet{couteron2002} to quantify canopy stucture in
+relation to biomass and biodiversity, and can be used across multiple
+scenes using normalization \citep{Barbier2010}.
 
@@
 
-For both site based and scene analysis we correlate PC values with
+For both site based and scene analysis we correlated PC values with
 permanent sample plot inventory data such as stem density, above ground
@@
 PC values. Due to the few available permanent sampling plots in both
-scenes we used a non-parametric paired signed rank (Wilcoxon) test to
-determine differences between the PC values of the Geo-Eye and
-historical orthomosaic image scenes across mono-dominant and mixed
-forest types. In all analysis, mono-dominant site 4 was removed from the
-analysis due to cloud contamination.
+scenes we used a non-parametric paired signed rank (Wilcoxon) test
+\citep{wilcoxon1945} to determine differences between the PC values of
+the Geo-Eye and historical orthomosaic image scenes across mono-dominant
+and mixed forest types. In all analysis, mono-dominant site 4 was
+removed from the analysis due to cloud contamination.
 
@@
 land-cover} within the Congo Basin. The use of high resolution
-historical images combined with structure from motion image processing
-techniques allowed us to mosaic old imagery across a large extent. The
-final orthomosaic composition of the Yangambi region \textbf{provided}
-an image scene covering approximately 733 million pixels across
+historical images combined with SfM image processing techniques allowed
+us to mosaic old imagery across a large extent. The final orthomosaic
+composition of the Yangambi region \textbf{provided} an image scene
+covering approximately 733 million pixels across
 \textbf{\textasciitilde{}82800 ha} (\textasciitilde{}23x36 km, Figure
@@
 \textbf{The CNN deep learning classifier reached an
-Intersection-over-Union (IoU) accuracy of 97\% on the detection of
-disturbed forest in the out-of-sample (validation) synthetic landscape
-data. Using all pixels within the validatoin polygons (Figure 5) showed
-a similar accuracy value of \textasciitilde{}98\%. Using all polygons
-across the scene, including those used in the generation of testing and
-training synthetic landscapes, increased the accuracy to
-\textasciitilde{}99\% (Table 2). A comparison with recent pan-chromatic
-Geo-Eye data shows good agreement, with an accuracy of
-\textasciitilde{}87\% across all pixels, between the landsat based GFC
-data and downscaled CNN results (Table 2 and Figure 7).}
+Intersection-over-Union accuracy of 97\% on the detection of disturbed
+forest in the out-of-sample (validation) synthetic landscape data. Using
+all pixels within the validatoin polygons (Figure 5) showed a similar
+accuracy value of \textasciitilde{}98\%. Using all polygons across the
+scene, including those used in the generation of testing and training
+synthetic landscapes, increased the accuracy to \textasciitilde{}99\%
+(Table 2). A comparison with recent pan-chromatic Geo-Eye data shows
+good agreement, with an accuracy of \textasciitilde{}87\% across all
+pixels, between the landsat based GFC data and downscaled CNN results
+(Table 2 and Figure 7).}
 
@@
 
-\hypertarget{long-term-lulc-changes-and-above-ground-carbon}{%
-\subsection{Long term LULC changes and Above Ground
-Carbon}\label{long-term-lulc-changes-and-above-ground-carbon}}
+\hypertarget{long-term-changes-in-lulc-and-above-ground-carbon}{%
+\subsection{Long term changes in LULC and Above Ground
+Carbon}\label{long-term-changes-in-lulc-and-above-ground-carbon}}
 
@@
 
-Recent deforestation follows a distincly different pattern compared to
+Recent deforestation follows a distinctly different pattern compared to
 historical patterns. Historical deforestation showed a classical
@@
 \textasciitilde{}921 \(\pm\) 362 m (Table 3). This shift in perimeter
-area ratio led to a similar change in the fractal index, increasing in
-value to 1.1 \(\pm\) 0.05 from 1.09 \(\pm\) 0.04 over time. Values
-closer to a fractal index of 2 suggest a more complex (fragmented)
-landscape.
+area ratio led to a similar change in the fractal index, slightly
+increasing in value to 1.1 \(\pm\) 0.05 from 1.09 \(\pm\) 0.04 over
+time. Values closer to a fractal index of 2 suggest a more complex
+(fragmented) landscape.
 
@@
 Changes in both land-use and land-cover led to concomitant changes in
-above ground carbon stocks. Recovery throughout the region was
-characterized for patches of forest and plantations. Assuming high
-density stands, based on previous work, this could amount to a carbon
-gains of 1592 Gg C across our study area, offsetting more recent losses
-of approximately 1408 Gg C. On the other hand, at the low end, if we
-assume a lower carbon density of 81.8 Mg C ha\textsuperscript{-1} this
-would result in a total carbon gain of 811 Gg C. Using our approach
-results indicate that overall deforestation around Yangambi has resulted
-in a loss of \textasciitilde{}2416 Gg C in AGC stocks.
+AGC stocks. Recovery throughout the region was characterized for patches
+of forest and plantations. Assuming high density stands, based on
+previous work, this could amount to a carbon gains of 1592 Gg C across
+our study area, offsetting more recent losses of approximately 1408 Gg
+C. On the other hand, at the low end, if we assume a lower carbon
+density of 81.8 Mg C ha\textsuperscript{-1} this would result in a total
+carbon gain of 811 Gg C. Using our approach results indicate that
+overall deforestation around Yangambi has resulted in a loss of
+\textasciitilde{}2416 Gg C in AGC stocks.
 
@@
 \cmidrule(l{3pt}r{3pt}){3-3}
-  & ha & (Gg C $ha\textsuperscript{-1}$\\
+  & ha & Gg C $ha\textsuperscript{-1}$\\
 \midrule
@@
 young-regrowth plots. One marked difference is noted in the
-mono-dominant site 6 (Appendix Table 1). Here, the current mono-dominant
-\emph{Brachystegia laurenttii} is a recent development, changing the
+mono-dominant plot 6 (Appendix Table 1). Here, the current mono-dominant
+\emph{Brachystegia laurentii} is a recent development, changing the
 canopy structure visibly during the last half century (\textbf{Figure
@@
 0.19 historically to its current value of 0.54. This historical value is
-similar to the mean of contemporary mono-dominant stands with PC
-averaging 0.34 \(\pm\) 0.1, and is only slightly higher than historical
-values for a mixed forest (0.18 \(\pm\) 0.08). The reverse pattern is
-seen in the contemporary PC values. Here, the value of 0.54 exceeds
-those of most mono-dominant stands (0.35 \(\pm\) 0.08), and is even
-further removed from the values noted for mixed forests (0.12 \(\pm\)
-0.03).
+similar to the mean of contemporary mono-dominant stands of
+\emph{Gilbertiodendron dewevrei} with PC averaging 0.34 \(\pm\) 0.1, and
+is only slightly higher than historical values for a mixed forest (0.18
+\(\pm\) 0.08). The reverse pattern is seen in the contemporary PC
+values. Here, the value of 0.54 exceeds those of most mono-dominant
+stands (0.35 \(\pm\) 0.08), and is even further removed from the values
+noted for mixed forests (0.12 \(\pm\) 0.03).
 
@@
 \textgreater{} 0.05). Any relationships between contemporary Geo-Eye
-data and permanent sampling plot measurements of Above Ground Biomass,
+data and permanent sampling plot measurements of Above Ground Carbon,
 stem density and species richness were non-significant (p \textgreater{}
@@
 
-\hypertarget{land-use-land-cover-change-classification-validation}{%
-\subsubsection{Land-use, Land-Cover change classification \&
-validation}\label{land-use-land-cover-change-classification-validation}}
+\hypertarget{lulc-classification-validation}{%
+\subsubsection{LULC classification \&
+validation}\label{lulc-classification-validation}}
 
@@
 characteristics (Figure 7), suggesting that the classifier consistently
-across works both space and time. We acknowledge that the use of
+works across both space and time. We acknowledge that the use of
 synthetic landscapes is limited by the available homogeneous areas to
@@
 
-\hypertarget{land-use-land-cover-change-above-ground-carbon}{%
-\subsection{Land-use, Land-Cover change \& Above Ground
-Carbon}\label{land-use-land-cover-change-above-ground-carbon}}
+\hypertarget{long-term-changes-in-lulc-and-above-ground-carbon-1}{%
+\subsubsection{Long term changes in LULC and Above Ground
+Carbon}\label{long-term-changes-in-lulc-and-above-ground-carbon-1}}
 
@@
 are located up to 1km away however most effects are pronounced within
-the first 300m from the edge \citep{gascon2000}. Our analysis of edge
-effects on AGC has shown that the influence is negligible 200m away from
-the edge. \citet{phillips2006} have shown similar weak responses to edge
-effects in the Amazon forest. Due to a lack of data on the extent
-(depth) of edge effects and their influence on AGC beyond 200m we did
+the first 300 m from the edge \citep{gascon2000}. Our analysis of edge
+effects on AGC has shown that the influence is negligible 200 m away
+from the edge. \citet{phillips2006} have shown similar weak responses to
+edge effects in the Amazon forest. Due to a lack of data on the extent
+(depth) of edge effects and their influence on AGC beyond 200 m we did
 not include any estimates of carbon loss or gain within these zones.
 However, it must be stated that the influence of edges throughout the
-landscape was not marginal. as these areas would account for
+landscape was not marginal as these areas would account for
 \textbf{13151 ha}. Thus edges could have a substantial negative
@@
 Similarly, uncertainties in how to explicitly correct for plantations in
-the landscape present a further challenge. Thus although our estimates
+the landscape present a further challenge. Thus, although our estimates
 are only indicative they do underscore the important influence of
@@
 \hypertarget{canopy-structure-foto-texture-analysis-2}{%
-\subsection{Canopy structure \& FOTO texture
+\subsubsection{Canopy structure \& FOTO texture
 analysis}\label{canopy-structure-foto-texture-analysis-2}}
@@
 that texture can be used to consistently capture canopy properties 60
-years apart. This suggests inflexibilit on part of the FOTO technique in
-dealing with non-standardized data.} We advise that future valorisation
-efforts should preferentially work from foto negatives (if available) to
-ensure optimal data quality in resolution, contrast and sharpness.
+years apart. This suggests inflexibility on part of the FOTO technique
+in dealing with non-standardized data.} We advise that future
+valorisation efforts should preferentially work from foto negatives (if
+available) to ensure optimal data quality in resolution, contrast and
+sharpness.
 
@@
 
-Given the impact of tropical forest disturbances on atmospheric
-CO\textsubscript{2} emissions, biodiversity and ecosystem productivity
-accurate long term reporting on land-cover and land-use change in the
-pre-satellite era is an imperative. Our analysis of historical aerial
-survey images of the Central Congo Basin provides a window into the
-state of the forest at the start of the anthropocene. Efforts to
-quantify canopy texture and their link to carbon storage had limited to
-no success. We have shown that historical aerial survey data can be used
-to quantify land-use and land-cover change using a highly automated
-workflow to quantify the structure and extent of forest cover change
-that can help assess the impact of fragmentation on above ground carbon
-stocks. We also show a shift from previously highly structured
-industrial deforestation of large areas \textbf{for plantation
-purposes}, to discrete smallholder clearing for farming, increasing
-landscape fragmentation but also opportuntiy for substantial regrowth.
-Efforts to quantify canopy texture and their link to carbon storage had
-limited to no success. As such, our analysis provides insights into the
-state of rarely studied tropical forests and the rate at which
-deforestation and reforestation has taken place over a multi-decadal
-scale in the central Congo basin providing a useful historical context
-to land-cover and land-use change spatial patterns for interpreting past
-and ongoing field research in the area.
+Given the impact of tropical forest disturbances on atmospheric carbon
+emissions, biodiversity and ecosystem productivity accurate long term
+reporting of LULCC is an imperative. Our analysis of historical aerial
+survey images (1958) of the Central Congo Basin provides a window into
+the state of the forest at the start of the anthropocene. The use of a
+CNN based LULC classifier, using synthetic landscapes based image
+augmentation, provides a robust semi-supervised solution which scales
+across space and time, even for image scenes with inconsistent
+illumination or sharpness. Combined with contemporary remote sensing
+data we have shown that historical aerial survey data can be used to
+quantify long-term changes in LULC and AGC. We showed a shift from
+previously highly structured industrial deforestation of large areas
+\textbf{for plantation purposes}, to discrete smallholder clearing for
+farming, increasing landscape fragmentation but also opportunties for
+substantial regrowth. Efforts to quantify canopy texture features and
+their link to AGC had limited to no success. Our analysis provides
+insights into the rate at which deforestation and reforestation has
+taken place over a multi-decadal scale in the central Congo basin. As
+such, it provides a useful historical context while interpreting past
+and ongoing forest research in the area.
 
@@
 Hufkens et al. (2019). A curated dataset of aerial survey images over
-the central Congo Basin, 1958. Zenodo: doi.org/10.5281/zenodo.3547767.
-All data not included in the latter repository can be found bundled with
-the analysis code as listed below. Proprietary datasets (i.e.~Geo-Eye
-data) are not shared, but purchase order numbers allow for acquisition
-of these datasets to ensure reproducibility.
+the central Congo Basin, 1958. Zenodo:
+\url{https://doi.org/10.5281/zenodo.3547767}. All data not included in
+the latter repository can be found bundled with the analysis code as
+listed below. Proprietary datasets (i.e.~Geo-Eye data) are not shared,
+but purchase order numbers allow for acquisition of these datasets to
+ensure reproducibility.
 
@@
 funding from the European Union Marie Skłodowska-Curie Action (project
-number 797668). I'm grateful for the support of Thales D'Haulleville
+number 797668). I'm grateful for the support of Thales de Haulleville
 volunteering his time in scanning the images.}
@@
 \authorcontributions{K.H. conceived and designed the study, analyzed the data, prepared
-figures and tables, authored, reviewed and approved the final draft of
-the manuscript. T.d.H. scanned all image data. E.K. and T.d.H. provided
-plot based AGC estimates. T.D., K.J., E.K, H.B., P.S., F.V.,S.M., M.A.,
-J.V.D.B., H.V and L.W. reviewed the final manuscript.}
+figures, tables and authored the final draft of the manuscript. T.d.H.
+scanned all image data. E.K. and T.d.H. provided plot based AGC
+estimates. T.D., K.J., E.K, H.B., P.S., F.V.,S.M., M.A., J.V.D.B., H.V
+and L.W. reviewed the final manuscript.}
 

### 2020-01-24 manuscript/main.Rmd
revision

@@
 
-Tropical ecosystem services are severely impacted by deforestation and forest degradation [@hansen2013; @houghton2012; @tyukavina2015].
+**Tropical ecosystem services are severely impacted by deforestation and forest degradation [@hansen2013; @houghton2012; @tyukavina2015].
 Not only does tropical forest Land-Use and Land-Cover Change (LULCC) constitute 10 to 15% of the total global carbon emissions [@vanderwerf2009], changes in forest fragmentation affect the forest structure and function [@fauset2017; @brinck2017; @didham1999].
 Strong fragmentation effects decreases the number of large trees along forest edges [@laurance2000; @magnago2015], while species composition and biodiversity are equally negatively affected [@poorter2006; @barlow2016; @vandeperre2018].
-Estimates show that 31% of carbon emissions are caused by edge effects alone [@brinck2017].
+Estimates show that 31% of carbon emissions are caused by edge effects alone [@brinck2017].**
 
-Accurate estimates of LULCC and forest canopy structure are therefore imperative to estimate carbon emissions and other ecosystem services [@houghton2012; @hansen2013].
+**Accurate estimates of LULCC and forest canopy structure are therefore imperative to estimate carbon emissions and other ecosystem services [@houghton2012; @hansen2013].
 Remote sensing products have been key inputs in LULCC assessments as they provide accurate spatial information to help estimate carbon emissions [@mitchard2018; @hansen2013].
 More so, high resolution aerial images provide scientists tools to monitor forest extent, structure and carbon emissions as canopy texture is linked to above ground biomass [@Ploton2012; @Couteron2005; @Barbier2010].
-Yet, most of these estimates are limited in time to recent decades [@defries2002; @houghton2012; @hansen2013; @achard2014].
+Yet, most of these estimates are limited in time to recent decades [@defries2002; @houghton2012; @hansen2013; @achard2014].**
 
-Historical estimates **of Land-Use and Land-Cover (LULC)**, in the pre-satellite era (< 1972), exist but generally rely on non-spatially explicit data (i.e socio-economic data) [@houghton2012; @ramankutty1999; @defries2002; @sader1988].
+**Historical estimates of Land-Use and Land-Cover (LULC), in the pre-satellite era (< 1972), exist but generally rely on non-spatially explicit data (i.e socio-economic data) [@houghton2012; @ramankutty1999; @defries2002; @sader1988].
 Efforts have been made to use other geo-spatial data sources such as historical maps [@willcock2016], declassified CORONA satellite surveillance data across the US and central Brazil [@song2015] as well as aerial surveys in post World War II German [@nita2018].
 Survey data across the African continent is less common, inaccessible or both.
-Some studies do exist, as @buitenwerf2012 and @hudak1998 used aerial survey images to map vegetation changes in South African savannas, whilst @frankl2015 and @nyssen2016 mapped the Ethiopian highlands of 1930's.
+Some studies do exist, as @buitenwerf2012 and @hudak1998 used aerial survey images to map vegetation changes in South African savannas, whilst @frankl2015 and @nyssen2016 mapped the Ethiopian highlands of 1930's.**
 
-Across the central Congo Basin most of these historical images were collected within the context of national cartographic efforts by the "Institut Géographique du Congo Belge" in Kinshasa (then Léopoldville), DR Congo.
+**Across the central Congo Basin most of these historical images were collected within the context of national cartographic efforts by the "Institut Géographique du Congo Belge" in Kinshasa (then Léopoldville), DR Congo.
 Despite the existence of large archives of aerial survey imagery (Figure 1, Appendix Figure 3), as of yet, no studies have valorized these data.
 The lack of a consistent valorization effort is unfortunate as the African rainforest is the second largest on Earth and covers ~630 million ha, representing up to 66 Pg of carbon storage [@Lewis2009b], and currently loses forest at an increasing pace [@butsic2015].
-Given the impact of LULCC on the structure and functioning of central African tropical forests, and their influence on both carbon dynamics [@Kearsley2013] and biodiversity [@vandeperre2018], accurate long term reporting of historical forest cover warrants more attention [@willcock2016].
+Given the impact of LULCC on the structure and functioning of central African tropical forests, and their influence on both carbon dynamics [@Kearsley2013] and biodiversity [@vandeperre2018], accurate long term reporting of historical forest cover warrants more attention [@willcock2016].**
 
-Here, we use a combination of historical aerial photography and contemporary remote sensing data to map long-term changes in the extent and structure of the tropical forest surrounding Yangambi (DR Congo), in the central Congo Basin, effectively linking the start of the antrophocene [@lewis2015] with current assessments.
+**Here, we use a combination of historical aerial photography and contemporary remote sensing data to map long-term changes in the extent and structure of the tropical forest surrounding Yangambi (DR Congo), in the central Congo Basin, effectively linking the start of the antrophocene [@lewis2015] with current assessments.
 Yangambi was, and remains, a focal center of forest and agricultural research and development in the central Congo Basin.
-Past research in the region allows for thorough assessment of land-use and land-cover change from a multi-disciplinary point of view, confronting us with complex deforestation and land-use patterns.
+Past research in the region allows for thorough assessment of land-use and land-cover change from a multi-disciplinary point of view, confronting us with complex deforestation and land-use patterns.**
 
-We leverage Structure-from-Motion to generate a large orthomosaic of historical imagery and develop a Convolutional Neural Network based forest cover mapping approach, based upon a semi-supervised generated dataset extensively leveraging data augmentation.
+**We leverage Structure-from-Motion to generate a large orthomosaic of historical imagery and develop a Convolutional Neural Network based forest cover mapping approach, based upon a semi-supervised generated dataset extensively leveraging data augmentation.
 Our methodology aims to provides a historical insight into important LULCC spatial patterns in Yangambi, such as fragmentation and edge complexity.
 We further contextualize the influence of changes in the forest's life history on past and current research into Above Ground Carbon (AGC) storage [@Kearsley2013] and biodiversity [@vandeperre2018] in the central Congo Basin.
-Our fast scalable mapping approach for historical aerial survey data, using limited supervised input, would further support long-term land-use and land-cover change analysis across the central Congo Basin.
+Our fast scalable mapping approach for historical aerial survey data, using limited supervised input, would further support long-term land-use and land-cover change analysis across the central Congo Basin.**
 
@@
 Tiles with mixed cover types were removed from the list of source tiles (Table 1).** Homogeneous **source** tiles were combined in synthetic landscapes using a random gaussian field based binary mask (Figure 3).
-We generated 5000 synthetic landscapes for training, while 500 landscapes were generated for both the validation and the testing datasets for a total of 6000 synthetic landscapes.
+We generated 5000 synthetic landscapes **(balancing forest and non-forest classes)** for training, while 500 landscapes were generated for both the validation and the testing datasets for a total of 6000 synthetic landscapes.
 In order to limit stitch line misclassifications, along the seams of mosaicked images, we created synthetic landscapes with different forest tiles to mimick forest texture transitions.
@@
 
-To quantify changes in the structure of forest cover and its disturbances we used spatial pattern analysis landscape fragmentation metrics [@dale1999].
-We report the ratio of edge to area and the fractal index to quantify landscape complexity of forest disturbances.
+To quantify changes in the structure of forest cover and its disturbances we used spatial pattern analysis landscape (fragmentation) metrics [@dale1999].
+**Landscape metrics provide a mathematical framework for the analysis of discrete land‐cover classes and allows us to capture their composition and configuration.
+These metrics are therefore commonly used to compare how landscapes change over time [@hesselbarth2019].
+In particular, fractals provide a way to quantify complex natural landscapes, including their self-similarity, across scales [@Li2000; @Mandelbrot1977]** We report the ratio of edge to area and the fractal dimension to quantify landscape complexity of forest disturbances.
+**A fractal dimension closer to 2 suggest a more complex (fragmented) landscape.**
+
 Statistics were calculated for all forest disturbance patches larger than 1 ha and smaller than the 95^th^ percentile of the patch size distribution using the R package landscapemetrics [@hesselbarth2019].
@@
 Land use has shifted away from for example a regular fishbone deforestation pattern that emerges when (large scale) agricultural interests dominate the landscape [@arima2015], to a more fragmented landscape (Figure 5D).
-The latter former?
-is consistent with historical land management at the time of the aerial survey [@bustillo2018].
+The former is consistent with historical land management at the time of the aerial survey [@bustillo2018].
 These regular patterns have since been reversed because of a decrease in large scale intensive agriculture, replaced by ad-hoc small scale subsistence farming with large perimeter to area relationships (i.e.
@@
 
-Given the impact of tropical forest disturbances on atmospheric carbon emissions, biodiversity and ecosystem productivity accurate long term reporting of LULCC is an imperative.
+**Given the impact of tropical forest disturbances on atmospheric carbon emissions, biodiversity and ecosystem productivity accurate long term reporting of LULCC is an imperative.
 Our analysis of historical aerial survey images (1958) of the Central Congo Basin provides a window into the state of the forest at the start of the anthropocene.
@@
 Combined with contemporary remote sensing data we have shown that historical aerial survey data can be used to quantify long-term changes in LULC and AGC.
-We showed a shift from previously highly structured industrial deforestation of large areas **for plantation purposes**, to discrete smallholder clearing for farming, increasing landscape fragmentation but also opportunties for substantial regrowth.
+We showed a shift from previously highly structured industrial deforestation of large areas for plantation purposes, to discrete smallholder clearing for farming, increasing landscape fragmentation but also opportunties for substantial regrowth.
 Efforts to quantify canopy texture features and their link to AGC had limited to no success.
 Our analysis provides insights into the rate at which deforestation and reforestation has taken place over a multi-decadal scale in the central Congo basin.
-As such, it provides a useful historical context while interpreting past and ongoing forest research in the area.
+As such, it provides a useful historical context while interpreting past and ongoing forest research in the area.**
 

### 2020-01-24 manuscript/main.tex
revision

@@
 
-Tropical ecosystem services are severely impacted by deforestation and
-forest degradation \citep{hansen2013, houghton2012, tyukavina2015}. Not
-only does tropical forest Land-Use and Land-Cover Change (LULCC)
-constitute 10 to 15\% of the total global carbon emissions
-\citep{vanderwerf2009}, changes in forest fragmentation affect the
-forest structure and function
+\textbf{Tropical ecosystem services are severely impacted by
+deforestation and forest degradation
+\citep{hansen2013, houghton2012, tyukavina2015}. Not only does tropical
+forest Land-Use and Land-Cover Change (LULCC) constitute 10 to 15\% of
+the total global carbon emissions \citep{vanderwerf2009}, changes in
+forest fragmentation affect the forest structure and function
 \citep{fauset2017, brinck2017, didham1999}. Strong fragmentation effects
@@
 31\% of carbon emissions are caused by edge effects alone
-\citep{brinck2017}.
-
-Accurate estimates of LULCC and forest canopy structure are therefore
-imperative to estimate carbon emissions and other ecosystem services
-\citep{houghton2012, hansen2013}. Remote sensing products have been key
-inputs in LULCC assessments as they provide accurate spatial information
-to help estimate carbon emissions \citep{mitchard2018, hansen2013}. More
-so, high resolution aerial images provide scientists tools to monitor
-forest extent, structure and carbon emissions as canopy texture is
-linked to above ground biomass
+\citep{brinck2017}.}
+
+\textbf{Accurate estimates of LULCC and forest canopy structure are
+therefore imperative to estimate carbon emissions and other ecosystem
+services \citep{houghton2012, hansen2013}. Remote sensing products have
+been key inputs in LULCC assessments as they provide accurate spatial
+information to help estimate carbon emissions
+\citep{mitchard2018, hansen2013}. More so, high resolution aerial images
+provide scientists tools to monitor forest extent, structure and carbon
+emissions as canopy texture is linked to above ground biomass
 \citep{Ploton2012, Couteron2005, Barbier2010}. Yet, most of these
 estimates are limited in time to recent decades
-\citep{defries2002, houghton2012, hansen2013, achard2014}.
-
-Historical estimates \textbf{of Land-Use and Land-Cover (LULC)}, in the
+\citep{defries2002, houghton2012, hansen2013, achard2014}.}
+
+\textbf{Historical estimates of Land-Use and Land-Cover (LULC), in the
 pre-satellite era (\textless{} 1972), exist but generally rely on
@@
 whilst \citet{frankl2015} and \citet{nyssen2016} mapped the Ethiopian
-highlands of 1930's.
-
-Across the central Congo Basin most of these historical images were
-collected within the context of national cartographic efforts by the
-``Institut Géographique du Congo Belge'' in Kinshasa (then
+highlands of 1930's.}
+
+\textbf{Across the central Congo Basin most of these historical images
+were collected within the context of national cartographic efforts by
+the ``Institut Géographique du Congo Belge'' in Kinshasa (then
 Léopoldville), DR Congo. Despite the existence of large archives of
@@
 \citep{vandeperre2018}, accurate long term reporting of historical
-forest cover warrants more attention \citep{willcock2016}.
-
-Here, we use a combination of historical aerial photography and
+forest cover warrants more attention \citep{willcock2016}.}
+
+\textbf{Here, we use a combination of historical aerial photography and
 contemporary remote sensing data to map long-term changes in the extent
@@
 multi-disciplinary point of view, confronting us with complex
-deforestation and land-use patterns.
-
-We leverage Structure-from-Motion to generate a large orthomosaic of
-historical imagery and develop a Convolutional Neural Network based
-forest cover mapping approach, based upon a semi-supervised generated
-dataset extensively leveraging data augmentation. Our methodology aims
-to provides a historical insight into important LULCC spatial patterns
-in Yangambi, such as fragmentation and edge complexity. We further
-contextualize the influence of changes in the forest's life history on
-past and current research into Above Ground Carbon (AGC) storage
-\citep{Kearsley2013} and biodiversity \citep{vandeperre2018} in the
-central Congo Basin. Our fast scalable mapping approach for historical
-aerial survey data, using limited supervised input, would further
-support long-term land-use and land-cover change analysis across the
-central Congo Basin.
+deforestation and land-use patterns.}
+
+\textbf{We leverage Structure-from-Motion to generate a large
+orthomosaic of historical imagery and develop a Convolutional Neural
+Network based forest cover mapping approach, based upon a
+semi-supervised generated dataset extensively leveraging data
+augmentation. Our methodology aims to provides a historical insight into
+important LULCC spatial patterns in Yangambi, such as fragmentation and
+edge complexity. We further contextualize the influence of changes in
+the forest's life history on past and current research into Above Ground
+Carbon (AGC) storage \citep{Kearsley2013} and biodiversity
+\citep{vandeperre2018} in the central Congo Basin. Our fast scalable
+mapping approach for historical aerial survey data, using limited
+supervised input, would further support long-term land-use and
+land-cover change analysis across the central Congo Basin.}
 
@@
 synthetic landscapes using a random gaussian field based binary mask
-(Figure 3). We generated 5000 synthetic landscapes for training, while
-500 landscapes were generated for both the validation and the testing
-datasets for a total of 6000 synthetic landscapes. In order to limit
-stitch line misclassifications, along the seams of mosaicked images, we
-created synthetic landscapes with different forest tiles to mimick
-forest texture transitions. We applied this technique to 10\% of the
-generated synthetic landscapes \textbf{(across training, validation and
-testing data)}.
+(Figure 3). We generated 5000 synthetic landscapes \textbf{(balancing
+forest and non-forest classes)} for training, while 500 landscapes were
+generated for both the validation and the testing datasets for a total
+of 6000 synthetic landscapes. In order to limit stitch line
+misclassifications, along the seams of mosaicked images, we created
+synthetic landscapes with different forest tiles to mimick forest
+texture transitions. We applied this technique to 10\% of the generated
+synthetic landscapes \textbf{(across training, validation and testing
+data)}.
 
@@
 To quantify changes in the structure of forest cover and its
-disturbances we used spatial pattern analysis landscape fragmentation
-metrics \citep{dale1999}. We report the ratio of edge to area and the
-fractal index to quantify landscape complexity of forest disturbances.
+disturbances we used spatial pattern analysis landscape (fragmentation)
+metrics \citep{dale1999}. \textbf{Landscape metrics provide a
+mathematical framework for the analysis of discrete land‐cover classes
+and allows us to capture their composition and configuration. These
+metrics are therefore commonly used to compare how landscapes change
+over time \citep{hesselbarth2019}. In particular, fractals provide a way
+to quantify complex natural landscapes, including their self-similarity,
+across scales \citep{Li2000, Mandelbrot1977}} We report the ratio of
+edge to area and the fractal dimension to quantify landscape complexity
+of forest disturbances. \textbf{A fractal dimension closer to 2 suggest
+a more complex (fragmented) landscape.}
+
 Statistics were calculated for all forest disturbance patches larger
@@
 agricultural interests dominate the landscape \citep{arima2015}, to a
-more fragmented landscape (Figure 5D). The latter former? is consistent
-with historical land management at the time of the aerial survey
+more fragmented landscape (Figure 5D). The former is consistent with
+historical land management at the time of the aerial survey
 \citep{bustillo2018}. These regular patterns have since been reversed
@@
 
-Given the impact of tropical forest disturbances on atmospheric carbon
-emissions, biodiversity and ecosystem productivity accurate long term
-reporting of LULCC is an imperative. Our analysis of historical aerial
-survey images (1958) of the Central Congo Basin provides a window into
-the state of the forest at the start of the anthropocene. The use of a
-CNN based LULC classifier, using synthetic landscapes based image
+\textbf{Given the impact of tropical forest disturbances on atmospheric
+carbon emissions, biodiversity and ecosystem productivity accurate long
+term reporting of LULCC is an imperative. Our analysis of historical
+aerial survey images (1958) of the Central Congo Basin provides a window
+into the state of the forest at the start of the anthropocene. The use
+of a CNN based LULC classifier, using synthetic landscapes based image
 augmentation, provides a robust semi-supervised solution which scales
@@
 quantify long-term changes in LULC and AGC. We showed a shift from
-previously highly structured industrial deforestation of large areas
-\textbf{for plantation purposes}, to discrete smallholder clearing for
-farming, increasing landscape fragmentation but also opportunties for
-substantial regrowth. Efforts to quantify canopy texture features and
-their link to AGC had limited to no success. Our analysis provides
-insights into the rate at which deforestation and reforestation has
-taken place over a multi-decadal scale in the central Congo basin. As
-such, it provides a useful historical context while interpreting past
-and ongoing forest research in the area.
+previously highly structured industrial deforestation of large areas for
+plantation purposes, to discrete smallholder clearing for farming,
+increasing landscape fragmentation but also opportunties for substantial
+regrowth. Efforts to quantify canopy texture features and their link to
+AGC had limited to no success. Our analysis provides insights into the
+rate at which deforestation and reforestation has taken place over a
+multi-decadal scale in the central Congo basin. As such, it provides a
+useful historical context while interpreting past and ongoing forest
+research in the area.}
 

### 2020-01-24 analysis/10_cnn_gfc_geo-eye_validation_map.R

@@ -85,6 +85,11 @@ spatial_map <- ggplot()+
                              "non-forest (CNN)",
                              "forest (GFC + CNN)")) +
   coord_fixed(ratio = 1) +
+  geom_sf(data = cloud_mask,
+          colour = "white",
+          fill = NA,
+          lty = 1,
+          size = 0.8) +
   geom_sf(data = box,
           colour = "white",
           fill = NA,

### 2020-01-30 manuscript/appendix.Rmd
minor revisions

@@
 
-```{r echo=FALSE, out.height='75%', fig.align="center",fig.cap="Shelves with aerial survey images at the Africa Museum archives."}
+```{r echo=FALSE, out.height='75%', fig.align="center",fig.cap="Shelves with aerial survey images at the Africa Museum archives, showing vasts amounts of historical data still mostly unexplored."}
 knitr::include_graphics('./figures/archives_aerial_photos_2.jpg')
-```
-
-```{r echo=FALSE, out.height='50%', fig.align="center",fig.cap="Boxplots comparing the first principal component (PC1) of a site based FOTO analysis across different forest types for both contemporary (Geo-Eye) and historical orthomosaic data."}
-knitr::include_graphics('./figures/foto_bplot_psp.png')
 ```

### 2020-01-30 manuscript/appendix.tex
minor revisions

@@
 
-\caption{Shelves with aerial survey images at the Africa Museum archives.}\label{fig:unnamed-chunk-5}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=14.58in,height=0.5\textheight]{./figures/foto_bplot_psp}
-
-}
-
-\caption{Boxplots comparing the first principal component (PC1) of a site based FOTO analysis across different forest types for both contemporary (Geo-Eye) and historical orthomosaic data.}\label{fig:unnamed-chunk-6}
+\caption{Shelves with aerial survey images at the Africa Museum archives, showing vasts amounts of historical data still mostly unexplored.}\label{fig:unnamed-chunk-5}
 \end{figure}
@@
 \caption{Scatterplot comparing Above Ground Carbon and the first principal component of the FOTO analysis for contemporary Geo-Eye (left) and the historical orthomosaic (right) data.
-Different forest types are plotted using closed circles, closed triangles, closed squares and this for fallow, mixed mon-dominant and young-regrowth forests, respectivelly }\label{fig:unnamed-chunk-7}
+Different forest types are plotted using closed circles, closed triangles, closed squares and this for fallow, mixed mon-dominant and young-regrowth forests, respectivelly }\label{fig:unnamed-chunk-6}
 \end{figure}
@@
 \caption{Scatterplot comparing tree species richness and the first principal component of the FOTO analysis for contemporary Geo-Eye (left) and the historical orthomosaic (right) data.
-Different forest types are plotted using closed circles, closed triangles, closed squares and this for fallow, mixed mon-dominant and young-regrowth forests, respectivelly}\label{fig:unnamed-chunk-8}
+Different forest types are plotted using closed circles, closed triangles, closed squares and this for fallow, mixed mon-dominant and young-regrowth forests, respectivelly}\label{fig:unnamed-chunk-7}
 \end{figure}
@@
 \caption{Scatterplot comparing stem density (per ha) and the first principal component of the FOTO analysis for contemporary Geo-Eye (left) and the historical orthomosaic (right) data.
-Different forest types are plotted using closed circles, closed triangles, closed squares and this for fallow, mixed mon-dominant and young-regrowth forests, respectivelly}\label{fig:unnamed-chunk-9}
+Different forest types are plotted using closed circles, closed triangles, closed squares and this for fallow, mixed mon-dominant and young-regrowth forests, respectivelly}\label{fig:unnamed-chunk-8}
 \end{figure}

### 2020-01-30 manuscript/main.Rmd
minor revisions

--- 2020-01-24 manuscript/main.Rmd
@@
   This research was supported through the Belgian Science Policy office COBECORE project (BELSPO; grant BR/175/A3/COBECORE).
-KH acknowledges funding from the European Union Marie Skłodowska-Curie Action (project number 797668).
-I'm grateful for the support of Thales de Haulleville volunteering his time in scanning the images.
+K.H.
+acknowledges funding from the European Union Marie Skłodowska-Curie Action (project number 797668).
+K.H.
+grateful for the support of T.d.H.
+volunteering his time in scanning the images.
 authorcontributions: |
@@
 
-  row.names(change_stats) <- c(
-   "Forest",
-   "  -  of which edges",
-   "Regrowth > 1958 loss",
-   "Loss > 2000",
-   "Loss > 1958 (permanent)")
-
   # drop square km values per reviewer request
   change_stats <- change_stats[,2:3]
+
+  change_stats <- rbind(c(sum(change_stats$ha[c(3,5)]), NA),
+                        change_stats)
+
+  row.names(change_stats) <- c(
+   "Non-forest cover (1958)",
+   "Forest cover",
+   "  -  of which forest edges",
+   "reforestation >1958",
+   "deforestation >2000",
+   "deforestation >1958 (permanent)")
 
@@
 Not only does tropical forest Land-Use and Land-Cover Change (LULCC) constitute 10 to 15% of the total global carbon emissions [@vanderwerf2009], changes in forest fragmentation affect the forest structure and function [@fauset2017; @brinck2017; @didham1999].
-Strong fragmentation effects decreases the number of large trees along forest edges [@laurance2000; @magnago2015], while species composition and biodiversity are equally negatively affected [@poorter2006; @barlow2016; @vandeperre2018].
+Strong fragmentation effects **decrease** the number of large trees along forest edges [@laurance2000; @magnago2015], while species composition and biodiversity are equally negatively affected [@poorter2006; @barlow2016; @vandeperre2018].
 Estimates show that 31% of carbon emissions are caused by edge effects alone [@brinck2017].
@@
 
-Historical estimates **of Land-Use and Land-Cover (LULC)**, in the pre-satellite era (< 1972), exist but generally rely on non-spatially explicit data (i.e socio-economic data) [@houghton2012; @ramankutty1999; @defries2002; @sader1988].
-Efforts have been made to use other geo-spatial data sources such as historical maps [@willcock2016], declassified CORONA satellite surveillance data across the US and central Brazil [@song2015] as well as aerial surveys in post World War II German [@nita2018].
+Historical estimates of Land-Use and Land-Cover (LULC), in the pre-satellite era (< 1972), exist but generally rely on non-spatially explicit data (i.e socio-economic data) [@houghton2012; @ramankutty1999; @defries2002; @sader1988].
+Efforts have been made to use other geo-spatial data sources such as historical maps [@willcock2016], declassified CORONA satellite surveillance data across the US and central Brazil [@song2015] as well as aerial surveys in post World War II Germany [@nita2018].
 Survey data across the African continent is less common, inaccessible or both.
-Some studies do exist, as @buitenwerf2012 and @hudak1998 used aerial survey images to map vegetation changes in South African savannas, whilst @frankl2015 and @nyssen2016 mapped the Ethiopian highlands of 1930's.
+Some studies do exist, as @buitenwerf2012 and @hudak1998 used aerial survey images to map vegetation changes in South African savannas, whilst @frankl2015 and @nyssen2016 mapped the Ethiopian highlands of **1930s**.
 
@@
 
-Here, we use a combination of historical aerial photography and contemporary remote sensing data to map long-term changes in the extent and structure of the tropical forest surrounding Yangambi (DR Congo), in the central Congo Basin, effectively linking the start of the antrophocene [@lewis2015] with current assessments.
+Here, we use a combination of historical aerial photography **(1958)** and contemporary remote sensing data **(2000-2018)** to map long-term changes in the extent and structure of the tropical forest surrounding Yangambi (DR Congo), in the central Congo Basin, effectively linking the start of the **anthrophocene** [@lewis2015] with current assessments.
 Yangambi was, and remains, a focal center of forest and agricultural research and development in the central Congo Basin.
@@
 We leverage Structure-from-Motion to generate a large orthomosaic of historical imagery and develop a Convolutional Neural Network based forest cover mapping approach, based upon a semi-supervised generated dataset extensively leveraging data augmentation.
-Our methodology aims to provides a historical insight into important LULCC spatial patterns in Yangambi, such as fragmentation and edge complexity.
+Our methodology aims to **provide** a historical insight into important LULCC spatial patterns in Yangambi, such as fragmentation and edge complexity.
 We further contextualize the influence of changes in the forest's life history on past and current research into Above Ground Carbon (AGC) storage [@Kearsley2013] and biodiversity [@vandeperre2018] in the central Congo Basin.
@@
 Data for the central Congo Basin region, surrounding Kisangani, were collected in several flights during the dry season of 1958 and 1959 (from 8/01/1958 to 20/02/1958 and from 28/12/1958 to 9/01/1959 respectively, see Appendix Figure 1) to generate topographic maps of the area, supervised by the "Institut Géographique du Congo Belge" in Kinshasa, DR Congo (then Léopoldville).
-Black-and-White infrared images (0.4 - 0.9 $\mu$m) were gathered along flight paths running mostly from west to east, between **09:00 and 11:00** local time.
+Black-and-White **panchromatic** images (0.4 - 0.9 $\mu$m) were gathered along flight paths running mostly from west to east, between 09:00 and 11:00 local time.
 Along a flight path continuous images were taken using a Wild Heerburgg RC5a (currently Leica Geosystems) with an Aviogon lens assembly (114.83mm / f 5.6, with a 90\degree view angle) resulting in square photo negatives of 180 by 180 mm.
@@
 Ground control points consisted of permanent structures which could be verified in both old and new aerial imagery (i.e.
-ESRI World Imagery) **and consisted of corner points of build structures (e.g.
-a building, bridge or swimming pool etc.).** Although clouds were removed during the SfM routine we did not mask all clouds in the final orthomosaic to maximize forest coverage.
-The final scene was cropped, to provide consistent wall to wall coverage of the reconstructed scene.
+ESRI World Imagery) and consisted of corner points of build structures (e.g.
+a building, bridge or swimming pool etc.).
+**Although most clouds were removed during the SfM routine some were retained to provide sufficient SfM tie points to maximize continuous forest coverage in the final orthomosaic.** The final scene was cropped, to provide consistent wall to wall coverage of the reconstructed scene.
 The orthomosaic was exported as a geotiff for further georeferencing in QGIS [@qgisdevelopmentteam2018] using the georeferencer plugin (version 3.1.9) and additional ESRI World Imagery high resolution reference data.
-We used 3^th^ degree polynomial and 16 ground control points to correct the final image.
+We used **3^rd^** degree polynomial and 16 ground control points to correct the final image.
 Ground control points, raw image data and final processed image are provided in addition to measures of uncertainty such as mean and median error across all ground control points.
@@
 
-We automatically delineated all natural forest in the historical data, thus excluding tree plantations, thinned or deteriorated forest stands which **showed visible canopy cover loss**, fields and buildings.
+We automatically delineated all natural forest in the historical data, thus excluding tree plantations, thinned or deteriorated forest stands which showed visible canopy cover loss, fields and buildings.
 We used the Unet Convolutional Neural Net (CNN, @ronneberger2015) architecture implemented in Keras [@chollet2015] with an efficientnetb3 pre-processing backbone [@yakubovskiy2019] running on TensorFlow [@martinabadi2015] to train a binary classifier (i.e.
 forest or non-forested).
-Training data were collected from the orthomosaic by randomly selecting 513 pixel square tiles from **locations within** homogeneous forested or non-forested polygons in the historical orthomosaic **(Figure 5).
+Training data were collected from the orthomosaic by randomly selecting 513 pixel square tiles from locations within homogeneous forested or non-forested polygons in the historical orthomosaic (Figure 5).
 Separate polygons were selected for training, testing and validation purposes.
@@
 Tiles extracted from locations close to the polygon border at times contained mixed cover types.
-Tiles with mixed cover types were removed from the list of source tiles (Table 1).** Homogeneous **source** tiles were combined in synthetic landscapes using a random gaussian field based binary mask (Figure 3).
-We generated 5000 synthetic landscapes for training, while 500 landscapes were generated for both the validation and the testing datasets for a total of 6000 synthetic landscapes.
+Tiles with mixed cover types were removed from the list of source tiles (Table 1).
+Homogeneous source tiles were combined in synthetic landscapes using a random gaussian field based binary mask (Figure 3).
+We generated 5000 synthetic landscapes (balancing forest and non-forest classes) for training, while 500 landscapes were generated for both the validation and the testing datasets for a total of 6000 synthetic landscapes.
 In order to limit stitch line misclassifications, along the seams of mosaicked images, we created synthetic landscapes with different forest tiles to mimick forest texture transitions.
-We applied this technique to 10% of the generated synthetic landscapes **(across training, validation and testing data)**.
+We applied this technique to 10% of the generated synthetic landscapes (across training, validation and testing data).
 
@@
 Data augmentation included random cropping to 320 pixel squares, random orientation, scaling, perspective, contrast and brightness shifts and image blurring.
-The optimized model was used to classify the complete orthomosaic using a moving window approach with a step size of 110 pixels and a majority vote **(> 50% agreement)** across overlapping areas to limit segmentation edge effects.
-**In addition, we provide raw pixel level classification agreement data for quality control purposes (see Data availability below).** We refer to Figure 4 for a synoptic overview of the full deep learning workflow.
+The optimized model was used to classify the complete orthomosaic using a moving window approach with a step size of 110 pixels and a majority vote (> 50% agreement) across overlapping areas to limit segmentation edge effects.
+In addition, we provide raw pixel level classification agreement data for quality control purposes (see Data availability below).
+We refer to Figure 4 for a synoptic overview of the full deep learning workflow.
 
@@
 
-**We report the CNN accuracy based upon the IoU of our out-of-sample validation dataset of synthetic landscapes.
+We report the CNN accuracy based upon the IoU of our out-of-sample validation dataset of synthetic landscapes.
 In addition, we report confusion matrices for all pixels across the homogeneous validation polygons, as well as the training and testing polygons (see Figure 5).
 Furthermore, we used the first acquisition of a recent pan-chromatic Geo-Eye 1 stereo pair (Geo-Eye, Thornton, Colorado, U.S.A., order 737537, 2011-11-11 08:55 GMT or 09:55 local time) to classify and assess the robustness of the CNN algorithm on contemporary remote sensing imagery with similar spectral and spatial characteristics.
-We used the Global Forest Change version 1.6 (GFC, tile 10N-020E) [@hansen2013] map data, capturing forest loss up to 2011, to quantify accuracy on downsampled CNN Geo-Eye classification results.
+We used the Global Forest Change version 1.6 (GFC, tile 10N-020E) [@hansen2013] map data, capturing **deforestation** up to 2011, to quantify accuracy on downsampled CNN Geo-Eye classification results.
 Once more, we report the confusion matrix of between the GFC and CNN derived forest cover maps, masking clouds and cloud shadows.
-To summarize confusion matrices we report accuracy as:**
+To summarize confusion matrices we report accuracy as:
 
@@
 
-**in which TP, TN, FP, FN are True Positive, True Negative, False Positive and False Negative, respectivelly.**
+in which TP, TN, FP, FN are True Positive, True Negative, False Positive and False Negative, respectivelly.
 
@@
 To map long term LULCC in the Yangambi region we used the contemporary Global Forest Change version 1.6 (GFC, tile 10N-020E) map data [@hansen2013].
-Using the **GFC** data we calculated the latest state of the forest with respect to the conditions at the start of 1958, 60 years earlier.
-In our analysis we only included **GFC** pixels which recorded no **forest** loss throughout the whole 2000 - 2018 period.
-As the resolution of the historical forest classification exceeds that of the GFC map we downsampled our historical forest cover data to 30 m GFC resolution **using a nearest neighbour approach**.
+Using the GFC data we calculated the latest state of the forest with respect to the conditions at the start of 1958, 60 years earlier.
+In our analysis we only included GFC pixels which recorded no forest loss throughout the whole 2000 - 2018 period, **as such locations which would see reforestation or deforestation between 2000 and 2018 would be marked as non-forest (i.e.
+disturbed)**.
+As the resolution of the historical forest classification exceeds that of the GFC map we downsampled our historical forest cover data to 30 m GFC resolution using a nearest neighbour approach.
 We masked out all water bodies using the Global Forest Change survey data mask layer, and limited the analysis to the right bank of the Congo river.
-We provide summary statistics of historical and contemporary forest loss and regrowth.
-**We map persistent forest loss after 1958, regrowth after loss in 1958, recent forest loss and long term (stable) forest cover.**
+We provide summary statistics of historical and contemporary **deforestation and reforestation**.
+We map permanent forest loss after 1958, **reforestation** after 1958, **recent deforestation** and long term (stable) forest cover.
+**All references to changes over time in the context of our analysis explicitly compare the historical and contemporary periods from hereon forth.**
 
@@
 
-To quantify changes in the structure of forest cover and its disturbances we used spatial pattern analysis landscape fragmentation metrics [@dale1999].
-We report the ratio of edge to area and the fractal index to quantify landscape complexity of forest disturbances.
+To quantify changes in the structure of forest cover and its disturbances we used spatial pattern analysis landscape (fragmentation) metrics [@dale1999].
+Landscape metrics provide a mathematical framework for the analysis of discrete land‐cover classes and allows us to capture their composition and configuration.
+These metrics are therefore commonly used to compare how landscapes change over time [@hesselbarth2019].
+In particular, fractals provide a way to quantify complex natural landscapes, including their self-similarity, across scales [@Li2000; @Mandelbrot1977] We report the ratio of edge to area and the fractal dimension to quantify landscape complexity of forest disturbances.
+A fractal dimension closer to 2 suggest a more complex (fragmented) landscape.
+
 Statistics were calculated for all forest disturbance patches larger than 1 ha and smaller than the 95^th^ percentile of the patch size distribution using the R package landscapemetrics [@hesselbarth2019].
@@
 
-We estimated above ground carbon (AGC) losses and gains over time using plot based averages of recent inventory data at permanent sampling plots in the area (Figure 2).
+We estimated above ground carbon (AGC) losses and gains, **due to deforestation and reforestation**, using plot based averages of recent inventory data at permanent sampling plots in the area (Figure 2).
 We refer to @Kearsley2013 for the survey method and allometric relations used to scale the survey data.
@@
 For example, the AGC values varied from 168.67 to 86.55 Mg C ha^-1^, for *Hevea brasiliensis* (rubber tree) and *Elaeis guineensis* (oil palm) plots respectively (@bustillo2018, personal communications).
-These higher values are in line with the mixed AGC estimates in the area, while the palm plantations resemble old-regrowth values (81.8 Mg C ha^-1^, see **Appendix Table 1**).
+These higher values are in line with the mixed AGC estimates in the area, while the palm plantations resemble old-regrowth values (81.8 Mg C ha^-1^, see Appendix Table 1).
 We therefore use both the estimates of old-regrowth and mixed forest to estimate AGC for regrowth.
@@
 
-Our analysis provides a first spatially explicity historical composite of aerial survey images **in support of mapping land-use and land-cover** within the Congo Basin.
+Our analysis provides a first spatially explicity historical composite of aerial survey images in support of mapping land-use and land-cover within the Congo Basin.
 The use of high resolution historical images combined with SfM image processing techniques allowed us to mosaic old imagery across a large extent.
-The final orthomosaic composition of the Yangambi region **provided** an image scene covering approximately 733 million pixels across **~82800 ha** (~23x36 km, Figure 2).
+The final orthomosaic composition of the Yangambi region provided an image scene covering approximately 733 million pixels across ~82800 ha (~23x36 km, Figure 2).
 The overall accuracy of the SfM orthomosaic composition was 0.88 m/pixel using the sparse cloud DEM for corrections at 45.8 m/pixel.
-The resulting georeferenced scene had a spatial accuracy of approximately **23 m**.
+The resulting georeferenced scene had a spatial accuracy of approximately 23 m.
 Further georeferencing outside the SfM workflow reduced the mean error at the ground control points to 5.3 $\pm$ 4.9 px (~4.7 $\pm$ 4.3 m), with a median error of 2.9 px (2.6 m).
@@
 
-**The CNN deep learning classifier reached an Intersection-over-Union accuracy of 97% on the detection of disturbed forest in the out-of-sample (validation) synthetic landscape data.
-Using all pixels within the validatoin polygons (Figure 5) showed a similar accuracy value of ~98%.
+The CNN deep learning classifier reached an Intersection-over-Union accuracy of 97% on the detection of disturbed forest in the out-of-sample (validation) synthetic landscape data.
+Using all pixels within the **validation** polygons (Figure 5) showed a similar accuracy value of ~98%.
 Using all polygons across the scene, including those used in the generation of testing and training synthetic landscapes, increased the accuracy to ~99% (Table 2).
-A comparison with recent pan-chromatic Geo-Eye data shows good agreement, with an accuracy of ~87% across all pixels, between the landsat based GFC data and downscaled CNN results (Table 2 and Figure 7).**
+A comparison with recent pan-chromatic Geo-Eye data shows good agreement, with an accuracy of ~87% across all pixels, between the landsat based GFC data and downscaled CNN results (Table 2 and Figure 7).
 
@@
    kable(
-    caption="Confusion matrix showing \\% agreement between forest / non-forest classes using a Convolutional Neural Network (CNN) across previously selected homogenous areas.",
+    caption="Confusion matrix showing \\% agreement between forest / non-forest classes using a Convolutional Neural Network (CNN) across previously selected homogenous areas.
+In addition, overall accuracy is reported for each confusion matrix.",
     booktabs = TRUE,
@@
 
-Scaling our classifier to the whole historical orthomosaic we detected **~16200 ha** (or ~20% of the scene) of disturbed forests.
-A large fraction of the disturbed area was restored in the period between the two **acquisition periods**.
-In total, **9918 ha**, or little over half of the affected forest was restored (Figure 5C-D, dark blue).
-Recent deforested areas, as registered through satellite remote sensing, approximate **8776 ha (Table 3, Figure 6 - light green)**.
+Scaling our classifier to the whole historical orthomosaic we detected ~16200 ha (or ~20% of the scene) of disturbed forests.
+A large fraction of the disturbed area was restored in the period between the two acquisition periods.
+In total, 9918 ha, or little over half of the affected forest was restored (Figure 5C-D, dark blue).
+Recent deforested areas, as registered through satellite remote sensing **(> 2000)**, approximate 8776 ha (Table 3, Figure 6 - light green).
 
 Recent deforestation follows a distinctly different pattern compared to historical patterns.
-Historical deforestation showed a classical fishbone pattern for forest clearing with very sharp edges, while current patterns are patchy and ad-hoc (**figure 6C**, dark blue and green colours respectively).
-These differences are reflected in the analysis of landscape metrics of forest loss.
+Historical deforestation showed a classical fishbone pattern for forest clearing with very sharp edges, while current patterns are patchy and ad-hoc (figure 6C, dark blue and green colours respectively).
+These differences are reflected in the analysis of landscape metrics of **deforestation**.
 Between the historical and contemporary LULCC maps we see an increase in small disturbances, as indicated by the decreasing area of the mean patch size, down to ~1.86 $\pm$ 0.75 ha from ~5.25 $\pm$ 5.02 ha historically.
-Perimeter lengths were longer historically, at 1451 $\pm$ 943 m,  compared to contemporary landscapes ~921 $\pm$ 362 m (Table 3).
+Perimeter lengths were longer historically, at 1451 $\pm$ 943 m,  compared to contemporary landscapes ~921 $\pm$ 362 m (Table **4**).
 This shift in perimeter area ratio led to a similar change in the fractal index, slightly increasing in value to 1.1 $\pm$ 0.05 from 1.09 $\pm$ 0.04 over time.
@@
 A comparison of forest edge plots with mixed forest plots showed no signficant difference in AGC, or other reported values such as species richness, basal area or stem density (Mann Whitney U test, p < 0.05).
-Edge influence did not extend beyond 200 m from a forest edge, but still represented an area of **13151 ha**.
+Edge influence did not extend beyond 200 m from a forest edge, but still represented an area of 13151 ha (Table 3).
 
@@
     add_header_above(c(" " = 2,"AGC" = 1)) %>%
+    #pack_rows("",6,6) %>%
     kable_styling(latex_options = c("hold_position"))
@@
 One marked difference is noted in the mono-dominant plot 6 (Appendix Table 1).
-Here, the current mono-dominant *Brachystegia laurentii* is a recent development, changing the canopy structure visibly during the last half century (**Figure 8**).
+Here, the current mono-dominant *Brachystegia laurentii* is a recent development, changing the canopy structure visibly during the last half century (Figure 8).
 The previous varied canopy structure gave way to a more dense and uniform canopy.
-This is reflected in a change of the FOTO PC value from 0.19 historically to its current value of 0.54.
-This historical value is similar to the mean of contemporary mono-dominant stands of *Gilbertiodendron dewevrei* with PC averaging 0.34 $\pm$ 0.1, and is only slightly higher than historical values for a mixed forest (0.18 $\pm$ 0.08).
+This is reflected in a change of the FOTO PC value from 0.19 historically to its current value of 0.54 **(Figure 9).** This historical value is similar to the mean of contemporary mono-dominant stands of *Gilbertiodendron dewevrei* with PC averaging 0.34 $\pm$ 0.1, and is only slightly higher than historical values for a mixed forest (0.18 $\pm$ 0.08, **Figure 9)**.
 The reverse pattern is seen in the contemporary PC values.
-Here, the value of 0.54 exceeds those of most mono-dominant stands (0.35 $\pm$ 0.08), and is even further removed from the values noted for mixed forests (0.12 $\pm$ 0.03).
+Here, the value of 0.54 exceeds those of most mono-dominant stands (0.35 $\pm$ 0.08), and is even further removed from the values noted for mixed forests (0.12 $\pm$ 0.03, **Figure 9**).
 
@@
 
-Using only small subsets around existing permanent sampling plots we show distinct differences between forest types, with PC values in both historical and contemporary imagery markedly higher for the mono-dominant forest types compared to all others (**Appendix Figure 4**).
+```{r echo=FALSE, out.width='50%', fig.align="center",fig.cap="Boxplots comparing the first principal component (PC1) of a site based FOTO analysis across different forest types for both contemporary (Geo-Eye) and historical orthomosaic data."}
+knitr::include_graphics('./figures/foto_bplot_psp.png')
+```
+
+Using only small subsets around existing permanent sampling plots we show distinct differences between forest types, with PC values in both historical and contemporary imagery markedly higher for the mono-dominant forest types compared to all others (Appendix Figure 4).
 Provided that the young-regrowth and fallow permanent sampling plots have seen recent disturbance the Wilcoxon signed rank test on the mixed and mono-dominant plots between the historical and contemporary PC values did not show a significant difference (p > 0.05).
 Similarly, no significant different using PC values extracted from the whole scene analysis was noted (p > 0.05).
-Any relationships between contemporary Geo-Eye data and permanent sampling plot measurements of Above Ground Carbon, stem density and species richness were non-significant (p > 0.05, Appendix **Figures 5-7**).
+Any relationships between contemporary Geo-Eye data and permanent sampling plot measurements of Above Ground Carbon, stem density and species richness were non-significant (p > 0.05, Appendix **Figures 4-6**).
 
-Furthermore, visual inspection of the scene wide analysis suggests historical scences do not show landscape wide canopy features (**Figure 9 A-B**), unlike the contemporary scene (**Figure 9 C-D**).
+Furthermore, visual inspection of the scene wide analysis suggests historical scences do not show landscape wide canopy features (Figure 10 A-B), unlike the contemporary scene (Figure 10 C-D).
 Where the FOTO algorithm picks up landscape features such as changes in texture across the contemporary Geo-Eye scene (e.g.
-the river valley as a diagonal line in **Figure 9D**), however, no corresponding landscape patterns are found by the FOTO algorithm in the historical orthomosaic.
+the river valley as a diagonal line in Figure 10D), however, no corresponding landscape patterns are found by the FOTO algorithm in the historical orthomosaic.
 
@@
 In our study the archive data recovered was limited to contact prints and therefore did not represent the true resolution of the original negative.
-In addition, analogue photography clearly produces a distinct softness compared to digital imagery (**Figure 8**).
+In addition, analogue photography clearly produces a distinct softness compared to digital imagery (Figure 8).
 Despite favourable nadir image acquistions [@verhoeven2017] image softness combined with illumination effects between flight paths, and the self-similar nature of vast canopy expanses [@park2019; @simini2010; @sole1995], limited our ability to provide wall-to-wall coverage of the entire dataset containing 334 images.
@@
 
-**When classifying the orthomosaic into forest and non-forest states we favoured a deep learning supervised classification using a CNN over manual segmentation to guarantee an "apples-to-apples" comparison between the historical and the contemporary GFC forest cover map used in our analysis.
+When classifying the orthomosaic into forest and non-forest states we favoured a deep learning supervised classification using a CNN over manual segmentation to guarantee an "apples-to-apples" comparison between the historical and the contemporary GFC forest cover map used in our analysis.
 We acknowledge that both the CNN and GFC land-use and land-cover maps use different underlying features, i.e.
@@
 Visual inspection of the classification data in Figure 7 suggests that the GFC map more often than not classifies non-forest areas as forest.
-Actual classification accuracy of our algorithm might therefore be higher than our reported value.**
+Actual classification accuracy of our algorithm might therefore be higher than our reported value.
 
@@
 
-**Our approach uses broadly defined homogeneous polygons to construct a balanced dataset of synthethic landscapes.
+Our approach uses broadly defined homogeneous polygons to construct a balanced dataset of synthethic landscapes.
 The methodology is analoguous to the use of sparse labelling as used by @buscombe2018 and contrasts with the standard methodologies which generally extract pixel (windows) [@song2015] or delineate land cover classes [@nita2018] to drive a classifier or analysis.
@@
 We acknowledge that the use of synthetic landscapes is limited by the available homogeneous areas to sample from and the number of classes.
-Yet, the latter should not be a constrained as previous research efforts have focussed on simple forest loss maps [@hansen2013].**
+Yet, the latter should not be a constrained as previous research efforts have focussed on simple forest cover maps [@hansen2013].
 
@@
 
-Our analysis shows that the majority of deforestation around Yangambi happened toward the late 1950's (**~16200 ha**).
-Considerable regrowth has occured since the aerial survey was executed (**~9918 ha**), and socio-economic instability prevented further large scale forest exploitation.
+Our analysis shows that the majority of deforestation around Yangambi happened toward the late 1950's (~16200 ha).
+Considerable **reforestation** has occured since the aerial survey was executed (~9918 ha), and socio-economic instability prevented further large scale forest exploitation.
 In particular, many plantations have reached maturity and forest has re-established in previously cleared or disturbed areas.
-The majority of this forest recovery takes the form of isolated patches of forest but is offset by further deforestation of previously untouched forest.
+The majority of this **reforestation** takes the form of isolated patches of forest but is offset by further deforestation of previously untouched forest.
 Generally, the function and structure of forests can be influenced by forest edges that are located up to 1km away however most effects are pronounced within the first 300 m from the edge [@gascon2000].
@@
 Due to a lack of data on the extent (depth) of edge effects and their influence on AGC beyond 200 m we did not include any estimates of carbon loss or gain within these zones.
-However, it must be stated that the influence of edges throughout the landscape was not marginal as these areas would account for **13151 ha**.
+However, it must be stated that the influence of edges throughout the landscape was not marginal as these areas would account for 13151 ha.
 Thus edges could have a substantial negative [@brinck2017] or positive [@reinmann2017] influence on AGC.
@@
 INERA concessions, @bustillo2018) on a multi-decadal time scale.
-Forest regrowth in non-continuous areas within Yangambi could increase landscape connectivity and help increase biodiversity [@vandeperre2018].
+**Reforestation** in non-continuous areas within Yangambi could increase landscape connectivity and help increase biodiversity [@vandeperre2018].
 
@@
 Unlike large scale studies by @Ploton2012 we could not scale this technique to historical data.
-**The successful use of our CNN classification model on a contemporary remote sensing data does suggest that texture can be used to consistently capture canopy properties 60 years apart.
-This suggests inflexibility on part of the FOTO technique in dealing with non-standardized data.** We advise that future valorisation efforts should preferentially work from foto negatives (if available) to ensure optimal data quality in resolution, contrast and sharpness.
+The successful use of our CNN classification model on a contemporary remote sensing data does suggest that texture can be used to consistently capture canopy properties 60 years apart.
+**Differences in PC between forest types (e.g.
+mono-dominant vs.
+mixed, Figure 9) corroborate that texture can serve as a basis for LULC classification.
+However, inflexibility on part of the FOTO technique in dealing with non-standardized (historical) data, or scaling these results to AGC values, limits its use case.** We advise that future valorisation efforts should preferentially work from foto negatives (if available) to ensure optimal data quality in resolution, contrast and sharpness.
 
@@
 Combined with contemporary remote sensing data we have shown that historical aerial survey data can be used to quantify long-term changes in LULC and AGC.
-We showed a shift from previously highly structured industrial deforestation of large areas **for plantation purposes**, to discrete smallholder clearing for farming, increasing landscape fragmentation but also opportunties for substantial regrowth.
+We showed a shift from previously highly structured industrial deforestation of large areas for plantation purposes, to discrete smallholder clearing for farming, increasing landscape fragmentation but also opportunties for substantial regrowth.
 Efforts to quantify canopy texture features and their link to AGC had limited to no success.

### 2020-01-30 manuscript/main.tex
minor revisions

--- 2020-01-24 manuscript/main.tex
@@
 \citep{fauset2017, brinck2017, didham1999}. Strong fragmentation effects
-decreases the number of large trees along forest edges
+\textbf{decrease} the number of large trees along forest edges
 \citep{laurance2000, magnago2015}, while species composition and
@@
 
-Historical estimates \textbf{of Land-Use and Land-Cover (LULC)}, in the
+Historical estimates of Land-Use and Land-Cover (LULC), in the
 pre-satellite era (\textless{} 1972), exist but generally rely on
@@
 data across the US and central Brazil \citep{song2015} as well as aerial
-surveys in post World War II German \citep{nita2018}. Survey data across
-the African continent is less common, inaccessible or both. Some studies
-do exist, as \citet{buitenwerf2012} and \citet{hudak1998} used aerial
-survey images to map vegetation changes in South African savannas,
-whilst \citet{frankl2015} and \citet{nyssen2016} mapped the Ethiopian
-highlands of 1930's.
+surveys in post World War II Germany \citep{nita2018}. Survey data
+across the African continent is less common, inaccessible or both. Some
+studies do exist, as \citet{buitenwerf2012} and \citet{hudak1998} used
+aerial survey images to map vegetation changes in South African
+savannas, whilst \citet{frankl2015} and \citet{nyssen2016} mapped the
+Ethiopian highlands of \textbf{1930s}.
 
@@
 
-Here, we use a combination of historical aerial photography and
-contemporary remote sensing data to map long-term changes in the extent
-and structure of the tropical forest surrounding Yangambi (DR Congo), in
-the central Congo Basin, effectively linking the start of the
-antrophocene \citep{lewis2015} with current assessments. Yangambi was,
-and remains, a focal center of forest and agricultural research and
-development in the central Congo Basin. Past research in the region
-allows for thorough assessment of land-use and land-cover change from a
-multi-disciplinary point of view, confronting us with complex
-deforestation and land-use patterns.
+Here, we use a combination of historical aerial photography
+\textbf{(1958)} and contemporary remote sensing data
+\textbf{(2000-2018)} to map long-term changes in the extent and
+structure of the tropical forest surrounding Yangambi (DR Congo), in the
+central Congo Basin, effectively linking the start of the
+\textbf{anthrophocene} \citep{lewis2015} with current assessments.
+Yangambi was, and remains, a focal center of forest and agricultural
+research and development in the central Congo Basin. Past research in
+the region allows for thorough assessment of land-use and land-cover
+change from a multi-disciplinary point of view, confronting us with
+complex deforestation and land-use patterns.
 
@@
 dataset extensively leveraging data augmentation. Our methodology aims
-to provides a historical insight into important LULCC spatial patterns
-in Yangambi, such as fragmentation and edge complexity. We further
-contextualize the influence of changes in the forest's life history on
-past and current research into Above Ground Carbon (AGC) storage
-\citep{Kearsley2013} and biodiversity \citep{vandeperre2018} in the
-central Congo Basin. Our fast scalable mapping approach for historical
-aerial survey data, using limited supervised input, would further
-support long-term land-use and land-cover change analysis across the
-central Congo Basin.
+to \textbf{provide} a historical insight into important LULCC spatial
+patterns in Yangambi, such as fragmentation and edge complexity. We
+further contextualize the influence of changes in the forest's life
+history on past and current research into Above Ground Carbon (AGC)
+storage \citep{Kearsley2013} and biodiversity \citep{vandeperre2018} in
+the central Congo Basin. Our fast scalable mapping approach for
+historical aerial survey data, using limited supervised input, would
+further support long-term land-use and land-cover change analysis across
+the central Congo Basin.
 
@@
 area, supervised by the ``Institut Géographique du Congo Belge'' in
-Kinshasa, DR Congo (then Léopoldville). Black-and-White infrared images
-(0.4 - 0.9 \(\mu\)m) were gathered along flight paths running mostly
-from west to east, between \textbf{09:00 and 11:00} local time. Along a
-flight path continuous images were taken using a Wild Heerburgg RC5a
-(currently Leica Geosystems) with an Aviogon lens assembly (114.83mm / f
-5.6, with a 90\degree view angle) resulting in square photo negatives of
-180 by 180 mm. Flights were flown at an average absolute altitude of
-\textasciitilde{}5200 m above sea level, covering roughly 18 530
-km\textsuperscript{2} at an approximate scale of 1/40 000. The use of
-the integrated autograph system ensured timely acquisition of pictures
-with a precise overlap (\textasciitilde{}1/3) between images. This large
-overlap between images together with flight parameters would allow
-post-processing, using stereographs, to create accurate topographic
-maps. Original data from this campaign are stored in the
-\href{https://www.africamuseum.be/en}{Royal Museum for Central Africa}
-in Tervuren, Belgium (Figure 1).
+Kinshasa, DR Congo (then Léopoldville). Black-and-White
+\textbf{panchromatic} images (0.4 - 0.9 \(\mu\)m) were gathered along
+flight paths running mostly from west to east, between 09:00 and 11:00
+local time. Along a flight path continuous images were taken using a
+Wild Heerburgg RC5a (currently Leica Geosystems) with an Aviogon lens
+assembly (114.83mm / f 5.6, with a 90\degree view angle) resulting in
+square photo negatives of 180 by 180 mm. Flights were flown at an
+average absolute altitude of \textasciitilde{}5200 m above sea level,
+covering roughly 18 530 km\textsuperscript{2} at an approximate scale of
+1/40 000. The use of the integrated autograph system ensured timely
+acquisition of pictures with a precise overlap (\textasciitilde{}1/3)
+between images. This large overlap between images together with flight
+parameters would allow post-processing, using stereographs, to create
+accurate topographic maps. Original data from this campaign are stored
+in the \href{https://www.africamuseum.be/en}{Royal Museum for Central
+Africa} in Tervuren, Belgium (Figure 1).
 
@@
 consisted of permanent structures which could be verified in both old
-and new aerial imagery (i.e.~ESRI World Imagery) \textbf{and consisted
-of corner points of build structures (e.g.~a building, bridge or
-swimming pool etc.).} Although clouds were removed during the SfM
-routine we did not mask all clouds in the final orthomosaic to maximize
-forest coverage. The final scene was cropped, to provide consistent wall
-to wall coverage of the reconstructed scene. The orthomosaic was
-exported as a geotiff for further georeferencing in QGIS
-\citep{qgisdevelopmentteam2018} using the georeferencer plugin (version
-3.1.9) and additional ESRI World Imagery high resolution reference data.
-We used 3\textsuperscript{th} degree polynomial and 16 ground control
-points to correct the final image. Ground control points, raw image data
-and final processed image are provided in addition to measures of
-uncertainty such as mean and median error across all ground control
-points. All subsequent analysis are executed on the final geo-referenced
-orthomosaic or subsets of it.
+and new aerial imagery (i.e.~ESRI World Imagery) and consisted of corner
+points of build structures (e.g.~a building, bridge or swimming pool
+etc.). \textbf{Although most clouds were removed during the SfM routine
+some were retained to provide sufficient SfM tie points to maximize
+continuous forest coverage in the final orthomosaic.} The final scene
+was cropped, to provide consistent wall to wall coverage of the
+reconstructed scene. The orthomosaic was exported as a geotiff for
+further georeferencing in QGIS \citep{qgisdevelopmentteam2018} using the
+georeferencer plugin (version 3.1.9) and additional ESRI World Imagery
+high resolution reference data. We used \textbf{3\textsuperscript{rd}}
+degree polynomial and 16 ground control points to correct the final
+image. Ground control points, raw image data and final processed image
+are provided in addition to measures of uncertainty such as mean and
+median error across all ground control points. All subsequent analysis
+are executed on the final geo-referenced orthomosaic or subsets of it.
 
@@
 thus excluding tree plantations, thinned or deteriorated forest stands
-which \textbf{showed visible canopy cover loss}, fields and buildings.
-We used the Unet Convolutional Neural Net (CNN, \citet{ronneberger2015})
+which showed visible canopy cover loss, fields and buildings. We used
+the Unet Convolutional Neural Net (CNN, \citet{ronneberger2015})
 architecture implemented in Keras \citep{chollet2015} with an
@@
 (i.e.~forest or non-forested). Training data were collected from the
-orthomosaic by randomly selecting 513 pixel square tiles from
-\textbf{locations within} homogeneous forested or non-forested polygons
-in the historical orthomosaic \textbf{(Figure 5). Separate polygons were
-selected for training, testing and validation purposes. Validation
-polygons were sampled 300 times, while both testing and validation
-polygons were sampled at 100 random locations. Tiles extracted from
-locations close to the polygon border at times contained mixed cover
-types. Tiles with mixed cover types were removed from the list of source
-tiles (Table 1).} Homogeneous \textbf{source} tiles were combined in
-synthetic landscapes using a random gaussian field based binary mask
-(Figure 3). We generated 5000 synthetic landscapes for training, while
+orthomosaic by randomly selecting 513 pixel square tiles from locations
+within homogeneous forested or non-forested polygons in the historical
+orthomosaic (Figure 5). Separate polygons were selected for training,
+testing and validation purposes. Validation polygons were sampled 300
+times, while both testing and validation polygons were sampled at 100
+random locations. Tiles extracted from locations close to the polygon
+border at times contained mixed cover types. Tiles with mixed cover
+types were removed from the list of source tiles (Table 1). Homogeneous
+source tiles were combined in synthetic landscapes using a random
+gaussian field based binary mask (Figure 3). We generated 5000 synthetic
+landscapes (balancing forest and non-forest classes) for training, while
 500 landscapes were generated for both the validation and the testing
@@
 forest texture transitions. We applied this technique to 10\% of the
-generated synthetic landscapes \textbf{(across training, validation and
-testing data)}.
+generated synthetic landscapes (across training, validation and testing
+data).
 
@@
 moving window approach with a step size of 110 pixels and a majority
-vote \textbf{(\textgreater{} 50\% agreement)} across overlapping areas
-to limit segmentation edge effects. \textbf{In addition, we provide raw
-pixel level classification agreement data for quality control purposes
-(see Data availability below).} We refer to Figure 4 for a synoptic
-overview of the full deep learning workflow.
+vote (\textgreater{} 50\% agreement) across overlapping areas to limit
+segmentation edge effects. In addition, we provide raw pixel level
+classification agreement data for quality control purposes (see Data
+availability below). We refer to Figure 4 for a synoptic overview of the
+full deep learning workflow.
 
@@
 
-\textbf{We report the CNN accuracy based upon the IoU of our
-out-of-sample validation dataset of synthetic landscapes. In addition,
-we report confusion matrices for all pixels across the homogeneous
-validation polygons, as well as the training and testing polygons (see
-Figure 5). Furthermore, we used the first acquisition of a recent
-pan-chromatic Geo-Eye 1 stereo pair (Geo-Eye, Thornton, Colorado,
-U.S.A., order 737537, 2011-11-11 08:55 GMT or 09:55 local time) to
-classify and assess the robustness of the CNN algorithm on contemporary
-remote sensing imagery with similar spectral and spatial
-characteristics. We used the Global Forest Change version 1.6 (GFC, tile
-10N-020E) \citep{hansen2013} map data, capturing forest loss up to 2011,
-to quantify accuracy on downsampled CNN Geo-Eye classification results.
-Once more, we report the confusion matrix of between the GFC and CNN
-derived forest cover maps, masking clouds and cloud shadows. To
-summarize confusion matrices we report accuracy as:}
+We report the CNN accuracy based upon the IoU of our out-of-sample
+validation dataset of synthetic landscapes. In addition, we report
+confusion matrices for all pixels across the homogeneous validation
+polygons, as well as the training and testing polygons (see Figure 5).
+Furthermore, we used the first acquisition of a recent pan-chromatic
+Geo-Eye 1 stereo pair (Geo-Eye, Thornton, Colorado, U.S.A., order
+737537, 2011-11-11 08:55 GMT or 09:55 local time) to classify and assess
+the robustness of the CNN algorithm on contemporary remote sensing
+imagery with similar spectral and spatial characteristics. We used the
+Global Forest Change version 1.6 (GFC, tile 10N-020E) \citep{hansen2013}
+map data, capturing \textbf{deforestation} up to 2011, to quantify
+accuracy on downsampled CNN Geo-Eye classification results. Once more,
+we report the confusion matrix of between the GFC and CNN derived forest
+cover maps, masking clouds and cloud shadows. To summarize confusion
+matrices we report accuracy as:
 
@@
 
-\textbf{in which TP, TN, FP, FN are True Positive, True Negative, False
-Positive and False Negative, respectivelly.}
+in which TP, TN, FP, FN are True Positive, True Negative, False Positive
+and False Negative, respectivelly.
 
@@
 Global Forest Change version 1.6 (GFC, tile 10N-020E) map data
-\citep{hansen2013}. Using the \textbf{GFC} data we calculated the latest
-state of the forest with respect to the conditions at the start of 1958,
-60 years earlier. In our analysis we only included \textbf{GFC} pixels
-which recorded no \textbf{forest} loss throughout the whole 2000 - 2018
-period. As the resolution of the historical forest classification
-exceeds that of the GFC map we downsampled our historical forest cover
-data to 30 m GFC resolution \textbf{using a nearest neighbour approach}.
-We masked out all water bodies using the Global Forest Change survey
-data mask layer, and limited the analysis to the right bank of the Congo
-river. We provide summary statistics of historical and contemporary
-forest loss and regrowth. \textbf{We map persistent forest loss after
-1958, regrowth after loss in 1958, recent forest loss and long term
-(stable) forest cover.}
+\citep{hansen2013}. Using the GFC data we calculated the latest state of
+the forest with respect to the conditions at the start of 1958, 60 years
+earlier. In our analysis we only included GFC pixels which recorded no
+forest loss throughout the whole 2000 - 2018 period, \textbf{as such
+locations which would see reforestation or deforestation between 2000
+and 2018 would be marked as non-forest (i.e.~disturbed)}. As the
+resolution of the historical forest classification exceeds that of the
+GFC map we downsampled our historical forest cover data to 30 m GFC
+resolution using a nearest neighbour approach. We masked out all water
+bodies using the Global Forest Change survey data mask layer, and
+limited the analysis to the right bank of the Congo river. We provide
+summary statistics of historical and contemporary \textbf{deforestation
+and reforestation}. We map permanent forest loss after 1958,
+\textbf{reforestation} after 1958, \textbf{recent deforestation} and
+long term (stable) forest cover. \textbf{All references to changes over
+time in the context of our analysis explicitly compare the historical
+and contemporary periods from hereon forth.}
 
@@
 To quantify changes in the structure of forest cover and its
-disturbances we used spatial pattern analysis landscape fragmentation
-metrics \citep{dale1999}. We report the ratio of edge to area and the
-fractal index to quantify landscape complexity of forest disturbances.
+disturbances we used spatial pattern analysis landscape (fragmentation)
+metrics \citep{dale1999}. Landscape metrics provide a mathematical
+framework for the analysis of discrete land‐cover classes and allows us
+to capture their composition and configuration. These metrics are
+therefore commonly used to compare how landscapes change over time
+\citep{hesselbarth2019}. In particular, fractals provide a way to
+quantify complex natural landscapes, including their self-similarity,
+across scales \citep{Li2000, Mandelbrot1977} We report the ratio of edge
+to area and the fractal dimension to quantify landscape complexity of
+forest disturbances. A fractal dimension closer to 2 suggest a more
+complex (fragmented) landscape.
+
 Statistics were calculated for all forest disturbance patches larger
@@
 
-We estimated above ground carbon (AGC) losses and gains over time using
-plot based averages of recent inventory data at permanent sampling plots
-in the area (Figure 2). We refer to \citet{Kearsley2013} for the survey
-method and allometric relations used to scale the survey data. Unlike
-standard square 1 ha plots, edge plots were set back 200 m from forest
-edges and were 50x200 m, with the 50 m side of the plot along the forest
-edge and continuing 200 m into the forest (Appendix Table 1). We further
-confirmed that forest edge plots did not show a significantly different
-AGC compared to those of non-edged / mixed forest plots (Mann Whitney U
-test, p \textless{} 0.05). Thus it was not necessary to explicitly
-quantify changes in AGC caused by edge effects. Moreover, we used the
-mean values of the mixed forest as representative for potential AGC
-losses. Despite the challenges inherent in quantifying AGC for forest
-edges we mapped the total extent of the edges in the contemporary
-landscape. To align our landscape analysis with exploratory analysis of
-the survey data we used a buffer of 200 m to estimate the extent of
-forest edges and patches, up to the location of forest edge plots.
+We estimated above ground carbon (AGC) losses and gains, \textbf{due to
+deforestation and reforestation}, using plot based averages of recent
+inventory data at permanent sampling plots in the area (Figure 2). We
+refer to \citet{Kearsley2013} for the survey method and allometric
+relations used to scale the survey data. Unlike standard square 1 ha
+plots, edge plots were set back 200 m from forest edges and were 50x200
+m, with the 50 m side of the plot along the forest edge and continuing
+200 m into the forest (Appendix Table 1). We further confirmed that
+forest edge plots did not show a significantly different AGC compared to
+those of non-edged / mixed forest plots (Mann Whitney U test, p
+\textless{} 0.05). Thus it was not necessary to explicitly quantify
+changes in AGC caused by edge effects. Moreover, we used the mean values
+of the mixed forest as representative for potential AGC losses. Despite
+the challenges inherent in quantifying AGC for forest edges we mapped
+the total extent of the edges in the contemporary landscape. To align
+our landscape analysis with exploratory analysis of the survey data we
+used a buffer of 200 m to estimate the extent of forest edges and
+patches, up to the location of forest edge plots.
 
@@
 plantations resemble old-regrowth values (81.8 Mg C
-ha\textsuperscript{-1}, see \textbf{Appendix Table 1}). We therefore use
-both the estimates of old-regrowth and mixed forest to estimate AGC for
-regrowth. We did not have sufficient data to account for individual
-changes in AGC across plantations.
+ha\textsuperscript{-1}, see Appendix Table 1). We therefore use both the
+estimates of old-regrowth and mixed forest to estimate AGC for regrowth.
+We did not have sufficient data to account for individual changes in AGC
+across plantations.
 
@@
 Our analysis provides a first spatially explicity historical composite
-of aerial survey images \textbf{in support of mapping land-use and
-land-cover} within the Congo Basin. The use of high resolution
-historical images combined with SfM image processing techniques allowed
-us to mosaic old imagery across a large extent. The final orthomosaic
-composition of the Yangambi region \textbf{provided} an image scene
-covering approximately 733 million pixels across
-\textbf{\textasciitilde{}82800 ha} (\textasciitilde{}23x36 km, Figure
-2). The overall accuracy of the SfM orthomosaic composition was 0.88
-m/pixel using the sparse cloud DEM for corrections at 45.8 m/pixel. The
-resulting georeferenced scene had a spatial accuracy of approximately
-\textbf{23 m}. Further georeferencing outside the SfM workflow reduced
-the mean error at the ground control points to 5.3 \(\pm\) 4.9 px
-(\textasciitilde{}4.7 \(\pm\) 4.3 m), with a median error of 2.9 px (2.6
-m). The orthomosaic served as input for all subsequent LULCC analysis
-with all derived maps provided with the manuscript repository (see data
-\& code availability statements below).
+of aerial survey images in support of mapping land-use and land-cover
+within the Congo Basin. The use of high resolution historical images
+combined with SfM image processing techniques allowed us to mosaic old
+imagery across a large extent. The final orthomosaic composition of the
+Yangambi region provided an image scene covering approximately 733
+million pixels across \textasciitilde{}82800 ha (\textasciitilde{}23x36
+km, Figure 2). The overall accuracy of the SfM orthomosaic composition
+was 0.88 m/pixel using the sparse cloud DEM for corrections at 45.8
+m/pixel. The resulting georeferenced scene had a spatial accuracy of
+approximately 23 m. Further georeferencing outside the SfM workflow
+reduced the mean error at the ground control points to 5.3 \(\pm\) 4.9
+px (\textasciitilde{}4.7 \(\pm\) 4.3 m), with a median error of 2.9 px
+(2.6 m). The orthomosaic served as input for all subsequent LULCC
+analysis with all derived maps provided with the manuscript repository
+(see data \& code availability statements below).
 
@@
 
-\textbf{The CNN deep learning classifier reached an
-Intersection-over-Union accuracy of 97\% on the detection of disturbed
-forest in the out-of-sample (validation) synthetic landscape data. Using
-all pixels within the validatoin polygons (Figure 5) showed a similar
+The CNN deep learning classifier reached an Intersection-over-Union
+accuracy of 97\% on the detection of disturbed forest in the
+out-of-sample (validation) synthetic landscape data. Using all pixels
+within the \textbf{validation} polygons (Figure 5) showed a similar
 accuracy value of \textasciitilde{}98\%. Using all polygons across the
@@
 pixels, between the landsat based GFC data and downscaled CNN results
-(Table 2 and Figure 7).}
+(Table 2 and Figure 7).
 
@@
 
-\caption{\label{tab:unnamed-chunk-7}Confusion matrix showing \% agreement between forest / non-forest classes using a Convolutional Neural Network (CNN) across previously selected homogenous areas.}
+\caption{\label{tab:unnamed-chunk-7}Confusion matrix showing \% agreement between forest / non-forest classes using a Convolutional Neural Network (CNN) across previously selected homogenous areas.
+In addition, overall accuracy is reported for each confusion matrix.}
 \centering
@@
 Scaling our classifier to the whole historical orthomosaic we detected
-\textbf{\textasciitilde{}16200 ha} (or \textasciitilde{}20\% of the
-scene) of disturbed forests. A large fraction of the disturbed area was
-restored in the period between the two \textbf{acquisition periods}. In
-total, \textbf{9918 ha}, or little over half of the affected forest was
-restored (Figure 5C-D, dark blue). Recent deforested areas, as
-registered through satellite remote sensing, approximate \textbf{8776 ha
-(Table 3, Figure 6 - light green)}.
+\textasciitilde{}16200 ha (or \textasciitilde{}20\% of the scene) of
+disturbed forests. A large fraction of the disturbed area was restored
+in the period between the two acquisition periods. In total, 9918 ha, or
+little over half of the affected forest was restored (Figure 5C-D, dark
+blue). Recent deforested areas, as registered through satellite remote
+sensing \textbf{(\textgreater{} 2000)}, approximate 8776 ha (Table 3,
+Figure 6 - light green).
 
@@
 fishbone pattern for forest clearing with very sharp edges, while
-current patterns are patchy and ad-hoc (\textbf{figure 6C}, dark blue
-and green colours respectively). These differences are reflected in the
-analysis of landscape metrics of forest loss. Between the historical and
-contemporary LULCC maps we see an increase in small disturbances, as
+current patterns are patchy and ad-hoc (figure 6C, dark blue and green
+colours respectively). These differences are reflected in the analysis
+of landscape metrics of \textbf{deforestation}. Between the historical
+and contemporary LULCC maps we see an increase in small disturbances, as
 indicated by the decreasing area of the mean patch size, down to
@@
 1451 \(\pm\) 943 m, compared to contemporary landscapes
-\textasciitilde{}921 \(\pm\) 362 m (Table 3). This shift in perimeter
-area ratio led to a similar change in the fractal index, slightly
-increasing in value to 1.1 \(\pm\) 0.05 from 1.09 \(\pm\) 0.04 over
-time. Values closer to a fractal index of 2 suggest a more complex
+\textasciitilde{}921 \(\pm\) 362 m (Table \textbf{4}). This shift in
+perimeter area ratio led to a similar change in the fractal index,
+slightly increasing in value to 1.1 \(\pm\) 0.05 from 1.09 \(\pm\) 0.04
+over time. Values closer to a fractal index of 2 suggest a more complex
 (fragmented) landscape.
@@
 0.05). Edge influence did not extend beyond 200 m from a forest edge,
-but still represented an area of \textbf{13151 ha}.
+but still represented an area of 13151 ha (Table 3).
 
@@
 \midrule
-Forest & 68455 & \\
--  of which edges & 13151 & \\
-Regrowth > 1958 loss & 9918 & 811 - 1592\\
-Loss > 2000 & 8776 & 1408\\
-Loss > 1958 (permanent) & 6282 & 1008\\
+Non-forest cover (1958) & 16200 & \\
+Forest cover & 68455 & \\
+-  of which forest edges & 13151 & \\
+reforestation >1958 & 9918 & 811 - 1592\\
+deforestation >2000 & 8776 & 1408\\
+\addlinespace
+deforestation >1958 (permanent) & 6282 & 1008\\
 \bottomrule
@@
 \emph{Brachystegia laurentii} is a recent development, changing the
-canopy structure visibly during the last half century (\textbf{Figure
-8}). The previous varied canopy structure gave way to a more dense and
-uniform canopy. This is reflected in a change of the FOTO PC value from
-0.19 historically to its current value of 0.54. This historical value is
-similar to the mean of contemporary mono-dominant stands of
-\emph{Gilbertiodendron dewevrei} with PC averaging 0.34 \(\pm\) 0.1, and
-is only slightly higher than historical values for a mixed forest (0.18
-\(\pm\) 0.08). The reverse pattern is seen in the contemporary PC
-values. Here, the value of 0.54 exceeds those of most mono-dominant
-stands (0.35 \(\pm\) 0.08), and is even further removed from the values
-noted for mixed forests (0.12 \(\pm\) 0.03).
+canopy structure visibly during the last half century (Figure 8). The
+previous varied canopy structure gave way to a more dense and uniform
+canopy. This is reflected in a change of the FOTO PC value from 0.19
+historically to its current value of 0.54 \textbf{(Figure 9).} This
+historical value is similar to the mean of contemporary mono-dominant
+stands of \emph{Gilbertiodendron dewevrei} with PC averaging 0.34
+\(\pm\) 0.1, and is only slightly higher than historical values for a
+mixed forest (0.18 \(\pm\) 0.08, \textbf{Figure 9)}. The reverse pattern
+is seen in the contemporary PC values. Here, the value of 0.54 exceeds
+those of most mono-dominant stands (0.35 \(\pm\) 0.08), and is even
+further removed from the values noted for mixed forests (0.12 \(\pm\)
+0.03, \textbf{Figure 9}).
 
@@
 
+\begin{figure}
+
+{\centering \includegraphics[width=0.5\linewidth]{./figures/foto_bplot_psp}
+
+}
+
+\caption{Boxplots comparing the first principal component (PC1) of a site based FOTO analysis across different forest types for both contemporary (Geo-Eye) and historical orthomosaic data.}\label{fig:unnamed-chunk-13}
+\end{figure}
+
 Using only small subsets around existing permanent sampling plots we
@@
 historical and contemporary imagery markedly higher for the
-mono-dominant forest types compared to all others (\textbf{Appendix
-Figure 4}). Provided that the young-regrowth and fallow permanent
-sampling plots have seen recent disturbance the Wilcoxon signed rank
-test on the mixed and mono-dominant plots between the historical and
-contemporary PC values did not show a significant difference (p
-\textgreater{} 0.05). Similarly, no significant different using PC
-values extracted from the whole scene analysis was noted (p
-\textgreater{} 0.05). Any relationships between contemporary Geo-Eye
-data and permanent sampling plot measurements of Above Ground Carbon,
-stem density and species richness were non-significant (p \textgreater{}
-0.05, Appendix \textbf{Figures 5-7}).
+mono-dominant forest types compared to all others (Appendix Figure 4).
+Provided that the young-regrowth and fallow permanent sampling plots
+have seen recent disturbance the Wilcoxon signed rank test on the mixed
+and mono-dominant plots between the historical and contemporary PC
+values did not show a significant difference (p \textgreater{} 0.05).
+Similarly, no significant different using PC values extracted from the
+whole scene analysis was noted (p \textgreater{} 0.05). Any
+relationships between contemporary Geo-Eye data and permanent sampling
+plot measurements of Above Ground Carbon, stem density and species
+richness were non-significant (p \textgreater{} 0.05, Appendix
+\textbf{Figures 4-6}).
 
 Furthermore, visual inspection of the scene wide analysis suggests
-historical scences do not show landscape wide canopy features
-(\textbf{Figure 9 A-B}), unlike the contemporary scene (\textbf{Figure 9
-C-D}). Where the FOTO algorithm picks up landscape features such as
-changes in texture across the contemporary Geo-Eye scene (e.g.~the river
-valley as a diagonal line in \textbf{Figure 9D}), however, no
-corresponding landscape patterns are found by the FOTO algorithm in the
-historical orthomosaic.
+historical scences do not show landscape wide canopy features (Figure 10
+A-B), unlike the contemporary scene (Figure 10 C-D). Where the FOTO
+algorithm picks up landscape features such as changes in texture across
+the contemporary Geo-Eye scene (e.g.~the river valley as a diagonal line
+in Figure 10D), however, no corresponding landscape patterns are found
+by the FOTO algorithm in the historical orthomosaic.
 
@@
 \caption{RGB visualizations of the first three principal components of scene wide FOTO texture analysis of historical and current (Geo-Eye) imagery.
-Current permanent sampling plots of mono-dominant, mixed, fallow and young (edge) forest plots are marked with open triangles, open circles, open squares and crosses, respectively.}\label{fig:unnamed-chunk-13}
+Current permanent sampling plots of mono-dominant, mixed, fallow and young (edge) forest plots are marked with open triangles, open circles, open squares and crosses, respectively.}\label{fig:unnamed-chunk-14}
 \end{figure}
@@
 negative. In addition, analogue photography clearly produces a distinct
-softness compared to digital imagery (\textbf{Figure 8}). Despite
-favourable nadir image acquistions \citep{verhoeven2017} image softness
-combined with illumination effects between flight paths, and the
-self-similar nature of vast canopy expanses
-\citep{park2019, simini2010, sole1995}, limited our ability to provide
-wall-to-wall coverage of the entire dataset containing 334 images. Few
-man made features in the scenes also made georeferencing challenging.
-Although the village of Yangambi provided a range of buildings as
-(hard-edge) references, other areas within the central Congo Basin might
-have fewer permanent structures and would require the use of soft-edged
-landscape features (e.g.~trees, river outflows). Research has shown that
-soft-edged features can help georeference scenes even when containing
-few man-made features \citep{hughes2006}. Our two step georeferencing
-approach resulted in a referencing accuracy of \textasciitilde{}4.7
-\(\pm\) 4.3 m across reference points. However, it shoud be noted that
-referencing accuracy of the final scene is less constrained toward the
-edges of the scene.
+softness compared to digital imagery (Figure 8). Despite favourable
+nadir image acquistions \citep{verhoeven2017} image softness combined
+with illumination effects between flight paths, and the self-similar
+nature of vast canopy expanses \citep{park2019, simini2010, sole1995},
+limited our ability to provide wall-to-wall coverage of the entire
+dataset containing 334 images. Few man made features in the scenes also
+made georeferencing challenging. Although the village of Yangambi
+provided a range of buildings as (hard-edge) references, other areas
+within the central Congo Basin might have fewer permanent structures and
+would require the use of soft-edged landscape features (e.g.~trees,
+river outflows). Research has shown that soft-edged features can help
+georeference scenes even when containing few man-made features
+\citep{hughes2006}. Our two step georeferencing approach resulted in a
+referencing accuracy of \textasciitilde{}4.7 \(\pm\) 4.3 m across
+reference points. However, it shoud be noted that referencing accuracy
+of the final scene is less constrained toward the edges of the scene.
 
@@
 
-\textbf{When classifying the orthomosaic into forest and non-forest
-states we favoured a deep learning supervised classification using a CNN
-over manual segmentation to guarantee an ``apples-to-apples'' comparison
+When classifying the orthomosaic into forest and non-forest states we
+favoured a deep learning supervised classification using a CNN over
+manual segmentation to guarantee an ``apples-to-apples'' comparison
 between the historical and the contemporary GFC forest cover map used in
@@
 accuracy of our algorithm might therefore be higher than our reported
-value.}
+value.
 
@@
 
-\textbf{Our approach uses broadly defined homogeneous polygons to
-construct a balanced dataset of synthethic landscapes. The methodology
-is analoguous to the use of sparse labelling as used by
-\citet{buscombe2018} and contrasts with the standard methodologies which
-generally extract pixel (windows) \citep{song2015} or delineate land
-cover classes \citep{nita2018} to drive a classifier or analysis. More
-so, the use of heavy image augmentation during model training sidesteps
-texture representation issues which affect classification of image
-scenes with inconsistent illumination or sharpness \citep{hudak1998} or
-ad-hoc feature engineering \citep{song2015}. The use of synthetic
-landscapes allowed us to account for most, but not all, of the
-variability within our orthomosaic. Our analysis has shown that despite
-being trained on historical data our model could map contemporary forest
-cover in remote sensing data with similar spatial and spectral
-characteristics (Figure 7), suggesting that the classifier consistently
-works across both space and time. We acknowledge that the use of
-synthetic landscapes is limited by the available homogeneous areas to
-sample from and the number of classes. Yet, the latter should not be a
-constrained as previous research efforts have focussed on simple forest
-loss maps \citep{hansen2013}.}
+Our approach uses broadly defined homogeneous polygons to construct a
+balanced dataset of synthethic landscapes. The methodology is analoguous
+to the use of sparse labelling as used by \citet{buscombe2018} and
+contrasts with the standard methodologies which generally extract pixel
+(windows) \citep{song2015} or delineate land cover classes
+\citep{nita2018} to drive a classifier or analysis. More so, the use of
+heavy image augmentation during model training sidesteps texture
+representation issues which affect classification of image scenes with
+inconsistent illumination or sharpness \citep{hudak1998} or ad-hoc
+feature engineering \citep{song2015}. The use of synthetic landscapes
+allowed us to account for most, but not all, of the variability within
+our orthomosaic. Our analysis has shown that despite being trained on
+historical data our model could map contemporary forest cover in remote
+sensing data with similar spatial and spectral characteristics (Figure
+7), suggesting that the classifier consistently works across both space
+and time. We acknowledge that the use of synthetic landscapes is limited
+by the available homogeneous areas to sample from and the number of
+classes. Yet, the latter should not be a constrained as previous
+research efforts have focussed on simple forest cover maps
+\citep{hansen2013}.
 
@@
 Our analysis shows that the majority of deforestation around Yangambi
-happened toward the late 1950's (\textbf{\textasciitilde{}16200 ha}).
-Considerable regrowth has occured since the aerial survey was executed
-(\textbf{\textasciitilde{}9918 ha}), and socio-economic instability
+happened toward the late 1950's (\textasciitilde{}16200 ha).
+Considerable \textbf{reforestation} has occured since the aerial survey
+was executed (\textasciitilde{}9918 ha), and socio-economic instability
 prevented further large scale forest exploitation. In particular, many
 plantations have reached maturity and forest has re-established in
-previously cleared or disturbed areas. The majority of this forest
-recovery takes the form of isolated patches of forest but is offset by
-further deforestation of previously untouched forest. Generally, the
-function and structure of forests can be influenced by forest edges that
-are located up to 1km away however most effects are pronounced within
-the first 300 m from the edge \citep{gascon2000}. Our analysis of edge
-effects on AGC has shown that the influence is negligible 200 m away
-from the edge. \citet{phillips2006} have shown similar weak responses to
-edge effects in the Amazon forest. Due to a lack of data on the extent
-(depth) of edge effects and their influence on AGC beyond 200 m we did
-not include any estimates of carbon loss or gain within these zones.
-However, it must be stated that the influence of edges throughout the
-landscape was not marginal as these areas would account for
-\textbf{13151 ha}. Thus edges could have a substantial negative
+previously cleared or disturbed areas. The majority of this
+\textbf{reforestation} takes the form of isolated patches of forest but
+is offset by further deforestation of previously untouched forest.
+Generally, the function and structure of forests can be influenced by
+forest edges that are located up to 1km away however most effects are
+pronounced within the first 300 m from the edge \citep{gascon2000}. Our
+analysis of edge effects on AGC has shown that the influence is
+negligible 200 m away from the edge. \citet{phillips2006} have shown
+similar weak responses to edge effects in the Amazon forest. Due to a
+lack of data on the extent (depth) of edge effects and their influence
+on AGC beyond 200 m we did not include any estimates of carbon loss or
+gain within these zones. However, it must be stated that the influence
+of edges throughout the landscape was not marginal as these areas would
+account for 13151 ha. Thus edges could have a substantial negative
 \citep{brinck2017} or positive \citep{reinmann2017} influence on AGC.
@@
 implicit protective policy measures (e.g.~INERA concessions,
-\citet{bustillo2018}) on a multi-decadal time scale. Forest regrowth in
-non-continuous areas within Yangambi could increase landscape
-connectivity and help increase biodiversity \citep{vandeperre2018}.
+\citet{bustillo2018}) on a multi-decadal time scale.
+\textbf{Reforestation} in non-continuous areas within Yangambi could
+increase landscape connectivity and help increase biodiversity
+\citep{vandeperre2018}.
 
@@
 Unlike large scale studies by \citet{Ploton2012} we could not scale this
-technique to historical data. \textbf{The successful use of our CNN
+technique to historical data. The successful use of our CNN
 classification model on a contemporary remote sensing data does suggest
 that texture can be used to consistently capture canopy properties 60
-years apart. This suggests inflexibility on part of the FOTO technique
-in dealing with non-standardized data.} We advise that future
-valorisation efforts should preferentially work from foto negatives (if
-available) to ensure optimal data quality in resolution, contrast and
-sharpness.
+years apart. \textbf{Differences in PC between forest types
+(e.g.~mono-dominant vs.~mixed, Figure 9) corroborate that texture can
+serve as a basis for LULC classification. However, inflexibility on part
+of the FOTO technique in dealing with non-standardized (historical)
+data, or scaling these results to AGC values, limits its use case.} We
+advise that future valorisation efforts should preferentially work from
+foto negatives (if available) to ensure optimal data quality in
+resolution, contrast and sharpness.
 
@@
 quantify long-term changes in LULC and AGC. We showed a shift from
-previously highly structured industrial deforestation of large areas
-\textbf{for plantation purposes}, to discrete smallholder clearing for
-farming, increasing landscape fragmentation but also opportunties for
-substantial regrowth. Efforts to quantify canopy texture features and
-their link to AGC had limited to no success. Our analysis provides
-insights into the rate at which deforestation and reforestation has
-taken place over a multi-decadal scale in the central Congo basin. As
-such, it provides a useful historical context while interpreting past
-and ongoing forest research in the area.
+previously highly structured industrial deforestation of large areas for
+plantation purposes, to discrete smallholder clearing for farming,
+increasing landscape fragmentation but also opportunties for substantial
+regrowth. Efforts to quantify canopy texture features and their link to
+AGC had limited to no success. Our analysis provides insights into the
+rate at which deforestation and reforestation has taken place over a
+multi-decadal scale in the central Congo basin. As such, it provides a
+useful historical context while interpreting past and ongoing forest
+research in the area.
 
@@
 \acknowledgments{This research was supported through the Belgian Science Policy office
-COBECORE project (BELSPO; grant BR/175/A3/COBECORE). KH acknowledges
+COBECORE project (BELSPO; grant BR/175/A3/COBECORE). K.H. acknowledges
 funding from the European Union Marie Skłodowska-Curie Action (project
-number 797668). I'm grateful for the support of Thales de Haulleville
-volunteering his time in scanning the images.}
+number 797668). K.H. grateful for the support of T.d.H. volunteering his
+time in scanning the images.}
 

### 2020-01-30 analysis/04_foto_summary_statistics.R

@@ -20,8 +20,7 @@ p <- ggplot(psp, aes(x=type, y=pc, fill = era)) +
         legend.position="bottom",
         text = element_text(size=20)) +
   scale_fill_manual(
-    values=c("#999999",
-              rgb(255, 173, 70, 100, maxColorValue = 255))) +
+    values=c("#1f78b4","#a6cee3")) +
   coord_flip()
 
 ggsave("manuscript/figures/foto_bplot_psp.png",

### 2020-01-30 analysis/06_lulcc_map.R

@@ -72,10 +72,10 @@ orthomosaic_subset_df <- orthomosaic_subset %>%
 forest_change_map <- ggplot()+
   geom_tile(data = forest_change_df, aes(x=x,y=y,fill=as.factor(val))) +
   scale_fill_brewer(palette = "Paired",
-                    labels=c(" no change",
-                             " forest regrowth > 1958",
-                             " forest loss > 2000",
-                             " forest loss > 1958")) +
+                    labels=c(" permanent forest cover",
+                             " reforestation >1958",
+                             " deforestation >2000",
+                             " deforestation >1958 (permanent)")) +
   coord_fixed(ratio = 1) +
   geom_sf(data = bb, colour = "white", fill = NA, size = 0.8) +
   ylim(c(0.725,1.07)) +
@@ -92,16 +92,17 @@ forest_change_subset_map <- ggplot()+
   geom_tile(data = forest_change_subset_df,
             aes(x=x,y=y,fill=as.factor(val))) +
   scale_fill_brewer(palette = "Paired",
-                    labels=c(" no change",
-                             " forest regrowth > 1958",
-                             " forest loss > 2000",
-                             " forest loss > 1958")) +
+                    labels=c(" permanent forest cover",
+                             " reforestation >1958",
+                             " deforestation >2000",
+                             " deforestation >1958 (permanent)")) +
   coord_fixed(ratio = 1) +
   geom_sf(data = bb, colour = NA, fill = NA) +
   theme_minimal() +
   theme(legend.title = element_blank(),
         legend.position="bottom",
         legend.direction="horizontal",
+        legend.text = element_text(size = 13),
         axis.text = element_text(size = 13),
         axis.text.x = element_text(angle = 45, hjust = 1),
         panel.grid.major = element_line(colour="grey"),
@@ -155,8 +156,10 @@ qa_map <- ggplot()+
   guides(fill = "legend", lty = "none") +
   theme(legend.position="bottom",
         legend.direction = "horizontal",
+        legend.text = element_text(size = 13),
         axis.text = element_text(size = 13),
         axis.text.x = element_text(angle = 45, hjust = 1),
+        axis.text.y = element_blank(),
         panel.grid.major = element_line(colour="grey"),
         panel.grid.minor = element_line(colour="grey")) +
   labs(
@@ -171,7 +174,7 @@ orthomosaic_subset_map <- ggplot()+
[…]

### 2020-01-30 analysis/07_foto_map.R

@@ -143,10 +143,11 @@ geoeye_foto_map <- ggplot()+
     x = "",
     y = "")
 
-p <- plot_grid(historical_map,
-               historical_foto_map,
+p <- plot_grid(historical_map + theme(axis.text.x = element_blank()),
+               historical_foto_map + theme(axis.text.x = element_blank(),
+                                           axis.text.y = element_blank()),
                geoeye_map,
-               geoeye_foto_map,
+               geoeye_foto_map  + theme(axis.text.y = element_blank()),
                nrow = 2,
                align = "hv",
                axis= "tblr",
@@ -213,7 +214,7 @@ geoeye_map <- ggplot()+
     y = "")
 
 p <- plot_grid(historical_map,
-               geoeye_map,
+               geoeye_map + theme(axis.text.y = element_blank()),
                ncol = 2,
                align = "hv",
                axis= "tblr",

### 2020-01-30 analysis/10_cnn_gfc_geo-eye_validation_map.R

@@ -99,25 +99,31 @@ spatial_map <- ggplot()+
   theme(legend.title = element_blank(),
         legend.position="bottom",
         legend.direction="horizontal",
-        legend.text = element_text(size = 8),
+        legend.text = element_text(size = 13),
         axis.text = element_text(size = 13),
         axis.text.x = element_text(angle = 45, hjust = 1),
+        axis.text.y = element_blank(),
         panel.grid.major = element_line(colour="grey"),
         panel.grid.minor = element_line(colour="grey")) +
   labs(
     x = "",
     y = "")
 
-p <- plot_grid(geoeye_map,
-               spatial_map,
+fig <- plot_grid(geoeye_map,
+               spatial_map + theme(legend.position = "none"),
                nrow = 1,
                align = "hv",
                axis= "tblr",
                labels = c("A", "B"),
                label_x = 0)
 
+p <- plot_grid(fig,
+               get_legend(spatial_map),
+               nrow = 2,
+               rel_heights = c(1, .1))
+
 save_plot("./manuscript/figures/visual_comparison_classifiers.png",
           p,
           base_width = 13,
-          base_height = 6,
+          base_height = 8,
           dpi = 350)

### 2020-02-06 manuscript/appendix.Rmd
AGC uncertainty + manuscript tweaks

@@
 ```{r echo = FALSE, eval = TRUE}
+flight_data %>%
+  kable(
+    caption="Flight path meta-data for the Isangi-Stanleyville aerial survey.
+Data provided consists of the flight path number, cardinal direction of the flight, the image numbers, and the duration of the flight provided by the start and end time of the acquisition.
+Data is sourced from Appendix Figure 2 and the sensor logs recorded in the margin of acquired images (see Figure 1 main manuscript).",
+    booktabs = TRUE,
+    escape = FALSE,
+    col.names = c("path",
+                  "direction",
+                  "image \\#",
+                  "start (H:M:S)",
+                  "end (H:M:S)")
+    ) %>%
+  kable_styling(latex_options = c("hold_position"))
+```
+
+```{r echo = FALSE, eval = TRUE}
 psp %>%
@@
 ```{r echo = FALSE, eval = TRUE}
-flight_data %>%
+agc <- psp %>%
+  group_by(type) %>%
+  summarize(mean_agc = mean(above.ground.C.stock..Mg.C.ha.1.),
+            sd_agc = sd(above.ground.C.stock..Mg.C.ha.1.),
+            mean_sd = mean(stem.density..ha.1.),
+            sd_sd = sd(stem.density..ha.1.),
+            mean_ba = mean(basal.area....m2.ha.1.),
+            sd_ba = sd(basal.area....m2.ha.1.),
+            n = length(above.ground.C.stock..Mg.C.ha.1.))
+
+mn_agc <- ifelse(!is.na(agc$sd_agc),
+                paste(round(agc$mean_agc,2), round(agc$sd_agc,2), sep = " $\\pm$ "),
+                round(agc$mean_agc,2))
+
+mn_sd <- ifelse(!is.na(agc$sd_sd),
+                paste(round(agc$mean_sd,2), round(agc$sd_sd,2), sep = " $\\pm$ "),
+                round(agc$mean_sd,2))
+
+mn_ba <- ifelse(!is.na(agc$sd_ba),
+                paste(round(agc$mean_ba,2), round(agc$sd_ba,2), sep = " $\\pm$ "),
+                round(agc$mean_ba,2))
+
+tb <- data.frame(type = agc$type,
+                 agc = mn_agc,
+                 sd = mn_sd,
+                 ba = mn_ba,
+                 n = agc$n)
+
+tb %>%
   kable(
-    caption="Flight path meta-data for the Isangi-Stanleyville aerial survey.
-Data provided consists of the flight path number, cardinal direction of the flight, the image numbers, and the duration of the flight provided by the start and end time of the acquisition.
-Data is sourced from Appendix Figure 2 and the sensor logs recorded in the margin of acquired images (see Figure 1 main manuscript).",
+    caption="Summary statistics for different forest plot types.
+We report the mean and standard deviation for Above Ground Carbon (AGC), basal area, stem density values and number of plots (as mentioned in Appendix Table 1).",
     booktabs = TRUE,
     escape = FALSE,
-    col.names = c("path",
-                  "direction",
-                  "image \\#",
-                  "start (H:M:S)",
-                  "end (H:M:S)")
+     col.names = c("forest type",
+                   "AGC (Mg C $ha\\textsuperscript{-1}$)",
+                   "stem density ($ha\\textsuperscript{-1}$)",
+                   "basal area ($m\\textsuperscript{2}$ $ha\\textsuperscript{-1}$)",
+                   "nr. plots"
+                   )
     ) %>%

### 2020-02-06 manuscript/appendix.tex
AGC uncertainty + manuscript tweaks

--- 2020-01-23 manuscript/appendix.tex
@@
 
-\caption{\label{tab:unnamed-chunk-1}Site description of all the current permanent sampling plots in the greater Yangambi region.
+\caption{\label{tab:unnamed-chunk-1}Flight path meta-data for the Isangi-Stanleyville aerial survey.
+Data provided consists of the flight path number, cardinal direction of the flight, the image numbers, and the duration of the flight provided by the start and end time of the acquisition.
+Data is sourced from Appendix Figure 2 and the sensor logs recorded in the margin of acquired images (see Figure 1 main manuscript).}
+\centering
+\begin{tabular}[t]{rllll}
+\toprule
+path & direction & image \# & start (H:M:S) & end (H:M:S)\\
+\midrule
+1 & W – E & 04 / 78 – 04 / 127 & 10:15:00 & 10:40:00\\
+2 & W – E & 05 / 01 – 05 / 56 & 10:15:00 & 10:40:00\\
+3 & W – E & 06 / 01 – 06 / 48 & 08:55:00 & 09:20:00\\
+4 & W – E & 07 / 01 – 07 / 20 & 09:30:00 & 09:45:00\\
+5 & W – E & 05 / 167 – 05 / 187 & 08:55:00 & 09:10:00\\
+\addlinespace
+6 & W – E & 07 / 67 – 07 / 82 & 10:50:00 & 11:00:00\\
+7 & E – W & 06 / 49 – 06 / 95 & 09:35:00 & 10:05:00\\
+8 & E – W & 05 / 57 – 05 / 103 & 10:20:00 & 10:50:00\\
+9 & E – W & 04 / 128 – 04 / 176 & 10:50:00 & 11:10:00\\
+10 & W – E & 08 / 01 – 08 / 27 & 09:05:00 & 09:20:00\\
+\addlinespace
+11 & W – E & 06 / 96 – 06 / 139 & 10:25:00 & 10:45:00\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\begin{table}[!h]
+
+\caption{\label{tab:unnamed-chunk-2}Site description of all the current permanent sampling plots in the greater Yangambi region.
 We list the forest type, plot number, geographic location (in decimal degrees), stem density, basal area Above Ground Carbon (AGC), species richness and Land-Use and Land-Cover Change (LULCC) classes in a radius of 100m around each plot location.}
@@
 
-\caption{\label{tab:unnamed-chunk-2}Flight path meta-data for the Isangi-Stanleyville aerial survey.
-Data provided consists of the flight path number, cardinal direction of the flight, the image numbers, and the duration of the flight provided by the start and end time of the acquisition.
-Data is sourced from Appendix Figure 2 and the sensor logs recorded in the margin of acquired images (see Figure 1 main manuscript).}
+\caption{\label{tab:unnamed-chunk-3}Summary statistics for different forest plot types.
+We report the mean and standard deviation for Above Ground Carbon (AGC), basal area, stem density values and number of plots (as mentioned in Appendix Table 1).}
 \centering
-\begin{tabular}[t]{rllll}
+\begin{tabular}[t]{llllr}
 \toprule
-path & direction & image \# & start (H:M:S) & end (H:M:S)\\
+forest type & AGC (Mg C $ha\textsuperscript{-1}$) & stem density ($ha\textsuperscript{-1}$) & basal area ($m\textsuperscript{2}$ $ha\textsuperscript{-1}$) & nr. plots\\
 \midrule
-1 & W – E & 04 / 78 – 04 / 127 & 10:15:00 & 10:40:00\\
-2 & W – E & 05 / 01 – 05 / 56 & 10:15:00 & 10:40:00\\
-3 & W – E & 06 / 01 – 06 / 48 & 08:55:00 & 09:20:00\\
-4 & W – E & 07 / 01 – 07 / 20 & 09:30:00 & 09:45:00\\
-5 & W – E & 05 / 167 – 05 / 187 & 08:55:00 & 09:10:00\\
+edge & 163.04 $\pm$ 19.39 & 404 $\pm$ 60.03 & 32.4 $\pm$ 2.39 & 5\\
+fallow & 4.37 $\pm$ 3.3 & 241 $\pm$ 154.15 & 3.72 $\pm$ 2.35 & 2\\
+mixed & 160.48 $\pm$ 23.84 & 467.25 $\pm$ 99.42 & 31.53 $\pm$ 3.26 & 8\\
+mono-dominant & 165.51 $\pm$ 12.75 & 354.17 $\pm$ 73.56 & 30.54 $\pm$ 2.64 & 6\\
+old-regrowth & 81.78 & 384 & 19.48 & 1\\
 \addlinespace
-6 & W – E & 07 / 67 – 07 / 82 & 10:50:00 & 11:00:00\\
-7 & E – W & 06 / 49 – 06 / 95 & 09:35:00 & 10:05:00\\
-8 & E – W & 05 / 57 – 05 / 103 & 10:20:00 & 10:50:00\\
-9 & E – W & 04 / 128 – 04 / 176 & 10:50:00 & 11:10:00\\
-10 & W – E & 08 / 01 – 08 / 27 & 09:05:00 & 09:20:00\\
-\addlinespace
-11 & W – E & 06 / 96 – 06 / 139 & 10:25:00 & 10:45:00\\
+young-regrowth & 37.16 $\pm$ 9.37 & 360.67 $\pm$ 74.9 & 18.29 $\pm$ 1.95 & 3\\
 \bottomrule
@@
 
-\caption{Overview of the complete flight plan of the survey around Kisangani (then Stanleyville) stored in the archives at the Africa Museum.}\label{fig:unnamed-chunk-3}
+\caption{Overview of the complete flight plan of the survey around Kisangani (then Stanleyville) stored in the archives at the Africa Museum.}\label{fig:unnamed-chunk-4}
 \end{figure}
@@
 
-{\centering \includegraphics[width=82.67in,height=0.75\textheight]{./figures/flight_paths_meta_data}
+{\centering \includegraphics[height=0.75\textheight]{./figures/flight_paths_meta_data}
 
@@
 
-\caption{Overview of the complete flight plan meta-data as stored in the archives at the Africa Museum.}\label{fig:unnamed-chunk-4}
+\caption{Overview of the complete flight plan meta-data as stored in the archives at the Africa Museum.}\label{fig:unnamed-chunk-5}
 \end{figure}
@@
 
-{\centering \includegraphics[width=14.39in,height=0.75\textheight]{./figures/archives_aerial_photos_2}
+{\centering \includegraphics[height=0.75\textheight]{./figures/archives_aerial_photos_2}
 
@@
 
-\caption{Shelves with aerial survey images at the Africa Museum archives.}\label{fig:unnamed-chunk-5}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=14.58in,height=0.5\textheight]{./figures/foto_bplot_psp}
-
-}
-
-\caption{Boxplots comparing the first principal component (PC1) of a site based FOTO analysis across different forest types for both contemporary (Geo-Eye) and historical orthomosaic data.}\label{fig:unnamed-chunk-6}
+\caption{Shelves with aerial survey images at the Africa Museum archives, showing vasts amounts of historical data still mostly unexplored.}\label{fig:unnamed-chunk-6}
 \end{figure}

### 2020-02-06 manuscript/main.Rmd
AGC uncertainty + manuscript tweaks

@@
 abstract: |
-   Given the impact of tropical forest disturbances on atmospheric carbon emissions, biodiversity and ecosystem productivity accurate long term reporting of Land-Use and Land-Cover (LULC) change in the pre-satellite era (< 1972) is an imperative.
-Here, we use a combination of historical (1958) aerial photography and contemporary remote sensing data to map long-term changes in the extent and structure of the tropical forest surrounding Yangambi (DR Congo), in the central Congo Basin.
-Our study leveraged Structure-from-Motion and a novel Convolutional Neural Network based LULC classifier, using synthetic landscapes based image augmentation, to map historical forest cover across a large sub meter resolution orthomosaic (~82800 ha) geo-referenced to ~4.7 $\pm$ 4.3 m.
+   Given the impact of tropical forest disturbances on atmospheric carbon emissions, biodiversity and ecosystem productivity, accurate long term reporting of Land-Use and Land-Cover (LULC) change in the pre-satellite era (< 1972) is an imperative.
+Here, we used a combination of historical (1958) aerial photography and contemporary remote sensing data to map long-term changes in the extent and structure of the tropical forest surrounding Yangambi (DR Congo), in the central Congo Basin.
+Our study leveraged Structure-from-Motion and a Convolutional Neural Network based LULC classifier, using synthetic landscapes based image augmentation, to map historical forest cover across a large orthomosaic (~82 800 ha) geo-referenced to ~4.7 $\pm$ 4.3 m at sub meter resolution.
 A comparison with contemporary LULC data showed a shift from previously highly regular industrial deforestation of large areas, to discrete smallholder farming clearing, increasing landscape fragmentation but also providing opportunties for substantial forest regrowth.
-Efforts to quantify long term canopy texture changes and a link to above ground carbon had limited to no success.
+We estimated above ground carbon gains through reforestation to range from 811 to 1592 Gg C, partially offseting historical deforestation (2416 Gg C), in our study area.
+Efforts to quantify long term canopy texture changes and their link to above ground carbon had limited to no success.
 Our analysis provides methods and insights into key spatial and temporal patterns of deforestation and reforestation at a multi-decadal scale, providing a historical context for past and ongoing forest research in the area.
@@
   knitr::opts_chunk$set(echo = TRUE)
-  library(orthodrc)
   library(raster)
@@
 # add a row if forest cover is 100% accurate
+# messes with table otherwise
 if(forest_stats$count[1] == 100){
@@
 Not only does tropical forest Land-Use and Land-Cover Change (LULCC) constitute 10 to 15% of the total global carbon emissions [@vanderwerf2009], changes in forest fragmentation affect the forest structure and function [@fauset2017; @brinck2017; @didham1999].
-Strong fragmentation effects **decrease** the number of large trees along forest edges [@laurance2000; @magnago2015], while species composition and biodiversity are equally negatively affected [@poorter2006; @barlow2016; @vandeperre2018].
+Strong fragmentation effects decrease the number of large trees along forest edges [@laurance2000; @magnago2015], while species composition and biodiversity are equally negatively affected [@poorter2006; @barlow2016; @vandeperre2018].
 Estimates show that 31% of carbon emissions are caused by edge effects alone [@brinck2017].
@@
 Survey data across the African continent is less common, inaccessible or both.
-Some studies do exist, as @buitenwerf2012 and @hudak1998 used aerial survey images to map vegetation changes in South African savannas, whilst @frankl2015 and @nyssen2016 mapped the Ethiopian highlands of **1930s**.
+Some studies do exist, as @buitenwerf2012 and @hudak1998 used aerial survey images to map vegetation changes in South African savannas, whilst @frankl2015 and @nyssen2016 mapped the Ethiopian highlands of 1930s.
 
 Across the central Congo Basin most of these historical images were collected within the context of national cartographic efforts by the "Institut Géographique du Congo Belge" in Kinshasa (then Léopoldville), DR Congo.
-Despite the existence of large archives of aerial survey imagery (Figure 1, Appendix Figure 3), as of yet, no studies have valorized these data.
+Despite the existence of large archives of aerial survey imagery of African rainforest (Figure 1, Appendix Figure 3), as of yet, no studies have valorized these data.
 The lack of a consistent valorization effort is unfortunate as the African rainforest is the second largest on Earth and covers ~630 million ha, representing up to 66 Pg of carbon storage [@Lewis2009b], and currently loses forest at an increasing pace [@butsic2015].
@@
 
-Here, we use a combination of historical aerial photography **(1958)** and contemporary remote sensing data **(2000-2018)** to map long-term changes in the extent and structure of the tropical forest surrounding Yangambi (DR Congo), in the central Congo Basin, effectively linking the start of the **anthrophocene** [@lewis2015] with current assessments.
+Here, we use a combination of historical aerial photography (1958) and contemporary remote sensing data (2000-2018) to map long-term changes in the extent and structure of the tropical forest surrounding Yangambi (DR Congo), in the central Congo Basin, effectively linking the start of the anthrophocene [@lewis2015] with current assessments.
 Yangambi was, and remains, a focal center of forest and agricultural research and development in the central Congo Basin.
-Past research in the region allows for thorough assessment of land-use and land-cover change from a multi-disciplinary point of view, confronting us with complex deforestation and land-use patterns.
+Past research in the region allows for thorough assessment of LULCC from a multi-disciplinary point of view, confronting us with complex deforestation and land-use patterns.
 
 We leverage Structure-from-Motion to generate a large orthomosaic of historical imagery and develop a Convolutional Neural Network based forest cover mapping approach, based upon a semi-supervised generated dataset extensively leveraging data augmentation.
-Our methodology aims to **provide** a historical insight into important LULCC spatial patterns in Yangambi, such as fragmentation and edge complexity.
+Our methodology aims to provide a historical insight into important LULCC spatial patterns in Yangambi, such as fragmentation and edge complexity.
 We further contextualize the influence of changes in the forest's life history on past and current research into Above Ground Carbon (AGC) storage [@Kearsley2013] and biodiversity [@vandeperre2018] in the central Congo Basin.
@@
 
-Data for the central Congo Basin region, surrounding Kisangani, were collected in several flights during the dry season of 1958 and 1959 (from 8/01/1958 to 20/02/1958 and from 28/12/1958 to 9/01/1959 respectively, see Appendix Figure 1) to generate topographic maps of the area, supervised by the "Institut Géographique du Congo Belge" in Kinshasa, DR Congo (then Léopoldville).
-Black-and-White **panchromatic** images (0.4 - 0.9 $\mu$m) were gathered along flight paths running mostly from west to east, between 09:00 and 11:00 local time.
+Data for the central Congo Basin region, surrounding Kisangani, were collected in several flights during the dry season of 1958 and 1959 (from 8/01/1958 to 20/02/1958 and from 28/12/1958 to 9/01/1959 respectively, see Appendix Figure 1 - 2 and Appendix Table 1) to generate topographic maps of the area, supervised by the "Institut Géographique du Congo Belge" in Kinshasa, DR Congo (then Léopoldville).
+Black-and-White panchromatic images (0.4 - 0.9 $\mu$m) were gathered along flight paths running mostly from west to east, between 09:00 and 11:00 local time.
 Along a flight path continuous images were taken using a Wild Heerburgg RC5a (currently Leica Geosystems) with an Aviogon lens assembly (114.83mm / f 5.6, with a 90\degree view angle) resulting in square photo negatives of 180 by 180 mm.
@@
 This large overlap between images together with flight parameters would allow post-processing, using stereographs, to create accurate topographic maps.
-Original data from this campaign are stored in the [Royal Museum for Central Africa](https://www.africamuseum.be/en) in Tervuren, Belgium (Figure 1).
+Original data from this campaign are stored in the [Royal Museum for Central Africa](https://www.africamuseum.be/en) in Tervuren, Belgium (Figure 1, Appendix Figure 3).
 
 ```{r echo=FALSE, out.width='75%', fig.align="center", fig.cap="A box of historical aerial photographs (left) and a single aerial photograph (right) showing part of the Congo river.
-Note the meta-data provided in the right side margin of the image, such as acquisition time and fleight height."}
+Note the meta-data provided in the right side margin of the image, such as acquisition time and flight height."}
 knitr::include_graphics('./figures/print_examples.jpg')
@@
 The bounding box of the orthomosaic data presented in this study is shown as a rectangle (23x36 km).
-The outline of a recent high-resolution Geo-eye panchromatic image is shown as a dashed dark grey rectangle (10x10km).
+The outline of a recent high-resolution Geo-eye panchromatic image is shown as a dashed dark grey rectangle (10x10 km).
 The location of various permanent sampling plots are shown as x, +, and open squares and triangles for the mixed, mono-dominant and edge plots respectively.
@@
 Using the SfM technique features, areas in images with a large degree of similarity, are matched across various images to reconstruct a three dimensional scene (topography) from two-dimensional image sequences.
-During the SfM analysis we masked clouds, glare or large water bodies such as the Congo river.
+During the SfM analysis we masked most clouds, glare or large water bodies such as the Congo river.
 
@@
 a building, bridge or swimming pool etc.).
-**Although most clouds were removed during the SfM routine some were retained to provide sufficient SfM tie points to maximize continuous forest coverage in the final orthomosaic.** The final scene was cropped, to provide consistent wall to wall coverage of the reconstructed scene.
+Although most clouds were removed during the SfM routine some were retained to provide sufficient SfM tie points to maximize continuous forest coverage in the final orthomosaic.
+The final scene was cropped, to provide consistent wall-to-wall coverage of the reconstructed scene.
 The orthomosaic was exported as a geotiff for further georeferencing in QGIS [@qgisdevelopmentteam2018] using the georeferencer plugin (version 3.1.9) and additional ESRI World Imagery high resolution reference data.
-We used **3^rd^** degree polynomial and 16 ground control points to correct the final image.
+We used 3^rd^ degree polynomial and 16 ground control points to correct the final image.
 Ground control points, raw image data and final processed image are provided in addition to measures of uncertainty such as mean and median error across all ground control points.
@@
 Furthermore, we used the first acquisition of a recent pan-chromatic Geo-Eye 1 stereo pair (Geo-Eye, Thornton, Colorado, U.S.A., order 737537, 2011-11-11 08:55 GMT or 09:55 local time) to classify and assess the robustness of the CNN algorithm on contemporary remote sensing imagery with similar spectral and spatial characteristics.
-We used the Global Forest Change version 1.6 (GFC, tile 10N-020E) [@hansen2013] map data, capturing **deforestation** up to 2011, to quantify accuracy on downsampled CNN Geo-Eye classification results.
-Once more, we report the confusion matrix of between the GFC and CNN derived forest cover maps, masking clouds and cloud shadows.
+We used the Global Forest Change version 1.6 (GFC, tile 10N-020E) [@hansen2013] map data, capturing deforestation up to 2011, to quantify accuracy on downsampled CNN Geo-Eye classification results.
+Once more, we report the confusion matrix between the GFC and CNN derived forest cover maps, masking clouds and cloud shadows.
 To summarize confusion matrices we report accuracy as:
@@
 
-To map long term LULCC in the Yangambi region we used the contemporary Global Forest Change version 1.6 (GFC, tile 10N-020E) map data [@hansen2013].
+To map long term LULCC in the Yangambi region we used the contemporary Global Forest Change version 1.6 (GFC, tile 10N-020E) (lossyear) map data [@hansen2013].
 Using the GFC data we calculated the latest state of the forest with respect to the conditions at the start of 1958, 60 years earlier.
-In our analysis we only included GFC pixels which recorded no forest loss throughout the whole 2000 - 2018 period, **as such locations which would see reforestation or deforestation between 2000 and 2018 would be marked as non-forest (i.e.
-disturbed)**.
+In our analysis we only included GFC pixels which recorded no forest loss throughout the whole 2000 - 2018 period.
+Forest loss in the context of GFC is defined as "a stand-replacement disturbance, or a change from a forest to non-forest state".
+As such, locations which would see reforestation or deforestation between 2000 and 2018 would be marked as non-forest (i.e.
+disturbed).
 As the resolution of the historical forest classification exceeds that of the GFC map we downsampled our historical forest cover data to 30 m GFC resolution using a nearest neighbour approach.
 We masked out all water bodies using the Global Forest Change survey data mask layer, and limited the analysis to the right bank of the Congo river.
-We provide summary statistics of historical and contemporary **deforestation and reforestation**.
-We map permanent forest loss after 1958, **reforestation** after 1958, **recent deforestation** and long term (stable) forest cover.
-**All references to changes over time in the context of our analysis explicitly compare the historical and contemporary periods from hereon forth.**
+We provide summary statistics of historical and contemporary deforestation and reforestation.
+We map permanent deforestation after 1958, reforestation after 1958, recent deforestation and long term (stable) forest cover.
+All references to changes over time in the context of our analysis explicitly compare the historical and contemporary periods from hereon forth.
 
@@
 
-To quantify changes in the structure of forest cover and its disturbances we used spatial pattern analysis landscape (fragmentation) metrics [@dale1999].
+To quantify changes in the structure of forest cover and its disturbances we used spatial landscape pattern analysis (i.e.
+fragmentation) metrics [@dale1999].
 Landscape metrics provide a mathematical framework for the analysis of discrete land‐cover classes and allows us to capture their composition and configuration.
 These metrics are therefore commonly used to compare how landscapes change over time [@hesselbarth2019].
-In particular, fractals provide a way to quantify complex natural landscapes, including their self-similarity, across scales [@Li2000; @Mandelbrot1977] We report the ratio of edge to area and the fractal dimension to quantify landscape complexity of forest disturbances.
+In particular, fractals provide a way to quantify complex natural landscapes, including their self-similarity across scales [@Li2000; @Mandelbrot1977].
+We report the ratio of edge to area and the fractal dimension to quantify landscape complexity of forest disturbances.
 A fractal dimension closer to 2 suggest a more complex (fragmented) landscape.
 
-Statistics were calculated for all forest disturbance patches larger than 1 ha and smaller than the 95^th^ percentile of the patch size distribution using the R package landscapemetrics [@hesselbarth2019].
+Statistics were calculated for all forest disturbance patches larger than 1 ha and smaller than the 95^th^ percentile of the patch size distribution using the R package "landscapemetrics" [@hesselbarth2019].
 We provide mean and standard deviation on edge, area, their ratio and fractal dimension for both the historical and contemporary @hansen2013 forest cover maps.
 
-We estimated above ground carbon (AGC) losses and gains, **due to deforestation and reforestation**, using plot based averages of recent inventory data at permanent sampling plots in the area (Figure 2).
+We estimated above ground carbon (AGC) losses and gains, due to deforestation and reforestation, using plot based averages of recent inventory data at permanent sampling plots in the area (Figure 2).
 We refer to @Kearsley2013 for the survey method and allometric relations used to scale the survey data.
-Unlike standard square 1 ha plots, edge plots were set back 200 m from forest edges and were 50x200 m, with the 50 m side of the plot along the forest edge and continuing 200 m into the forest (Appendix Table 1).
-We further confirmed that forest edge plots did not show a significantly different AGC compared to those of non-edged / mixed forest plots (Mann Whitney U test, p < 0.05).
+Unlike standard square 1 ha plots, edge plots (163.03 $\pm$ 19.39 Mg C ha^-1^, N = 5) were set back 200 m from forest edges and were 50x200 m, with the 50 m side of the plot along the forest edge and continuing 200 m into the forest (Appendix Table 2).
+We further confirmed that forest edge plots, as compared to mixed forest plots (160.48 $\pm$ 23.84 Mg C ha^-1^, N = 8, see Appendix Table 3 for all forest types), did not show a significantly different AGC (Mann Whitney U test, p < 0.05).
 Thus it was not necessary to explicitly quantify changes in AGC caused by edge effects.
-Moreover, we used the mean values of the mixed forest as representative for potential AGC losses.
+Moreover, we used the mean value, and its uncertainty (i.e.
+standard deviation), of the mixed forest as representative for potential AGC losses.
 Despite the challenges inherent in quantifying AGC for forest edges we mapped the total extent of the edges in the contemporary landscape.
@@
 Surveys of old plantations show a large variation in AGC, depending on age and the crop type.
-For example, the AGC values varied from 168.67 to 86.55 Mg C ha^-1^, for *Hevea brasiliensis* (rubber tree) and *Elaeis guineensis* (oil palm) plots respectively (@bustillo2018, personal communications).
-These higher values are in line with the mixed AGC estimates in the area, while the palm plantations resemble old-regrowth values (81.8 Mg C ha^-1^, see Appendix Table 1).
-We therefore use both the estimates of old-regrowth and mixed forest to estimate AGC for regrowth.
+For example, the AGC values varied from 86.55 to 168.67 Mg C ha^-1^, for *Elaeis guineensis* (oil palm) and *Hevea brasiliensis* (rubber tree) plots respectively (@bustillo2018, personal communications).
+These higher values are in line with the mixed plot AGC estimates (160.48 $\pm$ 23.84 Mg C ha^-1^, N = 8) in the area, while the palm plantations resemble old-regrowth values (81.87 Mg C ha^-1^, N = 1).
+To quantify AGC in reforested areas we therefore use both AGC estimates of old-regrowth and mixed forest, as lower and upper bounds.
 We did not have sufficient data to account for individual changes in AGC across plantations.
@@
 
-For both site based and scene analysis we correlated PC values with permanent sample plot inventory data such as stem density, above ground biomass and tree species richness.
+For both site based and scene analysis we correlated PC values with permanent sample plot inventory data such as stem density, above ground biomass and tree species richness (Appendix Table 2-3).
 Additional comparisons are made between contemporary Geo-Eye data and the historical orthomosaic derived PC values.
@@
 The use of high resolution historical images combined with SfM image processing techniques allowed us to mosaic old imagery across a large extent.
-The final orthomosaic composition of the Yangambi region provided an image scene covering approximately 733 million pixels across ~82800 ha (~23x36 km, Figure 2).
-The overall accuracy of the SfM orthomosaic composition was 0.88 m/pixel using the sparse cloud DEM for corrections at 45.8 m/pixel.
-The resulting georeferenced scene had a spatial accuracy of approximately 23 m.
+The final orthomosaic composition of the Yangambi region provided an image scene covering approximately 733 million pixels across ~82 800 ha with a resolution of 0.88 m / pixel (or ~23x36 km, Figure 2).
+The overall spatial accuracy of the SfM orthomosaic composition using the sparse cloud DEM (with a resolution of 45.8 m/pixel) was limited to approximately 23 m.
 Further georeferencing outside the SfM workflow reduced the mean error at the ground control points to 5.3 $\pm$ 4.9 px (~4.7 $\pm$ 4.3 m), with a median error of 2.9 px (2.6 m).
@@
 
-```{r echo=FALSE, out.width='100%', fig.align="center", fig.cap="orthomosaic", fig.cap="Overview of the final orthomosaic of the greater Yangambi region (A) as well as the forest cover classification uncertainty (B) used to generate the final Land Land Cover map (see Figure 6).
+```{r echo=FALSE, out.width='100%', fig.align="center", fig.cap="orthomosaic", fig.cap="Overview of the final orthomosaic of the greater Yangambi region (A) as well as the forest cover classification uncertainty (B) used to generate the final Land Cover map (see Figure 6).
 Homogeneous polygons used in the generation of the synthetic landscape for Convolutional Neural Network training, testing and validation are marked as dashed and full lines for forest and non-forest regions, respectively.
@@
 The CNN deep learning classifier reached an Intersection-over-Union accuracy of 97% on the detection of disturbed forest in the out-of-sample (validation) synthetic landscape data.
-Using all pixels within the **validation** polygons (Figure 5) showed a similar accuracy value of ~98%.
+Using all pixels within the validation polygons (Figure 5) showed a similar accuracy value of ~98%.
 Using all polygons across the scene, including those used in the generation of testing and training synthetic landscapes, increased the accuracy to ~99% (Table 2).
@@
 
-Scaling our classifier to the whole historical orthomosaic we detected ~16200 ha (or ~20% of the scene) of disturbed forests.
+Scaling our classifier to the whole historical orthomosaic we detected ~16 200 ha (or ~20% of the scene) of disturbed forests.
 A large fraction of the disturbed area was restored in the period between the two acquisition periods.
 In total, 9918 ha, or little over half of the affected forest was restored (Figure 5C-D, dark blue).
-Recent deforested areas, as registered through satellite remote sensing **(> 2000)**, approximate 8776 ha (Table 3, Figure 6 - light green).
+Recent deforested areas, as registered through satellite remote sensing (> 2000), approximate 8776 ha (Table 3, Figure 6 - light green).
 
@@
 Historical deforestation showed a classical fishbone pattern for forest clearing with very sharp edges, while current patterns are patchy and ad-hoc (figure 6C, dark blue and green colours respectively).
-These differences are reflected in the analysis of landscape metrics of **deforestation**.
+These differences are reflected in the analysis of landscape metrics of deforestation.
 Between the historical and contemporary LULCC maps we see an increase in small disturbances, as indicated by the decreasing area of the mean patch size, down to ~1.86 $\pm$ 0.75 ha from ~5.25 $\pm$ 5.02 ha historically.
-Perimeter lengths were longer historically, at 1451 $\pm$ 943 m,  compared to contemporary landscapes ~921 $\pm$ 362 m (Table **4**).
+Perimeter lengths were longer historically, at 1451 $\pm$ 943 m,  compared to contemporary landscapes ~921 $\pm$ 362 m (Table 4).
 This shift in perimeter area ratio led to a similar change in the fractal index, slightly increasing in value to 1.1 $\pm$ 0.05 from 1.09 $\pm$ 0.04 over time.
@@
 A comparison of forest edge plots with mixed forest plots showed no signficant difference in AGC, or other reported values such as species richness, basal area or stem density (Mann Whitney U test, p < 0.05).
-Edge influence did not extend beyond 200 m from a forest edge, but still represented an area of 13151 ha (Table 3).
+Edge influence did not extend beyond 200 m from a forest edge, but still represented an area of 13 151 ha (Table 3).
 
@@
 Recovery throughout the region was characterized for patches of forest and plantations.
-Assuming high density stands, based on previous work, this could amount to a carbon gains of 1592 Gg C across our study area, offsetting more recent losses of approximately 1408 Gg C.
+Assuming high density stands, based on previous work, this could amount to a carbon gains of up to 1592 Gg C across our study area, offsetting more recent losses of approximately 1408 $\pm$ 209 Gg C.
 On the other hand, at the low end, if we assume a lower carbon density of 81.8 Mg C ha^-1^ this would result in a total carbon gain of 811 Gg C.
-Using our approach results indicate that overall deforestation around Yangambi has resulted in a loss of ~2416 Gg C in AGC stocks.
+Using our approach results indicate that overall deforestation around Yangambi has resulted in a loss of ~2416 $\pm$ 358 Gg C in AGC stocks.
 
@@
 2013 based satellite remote sensing data.
-Spatial coverage statistics are provided in square kilometers (km) and hectares (ha), rounded to the nearest integer as well as Above Ground Carbon (AGC) scaled using recent survey measurements.",
+Spatial coverage statistics are provided hectares (ha), rounded to the nearest integer as well as Above Ground Carbon (AGC) scaled using recent survey measurements.",
        booktabs = TRUE,
@@
        col.names = c("ha",
-                     "Gg C $ha\\textsuperscript{-1}$")
+                     "Gg C")
          ) %>%
@@
 Visual interpretation of the scenes provide evidence that most locations do not change dramatically with respect to canopy composition, except for the large areas of disturbances in contemporary fallow or young-regrowth plots.
-One marked difference is noted in the mono-dominant plot 6 (Appendix Table 1).
+One marked difference is noted in the mono-dominant plot 6 (Appendix Table 2).
 Here, the current mono-dominant *Brachystegia laurentii* is a recent development, changing the canopy structure visibly during the last half century (Figure 8).
 The previous varied canopy structure gave way to a more dense and uniform canopy.
-This is reflected in a change of the FOTO PC value from 0.19 historically to its current value of 0.54 **(Figure 9).** This historical value is similar to the mean of contemporary mono-dominant stands of *Gilbertiodendron dewevrei* with PC averaging 0.34 $\pm$ 0.1, and is only slightly higher than historical values for a mixed forest (0.18 $\pm$ 0.08, **Figure 9)**.
+This is reflected in a change of the FOTO PC value from 0.19 historically to its current value of 0.54 (Figure 9).
+This historical value is similar to the mean of contemporary mono-dominant stands of *Gilbertiodendron dewevrei* with PC averaging 0.34 $\pm$ 0.1, and is only slightly higher than historical values for a mixed forest (0.18 $\pm$ 0.08, Figure 9).
 The reverse pattern is seen in the contemporary PC values.
-Here, the value of 0.54 exceeds those of most mono-dominant stands (0.35 $\pm$ 0.08), and is even further removed from the values noted for mixed forests (0.12 $\pm$ 0.03, **Figure 9**).
+Here, the value of 0.54 exceeds those of most mono-dominant stands (0.35 $\pm$ 0.08), and is even further removed from the values noted for mixed forests (0.12 $\pm$ 0.03, Figure 9).
 
@@
 Similarly, no significant different using PC values extracted from the whole scene analysis was noted (p > 0.05).
-Any relationships between contemporary Geo-Eye data and permanent sampling plot measurements of Above Ground Carbon, stem density and species richness were non-significant (p > 0.05, Appendix **Figures 4-6**).
+Any relationships between contemporary Geo-Eye data and permanent sampling plot measurements of Above Ground Carbon, stem density and species richness were non-significant (p > 0.05, Appendix Figures 4-6).
 
 Furthermore, visual inspection of the scene wide analysis suggests historical scences do not show landscape wide canopy features (Figure 10 A-B), unlike the contemporary scene (Figure 10 C-D).
-Where the FOTO algorithm picks up landscape features such as changes in texture across the contemporary Geo-Eye scene (e.g.
-the river valley as a diagonal line in Figure 10D), however, no corresponding landscape patterns are found by the FOTO algorithm in the historical orthomosaic.
+In the contemporary scene the FOTO algorithm picks up landscape features such as changes in texture along the river valley (the diagonal line in Figure 10D).
+However, no corresponding landscape patterns are found by the FOTO algorithm in the historical orthomosaic.
 
@@
 Few man made features in the scenes also made georeferencing challenging.
-Although the village of Yangambi provided a range of buildings as (hard-edge) references, other areas within the central Congo Basin might have fewer permanent structures and would require the use of soft-edged landscape features (e.g.
+Although the village of Yangambi provided a range of buildings as (hard-edge) references other areas within the central Congo Basin might have fewer permanent structures and would require the use of soft-edged landscape features (e.g.
 trees, river outflows).
@@
 
-Our analysis shows that the majority of deforestation around Yangambi happened toward the late 1950's (~16200 ha).
-Considerable **reforestation** has occured since the aerial survey was executed (~9918 ha), and socio-economic instability prevented further large scale forest exploitation.
+Our analysis shows that the majority of deforestation around Yangambi happened toward the late 1950s (~16 200 ha).
+Considerable reforestation has occured since the aerial survey was executed (~9918 ha), and socio-economic instability prevented further large scale forest exploitation.
 In particular, many plantations have reached maturity and forest has re-established in previously cleared or disturbed areas.
-The majority of this **reforestation** takes the form of isolated patches of forest but is offset by further deforestation of previously untouched forest.
-Generally, the function and structure of forests can be influenced by forest edges that are located up to 1km away however most effects are pronounced within the first 300 m from the edge [@gascon2000].
+The majority of this reforestation takes the form of isolated patches of forest but is offset by further deforestation of previously untouched forest.
+Generally, the function and structure of forests can be influenced by forest edges that are located up to 1 km away however most effects are pronounced within the first 300 m from the edge [@gascon2000].
 Our analysis of edge effects on AGC has shown that the influence is negligible 200 m away from the edge.
@@
 Due to a lack of data on the extent (depth) of edge effects and their influence on AGC beyond 200 m we did not include any estimates of carbon loss or gain within these zones.
-However, it must be stated that the influence of edges throughout the landscape was not marginal as these areas would account for 13151 ha.
+However, it must be stated that edges throughout the landscape make up a substantial area and account for 13 151 ha.
 Thus edges could have a substantial negative [@brinck2017] or positive [@reinmann2017] influence on AGC.
 Similarly, uncertainties in how to explicitly correct for plantations in the landscape present a further challenge.
-Thus, although our estimates are only indicative they do underscore the important influence of landscape structure when carbon accounting.
+Similarly, variability across mixed forest plots used in scaling above ground carbon estimates due to deforestation introduced additional uncertainty (see Appendix Table 3).
+Thus, although our estimates are only indicative they do underscore the important influence of landscape structure in carbon accounting.
 However, our findings do not indicate that deforestation in Congo basin is declining, on the contrary.
@@
 Over the past half century there has been a clear shift in land use in Yangambi (Figure 5).
-Land use has shifted away from for example a regular fishbone deforestation pattern that emerges when (large scale) agricultural interests dominate the landscape [@arima2015], to a more fragmented landscape (Figure 5D).
+Land use has shifted away from a regular (fishbone) deforestation pattern that emerges when (large scale) agricultural interests dominate the landscape [@arima2015], to a more fragmented landscape (Figure 5D).
 The former is consistent with historical land management at the time of the aerial survey [@bustillo2018].
-These regular patterns have since been reversed because of a decrease in large scale intensive agriculture, replaced by ad-hoc small scale subsistence farming with large perimeter to area relationships (i.e.
+These regular patterns reversed due to a decrease in large scale intensive agriculture and an increase in ad-hoc small scale subsistence farming with large perimeter to area relationships (i.e.
 ragged edges).
@@
 INERA concessions, @bustillo2018) on a multi-decadal time scale.
-**Reforestation** in non-continuous areas within Yangambi could increase landscape connectivity and help increase biodiversity [@vandeperre2018].
+Reforestation in non-continuous areas within Yangambi could increase landscape connectivity and help increase biodiversity [@vandeperre2018].
 
@@
 The successful use of our CNN classification model on a contemporary remote sensing data does suggest that texture can be used to consistently capture canopy properties 60 years apart.
-**Differences in PC between forest types (e.g.
+Differences in PC between forest types (e.g.
 mono-dominant vs.
 mixed, Figure 9) corroborate that texture can serve as a basis for LULC classification.
-However, inflexibility on part of the FOTO technique in dealing with non-standardized (historical) data, or scaling these results to AGC values, limits its use case.** We advise that future valorisation efforts should preferentially work from foto negatives (if available) to ensure optimal data quality in resolution, contrast and sharpness.
+However, inflexibility on part of the FOTO technique in dealing with non-standardized (historical) data, or scaling these results to AGC values, limits its use case.
+We advise that future valorisation efforts should preferentially work from foto negatives (if available) to ensure optimal data quality in resolution, contrast and sharpness.
 

### 2020-02-06 manuscript/main.tex
AGC uncertainty + manuscript tweaks

@@
 \abstract{Given the impact of tropical forest disturbances on atmospheric carbon
-emissions, biodiversity and ecosystem productivity accurate long term
+emissions, biodiversity and ecosystem productivity, accurate long term
 reporting of Land-Use and Land-Cover (LULC) change in the pre-satellite
-era (\textless{} 1972) is an imperative. Here, we use a combination of
+era (\textless{} 1972) is an imperative. Here, we used a combination of
 historical (1958) aerial photography and contemporary remote sensing
@@
 tropical forest surrounding Yangambi (DR Congo), in the central Congo
-Basin. Our study leveraged Structure-from-Motion and a novel
-Convolutional Neural Network based LULC classifier, using synthetic
-landscapes based image augmentation, to map historical forest cover
-across a large sub meter resolution orthomosaic (\textasciitilde{}82800
-ha) geo-referenced to \textasciitilde{}4.7 \(\pm\) 4.3 m. A comparison
+Basin. Our study leveraged Structure-from-Motion and a Convolutional
+Neural Network based LULC classifier, using synthetic landscapes based
+image augmentation, to map historical forest cover across a large
+orthomosaic (\textasciitilde{}82 800 ha) geo-referenced to
+\textasciitilde{}4.7 \(\pm\) 4.3 m at sub meter resolution. A comparison
 with contemporary LULC data showed a shift from previously highly
@@
 farming clearing, increasing landscape fragmentation but also providing
-opportunties for substantial forest regrowth. Efforts to quantify long
-term canopy texture changes and a link to above ground carbon had
-limited to no success. Our analysis provides methods and insights into
-key spatial and temporal patterns of deforestation and reforestation at
-a multi-decadal scale, providing a historical context for past and
-ongoing forest research in the area.}
+opportunties for substantial forest regrowth. We estimated above ground
+carbon gains through reforestation to range from 811 to 1592 Gg C,
+partially offseting historical deforestation (2416 Gg C), in our study
+area. Efforts to quantify long term canopy texture changes and their
+link to above ground carbon had limited to no success. Our analysis
+provides methods and insights into key spatial and temporal patterns of
+deforestation and reforestation at a multi-decadal scale, providing a
+historical context for past and ongoing forest research in the area.}
 
@@
 
+\usepackage{wrapfig} \usepackage{graphics} \usepackage{setspace} \doublespacing \usepackage{gensymb} \usepackage{booktabs}
 \usepackage{booktabs}
@@
 
-\usepackage{wrapfig} \usepackage{graphics} \usepackage{setspace} \doublespacing \usepackage{gensymb} \usepackage{booktabs}
-
 \begin{document}
@@
 \citep{fauset2017, brinck2017, didham1999}. Strong fragmentation effects
-\textbf{decrease} the number of large trees along forest edges
+decrease the number of large trees along forest edges
 \citep{laurance2000, magnago2015}, while species composition and
@@
 savannas, whilst \citet{frankl2015} and \citet{nyssen2016} mapped the
-Ethiopian highlands of \textbf{1930s}.
+Ethiopian highlands of 1930s.
 
@@
 Léopoldville), DR Congo. Despite the existence of large archives of
-aerial survey imagery (Figure 1, Appendix Figure 3), as of yet, no
-studies have valorized these data. The lack of a consistent valorization
-effort is unfortunate as the African rainforest is the second largest on
-Earth and covers \textasciitilde{}630 million ha, representing up to 66
-Pg of carbon storage \citep{Lewis2009b}, and currently loses forest at
-an increasing pace \citep{butsic2015}. Given the impact of LULCC on the
-structure and functioning of central African tropical forests, and their
-influence on both carbon dynamics \citep{Kearsley2013} and biodiversity
-\citep{vandeperre2018}, accurate long term reporting of historical
-forest cover warrants more attention \citep{willcock2016}.
-
-Here, we use a combination of historical aerial photography
-\textbf{(1958)} and contemporary remote sensing data
-\textbf{(2000-2018)} to map long-term changes in the extent and
-structure of the tropical forest surrounding Yangambi (DR Congo), in the
-central Congo Basin, effectively linking the start of the
-\textbf{anthrophocene} \citep{lewis2015} with current assessments.
-Yangambi was, and remains, a focal center of forest and agricultural
-research and development in the central Congo Basin. Past research in
-the region allows for thorough assessment of land-use and land-cover
-change from a multi-disciplinary point of view, confronting us with
-complex deforestation and land-use patterns.
+aerial survey imagery of African rainforest (Figure 1, Appendix Figure
+3), as of yet, no studies have valorized these data. The lack of a
+consistent valorization effort is unfortunate as the African rainforest
+is the second largest on Earth and covers \textasciitilde{}630 million
+ha, representing up to 66 Pg of carbon storage \citep{Lewis2009b}, and
+currently loses forest at an increasing pace \citep{butsic2015}. Given
+the impact of LULCC on the structure and functioning of central African
+tropical forests, and their influence on both carbon dynamics
+\citep{Kearsley2013} and biodiversity \citep{vandeperre2018}, accurate
+long term reporting of historical forest cover warrants more attention
+\citep{willcock2016}.
+
+Here, we use a combination of historical aerial photography (1958) and
+contemporary remote sensing data (2000-2018) to map long-term changes in
+the extent and structure of the tropical forest surrounding Yangambi (DR
+Congo), in the central Congo Basin, effectively linking the start of the
+anthrophocene \citep{lewis2015} with current assessments. Yangambi was,
+and remains, a focal center of forest and agricultural research and
+development in the central Congo Basin. Past research in the region
+allows for thorough assessment of LULCC from a multi-disciplinary point
+of view, confronting us with complex deforestation and land-use
+patterns.
 
@@
 dataset extensively leveraging data augmentation. Our methodology aims
-to \textbf{provide} a historical insight into important LULCC spatial
-patterns in Yangambi, such as fragmentation and edge complexity. We
-further contextualize the influence of changes in the forest's life
-history on past and current research into Above Ground Carbon (AGC)
-storage \citep{Kearsley2013} and biodiversity \citep{vandeperre2018} in
-the central Congo Basin. Our fast scalable mapping approach for
-historical aerial survey data, using limited supervised input, would
-further support long-term land-use and land-cover change analysis across
-the central Congo Basin.
+to provide a historical insight into important LULCC spatial patterns in
+Yangambi, such as fragmentation and edge complexity. We further
+contextualize the influence of changes in the forest's life history on
+past and current research into Above Ground Carbon (AGC) storage
+\citep{Kearsley2013} and biodiversity \citep{vandeperre2018} in the
+central Congo Basin. Our fast scalable mapping approach for historical
+aerial survey data, using limited supervised input, would further
+support long-term land-use and land-cover change analysis across the
+central Congo Basin.
 
@@
 (from 8/01/1958 to 20/02/1958 and from 28/12/1958 to 9/01/1959
-respectively, see Appendix Figure 1) to generate topographic maps of the
-area, supervised by the ``Institut Géographique du Congo Belge'' in
-Kinshasa, DR Congo (then Léopoldville). Black-and-White
-\textbf{panchromatic} images (0.4 - 0.9 \(\mu\)m) were gathered along
-flight paths running mostly from west to east, between 09:00 and 11:00
-local time. Along a flight path continuous images were taken using a
-Wild Heerburgg RC5a (currently Leica Geosystems) with an Aviogon lens
+respectively, see Appendix Figure 1 - 2 and Appendix Table 1) to
+generate topographic maps of the area, supervised by the ``Institut
+Géographique du Congo Belge'' in Kinshasa, DR Congo (then Léopoldville).
+Black-and-White panchromatic images (0.4 - 0.9 \(\mu\)m) were gathered
+along flight paths running mostly from west to east, between 09:00 and
+11:00 local time. Along a flight path continuous images were taken using
+a Wild Heerburgg RC5a (currently Leica Geosystems) with an Aviogon lens
 assembly (114.83mm / f 5.6, with a 90\degree view angle) resulting in
@@
 in the \href{https://www.africamuseum.be/en}{Royal Museum for Central
-Africa} in Tervuren, Belgium (Figure 1).
+Africa} in Tervuren, Belgium (Figure 1, Appendix Figure 3).
 
@@
 \caption{A box of historical aerial photographs (left) and a single aerial photograph (right) showing part of the Congo river.
-Note the meta-data provided in the right side margin of the image, such as acquisition time and fleight height.}\label{fig:unnamed-chunk-1}
+Note the meta-data provided in the right side margin of the image, such as acquisition time and flight height.}\label{fig:unnamed-chunk-1}
 \end{figure}
@@
 scene (topography) from two-dimensional image sequences. During the SfM
-analysis we masked clouds, glare or large water bodies such as the Congo
-river.
+analysis we masked most clouds, glare or large water bodies such as the
+Congo river.
 
@@
 points of build structures (e.g.~a building, bridge or swimming pool
-etc.). \textbf{Although most clouds were removed during the SfM routine
-some were retained to provide sufficient SfM tie points to maximize
-continuous forest coverage in the final orthomosaic.} The final scene
-was cropped, to provide consistent wall to wall coverage of the
+etc.). Although most clouds were removed during the SfM routine some
+were retained to provide sufficient SfM tie points to maximize
+continuous forest coverage in the final orthomosaic. The final scene was
+cropped, to provide consistent wall-to-wall coverage of the
 reconstructed scene. The orthomosaic was exported as a geotiff for
@@
 georeferencer plugin (version 3.1.9) and additional ESRI World Imagery
-high resolution reference data. We used \textbf{3\textsuperscript{rd}}
-degree polynomial and 16 ground control points to correct the final
-image. Ground control points, raw image data and final processed image
-are provided in addition to measures of uncertainty such as mean and
-median error across all ground control points. All subsequent analysis
-are executed on the final geo-referenced orthomosaic or subsets of it.
+high resolution reference data. We used 3\textsuperscript{rd} degree
+polynomial and 16 ground control points to correct the final image.
+Ground control points, raw image data and final processed image are
+provided in addition to measures of uncertainty such as mean and median
+error across all ground control points. All subsequent analysis are
+executed on the final geo-referenced orthomosaic or subsets of it.
 
@@
 Global Forest Change version 1.6 (GFC, tile 10N-020E) \citep{hansen2013}
-map data, capturing \textbf{deforestation} up to 2011, to quantify
-accuracy on downsampled CNN Geo-Eye classification results. Once more,
-we report the confusion matrix of between the GFC and CNN derived forest
-cover maps, masking clouds and cloud shadows. To summarize confusion
-matrices we report accuracy as:
+map data, capturing deforestation up to 2011, to quantify accuracy on
+downsampled CNN Geo-Eye classification results. Once more, we report the
+confusion matrix between the GFC and CNN derived forest cover maps,
+masking clouds and cloud shadows. To summarize confusion matrices we
+report accuracy as:
 
@@
 To map long term LULCC in the Yangambi region we used the contemporary
-Global Forest Change version 1.6 (GFC, tile 10N-020E) map data
-\citep{hansen2013}. Using the GFC data we calculated the latest state of
-the forest with respect to the conditions at the start of 1958, 60 years
-earlier. In our analysis we only included GFC pixels which recorded no
-forest loss throughout the whole 2000 - 2018 period, \textbf{as such
+Global Forest Change version 1.6 (GFC, tile 10N-020E) (lossyear) map
+data \citep{hansen2013}. Using the GFC data we calculated the latest
+state of the forest with respect to the conditions at the start of 1958,
+60 years earlier. In our analysis we only included GFC pixels which
+recorded no forest loss throughout the whole 2000 - 2018 period. Forest
+loss in the context of GFC is defined as ``a stand-replacement
+disturbance, or a change from a forest to non-forest state''. As such,
 locations which would see reforestation or deforestation between 2000
-and 2018 would be marked as non-forest (i.e.~disturbed)}. As the
+and 2018 would be marked as non-forest (i.e.~disturbed). As the
 resolution of the historical forest classification exceeds that of the
@@
 limited the analysis to the right bank of the Congo river. We provide
-summary statistics of historical and contemporary \textbf{deforestation
-and reforestation}. We map permanent forest loss after 1958,
-\textbf{reforestation} after 1958, \textbf{recent deforestation} and
-long term (stable) forest cover. \textbf{All references to changes over
-time in the context of our analysis explicitly compare the historical
-and contemporary periods from hereon forth.}
+summary statistics of historical and contemporary deforestation and
+reforestation. We map permanent deforestation after 1958, reforestation
+after 1958, recent deforestation and long term (stable) forest cover.
+All references to changes over time in the context of our analysis
+explicitly compare the historical and contemporary periods from hereon
+forth.
 
@@
 To quantify changes in the structure of forest cover and its
-disturbances we used spatial pattern analysis landscape (fragmentation)
-metrics \citep{dale1999}. Landscape metrics provide a mathematical
-framework for the analysis of discrete land‐cover classes and allows us
-to capture their composition and configuration. These metrics are
-therefore commonly used to compare how landscapes change over time
-\citep{hesselbarth2019}. In particular, fractals provide a way to
-quantify complex natural landscapes, including their self-similarity,
-across scales \citep{Li2000, Mandelbrot1977} We report the ratio of edge
-to area and the fractal dimension to quantify landscape complexity of
-forest disturbances. A fractal dimension closer to 2 suggest a more
+disturbances we used spatial landscape pattern analysis
+(i.e.~fragmentation) metrics \citep{dale1999}. Landscape metrics provide
+a mathematical framework for the analysis of discrete land‐cover classes
+and allows us to capture their composition and configuration. These
+metrics are therefore commonly used to compare how landscapes change
+over time \citep{hesselbarth2019}. In particular, fractals provide a way
+to quantify complex natural landscapes, including their self-similarity
+across scales \citep{Li2000, Mandelbrot1977}. We report the ratio of
+edge to area and the fractal dimension to quantify landscape complexity
+of forest disturbances. A fractal dimension closer to 2 suggest a more
 complex (fragmented) landscape.
@@
 than 1 ha and smaller than the 95\textsuperscript{th} percentile of the
-patch size distribution using the R package landscapemetrics
+patch size distribution using the R package ``landscapemetrics''
 \citep{hesselbarth2019}. We provide mean and standard deviation on edge,
@@
 
-We estimated above ground carbon (AGC) losses and gains, \textbf{due to
-deforestation and reforestation}, using plot based averages of recent
+We estimated above ground carbon (AGC) losses and gains, due to
+deforestation and reforestation, using plot based averages of recent
 inventory data at permanent sampling plots in the area (Figure 2). We
@@
 relations used to scale the survey data. Unlike standard square 1 ha
-plots, edge plots were set back 200 m from forest edges and were 50x200
-m, with the 50 m side of the plot along the forest edge and continuing
-200 m into the forest (Appendix Table 1). We further confirmed that
-forest edge plots did not show a significantly different AGC compared to
-those of non-edged / mixed forest plots (Mann Whitney U test, p
-\textless{} 0.05). Thus it was not necessary to explicitly quantify
-changes in AGC caused by edge effects. Moreover, we used the mean values
-of the mixed forest as representative for potential AGC losses. Despite
-the challenges inherent in quantifying AGC for forest edges we mapped
-the total extent of the edges in the contemporary landscape. To align
-our landscape analysis with exploratory analysis of the survey data we
-used a buffer of 200 m to estimate the extent of forest edges and
-patches, up to the location of forest edge plots.
+plots, edge plots (163.03 \(\pm\) 19.39 Mg C ha\textsuperscript{-1}, N =
+5) were set back 200 m from forest edges and were 50x200 m, with the 50
+m side of the plot along the forest edge and continuing 200 m into the
+forest (Appendix Table 2). We further confirmed that forest edge plots,
+as compared to mixed forest plots (160.48 \(\pm\) 23.84 Mg C
+ha\textsuperscript{-1}, N = 8, see Appendix Table 3 for all forest
+types), did not show a significantly different AGC (Mann Whitney U test,
+p \textless{} 0.05). Thus it was not necessary to explicitly quantify
+changes in AGC caused by edge effects. Moreover, we used the mean value,
+and its uncertainty (i.e.~standard deviation), of the mixed forest as
+representative for potential AGC losses. Despite the challenges inherent
+in quantifying AGC for forest edges we mapped the total extent of the
+edges in the contemporary landscape. To align our landscape analysis
+with exploratory analysis of the survey data we used a buffer of 200 m
+to estimate the extent of forest edges and patches, up to the location
+of forest edge plots.
 
 Surveys of old plantations show a large variation in AGC, depending on
-age and the crop type. For example, the AGC values varied from 168.67 to
-86.55 Mg C ha\textsuperscript{-1}, for \emph{Hevea brasiliensis} (rubber
-tree) and \emph{Elaeis guineensis} (oil palm) plots respectively
+age and the crop type. For example, the AGC values varied from 86.55 to
+168.67 Mg C ha\textsuperscript{-1}, for \emph{Elaeis guineensis} (oil
+palm) and \emph{Hevea brasiliensis} (rubber tree) plots respectively
 (\citet{bustillo2018}, personal communications). These higher values are
-in line with the mixed AGC estimates in the area, while the palm
-plantations resemble old-regrowth values (81.8 Mg C
-ha\textsuperscript{-1}, see Appendix Table 1). We therefore use both the
-estimates of old-regrowth and mixed forest to estimate AGC for regrowth.
-We did not have sufficient data to account for individual changes in AGC
-across plantations.
+in line with the mixed plot AGC estimates (160.48 \(\pm\) 23.84 Mg C
+ha\textsuperscript{-1}, N = 8) in the area, while the palm plantations
+resemble old-regrowth values (81.87 Mg C ha\textsuperscript{-1}, N = 1).
+To quantify AGC in reforested areas we therefore use both AGC estimates
+of old-regrowth and mixed forest, as lower and upper bounds. We did not
+have sufficient data to account for individual changes in AGC across
+plantations.
 
@@
 permanent sample plot inventory data such as stem density, above ground
-biomass and tree species richness. Additional comparisons are made
-between contemporary Geo-Eye data and the historical orthomosaic derived
-PC values. Due to the few available permanent sampling plots in both
-scenes we used a non-parametric paired signed rank (Wilcoxon) test
-\citep{wilcoxon1945} to determine differences between the PC values of
-the Geo-Eye and historical orthomosaic image scenes across mono-dominant
-and mixed forest types. In all analysis, mono-dominant site 4 was
-removed from the analysis due to cloud contamination.
+biomass and tree species richness (Appendix Table 2-3). Additional
+comparisons are made between contemporary Geo-Eye data and the
+historical orthomosaic derived PC values. Due to the few available
+permanent sampling plots in both scenes we used a non-parametric paired
+signed rank (Wilcoxon) test \citep{wilcoxon1945} to determine
+differences between the PC values of the Geo-Eye and historical
+orthomosaic image scenes across mono-dominant and mixed forest types. In
+all analysis, mono-dominant site 4 was removed from the analysis due to
+cloud contamination.
 
@@
 Yangambi region provided an image scene covering approximately 733
-million pixels across \textasciitilde{}82800 ha (\textasciitilde{}23x36
-km, Figure 2). The overall accuracy of the SfM orthomosaic composition
-was 0.88 m/pixel using the sparse cloud DEM for corrections at 45.8
-m/pixel. The resulting georeferenced scene had a spatial accuracy of
+million pixels across \textasciitilde{}82 800 ha with a resolution of
+0.88 m / pixel (or \textasciitilde{}23x36 km, Figure 2). The overall
+spatial accuracy of the SfM orthomosaic composition using the sparse
+cloud DEM (with a resolution of 45.8 m/pixel) was limited to
 approximately 23 m. Further georeferencing outside the SfM workflow
@@
 
-\caption{Overview of the final orthomosaic of the greater Yangambi region (A) as well as the forest cover classification uncertainty (B) used to generate the final Land Land Cover map (see Figure 6).
+\caption{Overview of the final orthomosaic of the greater Yangambi region (A) as well as the forest cover classification uncertainty (B) used to generate the final Land Cover map (see Figure 6).
 Homogeneous polygons used in the generation of the synthetic landscape for Convolutional Neural Network training, testing and validation are marked as dashed and full lines for forest and non-forest regions, respectively.
@@
 out-of-sample (validation) synthetic landscape data. Using all pixels
-within the \textbf{validation} polygons (Figure 5) showed a similar
-accuracy value of \textasciitilde{}98\%. Using all polygons across the
-scene, including those used in the generation of testing and training
-synthetic landscapes, increased the accuracy to \textasciitilde{}99\%
-(Table 2). A comparison with recent pan-chromatic Geo-Eye data shows
-good agreement, with an accuracy of \textasciitilde{}87\% across all
-pixels, between the landsat based GFC data and downscaled CNN results
-(Table 2 and Figure 7).
+within the validation polygons (Figure 5) showed a similar accuracy
+value of \textasciitilde{}98\%. Using all polygons across the scene,
+including those used in the generation of testing and training synthetic
+landscapes, increased the accuracy to \textasciitilde{}99\% (Table 2). A
+comparison with recent pan-chromatic Geo-Eye data shows good agreement,
+with an accuracy of \textasciitilde{}87\% across all pixels, between the
+landsat based GFC data and downscaled CNN results (Table 2 and Figure
+7).
 
@@
 Scaling our classifier to the whole historical orthomosaic we detected
-\textasciitilde{}16200 ha (or \textasciitilde{}20\% of the scene) of
+\textasciitilde{}16 200 ha (or \textasciitilde{}20\% of the scene) of
 disturbed forests. A large fraction of the disturbed area was restored
@@
 blue). Recent deforested areas, as registered through satellite remote
-sensing \textbf{(\textgreater{} 2000)}, approximate 8776 ha (Table 3,
-Figure 6 - light green).
+sensing (\textgreater{} 2000), approximate 8776 ha (Table 3, Figure 6 -
+light green).
 
@@
 colours respectively). These differences are reflected in the analysis
-of landscape metrics of \textbf{deforestation}. Between the historical
-and contemporary LULCC maps we see an increase in small disturbances, as
+of landscape metrics of deforestation. Between the historical and
+contemporary LULCC maps we see an increase in small disturbances, as
 indicated by the decreasing area of the mean patch size, down to
@@
 1451 \(\pm\) 943 m, compared to contemporary landscapes
-\textasciitilde{}921 \(\pm\) 362 m (Table \textbf{4}). This shift in
-perimeter area ratio led to a similar change in the fractal index,
-slightly increasing in value to 1.1 \(\pm\) 0.05 from 1.09 \(\pm\) 0.04
-over time. Values closer to a fractal index of 2 suggest a more complex
+\textasciitilde{}921 \(\pm\) 362 m (Table 4). This shift in perimeter
+area ratio led to a similar change in the fractal index, slightly
+increasing in value to 1.1 \(\pm\) 0.05 from 1.09 \(\pm\) 0.04 over
+time. Values closer to a fractal index of 2 suggest a more complex
 (fragmented) landscape.
@@
 0.05). Edge influence did not extend beyond 200 m from a forest edge,
-but still represented an area of 13151 ha (Table 3).
+but still represented an area of 13 151 ha (Table 3).
 
@@
 of forest and plantations. Assuming high density stands, based on
-previous work, this could amount to a carbon gains of 1592 Gg C across
-our study area, offsetting more recent losses of approximately 1408 Gg
-C. On the other hand, at the low end, if we assume a lower carbon
-density of 81.8 Mg C ha\textsuperscript{-1} this would result in a total
-carbon gain of 811 Gg C. Using our approach results indicate that
-overall deforestation around Yangambi has resulted in a loss of
-\textasciitilde{}2416 Gg C in AGC stocks.
+previous work, this could amount to a carbon gains of up to 1592 Gg C
+across our study area, offsetting more recent losses of approximately
+1408 \(\pm\) 209 Gg C. On the other hand, at the low end, if we assume a
+lower carbon density of 81.8 Mg C ha\textsuperscript{-1} this would
+result in a total carbon gain of 811 Gg C. Using our approach results
+indicate that overall deforestation around Yangambi has resulted in a
+loss of \textasciitilde{}2416 \(\pm\) 358 Gg C in AGC stocks.
 
@@
 2013 based satellite remote sensing data.
-Spatial coverage statistics are provided in square kilometers (km) and hectares (ha), rounded to the nearest integer as well as Above Ground Carbon (AGC) scaled using recent survey measurements.}
+Spatial coverage statistics are provided hectares (ha), rounded to the nearest integer as well as Above Ground Carbon (AGC) scaled using recent survey measurements.}
 \centering
@@
 \cmidrule(l{3pt}r{3pt}){3-3}
-  & ha & Gg C $ha\textsuperscript{-1}$\\
+  & ha & Gg C\\
 \midrule
@@
 reforestation >1958 & 9918 & 811 - 1592\\
-deforestation >2000 & 8776 & 1408\\
+deforestation >2000 & 8776 & 1408 $\pm$ 209\\
 \addlinespace
-deforestation >1958 (permanent) & 6282 & 1008\\
+deforestation >1958 (permanent) & 6282 & 1008 $\pm$ 150\\
 \bottomrule
@@
 young-regrowth plots. One marked difference is noted in the
-mono-dominant plot 6 (Appendix Table 1). Here, the current mono-dominant
+mono-dominant plot 6 (Appendix Table 2). Here, the current mono-dominant
 \emph{Brachystegia laurentii} is a recent development, changing the
@@
 canopy. This is reflected in a change of the FOTO PC value from 0.19
-historically to its current value of 0.54 \textbf{(Figure 9).} This
-historical value is similar to the mean of contemporary mono-dominant
-stands of \emph{Gilbertiodendron dewevrei} with PC averaging 0.34
-\(\pm\) 0.1, and is only slightly higher than historical values for a
-mixed forest (0.18 \(\pm\) 0.08, \textbf{Figure 9)}. The reverse pattern
-is seen in the contemporary PC values. Here, the value of 0.54 exceeds
-those of most mono-dominant stands (0.35 \(\pm\) 0.08), and is even
-further removed from the values noted for mixed forests (0.12 \(\pm\)
-0.03, \textbf{Figure 9}).
+historically to its current value of 0.54 (Figure 9). This historical
+value is similar to the mean of contemporary mono-dominant stands of
+\emph{Gilbertiodendron dewevrei} with PC averaging 0.34 \(\pm\) 0.1, and
+is only slightly higher than historical values for a mixed forest (0.18
+\(\pm\) 0.08, Figure 9). The reverse pattern is seen in the contemporary
+PC values. Here, the value of 0.54 exceeds those of most mono-dominant
+stands (0.35 \(\pm\) 0.08), and is even further removed from the values
+noted for mixed forests (0.12 \(\pm\) 0.03, Figure 9).
 
@@
 plot measurements of Above Ground Carbon, stem density and species
-richness were non-significant (p \textgreater{} 0.05, Appendix
-\textbf{Figures 4-6}).
+richness were non-significant (p \textgreater{} 0.05, Appendix Figures
+4-6).
 
@@
 historical scences do not show landscape wide canopy features (Figure 10
-A-B), unlike the contemporary scene (Figure 10 C-D). Where the FOTO
-algorithm picks up landscape features such as changes in texture across
-the contemporary Geo-Eye scene (e.g.~the river valley as a diagonal line
-in Figure 10D), however, no corresponding landscape patterns are found
-by the FOTO algorithm in the historical orthomosaic.
+A-B), unlike the contemporary scene (Figure 10 C-D). In the contemporary
+scene the FOTO algorithm picks up landscape features such as changes in
+texture along the river valley (the diagonal line in Figure 10D).
+However, no corresponding landscape patterns are found by the FOTO
+algorithm in the historical orthomosaic.
 
@@
 made georeferencing challenging. Although the village of Yangambi
-provided a range of buildings as (hard-edge) references, other areas
+provided a range of buildings as (hard-edge) references other areas
 within the central Congo Basin might have fewer permanent structures and
@@
 Our analysis shows that the majority of deforestation around Yangambi
-happened toward the late 1950's (\textasciitilde{}16200 ha).
-Considerable \textbf{reforestation} has occured since the aerial survey
-was executed (\textasciitilde{}9918 ha), and socio-economic instability
+happened toward the late 1950s (\textasciitilde{}16 200 ha).
+Considerable reforestation has occured since the aerial survey was
+executed (\textasciitilde{}9918 ha), and socio-economic instability
 prevented further large scale forest exploitation. In particular, many
@@
 previously cleared or disturbed areas. The majority of this
-\textbf{reforestation} takes the form of isolated patches of forest but
-is offset by further deforestation of previously untouched forest.
-Generally, the function and structure of forests can be influenced by
-forest edges that are located up to 1km away however most effects are
-pronounced within the first 300 m from the edge \citep{gascon2000}. Our
-analysis of edge effects on AGC has shown that the influence is
-negligible 200 m away from the edge. \citet{phillips2006} have shown
-similar weak responses to edge effects in the Amazon forest. Due to a
-lack of data on the extent (depth) of edge effects and their influence
-on AGC beyond 200 m we did not include any estimates of carbon loss or
-gain within these zones. However, it must be stated that the influence
-of edges throughout the landscape was not marginal as these areas would
-account for 13151 ha. Thus edges could have a substantial negative
-\citep{brinck2017} or positive \citep{reinmann2017} influence on AGC.
-Similarly, uncertainties in how to explicitly correct for plantations in
-the landscape present a further challenge. Thus, although our estimates
-are only indicative they do underscore the important influence of
-landscape structure when carbon accounting. However, our findings do not
-indicate that deforestation in Congo basin is declining, on the
-contrary.
+reforestation takes the form of isolated patches of forest but is offset
+by further deforestation of previously untouched forest. Generally, the
+function and structure of forests can be influenced by forest edges that
+are located up to 1 km away however most effects are pronounced within
+the first 300 m from the edge \citep{gascon2000}. Our analysis of edge
+effects on AGC has shown that the influence is negligible 200 m away
+from the edge. \citet{phillips2006} have shown similar weak responses to
+edge effects in the Amazon forest. Due to a lack of data on the extent
+(depth) of edge effects and their influence on AGC beyond 200 m we did
+not include any estimates of carbon loss or gain within these zones.
+However, it must be stated that edges throughout the landscape make up a
+substantial area and account for 13 151 ha. Thus edges could have a
+substantial negative \citep{brinck2017} or positive \citep{reinmann2017}
+influence on AGC. Similarly, uncertainties in how to explicitly correct
+for plantations in the landscape present a further challenge. Similarly,
+variability across mixed forest plots used in scaling above ground
+carbon estimates due to deforestation introduced additional uncertainty
+(see Appendix Table 3). Thus, although our estimates are only indicative
+they do underscore the important influence of landscape structure in
+carbon accounting. However, our findings do not indicate that
+deforestation in Congo basin is declining, on the contrary.
 
 Over the past half century there has been a clear shift in land use in
-Yangambi (Figure 5). Land use has shifted away from for example a
-regular fishbone deforestation pattern that emerges when (large scale)
-agricultural interests dominate the landscape \citep{arima2015}, to a
-more fragmented landscape (Figure 5D). The former is consistent with
-historical land management at the time of the aerial survey
-\citep{bustillo2018}. These regular patterns have since been reversed
-because of a decrease in large scale intensive agriculture, replaced by
-ad-hoc small scale subsistence farming with large perimeter to area
-relationships (i.e.~ragged edges). Consequently, edge effects in the
-current landscape are far more pronounced than in the historical scene.
+Yangambi (Figure 5). Land use has shifted away from a regular (fishbone)
+deforestation pattern that emerges when (large scale) agricultural
+interests dominate the landscape \citep{arima2015}, to a more fragmented
+landscape (Figure 5D). The former is consistent with historical land
+management at the time of the aerial survey \citep{bustillo2018}. These
+regular patterns reversed due to a decrease in large scale intensive
+agriculture and an increase in ad-hoc small scale subsistence farming
+with large perimeter to area relationships (i.e.~ragged edges).
+Consequently, edge effects in the current landscape are far more
+pronounced than in the historical scene.
 
@@
 implicit protective policy measures (e.g.~INERA concessions,
-\citet{bustillo2018}) on a multi-decadal time scale.
-\textbf{Reforestation} in non-continuous areas within Yangambi could
-increase landscape connectivity and help increase biodiversity
-\citep{vandeperre2018}.
+\citet{bustillo2018}) on a multi-decadal time scale. Reforestation in
+non-continuous areas within Yangambi could increase landscape
+connectivity and help increase biodiversity \citep{vandeperre2018}.
 
@@
 that texture can be used to consistently capture canopy properties 60
-years apart. \textbf{Differences in PC between forest types
-(e.g.~mono-dominant vs.~mixed, Figure 9) corroborate that texture can
-serve as a basis for LULC classification. However, inflexibility on part
-of the FOTO technique in dealing with non-standardized (historical)
-data, or scaling these results to AGC values, limits its use case.} We
-advise that future valorisation efforts should preferentially work from
-foto negatives (if available) to ensure optimal data quality in
-resolution, contrast and sharpness.
+years apart. Differences in PC between forest types (e.g.~mono-dominant
+vs.~mixed, Figure 9) corroborate that texture can serve as a basis for
+LULC classification. However, inflexibility on part of the FOTO
+technique in dealing with non-standardized (historical) data, or scaling
+these results to AGC values, limits its use case. We advise that future
+valorisation efforts should preferentially work from foto negatives (if
+available) to ensure optimal data quality in resolution, contrast and
+sharpness.
 
@@
 
-Hufkens et al. (2019). A curated dataset of aerial survey images over
+Hufkens et al.~(2019). A curated dataset of aerial survey images over
 the central Congo Basin, 1958. Zenodo:

### 2020-02-06 analysis/08_AGC_analysis.R

@@ -14,7 +14,8 @@ agc <- psp %>%
   group_by(type) %>%
   summarize(mean = mean(above.ground.C.stock..Mg.C.ha.1.),
             median = median(above.ground.C.stock..Mg.C.ha.1.),
-            sd = sd(above.ground.C.stock..Mg.C.ha.1.))
+            sd = sd(above.ground.C.stock..Mg.C.ha.1.),
+            n = length(above.ground.C.stock..Mg.C.ha.1.))
 
 # test if the two plot sets (edge and mixed) come from
 # significantly different populations or not
@@ -55,13 +56,22 @@ change_stats <- rbind(change_stats,c(5,freq(edge_values)[2,2]))
 change_stats$sq_km <- change_stats$count * 0.03^2
 change_stats$ha <- change_stats$sq_km / 0.01
 
+mean_mixed_agc <- agc$mean[agc$type=="mixed"]
+sd_mixed_agc <- agc$sd[agc$type=="mixed"]
+
+print(paste(mean_mixed_agc, sd_mixed_agc, sep = " +_ "))
+
 agc_high <-
   round(change_stats$ha * agc$mean[agc$type=="mixed"]/1000)
+
+agc_high_sd <-
+  round(change_stats$ha * agc$sd[agc$type=="mixed"]/1000)
 
 agc_low <-
   round(change_stats$ha * agc$mean[agc$type=="old-regrowth"]/1000)
 
 change_stats$agc <- agc_high
+change_stats$agc[3:4] <- paste(agc_high[3:4], agc_high_sd[3:4], sep = " $\\pm$ ")
 change_stats$agc[2] <- paste(agc_low[2], agc_high[2], sep = " - ")
 
 change_stats <-  change_stats %>%
@@ -75,10 +85,10 @@ change_stats <- change_stats %>%
 # reshuffle rows
 change_stats <- change_stats[c(1,5,2:4),]
 
-write.table(change_stats,
-            "./data/surveys/lulcc_change_stats.csv",
-            row.names = FALSE,
-            col.names = TRUE,
-            quote = TRUE,
-            sep = ",")
+# write.table(change_stats,
+#             "./data/surveys/lulcc_change_stats.csv",
+#             row.names = FALSE,
+#             col.names = TRUE,
+#             quote = TRUE,
+#             sep = ",")
 

### 2020-02-08 manuscript/main.Rmd
absolute total scene area

@@
 Here, we used a combination of historical (1958) aerial photography and contemporary remote sensing data to map long-term changes in the extent and structure of the tropical forest surrounding Yangambi (DR Congo), in the central Congo Basin.
-Our study leveraged Structure-from-Motion and a Convolutional Neural Network based LULC classifier, using synthetic landscapes based image augmentation, to map historical forest cover across a large orthomosaic (~82 800 ha) geo-referenced to ~4.7 $\pm$ 4.3 m at sub meter resolution.
+Our study leveraged Structure-from-Motion and a Convolutional Neural Network based LULC classifier, using synthetic landscapes based image augmentation, to map historical forest cover across a large orthomosaic (93 431 ha) geo-referenced to ~4.7 $\pm$ 4.3 m at sub meter resolution.
 A comparison with contemporary LULC data showed a shift from previously highly regular industrial deforestation of large areas, to discrete smallholder farming clearing, increasing landscape fragmentation but also providing opportunties for substantial forest regrowth.
@@
                             header = TRUE,
-                            sep = ",")
+                            sep = ",",
+                            stringsAsFactors = FALSE)
   change_stats[1:2,3] <- NA
@@
 
-  change_stats <- rbind(c(sum(change_stats$ha[c(3,5)]), NA),
+  # total non-forest 1958
+  change_stats <- rbind(
+    c(sum(change_stats$ha[c(3,5)]), NA),
+                        change_stats)
+  # total scene area
+  change_stats <- rbind(
+    c(sum(as.numeric(change_stats$ha[c(2,4:6)])), NA),
     change_stats)
@@
   row.names(change_stats) <- c(
+   "Total scene area",
    "Non-forest cover (1958)",
@@
 The use of high resolution historical images combined with SfM image processing techniques allowed us to mosaic old imagery across a large extent.
-The final orthomosaic composition of the Yangambi region provided an image scene covering approximately 733 million pixels across ~82 800 ha with a resolution of 0.88 m / pixel (or ~23x36 km, Figure 2).
+The final orthomosaic composition of the Yangambi region provided an image scene covering approximately 733 million pixels across 93 431 ha with a resolution of 0.88 m / pixel (or ~23x36 km, Figure 2).
 The overall spatial accuracy of the SfM orthomosaic composition using the sparse cloud DEM (with a resolution of 45.8 m/pixel) was limited to approximately 23 m.
@@
     add_header_above(c(" " = 2,"AGC" = 1)) %>%
-    #pack_rows("",6,6) %>%
+    #pack_rows("",1,1) %>%
     kable_styling(latex_options = c("hold_position"))

### 2020-02-08 manuscript/main.tex
absolute total scene area

@@
 image augmentation, to map historical forest cover across a large
-orthomosaic (\textasciitilde{}82 800 ha) geo-referenced to
-\textasciitilde{}4.7 \(\pm\) 4.3 m at sub meter resolution. A comparison
-with contemporary LULC data showed a shift from previously highly
-regular industrial deforestation of large areas, to discrete smallholder
-farming clearing, increasing landscape fragmentation but also providing
-opportunties for substantial forest regrowth. We estimated above ground
-carbon gains through reforestation to range from 811 to 1592 Gg C,
-partially offseting historical deforestation (2416 Gg C), in our study
-area. Efforts to quantify long term canopy texture changes and their
-link to above ground carbon had limited to no success. Our analysis
-provides methods and insights into key spatial and temporal patterns of
-deforestation and reforestation at a multi-decadal scale, providing a
-historical context for past and ongoing forest research in the area.}
+orthomosaic (93 431 ha) geo-referenced to \textasciitilde{}4.7 \(\pm\)
+4.3 m at sub meter resolution. A comparison with contemporary LULC data
+showed a shift from previously highly regular industrial deforestation
+of large areas, to discrete smallholder farming clearing, increasing
+landscape fragmentation but also providing opportunties for substantial
+forest regrowth. We estimated above ground carbon gains through
+reforestation to range from 811 to 1592 Gg C, partially offseting
+historical deforestation (2416 Gg C), in our study area. Efforts to
+quantify long term canopy texture changes and their link to above ground
+carbon had limited to no success. Our analysis provides methods and
+insights into key spatial and temporal patterns of deforestation and
+reforestation at a multi-decadal scale, providing a historical context
+for past and ongoing forest research in the area.}
 
@@
 Yangambi region provided an image scene covering approximately 733
-million pixels across \textasciitilde{}82 800 ha with a resolution of
-0.88 m / pixel (or \textasciitilde{}23x36 km, Figure 2). The overall
-spatial accuracy of the SfM orthomosaic composition using the sparse
-cloud DEM (with a resolution of 45.8 m/pixel) was limited to
-approximately 23 m. Further georeferencing outside the SfM workflow
-reduced the mean error at the ground control points to 5.3 \(\pm\) 4.9
-px (\textasciitilde{}4.7 \(\pm\) 4.3 m), with a median error of 2.9 px
-(2.6 m). The orthomosaic served as input for all subsequent LULCC
-analysis with all derived maps provided with the manuscript repository
-(see data \& code availability statements below).
+million pixels across 93 431 ha with a resolution of 0.88 m / pixel (or
+\textasciitilde{}23x36 km, Figure 2). The overall spatial accuracy of
+the SfM orthomosaic composition using the sparse cloud DEM (with a
+resolution of 45.8 m/pixel) was limited to approximately 23 m. Further
+georeferencing outside the SfM workflow reduced the mean error at the
+ground control points to 5.3 \(\pm\) 4.9 px (\textasciitilde{}4.7
+\(\pm\) 4.3 m), with a median error of 2.9 px (2.6 m). The orthomosaic
+served as input for all subsequent LULCC analysis with all derived maps
+provided with the manuscript repository (see data \& code availability
+statements below).
 
@@
 \midrule
+Total scene area & 93431 & \\
 Non-forest cover (1958) & 16200 & \\
@@
 reforestation >1958 & 9918 & 811 - 1592\\
+\addlinespace
 deforestation >2000 & 8776 & 1408 $\pm$ 209\\
-\addlinespace
 deforestation >1958 (permanent) & 6282 & 1008 $\pm$ 150\\

### 2020-02-08 analysis/08_AGC_analysis.R

@@ -56,6 +56,10 @@ change_stats <- rbind(change_stats,c(5,freq(edge_values)[2,2]))
 change_stats$sq_km <- change_stats$count * 0.03^2
 change_stats$ha <- change_stats$sq_km / 0.01
 
+# calcualte absolute surface area of the scene
+# not approximation
+total_ha_scene <- round(sum(change_stats$ha[1:4]))
+
 mean_mixed_agc <- agc$mean[agc$type=="mixed"]
 sd_mixed_agc <- agc$sd[agc$type=="mixed"]
 
@@ -84,11 +88,12 @@ change_stats <- change_stats %>%
 
 # reshuffle rows
 change_stats <- change_stats[c(1,5,2:4),]
+#change_stats <- rbind(as.numeric(c(NA, total_ha_scene, NA)), change_stats)
 
-# write.table(change_stats,
-#             "./data/surveys/lulcc_change_stats.csv",
-#             row.names = FALSE,
-#             col.names = TRUE,
-#             quote = TRUE,
-#             sep = ",")
+write.table(change_stats,
+            "./data/surveys/lulcc_change_stats.csv",
+            row.names = FALSE,
+            col.names = TRUE,
+            quote = TRUE,
+            sep = ",")
 

### 2020-02-09 manuscript/main.Rmd
tilde

@@
 Here, we used a combination of historical (1958) aerial photography and contemporary remote sensing data to map long-term changes in the extent and structure of the tropical forest surrounding Yangambi (DR Congo), in the central Congo Basin.
-Our study leveraged Structure-from-Motion and a Convolutional Neural Network based LULC classifier, using synthetic landscapes based image augmentation, to map historical forest cover across a large orthomosaic (93 431 ha) geo-referenced to ~4.7 $\pm$ 4.3 m at sub meter resolution.
+Our study leveraged Structure-from-Motion and a Convolutional Neural Network based LULC classifier, using synthetic landscapes based image augmentation, to map historical forest cover across a large orthomosaic (~93 431 ha) geo-referenced to ~4.7 $\pm$ 4.3 m at sub meter resolution.
 A comparison with contemporary LULC data showed a shift from previously highly regular industrial deforestation of large areas, to discrete smallholder farming clearing, increasing landscape fragmentation but also providing opportunties for substantial forest regrowth.
@@
 The use of high resolution historical images combined with SfM image processing techniques allowed us to mosaic old imagery across a large extent.
-The final orthomosaic composition of the Yangambi region provided an image scene covering approximately 733 million pixels across 93 431 ha with a resolution of 0.88 m / pixel (or ~23x36 km, Figure 2).
+The final orthomosaic composition of the Yangambi region provided an image scene covering approximately 733 million pixels across ~93 431 ha with a resolution of 0.88 m / pixel (or ~23x36 km, Figure 2).
 The overall spatial accuracy of the SfM orthomosaic composition using the sparse cloud DEM (with a resolution of 45.8 m/pixel) was limited to approximately 23 m.

### 2020-02-09 manuscript/main.tex
tilde

@@
 image augmentation, to map historical forest cover across a large
-orthomosaic (93 431 ha) geo-referenced to \textasciitilde{}4.7 \(\pm\)
-4.3 m at sub meter resolution. A comparison with contemporary LULC data
-showed a shift from previously highly regular industrial deforestation
-of large areas, to discrete smallholder farming clearing, increasing
-landscape fragmentation but also providing opportunties for substantial
-forest regrowth. We estimated above ground carbon gains through
-reforestation to range from 811 to 1592 Gg C, partially offseting
-historical deforestation (2416 Gg C), in our study area. Efforts to
-quantify long term canopy texture changes and their link to above ground
-carbon had limited to no success. Our analysis provides methods and
-insights into key spatial and temporal patterns of deforestation and
-reforestation at a multi-decadal scale, providing a historical context
-for past and ongoing forest research in the area.}
+orthomosaic (\textasciitilde{}93 431 ha) geo-referenced to
+\textasciitilde{}4.7 \(\pm\) 4.3 m at sub meter resolution. A comparison
+with contemporary LULC data showed a shift from previously highly
+regular industrial deforestation of large areas, to discrete smallholder
+farming clearing, increasing landscape fragmentation but also providing
+opportunties for substantial forest regrowth. We estimated above ground
+carbon gains through reforestation to range from 811 to 1592 Gg C,
+partially offseting historical deforestation (2416 Gg C), in our study
+area. Efforts to quantify long term canopy texture changes and their
+link to above ground carbon had limited to no success. Our analysis
+provides methods and insights into key spatial and temporal patterns of
+deforestation and reforestation at a multi-decadal scale, providing a
+historical context for past and ongoing forest research in the area.}
 
@@
 Yangambi region provided an image scene covering approximately 733
-million pixels across 93 431 ha with a resolution of 0.88 m / pixel (or
-\textasciitilde{}23x36 km, Figure 2). The overall spatial accuracy of
-the SfM orthomosaic composition using the sparse cloud DEM (with a
-resolution of 45.8 m/pixel) was limited to approximately 23 m. Further
-georeferencing outside the SfM workflow reduced the mean error at the
-ground control points to 5.3 \(\pm\) 4.9 px (\textasciitilde{}4.7
-\(\pm\) 4.3 m), with a median error of 2.9 px (2.6 m). The orthomosaic
-served as input for all subsequent LULCC analysis with all derived maps
-provided with the manuscript repository (see data \& code availability
-statements below).
+million pixels across \textasciitilde{}93 431 ha with a resolution of
+0.88 m / pixel (or \textasciitilde{}23x36 km, Figure 2). The overall
+spatial accuracy of the SfM orthomosaic composition using the sparse
+cloud DEM (with a resolution of 45.8 m/pixel) was limited to
+approximately 23 m. Further georeferencing outside the SfM workflow
+reduced the mean error at the ground control points to 5.3 \(\pm\) 4.9
+px (\textasciitilde{}4.7 \(\pm\) 4.3 m), with a median error of 2.9 px
+(2.6 m). The orthomosaic served as input for all subsequent LULCC
+analysis with all derived maps provided with the manuscript repository
+(see data \& code availability statements below).
 

### 2020-02-09 manuscript/main.Rmd
caption

@@
 2013 based satellite remote sensing data.
-Spatial coverage statistics are provided hectares (ha), rounded to the nearest integer as well as Above Ground Carbon (AGC) scaled using recent survey measurements.",
+Spatial coverage statistics are provided hectares (ha), rounded to the nearest integer as well as Above Ground Carbon (AGC) scaled using recent survey measurements.
+Both non-forest cover in 1958 and the forest edge area are subsets of the original data (see Figure 6) and do not count toward the total scene area.",
        booktabs = TRUE,

### 2020-02-09 manuscript/main.tex
caption

@@
 2013 based satellite remote sensing data.
-Spatial coverage statistics are provided hectares (ha), rounded to the nearest integer as well as Above Ground Carbon (AGC) scaled using recent survey measurements.}
+Spatial coverage statistics are provided hectares (ha), rounded to the nearest integer as well as Above Ground Carbon (AGC) scaled using recent survey measurements.
+Both non-forest cover in 1958 and the forest edge area are subsets of the original data (see Figure 6) and do not count toward the total scene area.}
 \centering
